### 2011-07-11 doc/report.tex
m

\documentclass[a4]{article}
\usepackage[utf8]{inputenc}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{clrscode}
\usepackage[draft]{fixme}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{textcomp}

\usepackage{prettyref}
\newcommand{\sectionname}{Section}
% \renewcommand{\figurename}{Figure}
\newcommand*{\pref}{\prettyref}
% Adds varioref to the prettyrefs.
% Replacing vref with ref to avoid page numbers
\newrefformat{cha}{\chaptername~\ref{#1}}
\newrefformat{sec}{\sectionname~\ref{#1}}
\newrefformat{fig}{\figurename~\ref{#1}}
\newrefformat{eq}{\equationname~\ref{#1}}
\newrefformat{tab}{\tablename~\ref{#1}}
\newrefformat{app}{\appendixname~\ref{#1}}
\newrefformat{alg}{Algorithm~\ref{#1}}

\title{Algorithms for Longest Common Extensions}

\author{Jesper Kristensen, s062397}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

% http://texblog.net/latex/layout/page/2/
\fontdimen2\font=0.25em% interword space
\fontdimen3\font=1em% interword stretch
\fontdimen4\font=0.10em% interword shrink

\begin{document}

\newif\ifarticle
\newif\ifreport

\reporttrue

\maketitle

\newcommand{\sortt}{\textit{sort}(n,\sigma)}
\newcommand{\LCE}{\textit{LCE}}
\newcommand{\NCA}{\textit{NCA}}
\newcommand{\RMQ}{\textit{RMQ}}
\newcommand{\SA}{\textit{SA}}
\newcommand{\SAinv}{\textit{SA}^{-1}} % math
\newcommand{\SAi}{SA$^{-1}$} % no math
\newcommand{\LCP}{\textit{LCP}}
\newcommand{\suff}{\textit{suff}}
\newcommand{\logceil}{\lceil\log n\rceil}
\newcommand{\fprint}[1][k]{\ensuremath{\proc{Fingerprint}_{#1}}}
\newcommand{\fprintk}{\fprint[k]}
\newcommand{\RMQpq}[2]{RMQ\textless$#1$, $#2$\textgreater}
\newcommand{\RMQn}{\RMQpq{1}{n}}
\newcommand{\RMQq}{\RMQpq{n}{1}}
\newcommand{\RMQlog}{\RMQpq{n}{\log n}}

\tableofcontents

\vspace{1cm}
% stuff on toc page

\listoffixmes

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Blah blah blah\fixme{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The LCE Problem}

Given a string $s$ of length $n$ and a pair of indexes $1\leq i\leq n$ and $1\leq j\leq n$, the Longest Common Extension of $i$ and $j$ on $s$, $\LCE_s(i,j)$, is the length of the longest common prefix of $\suff_i$ and $\suff_j$. If for example $s = abbababba$, $i = 4$ and $j=6$, then $\suff_4 = ababba$ and $\suff_6 = abba$. The longest common prefix of these two suffixes is $ab$, which has length $2$, and therefore we have $\LCE_s(4,6) = 2$.

The LCE problem is to preprocess a string in order to allow for a large number of LCE queries on different pairs $(i,j)$, such that the queries are efficient.

Given a string length $n$ and an alphabet size $\sigma$, we define average case query time as the average of query times over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Existing Results\label{sec:existing-results}}

One way of solving the LCE problem without any preprocessing uses $O(|\LCE(i,j)|)$ query time. We call this algorithm \proc{DirectComp}. For a query $\LCE(i,j)$, the algorithm compares $s[i]$ with $s[j]$, then $s[i+1]$ with $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached. The worst case query time is $O(n)$ on a string of length $n$, and the average case query time is $O(1)$, because the average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}.

The LCE problem can be solved optimally with $O(1)$ worst case query time, using $O(n)$ space and $O(\sortt)$ preprocessing time. Two different ways of doing this exists.

One method, which we call \proc{SuffixNca}, uses constant time Nearest Common Ancestor queries on a suffix tree. The LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$. In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$), and no two child edge labels of the same node will have the same first character. The longest common prefix of the two suffixes will therefore be the path label from the root to the nearest common ancestor of $L_i$ and $L_j$. This gives $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.

The other method, which we call \proc{LcpRmq}, uses constant time Range Minimum Queries on a Longest Common Prefix array. The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA. The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e. $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$. % An example is shown in \pref{fig:sa+lcp+min}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Existing Practical Results}

Ilie~et~al.~\cite{ilie-navarro-tinta} have looked at a number of real world texts as well as texts of randomly generated characters, and found that all the texts they examined each has an average LCE of at most one, over all $n^2$ input pairs. Therefore it is interesting to have LCE algorithms, which perform good on average when the LCE value is small.

Both \proc{SuffixNca} and \proc{LcpRmq} have the same asymptotic space and times. In practice, \proc{LcpRmq} is the best of the two~\cite{ilie-navarro-tinta}. The constant factor for average case query time of \proc{DirectComp} is much smaller than the constant factor for query time of \proc{LcpRmq}, thus \proc{DirectComp} is better in practice on average case inputs than the theoretically best \proc{LcpRmq}.

\ifreport
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Our Fingerprinting Algorithm}
\fi % report
\ifarticle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Our Results}
\fi % article

We present a new LCE algorithm based on string fingerprinting. We call our algorithm \fprintk, where $k$ is a parameter $1\leq k\leq\logceil$, which describes the number of levels used\footnote{All logarithms are base two.}.

\begin{theorem}
For a string $s$ of length $n$ and alphabet size $\sigma$, the \fprintk\ algorithm, where $k$ is a parameter $1 \leq k \leq \logceil$, can solve the LCE problem in $O(k\cdot n^{1/k})$ worst case query time and $O(1)$ average case query time using $O(k\cdot n)$ space and $O(\sortt + k\cdot n)$ preprocessing time.
\end{theorem}

The exact worst case performance of our solution depends on the amount of space you are willing to use.

\begin{corollary}
\fprint[1] is equivalent to \proc{DirectComp} with $O(n)$ space and $O(n)$ query time.
\end{corollary}

\begin{corollary}
\fprint[2] has $O(n)$ space and $O(\sqrt n)$ query time. It has two levels, where one uses a table of fingerprints and the other uses the original string.
\end{corollary}

\begin{corollary}
\fprint[\logceil] has $O(n\cdot\log n)$ space and $O(\log n)$ query time. The data structure is equivalent to the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, and a query only needs to do one comparison at each level.
\end{corollary}

To preprocess the $O(k\cdot n)$ fingerprints used by our algorithm, we can use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, which takes $O(n\log n)$ time. For $k=o(\log n)$, we can speed up preprocessing to $O(\sortt + k\cdot n)$ by using the Suffix Array and Longest Common Prefix array.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.5\textwidth,page=1]{wc-avg.pdf}
    \end{center}
    \caption{\label{fig:wc-avg}Graphical comparison of query times for different LCE algorithms.}
\end{figure}

Existing state of the art solutions are either good in worst case, while poor in average case (\proc{LcpRmq}), or good in average case while poor in worst case (\proc{DirectComp}). Our \fprintk\ solution targets a worst case vs. average case query time tradeoff between these two extremes. \pref{fig:wc-avg} illustrates the relative query time performances of these solutions. Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is a lot faster than \proc{DirectComp} on a worst case input. Compared to \proc{LcpRmq}, our solution has a significantly better performance on an average case input, but its worst case performance is not as good as \proc{LcpRmq}. The space usage for \proc{LcpRmq} and \fprintk\ are approximately the same when $k=6$.

Our algorithm is fairly simple. Though it is slightly more complicated than \proc{DirectComp}, it does not use any of the advanced algorithmic techniques required by \proc{LcpRmq} and \proc{SuffixNca}.

\ifreport

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Our Table Lookup Algorithm}

\fixme{Our Table Lookup Algorithm}

\fi % report

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Preliminaries}

\paragraph{String notation.} Let $s$ be a string of length $n$. Then $s[i]$ is the $i$'th character of $s$, and $s[i\twodots j]$ is a substring of $s$ containing characters $s[i]$ to $s[j]$, both inclusive. That is, $s[1]$ is the first character of $s$, $s[n]$ is the last character, and $s[1\twodots n]$ is the entire string. The suffix of $s$ starting at index $i$ is written $\suff_i = s[i\twodots n]$.

\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers of $\sigma$ distinct values is written as $\sortt$.

\paragraph{Suffix tree.} A suffix tree $\mathcal{T}$ encodes all suffixes of a string $s$ of length $n$ with alphabet $\sigma$. The tree has $n$ leaves named $L_1$ to $L_n$, one for each suffix of $s$. Each edge is labeled with a substring of $s$, such that for any $1\leq i\leq n$, the concatenation of labels on edges on the path from the root to $L_i$ gives $\suff_i$. Any internal node must have more than one child, and the labels of two child edges must not share the same first character. The string depth $D[v]$ of a node $v$ is the length of the string formed when concatenating the edge labels on the path from the root to $v$.
The tree uses $O(n)$ space, and building it takes $O(\sortt)$ time~\cite{sort-complexity}.

\ifreport

There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes. Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be reprecented by two indexes of $s$. The suffix tree therefore takes $O(n)$ space.

\fi % report

\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet $\sigma$, the Suffix Array, \SA, is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$. The lexicographically smallest suffix is $\suff_{\SA[1]}$, the lexicographically largest suffix is $\suff_{\SA[n]}$, and the lexicographically $i$'th smallest suffix is $\suff_{\SA[i]}$. The inverse Suffix Array, \SAi, describes where a given suffix is in the lexicographical order. Suffix $\suff_i$ is the lexicographically $\SAinv[i]$'th smallest suffix.

The Longest Common Prefix Array, LCP, describes the length of longest common prefixes of neighboring suffixes in SA. $\LCP[i]$ is the length of the longest common prefix of $\suff_{\SA[i-1]}$ and $\suff_{\SA[i]}$ for $2 \leq i \leq n$. The first element $\LCP[1]$ is always zero.

Building the SA, \SAi and LCP arrays takes $O(\sortt)$ time~\cite{sort-complexity}.

\paragraph{NCA.} The ancestors of a node $u$ is $u$ itself and any node, which is a parent of an ancestor of $u$. The common ancestors of two nodes $u$ and $v$ is the intersection between the ancestors of $u$ and the ancestors of $v$. The Nearest Common Ancestor of two nodes $u$ and $v$ is the common ancestor of $u$ and $v$ of greatest depth. An NCA query can be answered in $O(1)$ time with $O(n)$ space and preprocessing time in a static tree with $n$ nodes~\cite{nca}.

\paragraph{RMQ.} The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, or $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$. A Range Minimum Query on a static array of $n$ elements can be answered in $O(1)$ time with $O(n)$ space and preprocessing time~\cite{jf-rmq}.

\ifreport

\paragraph{I/O's and Cache-Oblivious Model.}
The I/O model describes the number of memory blocks an algorithm moves between two layers of a layered memory architecture, where the size of the internal memory layer is $M$ words, and data is moved between internal and external memory in blocks of $B$ words. In the cache-oblivious model, the algorithm has no knowledge of the values of $M$ and $B$.

\paragraph{Predecessor.}
\fixme{Preliminaries: Predecessor}

\paragraph{Level Ancestor.}
\fixme{Preliminaries: Level Ancestor}

\paragraph{Trees.}
\fixme{Preliminaries: Trees}
depth root = 0

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LCE Algorithms}
This section describes existing and new algorithms for solving the LCE problem. It contains a theoretical analysis of their asymptotic query times and space requirements, shown in summary in \pref{tab:algorithms}. \pref{sec:results} describes the results from experiments on practical implementations of these algorithms.

\begin{table}[ht]
\centering
% \begin{tabular}{l|p{160px}|p{160px}}
\begin{tabular}{l|l|l|l}
\hline\hline
Algorithm & Space & Query time & Preprocessing \\ [0.5ex] \hline
\proc{SuffixNca} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
\proc{LcpRmq} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
\proc{DirectComp} & $O(1)$ & $O(n)$ & $O(1)$ \\ \hline
\proc{DirectCompLookup}, $1 \leq t \leq n$ & $O(n^2/t)$ & $O(t)$ & $O(n^2)$ \\
$t=1$ & $O(n^2)$ & $O(1)$ & $O(n^2)$ \\ \hline
\fprintk, $1 \leq k \leq \logceil$ & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
$k=\logceil$ & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline
\end{tabular}
\caption{LCE algorithms with their space requirements, worst case query times and preprocessing times. Average case query times are $O(1)$ for all shown algorithms.}\label{tab:algorithms}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Existing Algorithms}

In this section we look at the existing LCE algorithms described in \pref{sec:existing-results} with a bit more details.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\proc{DirectComp}}

\proc{DirectComp} solves the LCE problem without any preprocessing in $O(|\LCE(i,j)|)$ query time. For a query $\LCE(i,j)$, the algorithm compares $s[i+v]$ with $s[j+v]$ in a loop, for $v$ starting at zero and incrementing, until the two characters differ, or the end of the string is reached. \proc{DirectComp} does not need to check if the end of the string is reached, i.e. $i+v>n$ or $j+v>n$. Instead we can preprocess the string by adding a special character \$ to the end of the string, which does not occur anywhere else in the string. When the algorithm compares $s[n] = \$$ to any other character, it will stop, as the characters differ. This can give better practical performance compared to checking $i+v>n$ and $j+v>n$ in each iteration of the algorithm. The query time is $O(|\LCE(i,j)|)$, which in worst case is $O(n)$ on a string of length $n$, and in average case is $O(1)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\proc{SuffixNca}: NCA on a Suffix Tree}

LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$. In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$). The path label from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes. Since two child edges of the same node never have the same first character on their edge labels, the path label from the root to the $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but longest common prefix of $\suff_i$ and $\suff_j$. We can therefore find the LCE value as $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$, where $\mathcal{T}$ is the suffix tree for $s$, and $D[v]$ is the label depth of node $v$ in the tree.

For a string of length $n$ with aplhabet size $\sigma$, we can preprocess \proc{SuffixNca} in $O(\sortt)$ time by building the suffix tree, preprocessing it for NCA and building the array of label depths. The query time is $O(1)$, since looking up $L_i$ and $L_j$, doing an NCA query, and looking up the label depth in $D$ all takes constant time.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\proc{LcpRmq}: RMQs on a LCP Array}

The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA. The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e. $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$. An example is shown in \pref{fig:sa+lcp+min}. Preprocessing time is $O(\sortt)$, space requirement is $O(n)$ and query time is $O(1)$.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.2\textwidth,page=1]{sa+lcp+min.pdf}
    \end{center}
    \caption{\label{fig:sa+lcp+min}For $s=abbababba$, $\LCE(2,3)$ is the longest common prefix of $bababba$ and $bbababba$. There are two other suffixes between these in the lexicographical ordering, and the longest common prefixes between each pair of neighbors are 3, 1 and 3. Therefore $\LCE(2,3)= \min ~ \{3, 1, 3\} = 1$.}
\end{figure}

Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space. But like for the LCE problem, the asymptotically best RMQ solution may not be the best RMQ solution in practice. We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
\begin{samepage}
\begin{description}
\item[\RMQn] walks through the array directly in $O(n)$ time and does not use any preprocessing (except for transforming LCE into RMQ).
\item[\RMQq] preprocesses the array in $O(n)$ space in order to do queries in $O(1)$ time. This is a twolevel data structure as shown in \pref{fig:sa+lcp+min-twolevel}. The array is split into blocks of size $O(\log n)$. The top level is an array of length $O(n/\log n)$, where each element contains the minimum value one block. The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct seach through a $O(\log n)$ length block.
\end{description}
\end{samepage}

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.8\textwidth,page=2]{sa+lcp+min.pdf}
    \end{center}
    \caption{\label{fig:sa+lcp+min-twolevel}Twolevel Range Minimum Query. The range minimum is calculated as the minimum of one query in the top level and two queries in the lower level.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Combining \proc{DirectComp} and \proc{LcpRmq}\label{sec:dc-lcprmq}}

Ilie~et~al.~\cite{ilie-navarro-tinta} suggests combining \proc{DirectComp} and \RMQn, by using \RMQn\ whenever the two suffixes we want to compare are close to each other in SA, and otherwise use \proc{DirectComp}. However that still gives $O(n)$ worst case query time. For example in the string $s=aaaaa...$, the query $\LCE(1,n/2)$ would require $\Omega(n)$ time for both \proc{DirectComp} and \RMQn.

We can instead combine \proc{DirectComp} and \RMQq, which we call \proc{DirectComp+LcpRmq}. \proc{DirectComp+LcpRmq} first uses \proc{DirectComp} until a specific cutoff and then switches over to \RMQq\ instead. This could have both the good average case query time of \proc{DirectComp} and the good worst case query time of \proc{LcpRmq}. However it still keeps the large space requirements of \proc{LcpRmq}, and it is not simple to implement like \proc{DirectComp}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The \proc{DirectCompLookup} Algorithm}

The input to the $LCE$ problem is two numbers $1 \leq i < j \leq n$ \footnote{We can assume that $i< j$ without loss of generality, since $\LCE(i,j)=\LCE(j,i)$}, so there are $n*(n-1)/2 = O(n^2)$ different input possibilities in total. Because of this limited space of possible query inputs, we can tabulate a number of LCE values, and use those to speed up queries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\proc{DirectLookup}}

We can store results for all $O(n^2)$ possible queries in $O(n^2)$ space, and a query will then be a single table lookup in $O(1)$ time. We call this algorithm \proc{DirectLookup}. A result space table for \proc{DirectLookup} is shown in \pref{fig:resultspace-empty}, and \pref{fig:resultspace-directcomp} shows how \proc{DirectComp} moves through that table.

We can preprocess the table in in $O(n^2)$ time, since each value can be calculated in constant time like follows, if $LCE(i+1,j+1)$ is always calculated before $LCE(i,j)$ for any $i$ and $j$:
\begin{align*}
LCE(i,j) &=
\begin{cases}
    0 & \textrm{if} ~ s[i] \neq s[j] \\
    1 & \textrm{if} ~ s[i] = s[j] \wedge (i = n \lor j = n) \\
    LCE(i+1, j+1) + 1 & \textrm{otherwise} \\
\end{cases}
\end{align*}

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.5\textwidth,page=1]{resultspace.pdf}
    \end{center}
    \caption{\label{fig:resultspace-empty}The result space for $LCE_s(i,j)$ for $1 \leq i < j \leq n$ where $s = abacbbacbacacabc$.}
\end{figure}

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.5\textwidth,page=2]{resultspace.pdf}
    \end{center}
    \caption{\label{fig:resultspace-directcomp}$LCE(3,7)$ using \proc{DirectComp}.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\proc{DirectComp} and \proc{DirectLookup} Combined}

\proc{DirectComp} and \proc{DirectLookup} preprocesses and stores no results and all results respectively. By only storing some results, we can use \proc{DirectComp} until we reach a stored result, and then retrieve this result using \proc{DirectLookup}. If we store every $t$'th row, as shown in \pref{fig:resultspace-t-row}, the space usage will be $O(n^2/t)$ and the worst case query time will be $O(t)$. We call this algorithm \proc{DirectCompLookup}.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.5\textwidth,page=3]{resultspace.pdf}
    \end{center}
    \caption{\label{fig:resultspace-t-row}$LCE(3,7)$ with distance $t=4$ between each stored row.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{The \fprintk\ Algorithm}

\fi % report

\ifarticle
\section{The \fprintk\ Algorithm}
\fi % article

Our \fprintk\ algorithm generalizes \proc{DirectComp}. It compares characters starting at positions $i$ and $j$, but instead of comparing individual characters, it compares fingerprints of substrings. Given fingerprints of all substrings of length $t$, our algorithm can compare two $t$-length substrings in constant time.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Definition of a Fingerprint}

Given a string $s$, the fingerprint $F_t[i]$ is a natural number identifying the substring $s[i\twodots i+t]$ among all $t$-length substrings of $s$. We assign fingerprints such that for any $i$, $j$ and $t$, $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t] = s[j\twodots j+t]$. In other words, if two substrings of $s$ have the same length, they have the same fingerprints if and only if the substrings themselves are the same.

At the end of a string when $i+t>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Data Structure\label{sec:fingerprint-ds}}

The \fprintk\ data structure for a string $s$ of length $n$, where $k$ is a parameter $1 \leq k \leq \logceil$, consists of $k$ natural numbers $t_0$, ..., $t_{k-1}$ and $k$ tables $H_0$, ..., $H_{k-1}$, each of length $n$. For each $\ell$ where $0\leq \ell\leq k-1$, $t_\ell = \Theta(n^{\ell/k})$ and table $H_\ell$ contains fingerprints of all $t_\ell$-length substrings of $s$, such that $H_\ell[i] = F_{t_\ell}[i]$. We always have $t_0 = n^{0/k} = 1$, such that $H_0$ is the original string $s$. The last character of the string must be a special character \$, which does not occur anywhere else in the string. An example is shown in \pref{fig:fingerprint-ds}.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.98\textwidth,page=1]{fingerprint.pdf}
    \end{center}
    \caption{\label{fig:fingerprint-ds}\fprintk\ data structure for $s=\textit{abbaabbababbaabbababaababa\$}$, $n=27$, $k=3$, $t_1=n^{1/3}=3$ and $t_2=n^{2/3}=9$. All substrings $\textit{bba}$ are highlighted with their $3$-length fingerprint $2$.}
\end{figure}

\begin{lemma}
The \fprintk\ data structure takes $O(k\cdot n)$ space.
\end{lemma}
\begin{proof}
Each of the $k$ tables stores $n$ fingerprints of constant size.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Query\label{sec:fingerprint-query}}

To perform an LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
\begin{samepage}
\begin{enumerate}
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
\item When $H_\ell[i+v] \neq H_\ell[j+v]$, decrement $\ell$ by one and repeat step 2.
\item Stop when $\ell = 0$ and $H_\ell[i+v] \neq H_\ell[j+v]$.
\end{enumerate}

The LCE value is now $v = \LCE(i,j)$. An example is shown in \pref{fig:fingerprint-query}.
\end{samepage}

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.98\textwidth,page=2]{fingerprint.pdf}
    \end{center}
    \caption{\label{fig:fingerprint-query}\fprintk\ query for $\LCE(3,12)$ on the data structure of \pref{fig:fingerprint-ds}. The top half shows how $H_\ell[i+v]$ moves through the data structure, and the bottom half shows $H_\ell[j+v]$.}
\end{figure}

\begin{lemma}
The \fprintk\ query algorithm given above finds the correct LCE value.
\end{lemma}
\begin{proof}
At each step of the algorithm $v \leq \LCE(i,j)$, since the algorithm only increments $v$ by $t_\ell$ when it has found two matching fingerprints, and fingerprints of two substrings of the same length are only equal if the substrings themselves are equal. When the algorithm stops, it has found two fingerprints, which are not equal, and the length of these substrings is $t_\ell = 1$, therefore $v = \LCE(i,j)$.

The algorithm never reads $H_\ell[x]$, where $x>n$, because the string contains a unique character \$ at the end. This character will be at different positions in the substrings whose fingerprints are the last $t_\ell$ elements of $H_\ell$. These $t_\ell$ fingerprints will therefore be unique, and the algorithm will not continue at level $\ell$ after reading one of them.
\end{proof}

\begin{lemma}
The worst case query time for \fprintk\ is $O(k\cdot n^{1/k})$, and the average case query time is $O(1)$.
\end{lemma}
\begin{proof}
Step 1 takes $O(k)$ time. In step 2 and 3, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$). Since we can check $t_\ell = \Theta(n^{\ell/k})$ characters in constant time at level $\ell$, the algorithm uses $O(n^{(\ell+1)/k})/\Theta(n^{\ell/k}) = O(n^{1/k})$ time at that level. Over all $k$ levels, $O(k\cdot n^{1/k})$ query time is used.

At each step except step 3, the algorithm increments $v$. Step 3 is executed the same number of times as step 1, in which $v$ is incremented. The query time is therefore linear to the number of times $v$ is incremented, and it is thereby $O(v)$. The query time is thus $O(1)$ in average case, where $v=O(1)$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Average Case Optimization\label{sec:improved-avg}}

We could have left out step 1 of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$, which would change the query pattern as shown in \pref{fig:fingerprint-updown}. This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, but it would increase our average case query time to $O(k)$. Our experiments have shown that whenever $k$ is small, keeping step 1 improves average case query time, while it does not have a measurable effect on worst case query times. When $k = \logceil$, keeping step 1 can double the worst case query time, while it can make the average case query time 40 times faster for an input string of ten million characters. We want to optimize our LCE query time for the average case where the LCE value is small, so our results in \pref{sec:results} does not include the variant of \fprintk\ with $O(k)$ average case query time.

In step 1 of our query algorithm we perform one comparison at each level. We could instead do up to $O(n^{1/k})$ comparisons at each level without affecting asymptotic times. Our experiments have shown that always doing one comparison at each level is the best in practice.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=1\textwidth,page=1]{fingerprint-updown.pdf}
    \end{center}
    \caption{\label{fig:fingerprint-updown}Two ways of querying the fingerprint levels, one starting from level $\ell=k-1$ and one starting at level $\ell=0$.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Preprocessing}

The tables of fingerprints use $O(k\cdot n)$ space. In the case with $k=\logceil$ levels, the data structure is the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}. This data structure can be constructed in $O(n\log n)$ time. With $k<\logceil$ levels, KMR can be adapted, but it still uses $O(n\log n)$ preprocessing time.

We can preprocess the data structure in $O(\sortt + k\cdot n)$ time using the SA and LCP arrays. First create the SA and LCP arrays. Then preprocess each of the $k$ levels using the following steps:
\begin{samepage}
\begin{enumerate}
\item Loop through the $n$ substrings of length $t_\ell$ in lexicographically sorted order by looping through the elements of SA.
\item Assign an arbitrary fingerprint to the first substring.
\item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell]$ is equal to the substring examined in the previous iteration of the loop, give the current substring the same fingerprint as the previous substring, otherwise give the current substring a new unused fingerprint. The two substrings are equal when $\LCE[i] \geq t_\ell$.
\end{enumerate}

An example is shown in \pref{fig:fingerprint-preproc}.
\end{samepage}

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.2\textwidth,page=1]{fingerprint-preproc.pdf}
    \end{center}
    \caption{\label{fig:fingerprint-preproc}The first column lists all substrings of $s=abbababba$ with length $t_\ell = 3$. The second column lists fingerprints assigned to each substring. The third column lists the position of each substring in $s$.}
\end{figure}

\begin{lemma}
The preprocessing algorithm described above generates the data structure described in \pref{sec:fingerprint-ds}.
\end{lemma}

\begin{proof}
We always assign two different fingerprints whenever two substrings are different, because whenever we see two differing substrings, we change the fingerprint to a value not previously assigned to any substring.

We always assign the same fingerprint whenever two substrings are equal, because all substrings, which are equal, are grouped next to each other, when we loop through them in lexicographical order.
\end{proof}

\begin{lemma}
The preprocessing algorithm described above takes $O(\sortt + k\cdot n)$ time.
\end{lemma}

\begin{proof}
We first construct the SA and LCP arrays, which takes $O(\sortt)$ time~\cite{sort-complexity}. We then preprocess each of the $k$ levels in $O(n)$ time, since we loop through $n$ substrings, and comparing neighboring substrings takes constant time when we use the LCE array. The total preprocessing time becomes $O(\sortt + k\cdot n)$.
\end{proof}

\ifreport

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Cache Optimization\label{sec:fingerprint-cache}}

The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$. However if we structure our tables of fingerprints differently, we can improve ths number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model. Instead of storing the fingerprint of $s[i\twodots i+t_\ell]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$. This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.

The size of each fingerprint table will grow from $|H_\ell| = n$ to $|H_\ell| = n+t_\ell$, because the rounding operations may introduce one-element gaps in the table after every $n/t_\ell$ elements. We achieve the greatest I/O improvement when $k$ is small. When $k=\logceil$, this cache optimization gives no difference in the amount of I/O's.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Space Usage}

\pref{tab:algorithms-space} shows the space requirement of each of our analyzed algorithms. The size of the input string of $n/4$ words is not included, and we ignore constant terms.

\begin{table}[ht]
\centering
\begin{minipage}{0.7\textwidth}
\centering
\begin{tabular}{l|l}
\hline\hline
Algorithm & Space \\ [0.5ex] \hline
\proc{DirectComp} & $1$ words \\ \hline
\proc{DirectLookup} & $2n^2$ words \footnote{The factor two comes from using shift instead of multiplication, which may double the size depending on $n$.} \\ \hline
\proc{DirectCompLookup} & $2n^2/t+n$ words \\ \hline
\proc{LcpRmq} & ca. $5.6n$ words \footnote{Space use for \proc{LcpRmq} is $2n + 2nb + (bs \cdot (bs+1) / 2 + 1)\cdot C_{bs} + (nb+1)\cdot \lfloor\log nb\rfloor \approx 5.6n$, where $bs = \lceil\log n / 4\rceil$, $nb = (n-1)/bs+1$ and $C_i$ is the $i$'t Catalan Number.} \\ \hline
\fprint[2] & $n$ words \\ \hline
\fprint[3] & $2n$ words \\ \hline
\fprint[\logceil] & $n\cdot\logceil$ words \\ \hline
\end{tabular}
\end{minipage}
\caption{Space requirements for the LCE algorithms, where $n$ is the number of characters, one character is byte and one word is four bytes.}\label{tab:algorithms-space}
\end{table}

\fi % report

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Experimental Results\label{sec:results}}

In this section we show results of actual performance measurements. The measurements were done on a Windows 23-bit machine with an Intel P8600 CPU (3 MB L2, 2.4 GHz) and 4 GB RAM. The code was compiled using GCC 4.5.0 with \texttt{-O3}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tested Algorithms}

\ifarticle

We implemented different variants of the \fprintk\ algorithm in C++ and compared them with optimized versions of the \proc{DirectComp} and \proc{LcpRmq} algorithms. The algorithms we compared are the following:
\begin{description}
\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$. The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater describe the exact size of fingerprinted substrings at each level.
\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
\end{description}

\fi % article

\ifreport

We implemented different variants \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other. We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}. The algorithms we compared are the following:
\begin{description}
\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$. The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater describe the exact size of fingerprinted substrings at each level.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater bc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_1$, ..., $u_{k-1}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
\item[\proc{RMQ}\textless$p$, $q$\textgreater$_{\textrm{virtual}}$] is the \proc{LcpRmq} algorithm using different RMQ algorithms. They all contain an additional virtual method call to make implementation simpler, but their results cannot be directly compared to other algorithms, which do not have this extra virtual method call.
\item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.

\end{description}

\fi % report

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Test Inputs and Setup}

We have tested the algorithms on different kinds of strings:
\begin{description}
\item[Average case strings] These strings have many small LCE values, such that the average LCE value over all $n^2$ query pairs is less than one. We use results on these strings as an indication average case query times over all input pairs $(i,j)$ in cases where most or all LCE values are small on expected input strings. We construct these strings by choosing each character uniformly at random from an alphabet of size 10
\item[Worst case strings] These strings have many large LCE values, such that the average LCE value over all $n^2$ query pairs is $n/2$. We use results on these strings as an indication of worst case query times, since the query times for all tested algorithms are asymptotically at their worst when the LCE value is large. We construct these strings with an alphabet size of one.
\item[Medium LCE value strings] These strings have an average LCE value over all $n^2$ query pairs of $n/2r$, where $r=0.73n^{0.42}$. We use results on these strings as an indication of query times somewhere between the average case and worst case. We construct these strings by repeating a substring of $r$ characters, where each character is unique.
\end{description}

For each kind of strings, we tested the algorithms using the following pattern:
\begin{enumerate}
\item Generate a string of length $n$ and generate a million random pairs $(i, j)$, where each of $i$ and $j$ is chosen between $1$ and $n$ uniformly at random.
\item For each tested algorithm, preprocess the given string, and run a query for each of the million pairs. Measure the time it takes to run all the queries combined.
\item Double the value of $n$ and repeat from step 1.
\end{enumerate}

\ifreport

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\proc{LcpRmq} Results\label{sec:test-rmq}}

We want to compare our new algortithms against the algortihm which is the best in practice of the asymptotically best $O(1)$ algorithms. The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at that one. We have looked at a number of RMQ algorithms to find out which works best for the LCE problem. The results are shown in \pref{fig:plot-rmq}. \RMQq\ is the best in practice, and we therefore use that when we test against other LCE algorithms. This result is different from that of Ilie~et~al.~\cite{ilie-navarro-tinta}, where \RMQpq{n}{4\log n} is the best. This could be due to hardware or compiler differences.

The performance of the implementations shown in \pref{fig:plot-rmq} cannot be compared with those shown on other figures, because they all contain an additional, slow virtual method call, which gives them an unfair disadvantage. We have optimized the best variant of \proc{LcpRmq} to avoid the virtual method call, when we compare it against other algorithms. We compare this optimized version to \proc{DirectComp} in \pref{fig:plot-rmq-directcomp}, which shows that \proc{DirectComp} is significantly better in average case, though both have constant asymptotic average case query times.

The RMQ algorithms show a significant jump in query times around $n=1.000.000$ on the plot with average case strings, but not on the plot with worst case strings. We have run the tests in Cachegrind, and found that the number of instructions executed and the number of data reads and writes are exactly the same for both average case strings and worst case strings. The cache miss rate for \RMQq\ on average case strings is 14\% and 9\% for the L1 and L2 caches, and on worst case strings the miss rate is 17\% and 13\%, which is the opposite of what could explain the jump we see in the plot.

% size = 3,5*2^30 = 3758096384
% assoc = (3,5*2^30)/(4*2^10) = 917504
% line = 4*2^10 = 4096
% --LL=3758096384,917504,4096

% size = 1*2^30 = 1073741824
% assoc = (1*2^30)/(4*2^10) = 262144
% line = 4*2^10 = 4096
% --LL=1073741824,262144,4096

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-alla.plt}
    \end{center}
    \caption{\label{fig:plot-rmq}Query times of \proc{LcpRmq} using different variants of RMQ.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{\proc{DirectComp+LcpRmq} Results}

\pref{fig:plot-rmq-directcomp} also shows \proc{DirectComp+LcpRmq}, where we first try \proc{DirectComp}, and if we have not found the LCE value after one iteration, we use \proc{LcpRmq} instead. We could increase the cutoff from one iteration to something bigger, which might improve average case performance. That would however decrease worst case performance. None of the two plots we measure represents a good worst case for \proc{DirectComp+LcpRmq} compared to the other algorithms we test. The worst case for \proc{DirectComp+LcpRmq} relative to other the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithm still work on LCE values as small as possible. As described in \pref{sec:dc-lcprmq}, \proc{DirectComp+LcpRmq} also uses a lot of space.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-alla.plt}
    \end{center}
    \caption{\label{fig:plot-rmq-directcomp}Query times of constant time \proc{LcpRmq}, \proc{DirectComp} and \proc{DirectComp+LcpRmq}.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\proc{DirectCompLookup} Results}

Figure \pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs $t$ together with \proc{LcpRmq}. In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra ckeck to find out if it has reached a position for which a result is stored. For short strings, \proc{DirectLookup} is faster than \proc{DirectCompLookup}, since it does not need to check if the value it needs is stored. However, it quickly grows to be as slow as \proc{LcpRmq}, while \proc{DirectCompLookup} does not experience this slowdown. We explain this with that \proc{DirectLookup} reads from its $O(n^2)$ table of results in every query, whereas \proc{DirectCompLookup} only reads from the $O(n)$ string in average case, and thus has a significantly smaller working set.

In the worst case, the \proc{DirectCompLookup} variants which use more space are faster in practice, just like in theory. If we are willing to use $O(n\log n)$ space, \proc{DirectCompLookup}\textless$n/\log n$\textgreater\ will only halve the worst case query time compared to \proc{DirectComp}. This makes \proc{DirectCompLookup} better than \proc{DirectComp}, but still not a very good candidate when worst case query time is important.

\proc{DirectCompLookup} uses $O(n^2)$ preprocessing time no matter what parameter $t$ we choose, which is significantly slower than any of the other algorithms we have tested, and it is also the reason why we have only tested this algorithm up until $n=100.000$. The only way we have found to improve this preprocessing time is to use one of the other algorithms we have covered to preprocess the data structure of \proc{DirectCompLookup}.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-alla.plt}
    \end{center}
    \caption{\label{fig:plot-complookup}Query times of our new DirectCompLookup algorithm versus the existing \proc{DirectComp} and \proc{LcpRmq} algorithms.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\fprintk\ Results}

\fi % report

\ifarticle

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Results}

\fi % article

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.7\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-rand10.plt}
        \includegraphics[width=0.7\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-alla.plt}
        \includegraphics[width=0.7\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-repeat-pow.plt}
    \end{center}
    \caption{\label{fig:test-fingerprint}Comparison of our new \fprintk\ algorithm for $k=2$, $k=3$ and $k=\logceil$ versus the existing \proc{DirectComp} and \proc{LcpRmq} algorithms.}
\end{figure}

\pref{fig:test-fingerprint} shows our experimental results on average case strings with a small average LCE value, worst case strings with a large average LCE value, and strings with a medium average LCE value.

On average case strings, our new \fprintk\ algorithm is approximately 20\% slower than \proc{DirectComp}, and it is more than five times faster than \proc{LcpRmq}.

On worst case strings, the \fprintk\ algorithms are signicicantly better than \proc{DirectComp} and somewhat worse than \proc{LcpRmq}. Up until $n=30.000$ the three measured \fprintk\ algorithms have nearly the same query times. Of the \fprintk\ algorithms, the $k=2$ variant has a slight advantage for small strings of length less than around $2.000$. For longer strings the $k=3$ variant performs the best up to strings of length $250.000$, at which point the $k=\logceil$ variant becomes the best. This indicates that for shorter strings, it is best to use fewer levels, and when the input size increases, the \fprintk\ variants with better asymptotic query times have better worst case times in practice.

On strings with medium aveage LCE values, we see that our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}

We conclude that our new \fprintk\ algorithm acieves a tradeoff between worst case times and average case times, which is better than the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, yet it is not strictly better than the existing algorithms on all inputs. \fprintk\ is therefore a good choice in cases where both average case and worst case performance is important.

%  \fprint[3]\ also shows such a jump, but for greater input sizes, as it uses less memory.

\ifarticle

\proc{LcpRmq} shows a significant jump in query times around $n=1.000.000$ on the plot with average case strings, but not on the plot with worst case strings. We have run the tests in Cachegrind, and found that the number of instructions executed and the number of data reads and writes are exactly the same for both average case strings and worst case strings. The cache miss rate for average case strings is 14\% and 9\% for the L1 and L2 caches, and for worst case strings the miss rate is 17\% and 13\%, which is the opposite of what could explain the jump we see in the plot.

\fi % article

\fixme{Investigate if we can explaing the jump in Fingerprint-log}
% \paragraph{\fprint[\log n] cache limit.}
% \proc{LcpRmq} uses more space than all of \proc{DirectComp}, \fprint[2] and \fprint[3], so it is expected that it reaches the cache limit before these. However \fprint[\log n] uses more space than \proc{LcpRmq} at the input sizes we measure, since \proc{LcpRmq} uses the same amount of memory as \fprint[6], and $\lceil\log(10.000.000)\rceil = 24$. However, the \fprintk\ algorithms only access the lowest levels of their data structure whenever the LCE values are small, and there is therefore no visible jump due to cache limits for \fprint[\log n] on the plot with average case strings. On the plot with worst case strings however, we see a jump in \fprint[\log n] between $n=20.000$ and $n=50.000$, since it uses all its $\logceil$ levels when the LCE values are large.

\ifreport

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Optimal Value of $t_\ell$}

\pref{fig:plot-fingerprint-tval} shows \fprint[2] for different values of $t_1$. On average case strings the value of $t_1$ makes no difference, and on worst case strings a value of $t_1$ between $2\sqrt n$ and $4\sqrt n$ is the best. The improvement relative to $\sqrt n$ is however insignificant, so we have chosen to use $t_1=\sqrt n$.

Our results indicate that reading values from $H_0$ is faster than reading values from $H_1$. Differences between these two tables include that elements of $H_0 = s$ use one byte, whereas elements of $H_1$ use four bytes, and that elements we read from $H_0$ are located next to each other in memory, whereas elements we read from $H_1$ are spaced $t_1$ elements apart in memory.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-2l-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-2l-alla.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-tval}Query times of \fprint[2] for different values of $t_1$ versus query times for \proc{DirectComp}.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Average Case Optimization\label{sec:test-average}}

In \pref{sec:improved-avg} we discussed how our \fprintk\ algorithm is optimized for the average case to get $O(1)$ instead of $O(k)$ average case query time, while it keeps the same asymptotic worst case query time. But in practice the worst case query time is slightly affected.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avglog-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avglog-alla.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-avglog}Query times of \fprint[\logceil] with $O(1)$ versus $O(k)$ average case query time.}
\end{figure}

In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the aveage case optimization. Both variations of \fprint[\logceil] show query times, whic are $O(\log n)$ as expected. On average case strings, the query time of \fprint[\logceil] improves significantly when using average case optimization, and both variants show their expected $O(1)$ and $O(\log n)$ average case query times.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avg3l-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avg3l-alla.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-avg3l}Query times of \fprint[3] with $O(1)$ versus $O(k)$ average case query time.}
\end{figure}

In \pref{fig:plot-fingerprint-avg3l} we see that doing one comparison at each level on the way up is better than doing $n^{1/k}$ comparisons. This is most visible on worst case strings, where the algorithm doing one comparison at each level uses less time to reach level $\ell=k-1$, which it will reach anyway. On average case strings, the code specialized in handling only one comparison at each level is simpler. On the question of doing average case optimization or not, we see the same results for \fprint[3], as we saw for \fprint[\logceil], but at a smaller scale.

We conclude that using average case optimization yields a query time improvement on average case strings, which is significantly greater than the query time degration on worst case strings, and that \fprintk\ therefore should use the average case optimization.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Cache Optimization}

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand10.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-alla.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-repeat-pow.plt}
        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand2.plt}
    \end{center}
    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without cache optimization.}
\end{figure}

\pref{fig:plot-fingerprint-cache-horiz} shows results of the cache optimization described in \pref{sec:fingerprint-cache}. We have implemented two cache optimized variants. One as described in \pref{sec:fingerprint-cache}, and one where multiplication and division is replaced with shift operations. To use shift operations, $t_\ell$ and $\lceil n/t_\ell\rceil$ must both be powers of two. This may double the size of the used address space.

On average case strings the cache optimization does not change the query times, while on worst case strings and strings with medium size LCE values, cache optimization gives a noticable improvement for large inputs. The cache optimized \fprint[3] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.

The last plot on \pref{fig:plot-fingerprint-cache-horiz} shows a variant of average case where the alphabet size is changed to two. This plot attempts to show the worst case for cache optimized \fprint[3]. LCE values in this plot are large enough to ensure that $H_1$ is used often, which should make the extra complexity of calculating indexes into $H_1$ visible. At the same time the LCE values are small enough to ensure, that the cache optimization has no effect. In this plot we see that the cache optimized variant of \fprint[3] has only slightly worse query time compared to the variant, which is not cache optimized.

We conclude that cache optimization does not visibly affect average case query times, while it improves worst case query times. However, due to late timing of this discovery, the cache optimized variant of \fprintk\ is not the main focus of our analysis.

\fixme{List these algorithms in the section on tested algorithms?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Conclutions on Experimental Results}

\fixme{Conclutions on Experimental Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Test Code}

Skal vi antage at $i < j$ eller ej, eller skal begge dele testes?

Skal vi antage at strengen ender i et specielt tegn eller ej, eller skal begge dele testes?

Hvor meget skal et tegn fylde? Altid 8 bit som i Tinta-artiklen eller varieres?

Er valg af random-generator vigtig?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Structure}

To test the practical performance of the algorithms, they are implemented in C++. The implementation is split into three parts. \pref{tab:test-variables} summarizes where the different variables for testing are implemented.

\begin{table}[ht]
\centering
\begin{tabular}{l l l}
Variable & Example & Component \\ \hline
alphabet size & $\sigma = 10$ & string \\
character pattern & uniformly at random & string \\ \hline
algorithm & Perfect Hashing & algorithm \\
alg. parameter & $t = 2\sqrt n$ & algorithm \\ \hline
string length & exponentially increasing from $1$ to $100.000$ & test \\
query pattern & uniformly at random & test \\ \hline
\end{tabular}
\caption{List of performance veriables tested.}\label{tab:test-variables}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Also tell about c++ bug hunting in tinta, cachegrind code}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Also tell about attempting by-LCE-value plot and why it fails}

\begin{algorithm}
\caption{Algorithm for plotting query time as a function of string length\label{alg:graph-length}}
\begin{codebox}
\li \For $\id{type} \in \{\textrm{string of random characters with}~\sigma=10,~\textrm{string with}~\sigma = 1\}$
    \Indentmore
\li     \For $n = 1$ \To $1e7$ \Comment Double $n$ at each iteration
        \Indentmore
\li         $s = \proc{MakeString}_{\id{type}}(n)$
\li         $\id{data} =$ a sequence of a million pairs $(i,j)$ where
            \Indentmore
\zi             $i\in[1\twodots n]$ and $j\in[1\twodots n]$ are chosen
\zi             uniformly at random.
            \End
\li         \For each $\id{alg} \in \id{Algorithms}$
            \Indentmore
\li             $\id{alg}.\proc{preprocess}(s)$
\li             start timing
\li             \For each $(i,j) \in \id{data}$
                \Indentmore
\li                 $\id{alg}.\proc{query}(i,j)$
                \End
\li             end timing
\li             \proc{Plot}$_{\id{type}}$($alg$, $n$, timing)
\li             $\id{alg}.\proc{cleanup}()$
            \End
        \End
    \End
\end{codebox}
\end{algorithm}

\begin{algorithm}
\caption{Algorithm for plotting query time as a function of LCE value\label{alg:graph-value}}
\begin{codebox}
\li $s = \proc{MakeString}(1000)$
\li \id{data} = an array where $\id{data}[v] \subseteq \{(i,j)~|~\LCE(i,j)=v\} \land |\id{data}[v]| =10.000$
\li \For each $\id{alg} \in \id{Algorithms}$
    \Indentmore
\li     $\id{alg}.\proc{preprocess}(s)$
\li     \For $v = 0$ \To $n$
        \Indentmore
\li         \Repeat 3 times
% \li     $A =$ generate 1000000 random pairs $(i,j)$ where $i<j$
\li             start timing
\li             \For each $(i,j) \in data[v]$
                \Indentmore
\li                 $\id{alg}.\proc{query}(i,j)$
                \End
\li             end timing
            \End
\li         \proc{Plot}($alg$, $v$, median of the three timings)
        \End
\li     $\id{alg}.\proc{cleanup}()$
    \End
\end{codebox}
\end{algorithm}

Where $\proc{Plot}(f,x,y)$ adds point $(x,y)$ to series $f$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Ziv-Lempel compression}

\newcommand{\zreference}{\textit{reference}}
\newcommand{\zlabel}{\textit{label}}
\newcommand{\zphrase}{\textit{phrase}}

Sometimes it can be useful to store a string in a compressed format and query the compressed string directly without using extra space and time to first decompress the string. In this section we explore the LCE problem on strings compressed using LZ78 compression. We look at solutions, which require space proportional to the compressed size of a string, and are able to answer LCE queries fast.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LZ78 definition\label{sec:lz-definition}}
A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$. Each compression element $z_x$ is a pair $(r_x, \alpha_x)$, where the reference, $\zreference_z(x)=r_x$, is an index of an earlier compression element in the compressed string or zero, i.e. $0 \leq \zreference_z(x) < x$, and the label, $\zlabel_z(x)=\alpha_z$, is a character from the original uncompressed string $s$. The \zphrase\ of a compression element is defined as:
\begin{align*}
\zphrase_z(x) =
\begin{cases}
    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference \neq 0\\
    \zlabel_z(x) & \textrm{otherwise}
\end{cases}
\end{align*}
where "$\cdot$" denotes string contatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.

We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions. To get the compressed string $z$ from the original string $s$ we can look at the compression elements as a tree, where the reference of each compression element corresponds to the parent pointer of each node in the tree. We consume $s$ from left to right while we build the tree:
\begin{enumerate}
\item Let the tree be single special root node $z_0$, let $i = 1$ and $c=z_0$.
\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create the child node $u$, set $\zlabel(u)=s[i]$ and set $c=z_0$.
\item Increment $i$ by one and repeat from step 2 until the entire string is consumed when $i = n$.
\end{enumerate}
The zompressed string $z$ is now the tree except the special root node $z_0$.

The size of the compressed string is at most $n=O(N)$ and at least $n=\Omega(\sqrt{N})$.
% N= 1+2+3+4+...+n=(n^2+n)/2

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Compressed \proc{DirectComp}\label{sec:lz-iterate-forward}}

We can adapt \proc{DirectComp} to work on a compressed string. To work directly on a compressed string, we first need to have ramdom access to the characters of the original string.

\begin{lemma}
\label{lem:lz-random-access}
Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$. For any $1 \leq i \leq N$, we can find $s[i]$ in $O(\log\log N)$ query time using $O(n)$ space.
\end{lemma}

\begin{proof}
To find the character $s[i]$, we can first use a Predecessor query to find out which compression element $s[i]$ is in. We can then use a Level Ancestor query to find the correct character from the found compression element.

% We can find the $i$'th character of $s$ as $s[i] = \zphrase(\proc{Pred}(i))[i - \textit{start}[\proc{Pred}(i)]]$ where $\proc{Pred}(i)$ is the index of the compression element containing $s[i]$, and $\textit{start}[x]$ is the position in $s$ of the first character of $\zphrase(x)$, i.e. $s[\textit{start}[x]] = \zphrase(x)[1]$.

% We can find \proc{Pred}(i) with a predecessor query over predecessor datastructure of the $n$ compression elemets, where each compression element $x$ is added using $\textit{start}[x] as its key, and the universe has size N. When we look at the compression elements as a tree, we have $\zlabel(x)[i]=\proc{Level-Ancestor}(i,x)$. We therefore have random access to che characters of a compressed string using one Predecessor query and one Level Ancestor query.

The query time for a random access query is $O(\log\log N)$, since we use $O(\log\log N)$ time on the Predecessor query and $O(1)$ time on the Level-Ancestor query. The space needed to support random access queries is $O(n)$, since both the predecessor data structure and the level ancestor data structure use $O(n)$ space.
\end{proof}

\begin{theorem}
Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$. Compressed \proc{DirectComp} can find $\LCE(i,j)$ in $O(\log\log N + |\LCE(i,j)|)$ query time using $O(n)$ space.
\end{theorem}

\begin{proof}
Whenever \proc{DirectComp} reads a character of the string $s$, we instead use the random access described in \pref{lem:lz-random-access} to read the character from the compressed string. The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$. In subsequent iterations of \proc{DirectComp} we know that we always need the same compression element or the following compression element as we used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the compression element. The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Reversed Compressed \proc{DirectComp}}

A simpler way of doing LCE on compressed a string is to reverse the string and then run \proc{DirectComp} in the reverse direction on the reversed string, which gives the same result since reversing a string twice gives the original string. Where \proc{DirectComp} compares $s[i+v]$ against $s[j+v]$ in each iteration of its loop, reversed \proc{DirectComp} compares $s[N+1-i-v]$ against $s[N+1-j-v]$ in each iteration. This is simpler than using \proc{DirectComp} directly on the compressed string, because Level Ancestor queries are no longer needed at each iteration, since we can use the reference of the current compression element as follows, where the position of each of $i$ and $j$ is stored as a pair $(x,y)$:
\begin{align*}
\proc{PreviousPosition}(x, y)=&
\begin{cases}
(\zreference(x), y) & \textrm{if} ~ \zreference(x) \neq 0 \\
(y-1, y-1) & \textrm{if} ~ y > 1\\
\textit{End of string} & \textrm{otherwise}
\end{cases}\\
\proc{Character}(x, y)=&~\zlabel(x)\\
\proc{Position}(i)=&~(x,y)~\textrm{where}\\
&~~y=\proc{Predecessor}(i)~\textrm{and}\\
&~~x=\proc{Level-Ancestor}(i-\proc{Start-Of}(y), y)
\end{align*}
We then only need one Predecessor and one Level Ancestor query to initialize each of the two pairs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Random Access Initialization}

We can try to completely avoid Predecessor and Level Ancestor when initializing the pair $(x,y)$. We could try to store all $N$ pairs, but that would require $O(N)$ space and defeat the purpose of compression. We could also just walk from the end of the string to the desired position, but that would bring the average case query time up to $O(N)$. Similarly to how \proc{DirectCompLookup} works, we can store some of these positions. If we store $t$ positions equally spaced across the string $s$, we would get $O(n+t)$ space and $O(N/t+|\LCE(i,j)|)$ query time. To keep the compressed space, we can set $t=n$, which gives $O(N/n+|\LCE(i,j)|) = O(\sqrt{N}+|\LCE(i,j)|)$ query time.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Length of the Reversed String}

Reversing the string can be done in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively. To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-definition} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.

But the question is how big $n_2$ is relative to $n_1$. From knowing how much compression is possible with LZ-compression, we can see that $n_2=O(n_1^{~2})$. It might be possible to improve that bound, but for the string $z=(0,a)$ $(1,b)$ $(2,c)$ $...$ $(n-1,z)$, we have experimentally determined that reversing $z$ gives $n_2=0.89n_1^{~1.51}$, so we cannot get an optimal bound of $n_2=O(n_1)$. This is a disadvantage of doing LCE on the reversed string compared to doing it on the original compressed string using Level Ancestor queries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Using Compression to Speed Up \proc{DirectComp}}

Our original goal of looking into compression was to see if using the techniques used in compression could improve the worst case query time of \proc{DirectComp} to something better than $O(N)$. Our idea is that whenever you have a long LCE value, you have a repetition within the string, and repetitions would typically mean good compression. If the compressed string can be searched over similarly to how \proc{DirectComp} searches the uncompressed string, it might be possible to establish a better worst case query time.

We have found one case where compressed \proc{DirectComp} is faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time. We explain this in terms of the reversed compressed \proc{DirectComp}, as it is simpler, but it also works without the string reversal. Reversed compressed \proc{DirectComp} starts with two positions $(x_i,y_i)$ and $(x_j,y_j)$ corresponding to the two positions $i$ and $j$ in $\LCE(i,j)$. At each iteration, it then updates $(x_i,y_i)$ and $(x_j,y_j)$ to find the previous two positions. If $x_i = x_j$ we know that we can add $\textit{depth}[x_i]$ to the LCE value, where \textit{depth} is the depth of a node in the tree of compression elements. \proc{DirectComp} can then compare $\textit{depth}[x_i]$ characters in constant time.

Through this technique can improve query time in some cases, it does not improve the worst case time of $\Theta(N)$. This is because we cannot be sure to hit the condition allowing us to compare many characters in constant time, even for long LCE values. As an example, $LCE(1,2)$ on a string where all $N$ characters are the same, takes $\Omega(N)$ time. For such a string, the tree representing the compressed string is always a path, and $x_i$ and $x_j$ will always be either a node and its parent or the root and some other node. They will therefore never end up being the same node at the same time, as shown in the example in \pref{fig:lz-no-jumps}.

\begin{figure}[tp]
    \begin{center}
        \includegraphics[width=0.9\textwidth]{lz.pdf}
    \end{center}
    \caption{\label{fig:lz-no-jumps}Reversed compressed \proc{DirectComp} for $LCE(1,2)$ on the string $s=aaaaaaaaaa\$$. None of the steps can be skipped, since $x_i$ and $x_j$ are never the same.}
\end{figure}

\fixme{But can we cache some results as we go even if pointers are not the same? No, that is Theta(n**2) = Omega(N) space. But do we really need it all?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Compression with Other Algorithms\label{sec:lz-other}}

We can look at the other LCE algorithms we have examined in addition to \proc{DirectComp}, and see if we can apply them to LZ-compressed strings. We have not found a way to apply \proc{SuffixNca} or \proc{LcpRmq}, as the string we want to find LCE values on still has N suffixes, and we have not found a way to relate the compression rate of the NCA structure or suffix array to the compression rate of the string.

Fingerprinting does not compress well either. We could LC-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each elvel, the compression ration becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.

Since we have not been able to find improvements to the $O(N)$ worst case query time in $O(n)$ space using existing algorithms, we have looked into new algorithms. The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree. The LCE value is the length of a substring occuring at two positions $i$ and $j$ in the input string. This substring spans a number of phrases, where each phrase is a path in the tree of compression elements. If we can make fast LCE queries on such a tree, we can make fast LCE queries on highly compressable strings.

\begin{theorem}
Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$. We can preprocess $z$ using $O(n)$ space to answer LCE queries $v=\LCE(i,j)$ in $O(\log\log N + \log n \cdot \textit{\#phrases})$ time, where \textit{\#phrases} is the number of phrases of $z$ used to store $s[i\twodots i+v]$ and $s[j\twodots j+v]$.
\end{theorem}

\begin{proof}
In \pref{sec:tree-lce} we describe how to perform LCE queries on a tree of $n$ nodes in $O(\log n)$ query time and $O(n)$ space. We can first find the positions of $i$ and $j$ within $z$ in $O(\log\log N)$ time, and then we can repeat the $O(\log n)$ tree LCE query \textit{\#phrases} times to find the LCE value.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Cache Performance}

We have looked into the cache performance of compressed \proc{DirectComp}. If the string does not compress at all, i.e. the compressed string is $z=(0,a)(0,b)(0,c)...$, \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case. When the string is LZ-compressable, the I/O pattern becomes much worse. After reading compression element $x_1$ with $\zreference(x_1) = x_2 > 0$, the code will read from position $x_2$, where all we know is that $1 \leq x_2 < x_1$. This gives $O(N)$ I/O's in worst case.

We could lay out the tree of compression elements in memory as a B-tree\fixme{hvad hedder dette layout? B-tree??} to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the number of I/O's.\fixme{Er det klart?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LCE on Trees\label{sec:tree-lce}}

\begin{verbatim}
* LCE on trees
  * Definition
  * Heavy paths
  * Hashing
  * Ladder attempt
\end{verbatim}

TREE-LCE: Vi kan finde LCE på et træ i $O(\log n)$ tid ved at lave tung/let-vej og lave streng-LCE på de tunge veje.

TREE-LCE: Vi kan lave hashing på et træ, ved at hver knude gemmer hash-værdien for de sidste t led af stien fra roden til knuden.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Definition}

Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $p$ to child node $c$ is labeled with a character $\zlabel(c)$. Let $x_i$ and $x_j$ be two nodes in T. Let $y_i$ be a decendent of $x_i$, and let $y_j$ be a decendent of $x_j$. Let $s_i$ be the string obtained from characters on the path from $x_i$ to $y_i$, and let $s_j$ be the string obtained from characters on the path from $x_j$ to $y_j$. Then $\LCE_T(x_i, y_i, x_j, y_j)$ is the length of the longest common prefix of $s_i$ and $s_j$.

This definition of LCE on a tree is chosen to match the tree of LZ compression elements and the nessecary LCE queries on it described in \pref{sec:lz-other}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Walking the Paths with Level Ancestor}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Using Fingerprinting}

Gem fingerprint for hver knude for forfæder sti. Samme princip som hash-LCE for strenge

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Using Constant Time String LCE in Heavy Paths}

Vi kan lave Træ-LCE i O(n) plads, O(log n) tid ved at lave streng-LCE på tunge stier + Level-Ancestor.

\begin{verbatim}
    Current nodes are i_c and j_c, and end nodes a i_d and j_d.
    while i_c != i_d && j_c != j_d
        Let m_i be the number of initial heavy edges on the path from i_c to i_d before any light edge.
        Let m_j be ...
        m_i and m_j can be calculated in constant time.
        Let m = min(m_i, m_j)
        if m = 0
            if the two next characters are the same
                advance one character
            else
                stop
        else
            Do heavy path LCE starting at i_c and j_c, and let v be the LCE value.
            If v >= m
                advance m characters using Level Ancestor
            else
                advance v characters
                stop

    O(log n) time, one step for each light edge encountered on each of the two paths.
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Directly Mapping Other Algorithms}

\begin{verbatim}
Directly apply SuffixNca or LcpRmq. Not possible.
A suffix tree would take too much space. Each leaf would have d suffixes, where d is the deapth of the leaf. That is more than n. (Only leafs, since for internal nodes we can choose a decendant leaf and cut off the excess later)
The input is bigger, on strings we had i (n possibilities) and j (n possibilities), and on trees we have (i_c,i_d) (>n possibilities) and (j_c,j_d) (n> possibilities). So directly using suffix trees or SA-LCP-RMQ would be >n space.
\end{verbatim}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Level Ancestor way?}

\begin{verbatim}
Kan vi lige som i Level Ancestor smække to O(log n) sammen til en O(1)?

Vi kan lave træ-LCE i O(hn) plads, O(1) tid ved at lave strenge for hver sti til et blad. Hver knude kender en sti den ligger på, og vi finder to stier for slut-knuderne og laver streng-LCE på dem.
\end{verbatim}

\fi % report

\begin{thebibliography}{9}

\bibitem{ilie-navarro-tinta} Lucian Ilie, Gonzalo Navarro, and Liviu Tinta. The longest common extension problem revisited and applications to approximate string searching. Journal of Discrete Algorithms, Volume 8, Issue 4, December 2010, pages 418-428.

% \bibitem{linear-lcp} Toru Kasai, Gunho Lee, Hiroki Arimura, Setsuo Arikawa, and Kunsoo Park. Linear-Time Longest-Common-Prefix Computation in Suffix Arrays and Its Applications. CPM 2001, LNCS 2089, 2001, pages 181-192.

\bibitem{sort-complexity} Martin Farach-Colton, Paolo Ferragina, and S. Muthukrishnan. On the sorting-complexity of suffix tree construction. J. ACM Vol. 47, No. 6, November 2000, pages 987-1011.

\bibitem{karp-miller-rosenberg} Richard M. Karp, Raymond E. Miller, and Arnold L. Rosenberg. 1972. Rapid identification of repeated patterns in strings, trees and arrays. In Proceedings of the fourth annual ACM symposium on Theory of computing (STOC '72). ACM, New York, NY, USA, 125-136.

\bibitem{nca} S. Alstrup, C. Gavoille, H. Kaplan, and T. Rauhe. Nearest Common Ancestors: A Survey and a New Algorithm for a Distributed Environment. Theory of Comput. Sys., 2004.

\bibitem{jf-rmq} Johannes Fischer, and Volker Heun. Theoretical and Practical Improvements on the RMQ-Problem, with Applications to LCA and LCE. Proceedings of the 17th Annual Symposium on Combinatorial Pattern Matching (CPM'06), Lecture Notes in Computer Science 4009, 36-48, Springer-Verlag, 2006.

\end{thebibliography}

\ifreport

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{THE END}

\section{Tinta tests}
The two tests of LCE query performance used in the Tinta paper are shown in \pref{alg:tinta-testlceonebyone} and \pref{alg:tinta-randomtestlce}.

% http://www.cs.dartmouth.edu/~thc/clrscode/clrscode.sty

\begin{algorithm}
\caption{Test algorithm for \texttt{test\_lce\_one\_by\_one}\label{alg:tinta-testlceonebyone}}
\begin{codebox}
\li \For each $LCE \in Algorithms$
    \Indentmore
\li     \For $k = 2$ \To $n$
        \Indentmore
\li         $A =\{(i,j) ~|~ j-i=k\}$
\li         \Repeat 3 times
\li             shuffle $A$
\li             start timing
\li             \For each $(i,j) \in A$
                \Indentmore
\li                 $LCE(i,j)$
                \End
\li             end timing
            \End
\li         $avg =$ average of the 3 timings
\li         log(each of the 3 timings)
\li         print($avg/length(A)$)
        \End
    \End
\end{codebox}
\end{algorithm}

\begin{algorithm}
\caption{Test algorithm for \texttt{random\_test\_lce}\label{alg:tinta-randomtestlce}}
\begin{codebox}
\li \Repeat 3 times
\li     $A =$ generate 1000000 random pairs $(i,j)$ where $i<j$
\li     \For each $\proc{LCE} \in \id{Algorithms}$
        \Indentmore
\li         start timing
\li         \For each $(i,j) \in A$
            \Indentmore
\li             $LCE(i,j)$
            \End
\li         end timing
        \End
    \End
\li \For each $LCE \in Algorithms$
    \Indentmore
\li     log(each of the 3 timings)
    \End
\li \For each $LCE \in Algorithms$
    \Indentmore
\li     print($($ average of the 3 timings $)/length(A)$)
    \End
\end{codebox}
\end{algorithm}

\section{DONE}
\begin{enumerate}
\item CODE:  test cache hops med cachegrind. STATUS: Samme instruction count, forskellige cache misses, men ser ud til at være modsat forventet.
("Cache set count is not a power of two.")
\item CODE: Undersøg præcis hvad der giver cache i kun average case i RMQ kode. STATUS: ikke cache, uopklaret
\item DOC: Mere diskussion af og konklusion på resultaterne, beskriv dem som Navarro gør.
\end{enumerate}

\section{TODO Article}
\begin{enumerate}
\item CODE: Test på real world tekster, i tabel som Navarro. STATUS: Pizza\&Chili-teksterne er for store. Brug nogle andre?
\item DOC: Opskriv funktioner for plads.
\end{enumerate}

\section{DONE Not for article}
\begin{enumerate}
\item DOC: Test on another machine. STATUS: Dropped
\item CODE: Cache optimized memory layout for fingerprinting: Initial results bad, maybe test for other $k$, and test vertical for $k=2$.
\item CODE: Plot results using LCE value as x-axis: Results looks wrong, why? Likely because of queries in a given pattern, v=max(i,j)
\end{enumerate}

\section{TODO Not for article}
\begin{enumerate}
\item CODE: Improve cache-vertical and make shift variant.
\item CODE: Non-random strings? Non-random queries? Different t-values for Hash-3L?
\item DOC: Explain why Karp-Miller-Rosenberg is $O(n\log n)$.
\end{enumerate}

\begin{comment}
\section{What is the time for KMR?}
\subsection{Original KMR}
"Rapid identification of repeated patterns in strings, trees and arrays"
says $O(n\cdot(\log n)^2)$.

http://portal.acm.org.globalproxy.cvt.dk/citation.cfm?id=804905

\subsection{RMR in $O(n \log n)$}
\begin{itemize}
\item
"A Theoretical and Experimental Study on the Construction of Suffix Arrays in External Memory"
page 10, 3.3 The Doubling Algorithm.

http://www.springerlink.com.globalproxy.cvt.dk/content/h1tejj3bamjj5qrg/
\item
"Compressing and indexing labeled trees, with applications"

http://portal.acm.org.globalproxy.cvt.dk/citation.cfm?id=1613680
\end{itemize}

KMR uses $O(sort(n,n^2)\cdot\log n) = O(n\log n)$ time.

At a given level:
\begin{enumerate}
\item create 2-touples of fingerprints from the previous level.
\item sort these touples lexicographically. (radix sort takes $O(kn)$ for $\sigma=n^k$, here $\sigma=n^2$)
\item scan through the sorted touples to assign new fingerprints.
\end{enumerate}
\end{comment}

\fi % report

\end{document}

### 2011-07-11 doc/compile.sh

@@ -1,0 +1,1 @@
+pdflatex -shell-escape report.tex

### 2011-07-11 src/RMQ_n_logn/RMQ.hpp

@@ -1,0 +1,18 @@
+#ifndef _RMQ_hpp_
+#define _RMQ_hpp_
+
+#include <math.h>
+
+typedef int DT;                 // use long for 64bit-version (but take care of fast log!)
+typedef unsigned int DTidx;     // for indexing in arrays
+
+/* Abstract class for RMQ-queries. Proprocessing is done
+   in constructor. */
+class RMQ {
+public:
+	// returns index of RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx) = 0;
+	virtual ~RMQ() = 0;
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_1_n.cpp

@@ -1,0 +1,26 @@
+#include "RMQ_1_n.hpp"
+
+DTidx RMQ_1_n::query(DTidx i, DTidx j) {
+	DTidx min = j;
+	for (DTidx x = i; x < j; x++)
+		if (a[x] < a[min]) min = x;
+	return min;
+}
+
+/**
+ * Standard Constructor. a is the array to be prepared for RMQ.
+ * n is the size of the array.
+ */
+RMQ_1_n::RMQ_1_n(DT* a, DTidx n) {
+	this->a = a;
+	this->n = n;
+}
+
+/**
+ * Destructor. Does nothing.
+ */
+RMQ_1_n::~RMQ_1_n() {
+	
+}
+
+RMQ::~RMQ() {} // TODO this is not the prettiest place

### 2011-07-11 src/RMQ_n_logn/RMQ_1_n.hpp

@@ -1,0 +1,29 @@
+#ifndef _RMQ_1_n_hpp_
+#define _RMQ_1_n_hpp_
+
+#include "RMQ.hpp"
+#include <stdlib.h>
+#include <iostream>
+using namespace std;
+
+
+/* Implements the most naive <O(1),O(n)>-method to compute RMQ.
+ */
+class RMQ_1_n : public RMQ {
+public:
+	// returns RMQ(i,j)
+	virtual DTidx query(DTidx, DTidx);
+
+	RMQ_1_n(DT* a, DTidx n);
+
+	virtual ~RMQ_1_n();
+
+protected:
+	// array
+	DT* a;
+
+	// size
+	DTidx n;
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_n_1_improved.hpp

@@ -1,0 +1,57 @@
+#ifndef _RMQ_n_1_improved_hpp_
+#define _RMQ_n_1_improved_hpp_
+
+#include "RMQ.hpp"
+#include "RMQ_nlogn_1.hpp"
+#include <stdlib.h>
+#include <iostream>
+using namespace std;
+
+class RMQ_n_1_improved : public RMQ {
+public:
+	// liefert RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx);
+
+	RMQ_n_1_improved(DT* a, DTidx n);
+
+	virtual ~RMQ_n_1_improved();
+
+protected:
+	// array
+	DT *a;
+
+	// index array for the out-of-block queries (contains indices of block-minima)
+	DTidx *c;
+
+	// type of blocks
+	DTidx *type;
+
+	// precomputed in-block queries
+	DTidx** Prec;
+
+	// number of different queries per block = bs*(bs+1)/2
+	DTidx qpb;
+
+	// size of array a
+	DTidx n;
+
+	// block size
+	DTidx bs;
+
+	// number of blocks (always n/bs)
+	DTidx nb;
+
+	// return block of entry i:
+	inline DTidx block(DTidx i) { return i/bs; }
+
+	// nlogn_1-Algo for out-of-block-block-queries:
+	RMQ_nlogn_1* RMQ_ST;
+
+	// precomputed Catalan triangle (17 is enough for 64bit computing):
+	static const DTidx Catalan[17][17];
+
+	// minus infinity (change for 64bit version)
+	static const DT minus_infinity;
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_n_logn.cpp

@@ -1,0 +1,75 @@
+#include "RMQ_n_logn.hpp"
+
+/*unsigned long RMQ_n_logn::getSize() {
+	return mem + RMQ_ST->getSize();
+}*/
+
+DTidx RMQ_n_logn::query(DTidx i, DTidx j) {
+	DTidx block_i = block(i);       // block where i is
+	DTidx block_j = block(j);       // j's block
+	DTidx min, tmp;	                // min: to be returned
+
+	if (block_j == block_i) {
+		min = j;
+		for (DTidx x = i; x < j; x++)
+			if (a[x] < a[min]) min = x;
+	}
+	else {
+		tmp = (block_i +1) * bs - 1; // end of i's block
+		min = tmp;
+		for (DTidx x = i; x < tmp; x++)  // calculate minimum in i's block
+			if (a[x] < a[min]) min = x;
+		for (DTidx x = block_j*bs; x <= j; x++) // calculate minimum in j's block
+			if (a[x] < a[min]) min = x;
+		if (block_j > block_i + 1) { // out-of-block query:
+			tmp = c[RMQ_ST->query(block_i+1, block_j-1)]; // index where minimum is
+			if (a[tmp] < a[min]) min = tmp;
+		}
+	}
+
+	return min;
+}
+
+/** Divide a into blocks of size log(n)/d. */
+RMQ_n_logn::RMQ_n_logn(DT* a, DTidx n, float d) {
+	this->a = a;
+	this->n = n;
+	bs = (DTidx) ceil(log2(n) / d); // calculate block size
+	nb = block(n-1)+1;              // number of blocks
+
+	// allocate and fill arrays for out-of-block-queries:
+	c = new DTidx[nb];
+	mem = sizeof(DTidx) * nb;
+
+	DTidx min;   // position of minimum in current block
+	DTidx z=0;   // index in array a
+	DTidx end;   // end of current block
+
+	for (DTidx i = 0; i < nb; i++) { // step through blocks
+		min = z;              // init minimum
+		end = z + bs;         // end of block
+		if (end > n) end = n; // last block could be smaller than bs!
+
+		while (++z < end)     // step through current block
+			if (a[z] < a[min]) min = z;   // update minimum
+		c[i] = min; // store index of minimum
+	}
+
+	RMQ_ST = new RMQ_nlogn_1(a, c, nb); // construct sparse table on block-minima
+}
+
[…]

### 2011-07-11 src/RMQ_n_logn/RMQ_n_logn.hpp

@@ -1,0 +1,52 @@
+#ifndef _RMQ_n_logn_hpp_
+#define _RMQ_n_logn_hpp_
+
+#include "RMQ.hpp"
+#include "RMQ_nlogn_1.hpp"
+#include <stdlib.h>
+#include <iostream>
+using namespace std;
+
+class RMQ_n_logn : public RMQ {
+public:
+	// liefert RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx);
+
+	// returns space occupied by the index (in bytes)
+	//virtual unsigned long getSize();
+
+	// this constructor divides the array into blocks of size log(n)
+	RMQ_n_logn(DT* a, DTidx n);
+
+	// this constructor divides the array into blocks of size log(n)/d
+	RMQ_n_logn(DT* a, DTidx n, float d);
+
+	virtual ~RMQ_n_logn();
+
+protected:
+	// array
+	DT *a;
+
+	// index array for the out-of-block queries (contains indices of block-minima)
+	DTidx *c;
+
+	// size of array a
+	DTidx n;
+
+	// block size
+	DTidx bs;
+
+	// number of blocks (always n/bs)
+	DTidx nb;
+
+	// return block of entry i:
+	inline DTidx block(DTidx i) { return i/bs; }
+
+	// nlogn_1-Algo for out-of-block-block-queries:
+	RMQ_nlogn_1* RMQ_ST;
+
+	// count allocated memory:
+	unsigned long mem;
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_n_logn_2.cpp

@@ -1,0 +1,139 @@
+#include "RMQ_n_logn_2.hpp"
+
+const DTidx RMQ_n_logn_2::Catalan[17][17] = {
+	{1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1},
+	{0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
+	{0,0,2,5,9,14,20,27,35,44,54,65,77,90,104,119,135},
+	{0,0,0,5,14,28,48,75,110,154,208,273,350,440,544,663,798},
+	{0,0,0,0,14,42,90,165,275,429,637,910,1260,1700,2244,2907,3705},
+	{0,0,0,0,0,42,132,297,572,1001,1638,2548,3808,5508,7752,10659,14364},
+	{0,0,0,0,0,0,132,429,1001,2002,3640,6188,9996,15504,23256,33915,48279},
+	{0,0,0,0,0,0,0,429,1430,3432,7072,13260,23256,38760,62016,95931,144210},
+	{0,0,0,0,0,0,0,0,1430,4862,11934,25194,48450,87210,149226,245157,389367},
+	{0,0,0,0,0,0,0,0,0,4862,16796,41990,90440,177650,326876,572033,961400},
+	{0,0,0,0,0,0,0,0,0,0,16796,58786,149226,326876,653752,1225785,2187185},
+	{0,0,0,0,0,0,0,0,0,0,0,58786,208012,534888,1188640,2414425,4601610},
+	{0,0,0,0,0,0,0,0,0,0,0,0,208012,742900,1931540,4345965,8947575},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,742900,2674440,7020405,15967980},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,2674440,9694845,25662825},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,9694845,35357670},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,35357670}
+};
+
+const DT RMQ_n_logn_2::minus_infinity = -2147483648;
+
+DTidx RMQ_n_logn_2::query(DTidx i, DTidx j) {
+	DTidx block_i = block(i);       // block where i is
+	DTidx block_j = block(j);       // j's block
+	DTidx i_pos = i - (block_i*bs); // position of i in block_i
[…]

### 2011-07-11 src/RMQ_n_logn/RMQ_n_logn_2.hpp

@@ -1,0 +1,60 @@
+#ifndef _RMQ_n_logn_2_improved_hpp_
+#define _RMQ_n_logn_2_improved_hpp_
+
+#include "RMQ.hpp"
+#include "RMQ_nlogn_1.hpp"
+#include <stdlib.h>
+#include <iostream>
+using namespace std;
+
+class RMQ_n_logn_2 : public RMQ {
+public:
+	// liefert RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx);
+
+	RMQ_n_logn_2(DT* a, DTidx n);
+
+	virtual ~RMQ_n_logn_2();
+
+protected:
+	// array
+	DT *a;
+
+	// index array for the out-of-block queries (contains indices of block-minima)
+	DTidx *c;
+
+	// type of blocks
+	DTidx *type;
+
+	// precomputed in-block queries
+	DTidx** Prec;
+
+	// number of different queries per block = bs*(bs+1)/2
+	DTidx qpb;
+
+	// size of array a
+	DTidx n;
+
+	// block size
+	DTidx bs;
+
+	// number of blocks (always n/bs)
+	DTidx nb;
+
+	// return block of entry i:
+	inline DTidx block(DTidx i) { return i/bs; }
+
+	// nlogn_1-Algo for out-of-block-block-queries:
+	RMQ_nlogn_1* RMQ_ST;
+
+	// precomputed Catalan triangle (17 is enough for 64bit computing):
+	static const DTidx Catalan[17][17];
+
+	// minus infinity (change for 64bit version)
+	static const DT minus_infinity;
+	
+	// compute in-bloc query
+	DTidx in_bloc_minimum(const DTidx start, const DTidx end)const;
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_n_logn_4.cpp

@@ -1,0 +1,139 @@
+#include "RMQ_n_logn_4.hpp"
+
+const DTidx RMQ_n_logn_4::Catalan[17][17] = {
+	{1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1},
+	{0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
+	{0,0,2,5,9,14,20,27,35,44,54,65,77,90,104,119,135},
+	{0,0,0,5,14,28,48,75,110,154,208,273,350,440,544,663,798},
+	{0,0,0,0,14,42,90,165,275,429,637,910,1260,1700,2244,2907,3705},
+	{0,0,0,0,0,42,132,297,572,1001,1638,2548,3808,5508,7752,10659,14364},
+	{0,0,0,0,0,0,132,429,1001,2002,3640,6188,9996,15504,23256,33915,48279},
+	{0,0,0,0,0,0,0,429,1430,3432,7072,13260,23256,38760,62016,95931,144210},
+	{0,0,0,0,0,0,0,0,1430,4862,11934,25194,48450,87210,149226,245157,389367},
+	{0,0,0,0,0,0,0,0,0,4862,16796,41990,90440,177650,326876,572033,961400},
+	{0,0,0,0,0,0,0,0,0,0,16796,58786,149226,326876,653752,1225785,2187185},
+	{0,0,0,0,0,0,0,0,0,0,0,58786,208012,534888,1188640,2414425,4601610},
+	{0,0,0,0,0,0,0,0,0,0,0,0,208012,742900,1931540,4345965,8947575},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,742900,2674440,7020405,15967980},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,2674440,9694845,25662825},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,9694845,35357670},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,35357670}
+};
+
+const DT RMQ_n_logn_4::minus_infinity = -2147483648;
+
+DTidx RMQ_n_logn_4::query(DTidx i, DTidx j) {
+	DTidx block_i = block(i);       // block where i is
+	DTidx block_j = block(j);       // j's block
+	DTidx i_pos = i - (block_i*bs); // position of i in block_i
[…]

### 2011-07-11 src/RMQ_n_logn/RMQ_n_logn_4.hpp

@@ -1,0 +1,60 @@
+#ifndef _RMQ_n_logn_4_improved_hpp_
+#define _RMQ_n_logn_4_improved_hpp_
+
+#include "RMQ.hpp"
+#include "RMQ_nlogn_1.hpp"
+#include <stdlib.h>
+#include <iostream>
+using namespace std;
+
+class RMQ_n_logn_4 : public RMQ {
+public:
+	// liefert RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx);
+
+	RMQ_n_logn_4(DT* a, DTidx n);
+
+	virtual ~RMQ_n_logn_4();
+
+protected:
+	// array
+	DT *a;
+
+	// index array for the out-of-block queries (contains indices of block-minima)
+	DTidx *c;
+
+	// type of blocks
+	DTidx *type;
+
+	// precomputed in-block queries
+	DTidx** Prec;
+
+	// number of different queries per block = bs*(bs+1)/2
+	DTidx qpb;
+
+	// size of array a
+	DTidx n;
+
+	// block size
+	DTidx bs;
+
+	// number of blocks (always n/bs)
+	DTidx nb;
+
+	// return block of entry i:
+	inline DTidx block(DTidx i) { return i/bs; }
+
+	// nlogn_1-Algo for out-of-block-block-queries:
+	RMQ_nlogn_1* RMQ_ST;
+
+	// precomputed Catalan triangle (17 is enough for 64bit computing):
+	static const DTidx Catalan[17][17];
+
+	// minus infinity (change for 64bit version)
+	static const DT minus_infinity;
+	
+	// compute in-bloc query
+	DTidx in_bloc_minimum(const DTidx start, const DTidx end)const;
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_nlogn_1.hpp

@@ -1,0 +1,46 @@
+#ifndef _RMQ_nlogn_1_hpp_
+#define _RMQ_nlogn_1_hpp_
+
+#include "RMQ.hpp"
+#include <stdlib.h>
+#include <iostream>
+using namespace std;
+
+
+/* Implements the <O(n log n), O(1)>-method for RMQ as described in
+ * Bender and Farach's paper, section 3.
+ */
+class RMQ_nlogn_1 : public RMQ {
+public:
+	// liefert RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx);
+
+	RMQ_nlogn_1(DT* a, DTidx* c, DTidx n);
+
+	virtual ~RMQ_nlogn_1();
+
+	// the following stuff is for fast base 2 logarithms:
+	// (currently only implemented for 32 bit numbers)
+	static const char LogTable256[256];
+
+	virtual DTidx log2fast(DTidx);
+
+protected:
+	// array
+	DT* a;
+
+	// index array for a:
+	DTidx* c;
+
+	// size of c
+	DTidx n;
+
+	// depth of table
+	DTidx depth;
+
+	// the precomputed table:
+	DTidx** M;
+
+};
+
+#endif

### 2011-07-11 src/RMQ_n_logn/RMQ_succinct.cpp

@@ -1,0 +1,349 @@
+// http://www-ab.informatik.uni-tuebingen.de/people/fischer
+// Systematic setting. [C++ source package] (ESCAPE'07; last update: 2010/03/17).
+// http://www-ab.informatik.uni-tuebingen.de/people/fischer/succinctRMQ.tgz
+#include "RMQ_succinct.hpp"
+
+const DTidx RMQ_succinct::Catalan[17][17] = {
+	{1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1},
+	{0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
+	{0,0,2,5,9,14,20,27,35,44,54,65,77,90,104,119,135},
+	{0,0,0,5,14,28,48,75,110,154,208,273,350,440,544,663,798},
+	{0,0,0,0,14,42,90,165,275,429,637,910,1260,1700,2244,2907,3705},
+	{0,0,0,0,0,42,132,297,572,1001,1638,2548,3808,5508,7752,10659,14364},
+	{0,0,0,0,0,0,132,429,1001,2002,3640,6188,9996,15504,23256,33915,48279},
+	{0,0,0,0,0,0,0,429,1430,3432,7072,13260,23256,38760,62016,95931,144210},
+	{0,0,0,0,0,0,0,0,1430,4862,11934,25194,48450,87210,149226,245157,389367},
+	{0,0,0,0,0,0,0,0,0,4862,16796,41990,90440,177650,326876,572033,961400},
+	{0,0,0,0,0,0,0,0,0,0,16796,58786,149226,326876,653752,1225785,2187185},
+	{0,0,0,0,0,0,0,0,0,0,0,58786,208012,534888,1188640,2414425,4601610},
+	{0,0,0,0,0,0,0,0,0,0,0,0,208012,742900,1931540,4345965,8947575},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,742900,2674440,7020405,15967980},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,2674440,9694845,25662825},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,9694845,35357670},
+	{0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,35357670}
+};
+
[…]

### 2011-07-11 src/RMQ_n_logn/RMQ_succinct.hpp

@@ -1,0 +1,105 @@
+#ifndef _RMQ_succinct_hpp_
+#define _RMQ_succinct_hpp_
+
+#define MEM_COUNT
+
+#include "RMQ.hpp"
+#include <stdlib.h>
+#include <limits.h>
+#include <iostream>
+using namespace std;
+
+typedef unsigned char DTsucc;
+typedef unsigned short DTsucc2;
+
+class RMQ_succinct : public RMQ {
+public:
+	// liefert RMQ[i,j]
+	virtual DTidx query(DTidx, DTidx);
+
+	RMQ_succinct(DT* a, DTidx n);
+
+	virtual ~RMQ_succinct();
+
+protected:
+	// array
+	DT *a;
+
+	// size of array a
+	DTidx n;
+
+	// table M for the out-of-block queries (contains indices of block-minima)
+	DTsucc** M;
+
+	// because M just stores offsets (rel. to start of block), this method
+	// re-calculates the true index:
+	inline DTidx m(DTidx k, DTidx block) { return M[k][block]+(block*sprime); }
+
+	// depth of table M:
+	DTidx M_depth;
+
+	// table M' for superblock-queries (contains indices of block-minima)
+	DTidx** Mprime;
+
+	// depth of table M':
+	DTidx Mprime_depth;
+
+	// type of blocks
+	DTsucc2 *type;
+
+	// precomputed in-block queries
+	DTsucc** Prec;
+
+	// microblock size
+	DTidx s;
+
+	// block size
+	DTidx sprime;
+
+	// superblock size
+	DTidx sprimeprime;
+
+	// number of blocks (always n/sprime)
+	DTidx nb;
+
+	// number of superblocks (always n/sprimeprime)
+	DTidx nsb;
+
+	// number of microblocks (always n/s)
+	DTidx nmb;
+
+	// return microblock-number of entry i:
+	inline DTidx microblock(DTidx i) { return i/s; }
+
+	// return block-number of entry i:
+	inline DTidx block(DTidx i) { return i/sprime; }
+
+	// return superblock-number of entry i:
+	inline DTidx superblock(DTidx i) { return i/sprimeprime; }
+
+	// precomputed Catalan triangle (17 is enough for 64bit computing):
+	static const DTidx Catalan[17][17];
+
+	// minus infinity (change for 64bit version)
+	static const DT minus_infinity;
+
+ 	// stuff for clearing the least significant x bits (change for 64-bit computing)
+	static const DTsucc HighestBitsSet[8];
[…]

### 2011-07-11 src/WinNTL-5_5_2.sh

@@ -1,0 +1,3 @@
+cd WinNTL-5_5_2/src/
+g++ FFT.cpp FacVec.cpp GF2.cpp GF2E.cpp GF2EX.cpp GF2EXFactoring.cpp GF2X.cpp GF2X1.cpp GF2XFactoring.cpp GF2XVec.cpp G_LLL_FP.cpp G_LLL_QP.cpp G_LLL_RR.cpp G_LLL_XD.cpp GetTime.cpp HNF.cpp LLL.cpp LLL_FP.cpp LLL_QP.cpp LLL_RR.cpp LLL_XD.cpp RR.cpp WordVector.cpp ZZ.cpp ZZVec.cpp ZZX.cpp ZZX1.cpp ZZXCharPoly.cpp ZZXFactoring.cpp ZZ_p.cpp ZZ_pE.cpp ZZ_pEX.cpp ZZ_pEXFactoring.cpp ZZ_pX.cpp ZZ_pX1.cpp ZZ_pXCharPoly.cpp ZZ_pXFactoring.cpp ctools.cpp fileio.cpp lip.cpp lzz_p.cpp lzz_pE.cpp lzz_pEX.cpp lzz_pEXFactoring.cpp lzz_pX.cpp lzz_pX1.cpp lzz_pXCharPoly.cpp lzz_pXFactoring.cpp mat_GF2.cpp mat_GF2E.cpp mat_RR.cpp mat_ZZ.cpp mat_ZZ_p.cpp mat_ZZ_pE.cpp mat_lzz_p.cpp mat_lzz_pE.cpp mat_poly_ZZ.cpp mat_poly_ZZ_p.cpp mat_poly_lzz_p.cpp pair_GF2EX_long.cpp pair_GF2X_long.cpp pair_ZZX_long.cpp pair_ZZ_pEX_long.cpp pair_ZZ_pX_long.cpp pair_lzz_pEX_long.cpp pair_lzz_pX_long.cpp quad_float.cpp tools.cpp vec_GF2.cpp vec_GF2E.cpp vec_GF2XVec.cpp vec_RR.cpp vec_ZZ.cpp vec_ZZVec.cpp vec_ZZ_p.cpp vec_ZZ_pE.cpp vec_double.cpp vec_long.cpp vec_lzz_p.cpp vec_lzz_pE.cpp vec_quad_float.cpp vec_ulong.cpp vec_vec_GF2.cpp vec_vec_GF2E.cpp vec_vec_RR.cpp vec_vec_ZZ.cpp 
[…]

### 2011-07-11 src/algDirectComp.cpp

@@ -1,0 +1,66 @@
+#include "algDirectComp.h"
+#include <stdio.h>
+
+AlgorithmDirectComp::AlgorithmDirectComp() {
+    this->s = NULL;
+}
+
+AlgorithmDirectComp::~AlgorithmDirectComp() {
+}
+
+bool AlgorithmDirectComp::preproc(TestString* s) {
+    this->s = s;
+    return true;
+}
+
+uint32 AlgorithmDirectComp::query(uint32 i, uint32 j) {
+    TestString* d = this->s;
+    uint32 t = 0;
+    if (i == j)
+        return d->n - i;
+    while (d->s[i + t] == d->s[j + t]) {
+        t++;
+    }
+    return t;
+}
+
+void AlgorithmDirectComp::cleanup() {
+    this->s = NULL;
+}
+
+void AlgorithmDirectComp::getName(char* name, uint32 size) {
+    snprintf(name, size, "DirectComp");
+}
+
+uint64 AlgorithmDirectComp::spaceUsage(uint32 n) {
+    return sizeof(TestString*);
+}
+
+// Their implementations
+/*
+
+
+// x, y from 1 to length of string
+inline int simple_lcp_2_suffixes(const char * const str, const unsigned int length, register int x, register int y){
+    register int lcp = 0;
+    while( (x-1 < length) && (y-1 < length) && (str[x-1] == str[y-1]) ){
+        ++lcp;
+        ++x;
+        ++y;
+    }
+    return lcp;
+}
+
+// this version is for zero terminated text strings (C strings)
+inline int simple_lcp_2_suffixes(const char * const str, const unsigned int x, const unsigned int y){
+    register int lcp = 0;
+    //const unsigned int *const int_str = (unsigned int*)str;
+    const char *p = &(str[x-1]);
+    const char *q = &(str[y-1]);
+    while( *p++ == *q++ ) ++lcp;
+
+    return lcp;
+}
+
+
+*/

### 2011-07-11 src/algDirectComp.h

@@ -1,0 +1,20 @@
+#ifndef _ALGDIRECTCOMP_H_
+#define _ALGDIRECTCOMP_H_
+
+#include "algorithm.h"
+class AlgorithmDirectComp : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    AlgorithmDirectComp();
+    ~AlgorithmDirectComp();
+
+protected:
+    TestString* s;
+};
+
+#endif

### 2011-07-11 src/algDirectCompLookup.cpp

@@ -1,0 +1,110 @@
+#include "algDirectCompLookup.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+AlgorithmDirectCompLookup::AlgorithmDirectCompLookup(NamedFunc* f) {
+    s = NULL;
+    table = NULL;
+    n = 0;
+    S_CLEAR(nn)
+    S_CLEAR_MOD(t)
+    this->f = f;
+    fp = 0;
+}
+
+AlgorithmDirectCompLookup::AlgorithmDirectCompLookup(NamedFunc* f, uint32 fp) {
+    s = NULL;
+    table = NULL;
+    n = 0;
+    S_CLEAR(nn)
+    S_CLEAR_MOD(t)
+    this->f = f;
+    this->fp = fp;
+}
+
+AlgorithmDirectCompLookup::~AlgorithmDirectCompLookup() {
+}
+
+bool AlgorithmDirectCompLookup::preproc(TestString* ts) {
+    uint32 t = f->f(ts->n);
+    if (fp)
+        t = f->f(fp);
+    uint32 i, j, k, cur;
+    uint64 size;
+
+    uint32 nn = ts->n;
+    S_PREP(nn)
+    S_PREP_MOD(t)
+    size = (uint64)ts->n * (S_DIV((uint64) nn, t) + 1u);
+    if (size > INT32_MAX/sizeof(uint32)) {
+        S_CLEAR(nn);
+        S_CLEAR_MOD(t)
+        return false;
+    }
+    table = new (std::nothrow) uint32[size];
+    if (!table) {
+        S_CLEAR(nn)
+        S_CLEAR_MOD(t)
+        return false;
+    }
+    s = ts->s;
+    n = ts->n;
+    for (k = 1; k < n; k++) {
+        cur = 0;
+        for (i = n-k-1; (int32)i >= 0; i--) {
+            j = i + k;
+            if (s[i] == s[j])
+                cur++;
+            else
+                cur = 0;
+            if (!S_MOD(i, t)) {
+                table[S_MUL(S_DIV(i, t), nn) + j] = cur;
+            }
+        }
+    }
+    return true;
+}
+
+uint32 AlgorithmDirectCompLookup::query(uint32 i, uint32 j) {
+    uint32 v = j;
+    if (i == j)
+        return n - i;
+    else if (i > j) {
+        j = i;
+        i = v;
+    }
+    v = 0;
+
[…]

### 2011-07-11 src/algDirectCompLookup.h

@@ -1,0 +1,29 @@
+#ifndef _ALGDIRECTCOMPLOOKUP_H_
+#define _ALGDIRECTCOMPLOOKUP_H_
+
+#include "algorithm.h"
+#include "bittricks.h"
+
+class AlgorithmDirectCompLookup : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    AlgorithmDirectCompLookup(NamedFunc*);
+    AlgorithmDirectCompLookup(NamedFunc*, uint32);
+    ~AlgorithmDirectCompLookup();
+
+protected:
+    uint8* s;
+    uint32* table;
+    uint32 n;
+    S_DECL(nn)
+    S_DECL_MOD(t)
+    NamedFunc* f;
+    uint32 fp;
+};
+
+#endif

### 2011-07-11 src/algDirectLookup.cpp

@@ -1,0 +1,75 @@
+#include "algDirectLookup.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+AlgorithmDirectLookup::AlgorithmDirectLookup() {
+    this->table = NULL;
+    this->n = 0;
+    this->sham = 0;
+}
+
+AlgorithmDirectLookup::~AlgorithmDirectLookup() {
+}
+
+// TODO kan det allokerede adresserum gøres mere kompakt?
+bool AlgorithmDirectLookup::preproc(TestString* s) {
+    uint32 i, j, k, cur;
+    uint64 size;
+    this->n = s->n;
+    this->sham = 0;
+    while (1u<<this->sham < s->n)
+        this->sham++;
+    size = (uint64)s->n*((uint64)1u<<this->sham);
+    if (size > INT32_MAX/sizeof(uint32)) {
+        this->n = 0;
+        this->sham = 0;
+        return false;
+    }
+    this->table = new (std::nothrow) uint32[size];
+    if (!this->table) {
+        this->n = 0;
+        this->sham = 0;
+        return false;
+    }
+    for (k = 1; k < s->n; k++) {
+        cur = 0;
+        for (i = s->n-k-1; (int32)i >= 0; i--) {
+            j = i + k;
+            if (s->s[i] == s->s[j])
+                cur++;
+            else
+                cur = 0;
+            this->table[(i<<this->sham) + j] = cur;
+        }
+    }
+    return true;
+}
+
+uint32 AlgorithmDirectLookup::query(uint32 i, uint32 j) {
+    if (i < j)
+        return this->table[(i<<this->sham) + j];
+    else if (i > j)
+        return this->table[(j<<this->sham) + i];
+    else
+        return this->n - i;
+}
+
+void AlgorithmDirectLookup::cleanup() {
+    delete[] this->table;
+    this->table = NULL;
+    this->n = 0;
+    this->sham = 0;
+}
+
+void AlgorithmDirectLookup::getName(char* name, uint32 size) {
+    snprintf(name, size, "DirectLookup");
+}
+
+uint64 AlgorithmDirectLookup::spaceUsage(uint32 n) {
+    uint32 sham = 0;
+    while (1u<<sham < n)
+        sham++;
+    return (uint64)n*((uint64)1u<<sham)*sizeof(uint32)
+        + sizeof(AlgorithmDirectLookup);
+}

### 2011-07-11 src/algDirectLookup.h

@@ -1,0 +1,22 @@
+#ifndef _ALGDIRECTLOOKUP_H_
+#define _ALGDIRECTLOOKUP_H_
+
+#include "algorithm.h"
+class AlgorithmDirectLookup : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    AlgorithmDirectLookup();
+    ~AlgorithmDirectLookup();
+
+protected:
+    uint32* table;
+    uint32 n;
+    uint32 sham;
+};
+
+#endif

### 2011-07-11 src/algHash.cpp

@@ -1,0 +1,73 @@
+#include "algHash.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+// http://cmph.sourceforge.net/
+// #include "cmph-1.0/src/cmph.h"
+
+Fingerprint2W::Fingerprint2W(NamedFunc* f) {
+    this->s = NULL;
+    this->n = 0;
+    this->H = NULL;
+    this->t = 0;
+    this->f = f;
+}
+
+Fingerprint2W::~Fingerprint2W() {
+}
+
+bool Fingerprint2W::preproc(TestString* s) {
+    uint32 t = this->f->f(s->n);
+    if (t < 1)
+        t = 1;
+
+    this->H = new (std::nothrow) uint32[s->n+1];
+    if (!this->H) {
+        return false;
+    }
+    this->s = s->s;
+    this->n = s->n;
+    this->t = t;
+
+    uint32 h;
+    // loop through each suffix s->sa[i] in sorted order.
+    for (uint32 i = 0; i < s->n; i++) {
+        // if suffix s->sa[i] does not have a prefix of length t in common with its predecessor, then update the current hash
+        if (s->lcp[i] < t)
+            h = i;
+        this->H[s->sa[i]] = h;
+    }
+    this->H[s->n] = s->n;
+
+    return true;
+}
+
+uint32 Fingerprint2W::query(uint32 i, uint32 j) {
+    uint32 t = 0;
+    if (i == j)
+        return this->n - i;
+    while (this->H[i + t] == this->H[j + t]) {
+        t += this->t;
+    }
+    while (this->s[i + t] == this->s[j + t]) {
+        t++;
+    }
+    return t;
+}
+
+void Fingerprint2W::cleanup() {
+    delete[] this->H;
+    this->s = NULL;
+    this->n = 0;
+    this->H = NULL;
+    this->t = 0;
+}
+
+void Fingerprint2W::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_2<%s>wc", this->f->name);
+}
+
+uint64 Fingerprint2W::spaceUsage(uint32 n) {
+    return sizeof(Fingerprint2W) + (n+1)*sizeof(uint32);
+}

### 2011-07-11 src/algHash.h

@@ -1,0 +1,25 @@
+#ifndef _ALGHASH_H_
+#define _ALGHASH_H_
+
+#include "algorithm.h"
+
+class Fingerprint2W : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint2W(NamedFunc*);
+    ~Fingerprint2W();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32* H;
+    uint32 t;
+    NamedFunc* f;
+};
+
+#endif

### 2011-07-11 src/algHash2LCache.cpp

@@ -1,0 +1,106 @@
+#include "algHash2LCache.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+Fingerprint2AQCacheHorizShift::Fingerprint2AQCacheHorizShift(NamedFunc* f) {
+    s = NULL;
+    n = 0;
+    H = NULL;
+    t = 0;
+    this->f = f;
+    S_CLEAR(t)
+    S_CLEAR(ndt);
+}
+
+// H[(i mod t)*ndt + i div t]
+#define HH(i) S_MUL(S_MOD(i, t), ndt) + S_DIV(i, t)
+
+Fingerprint2AQCacheHorizShift::~Fingerprint2AQCacheHorizShift() {
+}
+
+bool Fingerprint2AQCacheHorizShift::preproc(TestString* ts) {
+    s = ts->s;
+    n = ts->n;
+
+    t = f->f(n);
+    if (t < 1)
+        t = 1;
+    S_PREP_MOD(t)
+
+    // ndt = ceil((n+1)/t)
+    uint32 ndt = (n+1+t)/t;
+    S_PREP(ndt)
+
+    uint32 max = S_MUL(t - 1, ndt) + S_DIV(n, t);
+
+    H = new (std::nothrow) uint32[max+1];
+    if (!H) {
+        s = NULL;
+        n = 0;
+        t = 0;
+        S_CLEAR(t)
+        S_CLEAR(ndt)
+        return false;
+    }
+
+    uint32 h;
+    // loop through each suffix s->sa[i] in sorted order.
+    for (uint32 i = 0; i < n; i++) {
+        // if suffix s->sa[i] does not have a prefix of length t in common with its predecessor, then update the current hash
+        if (ts->lcp[i] < t)
+            h = i;
+        H[HH(ts->sa[i])] = h;
+    }
+    H[HH(n)] = n;
+
+    return true;
+}
+
+uint32 Fingerprint2AQCacheHorizShift::query(uint32 i, uint32 j) {
+    uint32 v = 0;
+    if (i == j)
+        return n - i;
+    if (s[i + v] == s[j + v]) {
+        uint32 ihh = HH(i);
+        uint32 jhh = HH(j);
+        while (H[ihh] == H[jhh]) {
+            v += t;
+            ihh++;
+            jhh++;
+        }
+        while (s[i + v] == s[j + v]) {
[…]

### 2011-07-11 src/algHash2LCache.h

@@ -1,0 +1,28 @@
+#ifndef _ALGHASH2LCACNE_H_
+#define _ALGHASH2LCACNE_H_
+
+#include "algorithm.h"
+#include "bittricks.h"
+
+class Fingerprint2AQCacheHorizShift : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint2AQCacheHorizShift(NamedFunc*);
+    ~Fingerprint2AQCacheHorizShift();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32* H;
+    uint32 t;
+    S_DECL_MOD(t)
+    S_DECL(ndt)
+    NamedFunc* f;
+};
+
+#endif

### 2011-07-11 src/algHash2LCacheMul.cpp

@@ -1,0 +1,96 @@
+#include "algHash2LCacheMul.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+Fingerprint2AQCacheHorizMult::Fingerprint2AQCacheHorizMult(NamedFunc* f) {
+    s = NULL;
+    n = 0;
+    H = NULL;
+    t = 0;
+    this->f = f;
+    ndt = 0;
+}
+
+#define HH(i) ((i)%t)*ndt + (i)/t
+
+Fingerprint2AQCacheHorizMult::~Fingerprint2AQCacheHorizMult() {
+}
+
+bool Fingerprint2AQCacheHorizMult::preproc(TestString* ts) {
+    s = ts->s;
+    n = ts->n;
+
+    t = f->f(n);
+    if (t < 1)
+        t = 1;
+
+    // ndt = ceil((n+1)/t)
+    ndt = (n+1+t)/t;
+
+    uint32 max = (t - 1)*ndt + n/t;
+
+    H = new (std::nothrow) uint32[max+1];
+    if (!H) {
+        s = NULL;
+        n = 0;
+        t = 0;
+        ndt = 0;
+        return false;
+    }
+
+    uint32 h;
+    // loop through each suffix s->sa[i] in sorted order.
+    for (uint32 i = 0; i < n; i++) {
+        // if suffix s->sa[i] does not have a prefix of length t in common with its predecessor, then update the current hash
+        if (ts->lcp[i] < t)
+            h = i;
+        H[HH(ts->sa[i])] = h;
+    }
+    H[HH(n)] = n;
+
+    return true;
+}
+
+uint32 Fingerprint2AQCacheHorizMult::query(uint32 i, uint32 j) {
+    uint32 v = 0;
+    if (i == j)
+        return n - i;
+    if (s[i + v] == s[j + v]) {
+        uint32 ihh = HH(i);
+        uint32 jhh = HH(j);
+        while (H[ihh] == H[jhh]) {
+            v += t;
+            ihh++;
+            jhh++;
+        }
+        while (s[i + v] == s[j + v]) {
+            v++;
+        }
+    }
+    return v;
+}
+
+void Fingerprint2AQCacheHorizMult::cleanup() {
+    delete[] H;
+    s = NULL;
+    n = 0;
+    H = NULL;
+    t = 0;
+    ndt = 0;
+}
+
[…]

### 2011-07-11 src/algHash2LCacheMul.h

@@ -1,0 +1,26 @@
+#ifndef _ALGHASH2LCACNEMUL_H_
+#define _ALGHASH2LCACNEMUL_H_
+
+#include "algorithm.h"
+
+class Fingerprint2AQCacheHorizMult : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint2AQCacheHorizMult(NamedFunc*);
+    ~Fingerprint2AQCacheHorizMult();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32* H;
+    uint32 t;
+    uint32 ndt;
+    NamedFunc* f;
+};
+
+#endif

### 2011-07-11 src/algHash2Ldq.cpp

@@ -1,0 +1,26 @@
+#include "algHash2Ldq.h"
+#include <stdlib.h>
+#include <stdio.h>
+
+Fingerprint2AQ::Fingerprint2AQ(NamedFunc* f) : Fingerprint2W(f) {
+}
+
+Fingerprint2AQ::~Fingerprint2AQ() {
+}
+
+uint32 Fingerprint2AQ::query(uint32 i, uint32 j) {
+    uint32 v = 0;
+    if (i == j)
+        return n - i;
+    if (s[i + v] == s[j + v]) {
+        while (H[i + v] == H[j + v])
+            v += t;
+        while (s[i + v] == s[j + v])
+            v++;
+    }
+    return v;
+}
+
+void Fingerprint2AQ::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_2<%s>ac", f->name);
+}

### 2011-07-11 src/algHash2Ldq.h

@@ -1,0 +1,15 @@
+#ifndef _ALGHASH2LDQ_H_
+#define _ALGHASH2LDQ_H_
+
+#include "algHash.h"
+
+class Fingerprint2AQ : public Fingerprint2W {
+public:
+    virtual uint32 query(uint32, uint32);
+    virtual void getName(char*, uint32);
+
+    Fingerprint2AQ(NamedFunc*);
+    ~Fingerprint2AQ();
+};
+
+#endif

### 2011-07-11 src/algHash3L.cpp

@@ -1,0 +1,96 @@
+#include "algHash3L.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+// http://cmph.sourceforge.net/
+// #include "cmph-1.0/src/cmph.h"
+
+Fingerprint3W::Fingerprint3W(NamedFunc* f2, NamedFunc* f1) {
+    this->s = NULL;
+    this->n = 0;
+    this->H1 = NULL;
+    this->t1 = 0;
+    this->f1 = f1;
+    this->H2 = NULL;
+    this->t2 = 0;
+    this->f2 = f2;
+}
+
+Fingerprint3W::~Fingerprint3W() {
+}
+
+bool Fingerprint3W::preproc(TestString* s) {
+    uint32 t1 = this->f1->f(s->n);
+    uint32 t2 = this->f2->f(s->n);
+    if (t1 < 1)
+        t1 = 1;
+    if (t2 < t1)
+        t2 = t1;
+
+    this->H1 = new (std::nothrow) uint32[s->n+1];
+    if (!this->H1) {
+        return false;
+    }
+    this->H2 = new (std::nothrow) uint32[s->n+1];
+    if (!this->H2) {
+        delete[] this->H1;
+        this->H1 = NULL;
+        return false;
+    }
+    this->s = s->s;
+    this->n = s->n;
+    this->t1 = t1;
+    this->t2 = t2;
+
+    uint32 h1, h2;
+    // loop through each suffix s->sa[i] in sorted order.
+    for (uint32 i = 0; i < s->n; i++) {
+        // if suffix s->sa[i] does not have a prefix of length t in common with its predecessor, then update the current hash
+        if (s->lcp[i] < t1)
+            h1 = i;
+        if (s->lcp[i] < t2)
+            h2 = i;
+        this->H1[s->sa[i]] = h1;
+        this->H2[s->sa[i]] = h2;
+    }
+    this->H1[s->n] = s->n;
+    this->H2[s->n] = s->n;
+
+    return true;
+}
+
+uint32 Fingerprint3W::query(uint32 i, uint32 j) {
+    uint32 t = 0;
+    if (i == j)
+        return this->n - i;
+    while (this->H2[i + t] == this->H2[j + t]) {
+        t += this->t2;
+    }
+    while (this->H1[i + t] == this->H1[j + t]) {
+        t += this->t1;
+    }
[…]

### 2011-07-11 src/algHash3L.h

@@ -1,0 +1,28 @@
+#ifndef _ALGHASH3L_H_
+#define _ALGHASH3L_H_
+
+#include "algorithm.h"
+
+class Fingerprint3W : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint3W(NamedFunc*, NamedFunc*);
+    ~Fingerprint3W();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32* H1;
+    uint32 t1;
+    NamedFunc* f1;
+    uint32* H2;
+    uint32 t2;
+    NamedFunc* f2;
+};
+
+#endif

### 2011-07-11 src/algHash3Ld.cpp

@@ -1,0 +1,71 @@
+#include "algHash3Ld.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+Fingerprint3AP::Fingerprint3AP(
+        NamedFunc* fd0,
+        NamedFunc* fd1,
+        NamedFunc* f2,
+        NamedFunc* f1)
+        : Fingerprint3W(f2, f1) {
+    this->td0 = 0;
+    this->fd0 = fd0;
+    this->td1 = 0;
+    this->fd1 = fd1;
+}
+
+Fingerprint3AP::~Fingerprint3AP() {
+}
+
+bool Fingerprint3AP::preproc(TestString* s) {
+    if (!Fingerprint3W::preproc(s))
+        return false;
+    td0 = fd0->f(s->n);
+    if (td0 < 1)
+        td0 = 1;
+    td1 = fd1->f(s->n);
+    if (td1 < td0)
+        td1 = td0;
+    return true;
+}
+
+uint32 Fingerprint3AP::query(uint32 i, uint32 j) {
+    uint32 t = 0;
+    if (i == j)
+        return this->n - i;
+
+    while (this->s[i + t] == this->s[j + t] && t < this->td0)
+        t++;
+    if (t >= this->td0) {
+
+        while (this->H1[i + t] == this->H1[j + t] && t < this->td1)
+            t += this->t1;
+        if (t >= this->td1) {
+
+            while (this->H2[i + t] == this->H2[j + t])
+                t += this->t2;
+
+            while (this->H1[i + t] == this->H1[j + t])
+                t += this->t1;
+        }
+
+        while (this->s[i + t] == this->s[j + t])
+            t++;
+    }
+    return t;
+}
+
+void Fingerprint3AP::cleanup() {
+    Fingerprint3W::cleanup();
+    td0 = 0;
+    td1 = 0;
+}
+
+void Fingerprint3AP::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_3<%s;%s>ac<%s;%s>", this->f2->name, this->f1->name, this->fd0->name, this->fd1->name);
+}
+
+uint64 Fingerprint3AP::spaceUsage(uint32 n) {
+    return sizeof(Fingerprint3AP) - sizeof(Fingerprint3W) + Fingerprint3W::spaceUsage(n);
+}

### 2011-07-11 src/algHash3Ld.h

@@ -1,0 +1,24 @@
+#ifndef _ALGHASH3LD_H_
+#define _ALGHASH3LD_H_
+
+#include "algHash3L.h"
+
+class Fingerprint3AP : public Fingerprint3W {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint3AP(NamedFunc*, NamedFunc*, NamedFunc*, NamedFunc*);
+    ~Fingerprint3AP();
+
+protected:
+    uint32 td0;
+    NamedFunc* fd0;
+    uint32 td1;
+    NamedFunc* fd1;
+};
+
+#endif

### 2011-07-11 src/algHash3Ldq.cpp

@@ -1,0 +1,40 @@
+#include "algHash3Ldq.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+Fingerprint3AQ::Fingerprint3AQ(
+        NamedFunc* f2,
+        NamedFunc* f1)
+        : Fingerprint3W(f2, f1) {
+}
+
+Fingerprint3AQ::~Fingerprint3AQ() {
+}
+
+uint32 Fingerprint3AQ::query(uint32 i, uint32 j) {
+    uint32 v = 0;
+    if (i == j)
+        return n - i;
+    if (s[i + v] == s[j + v]) {
+        v++;
+        if (H1[i + v] == H1[j + v]) {
+            v += t1;
+            while (H2[i + v] == H2[j + v])
+                v += t2;
+            while (H1[i + v] == H1[j + v])
+                v += t1;
+        }
+        while (s[i + v] == s[j + v])
+            v++;
+    }
+    return v;
+}
+
+void Fingerprint3AQ::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_3<%s;%s>ac", f2->name, f1->name);
+}
+
+uint64 Fingerprint3AQ::spaceUsage(uint32 n) {
+    return sizeof(Fingerprint3AQ) - sizeof(Fingerprint3W) + Fingerprint3W::spaceUsage(n);
+}

### 2011-07-11 src/algHash3Ldq.h

@@ -1,0 +1,16 @@
+#ifndef _ALGHASH3LDQ_H_
+#define _ALGHASH3LDQ_H_
+
+#include "algHash3L.h"
+
+class Fingerprint3AQ : public Fingerprint3W {
+public:
+    virtual uint32 query(uint32, uint32);
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint3AQ(NamedFunc*, NamedFunc*);
+    ~Fingerprint3AQ();
+};
+
+#endif

### 2011-07-11 src/algHashLog.cpp

@@ -1,0 +1,73 @@
+#include "algHashLog.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+#include <math.h>
+
+FingerprintLW::FingerprintLW() {
+    this->s = NULL;
+    this->n = 0;
+    this->H = NULL;
+    this->k = 0;
+}
+
+FingerprintLW::~FingerprintLW() {
+}
+
+bool FingerprintLW::preproc(TestString* s) {
+    uint32 k = (uint32) ceil(log2(s->n+1));
+
+    H = new (std::nothrow) uint32[(s->n+1)*k];
+    if (!H) {
+        return false;
+    }
+    this->s = s->s;
+    this->n = s->n;
+    this->k = k;
+
+    uint32 h;
+    for (uint32 c = 0; c < k; c++) {
+        // loop through each suffix s->sa[i] in sorted order.
+        for (uint32 i = 0; i < n; i++) {
+            // if suffix s->sa[i] does not have a prefix of length 1<<c in common with its predecessor, then update the current hash
+            if (s->lcp[i] < 1<<c)
+                h = i;
+            H[s->sa[i]] = h;
+        }
+        H[n] = n;
+        H += n+1;
+    }
+
+    return true;
+}
+
+uint32 FingerprintLW::query(uint32 i, uint32 j) {
+    if (i == j)
+        return this->n - i;
+    uint32 t = 0;
+    uint32* H = this->H;
+    for (int32 c = k-1; c >= 0; c--) {
+        H -= (n+1);
+        if (H[i + t] == H[j + t])
+            t += 1<<c;
+    }
+    return t;
+}
+
+void FingerprintLW::cleanup() {
+    H -= k*(n+1);
+    delete[] this->H;
+    this->s = NULL;
+    this->n = 0;
+    this->H = NULL;
+    this->k = 0;
+}
+
+void FingerprintLW::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_{log n}wc");
+}
+
+uint64 FingerprintLW::spaceUsage(uint32 n) {
+    uint32 k = (uint32) ceil(log2(n+1));
+    return sizeof(FingerprintLW) + (n+1)*k*sizeof(uint32);
+}

### 2011-07-11 src/algHashLog.h

@@ -1,0 +1,24 @@
+#ifndef _ALGHASHLOG_H_
+#define _ALGHASHLOG_H_
+
+#include "algorithm.h"
+
+class FingerprintLW : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    FingerprintLW();
+    ~FingerprintLW();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32* H;
+    uint32 k;
+};
+
+#endif

### 2011-07-11 src/algHashLogD.cpp

@@ -1,0 +1,47 @@
+#include "algHashLogD.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+#include <math.h>
+
+FingerprintLA::FingerprintLA() : FingerprintLW() {
+}
+
+FingerprintLA::~FingerprintLA() {
+}
+
+bool FingerprintLA::preproc(TestString* s) {
+    if (!FingerprintLW::preproc(s))
+        return false;
+    H -= k*(n+1);
+    return true;
+}
+
+uint32 FingerprintLA::query(uint32 i, uint32 j) {
+    if (i == j)
+        return this->n - i;
+    uint32 t = 0;
+    uint32* H = this->H;
+    int32 c;
+    for (c = 0; c < k; c++) {
+        if (H[i + t] == H[j + t])
+            t += 1<<c;
+        else break;
+        H += (n+1);
+    }
+    for (c--; c >= 0; c--) {
+        H -= (n+1);
+        if (H[i + t] == H[j + t])
+            t += 1<<c;
+    }
+    return t;
+}
+
+void FingerprintLA::cleanup() {
+    H += k*(n+1);
+    FingerprintLW::cleanup();
+}
+
+void FingerprintLA::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_{log n}ac");
+}

### 2011-07-11 src/algHashLogD.h

@@ -1,0 +1,17 @@
+#ifndef _ALGHASHLOGD_H_
+#define _ALGHASHLOGD_H_
+
+#include "algHashLog.h"
+
+class FingerprintLA : public FingerprintLW {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+
+    FingerprintLA();
+    ~FingerprintLA();
+};
+
+#endif

### 2011-07-11 src/algHashLogDCache.cpp

@@ -1,0 +1,75 @@
+#include "algHashLogDCache.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+#include <math.h>
+
+FingerprintLACacheVertiMult::FingerprintLACacheVertiMult() {
+    this->s = NULL;
+    this->n = 0;
+    this->H = NULL;
+    this->k = 0;
+}
+
+FingerprintLACacheVertiMult::~FingerprintLACacheVertiMult() {
+}
+
+bool FingerprintLACacheVertiMult::preproc(TestString* s) {
+    uint32 k = (uint32) ceil(log2(s->n+1));
+
+    H = new (std::nothrow) uint32[(s->n+1)*k];
+    if (!H) {
+        return false;
+    }
+    this->s = s->s;
+    this->n = s->n;
+    this->k = k;
+
+    uint32 h;
+    for (uint32 c = 0; c < k; c++) {
+        // loop through each suffix s->sa[i] in sorted order.
+        for (uint32 i = 0; i < n; i++) {
+            // if suffix s->sa[i] does not have a prefix of length 1<<c in common with its predecessor, then update the current hash
+            if (s->lcp[i] < 1<<c)
+                h = i;
+            H[c + (s->sa[i])*k] = h;
+        }
+        H[c + n*k] = n;
+    }
+
+    return true;
+}
+
+uint32 FingerprintLACacheVertiMult::query(uint32 i, uint32 j) {
+    if (i == j)
+        return this->n - i;
+    uint32 t = 0;
+    int32 c;
+    for (c = 0; c < k; c++) {
+        if (H[c + (i + t)*k] == H[c + (j + t)*k])
+            t += 1<<c;
+        else break;
+    }
+    for (c--; c >= 0; c--) {
+        if (H[c + (i + t)*k] == H[c + (j + t)*k])
+            t += 1<<c;
+    }
+    return t;
+}
+
+void FingerprintLACacheVertiMult::cleanup() {
+    delete[] this->H;
+    this->s = NULL;
+    this->n = 0;
+    this->H = NULL;
+    this->k = 0;
+}
+
+void FingerprintLACacheVertiMult::getName(char* name, uint32 size) {
+    snprintf(name, size, "Fingerprint_{log n}ac_{cache-vertical-mult}");
+}
+
+uint64 FingerprintLACacheVertiMult::spaceUsage(uint32 n) {
[…]

### 2011-07-11 src/algHashLogDCache.h

@@ -1,0 +1,24 @@
+#ifndef _ALGHASHLOGDCACHE_H_
+#define _ALGHASHLOGDCACHE_H_
+
+#include "algHashLog.h"
+
+class FingerprintLACacheVertiMult : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    FingerprintLACacheVertiMult();
+    ~FingerprintLACacheVertiMult();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32* H;
+    uint32 k;
+};
+
+#endif

### 2011-07-11 src/algRmq.cpp

@@ -1,0 +1,97 @@
+#include "algRmq.h"
+
+#include <stdio.h>
+#include <new>
+
+#include "RMQ_n_logn/RMQ_1_n.hpp"
+#include "RMQ_n_logn/RMQ_n_1_improved.hpp"
+#include "RMQ_n_logn/RMQ_n_logn_2.hpp"
+#include "RMQ_n_logn/RMQ_n_logn_4.hpp"
+//#include "RMQ_n_logn/RMQ_nlogn_1.hpp"
+#include "RMQ_n_logn/RMQ_succinct.hpp"
+#include "RMQ_n_logn/RMQ_n_logn.hpp"
+//#include "jf/optimalRMQ/optimalRMQ.hpp"
+
+AlgorithmRmq::AlgorithmRmq(enum RMQ_ALG alg) {
+    this->s = NULL;
+    this->rmq = NULL;
+    this->alg = alg;
+}
+
+AlgorithmRmq::~AlgorithmRmq() {
+}
+
+bool AlgorithmRmq::preproc(TestString* s) {
+    this->s = s;
+
+    switch (alg) {
+    case RMQ_1_N:
+        this->rmq = new (std::nothrow) RMQ_1_n((int*)s->lcp, s->n);
+        break;
+    case RMQ_N_1:
+        this->rmq = new (std::nothrow) RMQ_n_1_improved((int*)s->lcp, s->n);
+        break;
+    case RMQ_N_LOGN_2:
+        this->rmq = new (std::nothrow) RMQ_n_logn_2((int*)s->lcp, s->n);
+        break;
+    case RMQ_N_LOGN_4:
+        this->rmq = new (std::nothrow) RMQ_n_logn_4((int*)s->lcp, s->n);
+        break;
+    //case RMQ_NLOGN_1:
+    //    d->rmq = new RMQ_nlogn_1((int*)s->lcp, ???, s->n);
+    //    break;
+    case RMQ_SUCCINCT:
+        this->rmq = new (std::nothrow) RMQ_succinct((int*)s->lcp, s->n);
+        break;
+    case RMQ_N_LOGN_D2:
+        this->rmq = new (std::nothrow) RMQ_n_logn((int*)s->lcp, s->n, 2);
+        break;
+    case RMQ_N_LOGN_D4:
+        this->rmq = new (std::nothrow) RMQ_n_logn((int*)s->lcp, s->n, 4);
+        break;
+    case RMQ_OPTIMAL:
[…]

### 2011-07-11 src/algRmq.h

@@ -1,0 +1,25 @@
+#ifndef _ALGRMQ_H_
+#define _ALGRMQ_H_
+
+#include "algorithm.h"
+#include "RMQ_n_logn/RMQ.hpp"
+enum RMQ_ALG { RMQ_1_N, RMQ_N_1, RMQ_N_LOGN_2, RMQ_N_LOGN_4/*, RMQ_NLOGN_1*/, RMQ_SUCCINCT, RMQ_N_LOGN_D2, RMQ_N_LOGN_D4, RMQ_OPTIMAL};
+
+class AlgorithmRmq : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    AlgorithmRmq(enum RMQ_ALG);
+    ~AlgorithmRmq();
+
+protected:
+    TestString* s;
+    RMQ* rmq;
+    enum RMQ_ALG alg;
+};
+
+#endif

### 2011-07-11 src/algRmq1.cpp

@@ -1,0 +1,271 @@
+#include "algRmq1.h"
+
+#include <stdio.h>
+#include <new>
+#include <math.h>
+
+AlgorithmRmq1::AlgorithmRmq1() {
+}
+
+AlgorithmRmq1::~AlgorithmRmq1() {
+}
+
+void AlgorithmRmq1::getName(char* name, uint32 size) {
+    snprintf(name, size, "RMQ<n;1>");
+}
+
+uint64 AlgorithmRmq1::spaceUsage(uint32 n) {
+    uint32 bs = (uint32) ceil(log2(n) / 4);
+    uint32 nb = (n-1)/bs+1;
+    uint32 qpb = bs * (bs+1) / 2;
+    // log part
+    uint32 depth = log2fast(nb);
+
+    return sizeof(AlgorithmRmq1)
+        + n*sizeof(uint32) // sainv
+        + n*sizeof(uint32) // a
+        + sizeof(uint32)*nb // c
+        + sizeof(uint32)*nb // type
+        + sizeof(uint32*)*Catalan[bs][bs] // Prec
+        + sizeof(uint32)*qpb*Catalan[bs][bs] // Prec[i]
+        // log part
+        + sizeof(uint32*)*depth // logM
+        + sizeof(uint32)*nb*depth; // logM[i]
+
+/*
+    bs = ceil(log2(n) / 4);
+    nb = (n-1)/bs+1;
+    words = 2*n + 2*nb + (bs * (bs+1) / 2 + 1)*Catalan_bs + (nb+1)*log2fast(nb);
+*/
+}
+
+const uint32 AlgorithmRmq1::Catalan[17][17] = {
+    {1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1},
+    {0,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16},
+    {0,0,2,5,9,14,20,27,35,44,54,65,77,90,104,119,135},
+    {0,0,0,5,14,28,48,75,110,154,208,273,350,440,544,663,798},
+    {0,0,0,0,14,42,90,165,275,429,637,910,1260,1700,2244,2907,3705},
+    {0,0,0,0,0,42,132,297,572,1001,1638,2548,3808,5508,7752,10659,14364},
+    {0,0,0,0,0,0,132,429,1001,2002,3640,6188,9996,15504,23256,33915,48279},
[…]

### 2011-07-11 src/algRmq1.h

@@ -1,0 +1,83 @@
+#ifndef _ALGRMQ1_H_
+#define _ALGRMQ1_H_
+
+#include "algorithm.h"
+
+class AlgorithmRmq1 : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    AlgorithmRmq1();
+    ~AlgorithmRmq1();
+
+protected:
+    TestString* s;
+
+    // array
+    int32 *a;
+
+    // index array for the out-of-block queries (contains indices of block-minima)
+    uint32 *c;
+
+    // type of blocks
+    uint32 *type;
+
+    // precomputed in-block queries
+    uint32** Prec;
+
+    // number of different queries per block = bs*(bs+1)/2
+    uint32 qpb;
+
+    // size of array a
+    uint32 n;
+
+    // block size
+    uint32 bs;
+
+    // number of blocks (always n/bs)
+    uint32 nb;
+
+    // return block of entry i:
+    inline uint32 block(uint32 i) { return i/bs; }
+
+    // precomputed Catalan triangle (17 is enough for 64bit computing):
+    static const uint32 Catalan[17][17];
+
+    // minus infinity (change for 64bit version)
+    static const int32 minus_infinity;
+
+
+
+	void logpreproc(int32* a, uint32* c, uint32 n);
+
+	void logcleanup();
+
+	// the following stuff is for fast base 2 logarithms:
+	// (currently only implemented for 32 bit numbers)
+	static const char LogTable256[256];
+
+	virtual uint32 log2fast(uint32);
+
+	// array
+	int32* loga;
+
+	// index array for a:
+	uint32* logc;
+
+	// size of c
+	uint32 logn;
+
+	// depth of table
+	uint32 logdepth;
+
+	// the precomputed table:
+	uint32** logM;
+
+
+};
+
+#endif

### 2011-07-11 src/algRmq11.cpp

@@ -1,0 +1,76 @@
+#include "algRmq11.h"
+
+#include <stdio.h>
+#include <new>
+#include <math.h>
+
+AlgorithmRmq11::AlgorithmRmq11() {
+}
+
+AlgorithmRmq11::~AlgorithmRmq11() {
+}
+
+void AlgorithmRmq11::getName(char* name, uint32 size) {
+    snprintf(name, size, "DirectComp<1>RMQ<n;1>");
+}
+
+/*uint64 AlgorithmRmq11::spaceUsage(uint32 n) {
+    return sizeof(AlgorithmRmq1) + (-100); // TODO find out how to measure the RMQ structure
+}*/
+
+uint32 AlgorithmRmq11::query(uint32 i, uint32 j) {
+    if (s->s[i] != s->s[j])
+        return 0;
+
+    if (i == j)
+        return this->n - i;
+
+    i = this->s->sainv[i];
+    j = this->s->sainv[j];
+
+    if (i > j) {
+        uint32 m = i;
+        i = j;
+        j = m;
+    }
+
+    i++;
+
+
+    uint32 block_i = block(i);       // block where i is
+    uint32 block_j = block(j);       // j's block
+    uint32 i_pos = i - (block_i*bs); // position of i in block_i
+    uint32 j_pos = j - (block_j*bs); // pos. of j in block_j
+    uint32 min, min_i, min_j; 		// min: to be returned
+
+    if (block_j == block_i)
+        min = (block_i*bs) + Prec[type[block_i]][i_pos*(bs-1)+j_pos-((i_pos-1)*i_pos/2)];
+    else {
+        min_i = (block_i*bs) + Prec[type[block_i]][(i_pos+1)*(bs-1)-((i_pos-1)*i_pos/2)];
+        min_j = (block_j*bs) + Prec[type[block_j]][j_pos];
+        min = a[min_i] < a[min_j] ? min_i : min_j;
+        if (block_j > block_i + 1) { // out-of-block query:
+            i = block_i+1;
+            j = block_j-1;
+
+            if (j-i == 0) min_i = c[j];
+            else if (j-i == 1) min_i = c[logM[0][i]];
+            else {
+                uint32 k = log2fast(j-i);
+                uint32 twotothek = 1 << k; // 2^k
[…]

### 2011-07-11 src/algRmq11.h

@@ -1,0 +1,18 @@
+#ifndef _ALGRMQ11_H_
+#define _ALGRMQ11_H_
+
+#include "algRmq1.h"
+
+class AlgorithmRmq11 : public AlgorithmRmq1 {
+public:
+    //virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    //virtual void cleanup();
+    virtual void getName(char*, uint32);
+    //virtual uint64 spaceUsage(uint32);
+
+    AlgorithmRmq11();
+    ~AlgorithmRmq11();
+};
+
+#endif

### 2011-07-11 src/algRmqN.cpp

@@ -1,0 +1,51 @@
+#include "algRmqN.h"
+#include <stdio.h>
+
+AlgorithmRmqN::AlgorithmRmqN() {
+    this->s = NULL;
+}
+
+AlgorithmRmqN::~AlgorithmRmqN() {
+}
+
+bool AlgorithmRmqN::preproc(TestString* s) {
+    this->s = s;
+    return true;
+}
+
+uint32 AlgorithmRmqN::query(uint32 i, uint32 j) {
+    TestString* s = this->s;
+    uint32 min, x;
+
+    if (i == j)
+        return s->n - i;
+
+    i = s->sainv[i];
+    j = s->sainv[j];
+
+    if (i > j) {
+        min = i;
+        i = j;
+        j = min;
+    }
+
+    min = s->lcp[j];
+    for (x = i + 1; x < j; x++) {
+        if (s->lcp[x] < min) {
+            min = s->lcp[x];
+        }
+    }
+    return min;
+}
+
+void AlgorithmRmqN::cleanup() {
+    this->s = NULL;
+}
+
+void AlgorithmRmqN::getName(char* name, uint32 size) {
+    snprintf(name, size, "RMQ<1;n>");
+}
+
+uint64 AlgorithmRmqN::spaceUsage(uint32 n) {
+    return 2*n*sizeof(uint32); // from sainv and lcp
+}

### 2011-07-11 src/algRmqN.h

@@ -1,0 +1,20 @@
+#ifndef _ALGRMQN_H_
+#define _ALGRMQN_H_
+
+#include "algorithm.h"
+class AlgorithmRmqN : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    AlgorithmRmqN();
+    ~AlgorithmRmqN();
+
+protected:
+    TestString* s;
+};
+
+#endif

### 2011-07-11 src/algorithm.h

@@ -1,0 +1,57 @@
+#ifndef _ALGORITHM_H_
+#define _ALGORITHM_H_
+
+#include "string.h"
+
+# define ALGORITHM_NAME_SIZE 60
+
+class Algorithm {
+public:
+    /**
+     * Preprocess a string for fast LCE-queries.
+     *
+     * @param s A string
+     * @return True iff preprocessing was successful
+     */
+    virtual bool preproc(TestString*) = 0;
+
+    /**
+     * Query some preprocessed string for LCE.
+     *
+     * @param i The first index of interest in the query
+     * @param j The second index of interest in the query
+     * @return LCE(i,j) for the string represented by d
+     */
+    virtual uint32 query(uint32, uint32) = 0;
+
+    /**
+     * Free any resources allocated by LCE-preprocessing.
+     */
+    virtual void cleanup() = 0;
+    
+    /**
+     * A string representing the name of the algorithm.
+     *
+     * @param name A buffer in which to store the name
+     * @param size The size of the buffer in bytes
+     */
+    virtual void getName(char*, uint32) = 0;
+
+    /**
+     * Calculate how much space the preprocessed datastructure would use.
+     *
+     * @param n Length of the string, not including ending null
+     * @return Size of the preprocessed datastructure, in bytes
+     */
+    virtual uint64 spaceUsage(uint32) = 0;
+};
+
+// For used as parameter to an algorithm
+typedef uint32 (*IntIntFunc)(uint32);
+typedef struct {
+    IntIntFunc f;
+    const char* name;
+} NamedFunc;
+
+
+#endif

### 2011-07-11 src/bittricks.h

@@ -1,0 +1,41 @@
+#ifndef _BITTRICKS_H_
+#define _BITTRICKS_H_
+
+// multiplication, division and modulo bittricks for uint32
+
+// Declare auxiliary variables for variable "b", supporting MUL and DIV.
+#define S_DECL(b) uint32 b##_sham;
+
+// Declare auxiliary variables for variable "b", supporting MUL, DIV and MOD.
+#define S_DECL_MOD(b) S_DECL(b) uint32 b##_mask;
+
+// Change "b" to the smallest power of two, which is greater than "b", and
+// initialize auxiliary variables declared with S_DECL.
+#define S_PREP(b) b##_sham = 0;\
+    while (1u<<b##_sham < b)\
+        b##_sham++;\
+    b = 1u<<b##_sham;
+
+// Change "b" to the smallest power of two, which is greater than "b", and
+// initialize auxiliary variables declared with S_DECL_MOD.
+#define S_PREP_MOD(b) S_PREP(b) b##_mask = b - 1;
+
+// Modulo operation, where "a" is any arithmetic expression, and "b" is a
+// variable prepared with S_PREP_MOD.
+#define S_MOD(a, b) ((a)&b##_mask)
+
+// Multiplication, where "a" is any arithmetic expression, and "b" is a
+// variable prepared with S_PREP or S_PREP_MOD.
+#define S_MUL(a, b) ((a)<<b##_sham)
+
+// Division, where "a" is any arithmetic expression, and "b" is a
+// variable prepared with S_PREP or S_PREP_MOD.
+#define S_DIV(a, b) ((a)>>b##_sham)
+
+// Clear auxiliary variables initialized with S_DECL.
+#define S_CLEAR(b) b##_sham = 0;
+
+// Clear auxiliary variables initialized with S_DECL_MOD.
+#define S_CLEAR_MOD(b) S_CLEAR(b) b##_mask = 0;
+
+#endif

### 2011-07-11 src/buildSa.cpp

@@ -1,0 +1,56 @@
+#include <limits.h>
+#include "qsufsort.h"
+
+void TestString::buildSa() {
+    MY_INT *my_sa = new (std::nothrow) MY_INT[n + 1];
+    MY_INT *my_sainv = new (std::nothrow) MY_INT[n + 1];
+    sa = new (std::nothrow) uint32[n];
+    sainv = new (std::nothrow) uint32[n];
+    lcp = new (std::nothrow) uint32[n];
+    if (!(my_sa && my_sainv && sa && sainv && lcp)) {
+        if (my_sa) delete[] my_sa;
+        if (my_sainv) delete[] my_sainv;
+        if (sa) delete[] sa;
+        if (sainv) delete[] sainv;
+        if (lcp) delete[] lcp;
+        sa = NULL;
+        sainv = NULL;
+        lcp = NULL;
+        return;
+    }
+
+    MY_INT k = UCHAR_MAX + 1;
+    MY_INT l = 0;
+
+    for (uint32 j = 0; j < n; j++)
+        my_sainv[j] = s[j];
+    suffixsort(my_sainv, my_sa, MY_INT(n), k, l); // qsufsort.h
+
+    // transform from MY_INT to uint32
+    for (uint32 i = 0; i < n; i++) {
+        sa[i] = (uint32) my_sa[i + 1];
+        sainv[(uint32) my_sa[i + 1]] = i;
+    }
+    delete[] my_sa;
+    delete[] my_sainv;
+
+    // from code_LiviuTinta/LCE/manzini_ds/lcp_aux.c
+    // Compute the lcp array from the suffix array
+    // using the algorithm by Kasai et al. (CPM '01)
+
+    // rank = sa^{-1}
+    // traverse suffixes in rank order
+    uint32 h = 0;
+    for (uint32 i = 0; i < n; i++) {
+        uint32 k = sainv[i];          // rank of s[i ... n-1]
+        if (k > 0) {
+            uint32 j = sa[k-1];        // predecessor of s[i ... n-1]
+            while (i + h < n && j + h < n && s[i + h] == s[j + h])
+                h++;
+            lcp[k] = h;
+        }
+        if (h > 0) h--;
+    }
+
+    lcp[0] = 0;
+}

### 2011-07-11 src/cacheTest.cpp

@@ -1,0 +1,35 @@
+#include "cacheTest.h"
+#include "algorithm.h"
+#include "algRmq1.h"
+#include "testRandom.h"
+#include <iostream>
+#include <stdlib.h>
+
+void cacheTest(uint32 n, enum string_type type) {
+    std::cout << "Cachegrind test for string " << nameStr(type) << " length " << n << "\n";
+    Algorithm* rmq1 = new AlgorithmRmq1;
+
+    uint32 paircount = 1000000;
+    uint32* pairs = new (std::nothrow) uint32[paircount*2];
+    if (!pairs) {
+        delete rmq1;
+        std::cout << "Cannot allocate pairs\n";
+        return;
+    }
+
+    for (uint32 i = 0; i < paircount*2; i += 2) {
+        pairs[i] = rand() % n;
+        pairs[i + 1] = rand() % n;
+    }
+
+    TestString* s = new (std::nothrow) TestString(n, type);
+    if (!checkStr(s)) {
+        delete rmq1;
+        delete pairs;
+        std::cout << "Cannot allocate string\n";
+        return;
+    }
+    
+    double time = testRandom(s, pairs, paircount, rmq1, false, 10000);
+    std::cout << "Time was " << time << ".\n";
+}

### 2011-07-11 src/cacheTest.h

@@ -1,0 +1,7 @@
+#ifndef _CACHETEST_H_
+#define _CACHETEST_H_
+
+#include "string.h"
+void cacheTest(uint32 n, enum string_type type);
+
+#endif

### 2011-07-11 src/cachegrind.sh

@@ -1,0 +1,13 @@
+# http://www.majorgeeks.com/CPU-Z_d425.html
+# http://www.cpuid.com/softwares/cpu-z/versions-history.html
+# Caches: L1 D: 32KB,8-way,64B lines ; L1 I: 32KB,8-way,64B lines ; L2: 3072KB,12-way,64B lines
+CXXFLAGS="-DCACHE_ALLA=1" make main -s
+../valgrind/valgrind-3.6.1/vg-in-place --tool=cachegrind --D1=32768,8,64 --I1=32768,8,64 --LL=3145728,12,64 --cachegrind-out-file=cachegrind-alla.out main.exe > cachegrind-alla.txt 2> cachegrind-alla-err.txt
+#../valgrind/valgrind-3.6.1/vg-in-place --tool=cachegrind --D1=32768,8,64 --I1=32768,8,64 --LL=1073741824,262144,4096 --cachegrind-out-file=cachegrind-swap-alla.out main.exe > cachegrind-swap-alla.txt 2> cachegrind-swap-alla-err.txt
+CXXFLAGS="-DCACHE_RAND10=1" make main -s
+../valgrind/valgrind-3.6.1/vg-in-place --tool=cachegrind --D1=32768,8,64 --I1=32768,8,64 --LL=3145728,12,64 --cachegrind-out-file=cachegrind-rand10.out main.exe > cachegrind-rand10.txt 2> cachegrind-rand10-err.txt
+#../valgrind/valgrind-3.6.1/vg-in-place --tool=cachegrind --D1=32768,8,64 --I1=32768,8,64 --LL=1073741824,262144,4096 --cachegrind-out-file=cachegrind-swap-rand10.out main.exe > cachegrind-swap-rand10.txt 2> cachegrind-swap-rand10-err.txt
+../valgrind/valgrind-3.6.1/cachegrind/cg_annotate cachegrind-alla.out --auto=yes > cachegrind-alla-annotate.txt
+../valgrind/valgrind-3.6.1/cachegrind/cg_annotate cachegrind-rand10.out --auto=yes > cachegrind-rand10-annotate.txt
+#../valgrind/valgrind-3.6.1/cachegrind/cg_annotate cachegrind-swap-alla.out --auto=yes > cachegrind-swap-alla-annotate.txt
+#../valgrind/valgrind-3.6.1/cachegrind/cg_annotate cachegrind-swap-rand10.out --auto=yes > cachegrind-swap-rand10-annotate.txt

### 2011-07-11 src/code_LiviuTinta.sh

@@ -1,0 +1,23 @@
+cd manzini_ds/
+make da_ssort.a 
+cp ds_ssort.a ../ds_ssort.a
+make bwtlcp.a
+cp bwtlcp.a ../bwtlcp.a
+cd ..
+cd RMQ_n_logn/
+make RMQ.a
+cp RMQ.a ../RMQ.a
+cd ..
+cd qsufsort/
+make qsufsort.a
+cp qsufsort.a ../qsufsort.a
+cd ..
+cd compute_avg_lcp
+make compute_avg_lcp.a
+cp compute_avg_lcp.a ../compute_avg_lcp.a
+cd ..
+cd SA_Bentley\&Sedgewick/
+make SABentleySedgewick.a
+cp SABentleySedgewick.a ../SABentleySedgewick.a
+cd ..
+make

### 2011-07-11 src/correctness.cpp

@@ -1,0 +1,43 @@
+#include "correctness.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+#include "string.h"
+
+void correctness(Algorithm** algs, enum string_type type) {
+    uint32 x;
+    uint32 alg;
+
+    TestString* s = new (std::nothrow) TestString(100, type);
+    if (!checkStr(s))
+        return;
+    for (alg = 0; algs[alg]; alg++) {
+        bool success = algs[alg]->preproc(s);
+        if (!success) {
+            printf("CORRECTNESS ERROR alg=%i, OOM\n", alg);
+            for (x = 0; x < alg; x++) {
+                algs[x]->cleanup();
+            }
+            return;
+        }
+    }
+
+    for (x = 0; x < 10000; x++) {
+        uint32 i = rand() % s->n;
+        uint32 j = rand() % s->n;
+        uint32 lce0 = algs[0]->query(i, j);
+        for (alg = 1; algs[alg]; alg++) {
+            uint32 lce1 = algs[alg]->query(i, j);
+            if (lce0 != lce1) {
+                printf("CORRECTNESS ERROR alg=%i, i=%i, j=%i, lce0=%i, lce1=%i\n", alg, i, j, lce0, lce1);
+                return;
+            }
+        }
+    }
+    for (alg = 0; algs[alg]; alg++) {
+        algs[alg]->cleanup();
+    }
+    delete s;
+    printf("correct\n");
+}

### 2011-07-11 src/correctness.h

@@ -1,0 +1,7 @@
+#ifndef _CORRECTNESS_H_
+#define _CORRECTNESS_H_
+
+#include "algorithm.h"
+void correctness(Algorithm** algs, enum string_type type);
+
+#endif

### 2011-07-11 src/functionalTest.cpp

@@ -1,0 +1,37 @@
+#include "functionalTest.h"
+#include "string.h"
+#include "testRandom.h"
+#include <new>
+#include <stdlib.h>
+
+void functionalTest(Algorithm** algs) {
+    uint32 paircount = 1000000;
+    uint32 n = 1000;
+    uint32 j;
+
+    uint32* pairs = new (std::nothrow) uint32[paircount*2];
+    if (!pairs) {
+        return;
+    }
+    for (uint32 i = 0; i < paircount*2; i += 2) {
+        pairs[i] = rand() % n;
+        pairs[i + 1] = rand() % n;
+    }
+
+    TestString* s = new (std::nothrow) TestString(n, ALL_A_STRING);
+    if (!checkStr(s)) {
+        delete[] pairs;
+        return;
+    }
+    for (j = 0; algs[j]; j++)
+        testRandom(s, pairs, paircount, algs[j], true, 1);
+    delete s;
+
+    s = new (std::nothrow) TestString(n, RANDOM10_STRING);
+    if (!checkStr(s))
+        return;
+    for (j = 0; algs[j]; j++)
+        testRandom(s, pairs, paircount, algs[j], true, 1);
+    delete s;
+    delete[] pairs;
+}

### 2011-07-11 src/functionalTest.h

@@ -1,0 +1,7 @@
+#ifndef _FUNCTIONALTEST_H_
+#define _FUNCTIONALTEST_H_
+
+#include "algorithm.h"
+void functionalTest(Algorithm** algs);
+
+#endif

### 2011-07-11 src/graphBySaDistance.cpp

@@ -1,0 +1,80 @@
+#include "graphBySaDistance.h"
+#include "timer.h"
+
+#include <iostream>
+#include <string>
+#include <vector>
+#include <algorithm>
+#include <functional>
+#include <iterator>
+
+void shuffle_array (std::pair<uint32, uint32>* const a, const uint32 length) {
+    for (int32 i = length - 1; i >= 0; i--) {
+        const uint32 index = rand() % (i + 1);
+        const std::pair<uint32, uint32> tmp = a[index];
+        a[index] = a[i];
+        a[i] = tmp;
+    }
+}
+
+void testBySaDistance(TestString* s, Algorithm* a) {
+    const uint32 TIMES = 3;
+
+    double start, end;
+    bool success = a->preproc(s);
+    if (!success)
+        return;
+
+    // now compute lce at each step distance
+    char algName[ALGORITHM_NAME_SIZE];
+    a->getName(algName, ALGORITHM_NAME_SIZE);
+    std::cout << algName << ",";
+
+    for (uint32 step = 2; step < s->n; step = step << 1) {
+
+        std::vector< std::pair<uint32, uint32> > all_pairs;
+        uint32 i;
+
+        for (i = 0; i + step < s->n; i++) {
+            std::pair<uint32, uint32> p;
+            p.first = s->sa[i];
+            p.second= s->sa[i + step];
+            if (p.first > p.second)
+                std::swap(p.first, p.second);
+            all_pairs.push_back(p);
+        }
+
+        double avg = 0;
+
+        for (uint32 k = 0; k < TIMES; k++) {
+            shuffle_array(&(all_pairs[0]), all_pairs.size());
+
+            start = gettime();
+            for (uint32 j = 0; j < i; j++) {
+                a->query(all_pairs[j].first, all_pairs[j].second);
+            }
+            end = gettime();
+            avg += end - start;
+        }
+        std::cout << avg / TIMES / i << ",";
+    }
+
+    a->cleanup();
+    std::cout << std::endl;
+}
+
+void graphBySaDistance(Algorithm** algs, enum string_type type) {
+    TestString* s = new (std::nothrow) TestString(1000, type);
+    if (!checkStr(s))
+        return;
+
+    std::cout << std::endl << nameStr(type) << " n=1000 string by SA distance,";
+    for (uint32 step = 2; step < s->n; step = step << 1) {
+        std::cout << step << ",";
+    }
[…]

### 2011-07-11 src/graphBySaDistance.h

@@ -1,0 +1,8 @@
+#ifndef _GRAPHBYSADISTANCE_H_
+#define _GRAPHBYSADISTANCE_H_
+
+#include "algorithm.h"
+#include "string.h"
+void graphBySaDistance(Algorithm** algs, enum string_type type);
+
+#endif

### 2011-07-11 src/graphIncreasingLength.cpp

@@ -1,0 +1,79 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include "graphIncreasingLength.h"
+#include "testRandom.h"
+#include <new>
+#include <iostream>
+#include <fstream>
+
+void graphIncreasingLength(Algorithm** algs, enum string_type type, const char* output, double preproc_time_limit, double query_time_limit, uint32 max_n) {
+    uint32 paircount = 1000000;
+    uint32 algcount;
+    for (algcount = 0; algs[algcount]; algcount++);
+    bool* ban = new (std::nothrow) bool[algcount];
+    double time;
+    char algName[ALGORITHM_NAME_SIZE];
+
+    if (!ban)
+        return;
+
+    uint32* pairs = new (std::nothrow) uint32[paircount*2];
+    if (!pairs) {
+        delete[] ban;
+        return;
+    }
+
+    std::ofstream f(output);
+
+    f << nameStr(type) << " string increasing length,";
+    std::cout <<"\n" << nameStr(type) << " string increasing length,";
+
+    for (uint32 alg = 0; algs[alg]; alg++) {
+        algs[alg]->getName(algName, ALGORITHM_NAME_SIZE);
+        f << algName << ",";
+        std::cout << algName << ",";
+        ban[alg] = false;
+    }
+    f << "\n";
+    std::cout << "\n";
+
+    for (uint32 n = 2; n <= max_n; n = n<<1) {
+        TestString* s = new (std::nothrow) TestString(n, type);
+        if (!checkStr(s)) {
+            delete[] ban;
+            delete[] pairs;
+            f.close();
+            return;
+        }
+        for (uint32 i = 0; i < paircount*2; i += 2) {
+            pairs[i] = rand() % n;
+            pairs[i + 1] = rand() % n;
+        }
+        f << n << ",";
+        std::cout << n << ",";
+        for (uint32 alg = 0; algs[alg]; alg++) {
+            if (ban[alg]) {
+                f << ",";
+                std::cout << ",";
+            } else {
+                time = testRandom(s, pairs, paircount, algs[alg], false, preproc_time_limit);
+                if (time < 0.0 || time > query_time_limit) {
+                    ban[alg] = true;
+                }
+                if (time < 0.0) {
+                    f << ",";
+                    std::cout << ",";
+                } else {
+                    f << time << ",";
+                    std::cout << time << ",";
+                }
+            }
+        }
+        f << "\n";
[…]

### 2011-07-11 src/graphIncreasingLength.h

@@ -1,0 +1,8 @@
+#ifndef _GRAPHINCREASINGLENGTH_H_
+#define _GRAPHINCREASINGLENGTH_H_
+
+#include "algorithm.h"
+#include "string.h"
+void graphIncreasingLength(Algorithm** algs, enum string_type type, const char* output, double preproc_time_limit, double query_time_limit, uint32 max_n);
+
+#endif

### 2011-07-11 src/graphSpaceIncreasingLength.cpp

@@ -1,0 +1,31 @@
+#include "graphSpaceIncreasingLength.h"
+#include <iostream>
+#include <fstream>
+
+void graphSpaceIncreasingLength(Algorithm** algs, const char* output, uint32 max_n) {
+    std::ofstream f(output);
+
+    f << "Space usage increasing length,";
+    std::cout << "\nSpace usage increasing length,";
+
+    for (uint32 alg = 0; algs[alg]; alg++) {
+        char algName[ALGORITHM_NAME_SIZE];
+        algs[alg]->getName(algName, ALGORITHM_NAME_SIZE);
+        f << algName << ",";
+        std::cout << algName << ",";
+    }
+    f << "\n";
+    std::cout << "\n";
+
+    for (uint32 i = 2; i <= max_n; i = i<<1) {
+        f << i << ",";
+        std::cout << i << ",";
+        for (uint32 alg = 0; algs[alg]; alg++) {
+            f << (double) algs[alg]->spaceUsage(i) << ",";
+            std::cout << (double) algs[alg]->spaceUsage(i) << ",";
+        }
+        f << "\n";
+        std::cout << "\n";
+    }
+    f.close();
+}

### 2011-07-11 src/graphSpaceIncreasingLength.h

@@ -1,0 +1,7 @@
+#ifndef _GRAPHSPACEINCREASINGLENGTH_H_
+#define _GRAPHSPACEINCREASINGLENGTH_H_
+
+#include "algorithm.h"
+void graphSpaceIncreasingLength(Algorithm** algs, const char* output, uint32 max_n);
+
+#endif

### 2011-07-11 src/graphValue.cpp

@@ -1,0 +1,119 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include "graphValue.h"
+#include <new>
+#include <iostream>
+#include <fstream>
+#include "timer.h"
+#include <vector>
+
+void sort(double* array, uint32 length) {
+    for (uint32 i = 0; i < length; i++) {
+        uint32 min = i;
+        for (uint32 j = i + 1; j < length; j++) {
+            if (array[j] > array[min])
+                min = j;
+        }
+        double temp = array[min];
+        array[min] = array[i];
+        array[i] = temp;
+    }
+}
+
+void graphValue(Algorithm** algs, enum string_type type, uint32 n, const char* output, uint32 repeats) {
+    uint32 count = 100000000;
+
+    uint32 algcount;
+    for (algcount = 0; algs[algcount]; algcount++);
+
+    double* valueSum = new (std::nothrow) double[algcount*(n+1)];
+    if (!valueSum)
+        return;
+
+    std::ofstream f(output);
+
+    f << nameStr(type) << " string length " << n << " by LCE value,";
+    std::cout <<"\n" << nameStr(type) << " string length " << n << " by LCE value,";
+
+    TestString* s = new (std::nothrow) TestString(n, type);
+    if (!checkStr(s)) {
+        delete[] valueSum;
+        f.close();
+        return;
+    }
+
+    bool success = algs[0]->preproc(s);
+    if (!success) {
+        delete s;
+        delete[] valueSum;
+        f.close();
+        return;
+    }
+    // forall (i,j): (i,j) isin all_pairs[v] => LCE(i,j) = v
+    std::vector< std::pair<uint32, uint32> >* all_pairs = new (std::nothrow) std::vector< std::pair<uint32, uint32> >[n+1];
+    if (!all_pairs) {
+        algs[0]->cleanup();
+        delete s;
+        delete[] valueSum;
+        f.close();
+        return;
+    }
+    for (uint32 i = 0; i < count; i++) {
+        std::pair<uint32, uint32> p;
+        p.first = rand() % n;
+        p.second = rand() % n;
+        uint32 v = algs[0]->query(p.first, p.second);
+        if (all_pairs[v].size() < repeats)
+            all_pairs[v].push_back(p);
+    }
+    algs[0]->cleanup();
+
+    for (uint32 alg = 0; algs[alg]; alg++) {
+        Algorithm* a = algs[alg];
[…]

### 2011-07-11 src/graphValue.h

@@ -1,0 +1,8 @@
+#ifndef _GRAPHVALUE_H_
+#define _GRAPHVALUE_H_
+
+#include "algorithm.h"
+#include "string.h"
+void graphValue(Algorithm** algs, enum string_type type, uint32 n, const char* output, uint32 repeats);
+
+#endif

### 2011-07-11 src/main.cpp

@@ -1,0 +1,165 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <math.h>
+
+#include "algorithm.h"
+
+#include "algDirectComp.h"
+#include "algDirectLookup.h"
+#include "algDirectCompLookup.h"
+#include "algHash.h"
+#include "algHash2Ldq.h"
+#include "algHash2LCache.h"
+#include "algHash2LCacheMul.h"
+#include "algHash3L.h"
+#include "algHash3Ld.h"
+#include "algHash3Ldq.h"
+#include "algHashLog.h"
+#include "algHashLogD.h"
+#include "algHashLogDCache.h"
+#include "algRmqN.h"
+#include "algRmq1.h"
+#include "algRmq11.h"
+#ifndef _MSC_VER
+#include "algRmq.h"
+#endif
+
+#include "correctness.h"
+#include "functionalTest.h"
+#include "graphIncreasingLength.h"
+#include "graphSpaceIncreasingLength.h"
+#include "graphValue.h"
+#include "tableFiles.h"
+#include "graphBySaDistance.h"
+#include "cacheTest.h"
+#include "plotDirectCompLookup.h"
+
+#ifdef _MSC_VER
+double log2(double n) {
+    return log(n) / log(2.0);  
+}
+#endif
+
+uint32 const1f(uint32 _) { return 1; } NamedFunc const1 = {const1f, "1"};
+uint32 const10f(uint32 _) { return 10; } NamedFunc const10 = {const10f, "10"};
+uint32 const20f(uint32 _) { return 20; } NamedFunc const20 = {const20f, "20"};
+uint32 const100f(uint32 _) { return 100; } NamedFunc const100 = {const100f, "100"};
+uint32 sqrt_nf(uint32 n) { return (uint32) sqrt((double)n); } NamedFunc sqrt_n = {sqrt_nf, "sqrt n"};
+uint32 sqrt_n_d2f(uint32 n) { return (uint32) (.5*sqrt((double)n)); } NamedFunc sqrt_n_d2 = {sqrt_n_d2f, "1/2sqrt n"};
+uint32 sqrt_n_d4f(uint32 n) { return (uint32) (.25*sqrt((double)n)); } NamedFunc sqrt_n_d4 = {sqrt_n_d4f, "1/4sqrt n"};
+uint32 sqrt_n_2f(uint32 n) { return (uint32) (2*sqrt((double)n)); } NamedFunc sqrt_n_2 = {sqrt_n_2f, "2sqrt n"};
[…]

### 2011-07-11 src/msvc.sh

@@ -1,0 +1,31 @@
+cl -c qsufsort.c -Fo -Zi -Yd
+cl -Zi -Yd \
+		main.cpp \
+		correctness.cpp \
+		timer.cpp \
+		algDirectComp.cpp \
+		algDirectLookup.cpp \
+		algDirectCompLookup.cpp \
+		algHash.cpp \
+		algHash2Ldq.cpp \
+		algHash2LCache.cpp \
+		algHash2LCacheMul.cpp \
+		algHash3L.cpp \
+		algHash3Ld.cpp \
+		algHash3Ldq.cpp \
+		algHashLog.cpp \
+		algHashLogD.cpp \
+		algHashLogDCache.cpp \
+		algRmqN.cpp \
+		algRmq1.cpp \
+		algRmq11.cpp \
+		testRandom.cpp \
+		graphIncreasingLength.cpp \
+		graphSpaceIncreasingLength.cpp \
+		graphValue.cpp \
+		functionalTest.cpp \
+		str.cpp \
+		graphBySaDistance.cpp \
+		qsufsort.obj
+
+# VCExpress.exe -debugexe main.exe

### 2011-07-11 src/plot2.sh

@@ -1,0 +1,219 @@
+
+oneplot()
+{
+    cat plotheader.txt > $plotfile
+    echo "set out '$plotfile.eps'" >> $plotfile
+    echo "set xrange [10:]" >> $plotfile
+    title=$(head -1 $datafile | sed -e 's/,/\n/g' | head -1)
+    # echo "set title '$title'" >> $plotfile
+    echo -e "$xtra" >> $plotfile
+    echo "plot \\" >> $plotfile
+    first=1
+    for algo in "$@"
+    do
+        if test $first = 0
+        then
+            echo ", \\" >> $plotfile
+        fi
+        first=0
+        algoidx=$(head -1 $datafile | sed -e 's/,/,\n/g' | grep -nF "$algo" -m1 | cut -d':' -f1)
+        echo -n "  '$datafile' using 1:$algoidx with $style" >> $plotfile
+    done
+    echo "Plotting $plotfile";
+    gnuplot $plotfile; ps2pdf -dEPSCrop $plotfile.eps $plotfile.pdf
+}
+
+listdata()
+{
+    head -1 $datafile | sed -e 's/,/,\n/g'
+    echo ""
+}
+
+rm results/*.pdf
+
+datafile="results/length-rand10.txt"
+listdata
+
+yavg="set logscale y\nset yrange [7e-9:1e-6]\nset title 'Average case strings'"
+ywc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Worst case strings'"
+ymc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Medium LCE-size strings'"
+
+style="lines"
+
+# x-axis = LCE value
+xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:10]"
+datafile="results/value-rand4.txt"
+plotfile="results/value-article-rand4.plt"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>wc," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}wc," "Fingerprint_{log n}ac," "RMQ<n;1>," "DirectComp<1>RMQ<n;1>,"
+
+xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:1000]"
[…]

### 2011-07-11 src/plotDirectCompLookup.cpp

@@ -1,0 +1,42 @@
+#include "plotDirectCompLookup.h"
+#include "algorithm.h"
+#include "algDirectCompLookup.h"
+#include "testRandom.h"
+#include <new>
+#include <stdlib.h>
+#include <iostream>
+#include <fstream>
+
+uint32 idf(uint32 p) { return p; } NamedFunc id = {idf, "id"};
+
+void plotDirectCompLookup(enum string_type type, const char* output, uint32 n, uint32 t_step, uint32 preproc_time_limit) {
+    uint32 paircount = 1000000;
+    uint32* pairs = new (std::nothrow) uint32[paircount*2];
+    if (!pairs) {
+        return;
+    }
+
+    std::ofstream f(output);
+
+    f << "DirectCompLookup tests,DirectCompLookup<t>,\n";
+    std::cout <<"\nDirectCompLookup tests,DirectCompLookup<t>,\n";
+
+    for (uint32 i = 0; i < paircount*2; i += 2) {
+        pairs[i] = rand() % n;
+        pairs[i + 1] = rand() % n;
+    }
+    for (uint32 t = 1; t <= n; t += t_step) {
+        TestString* s = new (std::nothrow) TestString(n, type);
+        if (!checkStr(s)) {
+            delete[] pairs;
+            f.close();
+            return;
+        }
+        AlgorithmDirectCompLookup alg(&id, t);
+        double time = testRandom(s, pairs, paircount, &alg, false, preproc_time_limit);
+        f << t << "," << time << ",\n";
+        std::cout << t << "," << time << ",\n";
+        delete s;
+    }
+    f.close();
+}

### 2011-07-11 src/plotDirectCompLookup.h

@@ -1,0 +1,7 @@
+#ifndef _PLOTDIRECTCOMPLOOKUP_H_
+#define _PLOTDIRECTCOMPLOOKUP_H_
+
+#include "string.h"
+void plotDirectCompLookup(enum string_type type, const char* output, uint32 n, uint32 t_step, uint32 preproc_time_limit);
+
+#endif

### 2011-07-11 src/qsufsort.c

@@ -1,0 +1,307 @@
+/* qsufsort.c
+   Copyright 1999, N. Jesper Larsson, all rights reserved.
+
+   This file contains an implementation of the algorithm presented in "Faster
+   Suffix Sorting" by N. Jesper Larsson (jesper@cs.lth.se) and Kunihiko
+   Sadakane (sada@is.s.u-tokyo.ac.jp).
+
+   This software may be used freely for any purpose. However, when distributed,
+   the original source must be clearly stated, and, when the source code is
+   distributed, the copyright notice must be retained and any alterations in
+   the code must be clearly marked. No warranty is given regarding the quality
+   of this software.*/
+
+#include <limits.h>
+
+// included qsufsort
+#include "qsufsort.h" 
+
+
+
+static MY_INT *I,                  /* group array, ultimately suffix array.*/
+   *V,                          /* inverse array, ultimately inverse of I.*/
+   r,                           /* number of symbols aggregated by transform.*/
+   h;                           /* length of already-sorted prefixes.*/
+
+#define KEY(p)          (V[*(p)+(h)])
+#define SWAP(p, q)      (tmp=*(p), *(p)=*(q), *(q)=tmp)
+#define MED3(a, b, c)   (KEY(a)<KEY(b) ?                        \
+        (KEY(b)<KEY(c) ? (b) : KEY(a)<KEY(c) ? (c) : (a))       \
+        : (KEY(b)>KEY(c) ? (b) : KEY(a)>KEY(c) ? (c) : (a)))
+
+/* Subroutine for select_sort_split and sort_split. Sets group numbers for a
+   group whose lowest position in I is pl and highest position is pm.*/
+
+static void update_group(MY_INT *pl, MY_INT *pm)
+{
+   MY_INT g;
+
+   g=pm-I;                      /* group number.*/
+   V[*pl]=g;                    /* update group number of first position.*/
+   if (pl==pm)
+      *pl=-1;                   /* one element, sorted group.*/
+   else
+      do                        /* more than one element, unsorted group.*/
+         V[*++pl]=g;            /* update group numbers.*/
+      while (pl<pm);
+}
+
+/* Quadratic sorting method to use for small subarrays. To be able to update
+   group numbers consistently, a variant of selection sorting is used.*/
+
[…]

### 2011-07-11 src/qsufsort.h

@@ -1,0 +1,36 @@
+/*
+ Written by: Liviu Tinta
+ Purpose: to access function suffixsort from the qsufsort.a library
+*/
+
+#ifndef _QSUFSORT_H_
+#define _QSUFSORT_H_
+
+#ifdef __cplusplus
+extern "C" {
+#endif
+
+// Modified on April 26, 2009 by Liviu Tinta
+
+// Original Jesper Larsson values
+//typedef int MY_INT;
+//#define MY_INT_MAX INT_MAX 
+
+#include "types.h" // includes <stdint.h>
+typedef intptr_t MY_INT;
+#define MY_INT_MAX INTPTR_MAX
+
+// End modification
+
+/* Makes suffix array p of x. x becomes inverse of p. p and x are both of size
+   n+1. Contents of x[0...n-1] are integers in the range l...k-1. Original
+   contents of x[n] is disregarded, the n-th symbol being regarded as
+   end-of-string smaller than all other symbols.*/
+
+void suffixsort(MY_INT *x, MY_INT *p, MY_INT n, MY_INT k, MY_INT l);
+
+#ifdef  __cplusplus
+}
+#endif
+
+#endif

### 2011-07-11 src/str.cpp

@@ -1,0 +1,145 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include "string.h"
+#include <new>
+#include <iostream>
+#include <math.h>
+
+#include "buildSa.cpp"
+TestString::~TestString() {
+    if (s)
+        delete[] s;
+    if (sa)
+        delete[] sa;
+    if (sa)
+        delete[] sainv;
+    if (lcp)
+        delete[] lcp;
+}
+
+const char* nameStr(enum string_type type) {
+    switch (type) {
+    case RANDOM2_STRING:
+        return "Random-2";
+    case RANDOM4_STRING:
+        return "Random-4";
+    case RANDOM10_STRING:
+        return "Random-10";
+    case RANDOM255_STRING:
+        return "Random-255";
+    case ALL_A_STRING:
+        return "All ´a´";
+    case RANDOM10_REPEAT10_STRING:
+        return "Random-10-Repeat-10";
+    case REPEAT_POW_STRING:
+        return "Repeat-pow";
+    default:
+        return "UNKNOWN";
+    }
+}
+
+/**
+ * Initializes a pre-allocated buffer s->s of size n+1 with a string of
+ * length s->n with characters from the alphabet [1..mod], where each character
+ * is chosen uniformly at random.
+ */
+void strRandom(TestString* s, uint8 mod, bool print) {
+    uint32 i;
+    if (print) std::cout << "String: ";
+    for (i = 0; i < s->n; i++) {
+        s->s[i] = rand() % mod + 1;
+        if (print) std::cout << std::hex << s->s[i];
+    }
+    s->s[s->n] = 0;
+    if (print) std::cout << ", n=" << s->n << "\n";
+    snprintf(s->name, 20, "random n=%i c=%i", s->n, mod);
+}
+
+void strAllA(TestString* s) {
+    uint32 i;
+    for (i = 0; i < s->n; i++) s->s[i] = 'a';
+    s->s[s->n] = 0;
+    snprintf(s->name, 20, "all a n=%i", s->n);
+}
+
+void strRepeat(TestString* s, uint32 len) {
+    uint32 i;
+    for (i = 0; i < s->n; i++) s->s[i] = '1' + (i % len);
+    s->s[s->n] = 0;
+    snprintf(s->name, 20, "repeat %i n=%i", len, s->n);
+}
+
+TestString::TestString(uint32 n, enum string_type type) {
+    this->n = n;
[…]

### 2011-07-11 src/string.h

@@ -1,0 +1,67 @@
+#ifndef _TESTSTRING_H_
+#define _TESTSTRING_H_
+
+#include "types.h"
+
+enum string_type {RANDOM2_STRING, RANDOM4_STRING, RANDOM10_STRING, RANDOM255_STRING, ALL_A_STRING, RANDOM10_REPEAT10_STRING, REPEAT_POW_STRING};
+
+class TestString {
+public:
+    /**
+     * Create a new string
+     *
+     * Immediately call checkStr to check if everything went OK.
+     *
+     * @param n The length of the string to make (terminating \0 not included).
+     * @param type The pattern used to create the string.
+     */
+    TestString(uint32, enum string_type);
+
+    /**
+     * Create a new string from a file
+     *
+     * Immediately call checkStr to check if everything went OK.
+     *
+     * @param fileName The path of the file to load.
+     */
+    TestString(const char* fileName);
+
+    /**
+     * Free all memory related to the string
+     */
+    ~TestString();
+
+    /**
+     * The string, a buffer of size n+1, where s[0..n-1] is the string of
+     * characters >0, and s[n]=0.
+     */
+    uint8* s;
+
+    /**
+     * Length of the string, not counting the last 0 byte.
+     */
+    uint32 n;
+
+    /**
+     * A name for the string indictaing how it was built.
+     */
+    char name[20];
+
+    // sa, sainv and lcp are lazily initialized 
+    uint32* sa;
+    uint32* sainv;
+    uint32* lcp;
+
+private:
+    void buildSa();
+};
+
+/**
+ * Returns true if s is a successfully initialized string.
+ * Returns false and cleans up s if s was not successfully initialized.
+ */
+bool checkStr(TestString* s);
+
+const char* nameStr(enum string_type type);
+
+#endif

### 2011-07-11 src/tableFiles.cpp

@@ -1,0 +1,64 @@
+#include "tableFiles.h"
+#include "string.h"
+#include "testRandom.h"
+#include <stdlib.h>
+#include <new>
+#include <iostream>
+#include <fstream>
+
+void tableFiles(Algorithm** algs, const char** files, const char* output, double preproc_time_limit) {
+    uint32 algcount;
+    for (algcount = 0; algs[algcount]; algcount++);
+    uint32 filecount;
+    for (filecount = 0; files[filecount]; filecount++);
+
+    uint32 paircount = 1000000;
+    uint32* pairs = new (std::nothrow) uint32[paircount*2];
+    if (!pairs) {
+        return;
+    }
+
+    std::ofstream f(output);
+
+    f << "Files,";
+    std::cout <<"\nFiles,";
+
+    for (uint32 alg = 0; algs[alg]; alg++) {
+        char algName[ALGORITHM_NAME_SIZE];
+        algs[alg]->getName(algName, ALGORITHM_NAME_SIZE);
+        f << algName << ",";
+        std::cout << algName << ",";
+    }
+    f << "\n";
+    std::cout << "\n";
+
+    for (uint32 file = 0; file < filecount; file++) {
+        TestString* s = new (std::nothrow) TestString(files[file]);
+        if (!checkStr(s)) {
+            f << files[file] << " file too big\n";
+            std::cout << files[file] << " file too big\n";
+            continue;
+        }
+        f << files[file] << ",";
+        std::cout << files[file] << ",";
+        for (uint32 i = 0; i < paircount*2; i += 2) {
+            pairs[i] = rand() % s->n;
+            pairs[i + 1] = rand() % s->n;
+        }
+        for (uint32 alg = 0; alg < algcount; alg++) {
+            double time = testRandom(s, pairs, paircount, algs[alg], false, preproc_time_limit);
+            if (time < 0.0) {
+                f << ",";
+                std::cout << ",";
+            } else {
+                f << time << ",";
+                std::cout << time << ",";
+            }
+        }
+        f << "\n";
+        std::cout << "\n";
+        delete s;
+    }
+    f.close();
+    delete[] pairs;
+}

### 2011-07-11 src/tableFiles.h

@@ -1,0 +1,7 @@
+#ifndef _TABLEFILES_H_
+#define _TABLEFILES_H_
+
+#include "algorithm.h"
+void tableFiles(Algorithm** algs, const char** files, const char* output, double preproc_time_limit);
+
+#endif

### 2011-07-11 src/testRandom.cpp

@@ -1,0 +1,42 @@
+#include "testRandom.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include "timer.h"
+
+/**
+ * Tests a given agorithm on a given string using random query pairs.
+ */
+double testRandom(TestString* s, uint32* pairs, uint32 paircount, Algorithm* a, bool print, double preproc_time_limit) {
+    uint32 x;
+    double start;
+    double end;
+    char algName[ALGORITHM_NAME_SIZE];
+    a->getName(algName, ALGORITHM_NAME_SIZE);
+    if (a->spaceUsage(s->n) > 1e9) {
+        if (print) printf("\ntestRandom,%s,too much memory,%s\n", algName, s->name);
+        return -1;
+    }
+    start = gettime();
+    bool success = a->preproc(s);
+    end = gettime();
+    if (!success) {
+        if (print) printf("\ntestRandom,%s,out of memory,%s\n", algName, s->name);
+        return -1;
+    }
+    if (end - start > preproc_time_limit) {
+        a->cleanup();
+        if (print) printf("\ntestRandom,%s,preprocessing timed out,%s\n", algName, s->name);
+        return -1;
+    }
+    start = gettime();
+    for (x = 0; x < paircount*2; x += 2) {
+        uint32 i = pairs[x];
+        uint32 j = pairs[x + 1];
+        uint32 lce = a->query(i, j);
+        //printf("i=%i, j=%i, lce=%i\n", i, j, lce);
+    }
+    end = gettime();
+    if (print) printf("\ntestRandom,%s,%e,%s\n", algName, (end - start) / (double) paircount, s->name);
+    a->cleanup();
+    return (end - start) / (double) paircount;
+}

### 2011-07-11 src/testRandom.h

@@ -1,0 +1,7 @@
+#ifndef _TESTRANDOM_H_
+#define _TESTRANDOM_H_
+
+#include "algorithm.h"
+double testRandom(TestString* s, uint32* pairs, uint32 paircount, Algorithm* a, bool print, double preproc_time_limit);
+
+#endif

### 2011-07-11 src/timer.cpp

@@ -1,0 +1,20 @@
+#include "timer.h"
+#ifdef _WIN32
+#include <windows.h>
+#else
+#include <sys/time.h>
+#endif
+
+// returns seconds.
+double gettime() {
+#ifdef _WIN32
+    __int64 t, f;
+    QueryPerformanceFrequency((LARGE_INTEGER *)&f);
+    QueryPerformanceCounter((LARGE_INTEGER *)&t);
+    return (double)t / (double)f;
+#else
+    struct timeval s;
+    gettimeofday(&s, 0);
+    return (double)s.tv_sec + (double)s.tv_usec / 1000000;
+#endif
+}

### 2011-07-11 src/timer.h

@@ -1,0 +1,6 @@
+#ifndef _TIMER_H_
+#define _TIMER_H_
+
+double gettime();
+
+#endif

### 2011-07-11 src/types.h

@@ -1,0 +1,28 @@
+#ifndef _TYPES_H
+#define _TYPES_H
+
+#if _MSC_VER
+
+typedef unsigned __int8 uint8_t;
+typedef unsigned __int16 uint16_t;
+typedef unsigned __int32 uint32_t;
+typedef unsigned __int64 uint64_t;
+typedef __int32 int32_t;
+#define INT32_MAX _I32_MAX
+#define snprintf sprintf_s
+double log2(double n);
+
+TODO: define what intptr_t and INTPTR_MAX is
+
+#else
+#define __STDC_LIMIT_MACROS
+#include <stdint.h>
+#endif
+
+typedef uint8_t uint8;
+typedef uint16_t uint16;
+typedef uint32_t uint32;
+typedef uint64_t uint64;
+typedef int32_t int32;
+
+#endif

### 2011-07-11 src/valgrind.sh

@@ -1,0 +1,9 @@
+# http://valgrind.org/docs/manual/dist.readme-developers.html
+cd ~/thesis/valgrind
+tar -xjf valgrind-3.6.1.tar.bz2
+cd valgrind-3.6.1
+patch < ../../src/valgrind.diff
+./configure
+make
+# cd ~/thesis/src
+# ../valgrind/valgrind-3.6.1/vg-in-place --tool=cachegrind --LL=<size>,<assoc>,<linesize> main.exe

### 2011-07-11 doc/report.tex
m

@@
 
-\fixme{Our Table Lookup Algorithm}
+\fixme{Introduction: Our Table Lookup Algorithm}
 
@@
 
+\paragraph{Horizontal, intra-level optimization}
+
 The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$.
@@
 When $k=\logceil$, this cache optimization gives no difference in the amount of I/O's.
+
+\paragraph{Vertical, inter-level optimization}
+
+For larger values of $k$, we can attempt to optimize memory access accros different levels.
+When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$.
+Neither of these positions are next to $H_\ell[i]$ in memory.
+We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other.
+This will not improve the asymptotic I/O bound, but could impove practical query times.
 
@@
 \item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
-
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations are used.
 \end{description}
@@
 \ifarticle
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Results}
-
 \fi % article
@@
 
-\fixme{Investigate if we can explaing the jump in Fingerprint-log}
+\fixme{Results: Add some results on real world files, in a table like Navarro. STATUS: Pizza\&Chili-teksterne er for store. Brug nogle andre?}
+\fixme{Results: Investigate if we can explain the jump in Fingerprint-log}
 % \paragraph{\fprint[\log n] cache limit.}
@@
 
-\pref{fig:plot-fingerprint-cache-horiz} shows results of the cache optimization described in \pref{sec:fingerprint-cache}.
-We have implemented two cache optimized variants.
+\paragraph{Horizontal, intra-level optimization}
+
+\pref{fig:plot-fingerprint-cache-horiz} shows results of the intra-level cache optimization described in \pref{sec:fingerprint-cache}.
+We have implemented two intra-level cache optimized variants.
 One as described in \pref{sec:fingerprint-cache}, and one where multiplication and division is replaced with shift operations.
@@
 
-We conclude that cache optimization does not visibly affect average case query times, while it improves worst case query times.
+We conclude that this cache optimization does not visibly affect average case query times, while it improves worst case query times.
 However, due to late timing of this discovery, the cache optimized variant of \fprintk\ is not the main focus of our analysis.
 
-\fixme{List these algorithms in the section on tested algorithms?}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Conclutions on Experimental Results}
-
-\fixme{Conclutions on Experimental Results}
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-alla.plt}
+    \end{center}
+    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without cache optimization.}
+\end{figure}
+
+\paragraph{Vertical, inter-level optimization}
+
+\pref{fig:plot-fingerprint-cache-vert} shows results of our inter-level cache optimization.
+We see no improvement on worst case strings, and we see increased query times on average case strings.
+We conclude that this attempt at cache optimization does not improve practical query times, which could be expected, since it does not improve the asymptotic I/O bund either.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Conclusions on Experimental Results}
+
+\fixme{Results: Conclusions on Experimental Results}
 
@@
 
+("Cache set count is not a power of two.")
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsubsection{Also tell about attempting by-LCE-value plot and why it fails}
+
+Plot results using LCE value as x-axis: Results looks wrong, why? Likely because of queries in a given pattern, v=max(i,j)
 
@@
 
-\fixme{But can we cache some results as we go even if pointers are not the same? No, that is Theta(n**2) = Omega(N) space. But do we really need it all?}
+\fixme{Compression: But can we cache some results as we go even if pointers are not the same? No, that is Theta(n**2) = Omega(N) space. But do we really need it all?}
 
@@
 
-We could lay out the tree of compression elements in memory as a B-tree\fixme{hvad hedder dette layout?
-B-tree??} to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the number of I/O's.\fixme{Er det klart?}
+We could lay out the tree of compression elements in memory as a B-tree\fixme{Compression: hvad hedder dette layout?
+B-tree??} to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the number of I/O's.\fixme{Compression: Er det klart?}
 
@@
 
-\section{Tinta tests}
+\subsection{Tinta tests}
 The two tests of LCE query performance used in the Tinta paper are shown in \pref{alg:tinta-testlceonebyone} and \pref{alg:tinta-randomtestlce}.
@@
 
-\section{DONE}
+\ifarticle
+\fxfatal{Opskriv funktioner for plads.}
+\fi % article
+
+\subsection{Dropped}
 \begin{enumerate}
-\item CODE:  test cache hops med cachegrind. STATUS: Samme instruction count, forskellige cache misses, men ser ud til at være modsat forventet.
-("Cache set count is not a power of two.")
-\item CODE: Undersøg præcis hvad der giver cache i kun average case i RMQ kode. STATUS: ikke cache, uopklaret
-\item DOC: Mere diskussion af og konklusion på resultaterne, beskriv dem som Navarro gør.
-\end{enumerate}
-
-\section{TODO Article}
-\begin{enumerate}
-\item CODE: Test på real world tekster, i tabel som Navarro. STATUS: Pizza\&Chili-teksterne er for store. Brug nogle andre?
-\item DOC: Opskriv funktioner for plads.
-\end{enumerate}
-
-\section{DONE Not for article}
-\begin{enumerate}
-\item DOC: Test on another machine. STATUS: Dropped
-\item CODE: Cache optimized memory layout for fingerprinting: Initial results bad, maybe test for other $k$, and test vertical for $k=2$.
-\item CODE: Plot results using LCE value as x-axis: Results looks wrong, why? Likely because of queries in a given pattern, v=max(i,j)
-\end{enumerate}
-
-\section{TODO Not for article}
-\begin{enumerate}
+\item DOC: Test on another machine.
 \item CODE: Improve cache-vertical and make shift variant.

### 2011-07-11 src/main.cpp
m

@@ -129,7 +129,6 @@ int main(int argc, char* argv[]) {
     graphValue(algs, RANDOM10_REPEAT10_STRING, 1000, "results/value-repeat10-10.txt", 1000);
     graphValue(algs, ALL_A_STRING, 1000, "results/value-alla.txt", 10000);
 
-    /*
     Algorithm* filealgs[] = {
         new AlgorithmDirectComp,
         new Fingerprint2AQ(&sqrt_n),
@@ -140,16 +139,15 @@ int main(int argc, char* argv[]) {
     };
     const char* files[] = {
         "tests/test.txt",
-        "tests/pizzachili/sources",
-        "tests/pizzachili/pitches",
-        "tests/pizzachili/proteins",
-        "tests/pizzachili/dna",
-        "tests/pizzachili/english",
-        "tests/pizzachili/dblp.xml",
+        "tests/canterbury/book1",
+        "tests/canterbury/kennedy.xls",
+        "tests/canterbury/E.coli",
+        "tests/canterbury/bible.txt",
+        "tests/canterbury/world192.txt",
         NULL
     };
     tableFiles(filealgs, files, "results/files.txt", 1000);
-    */
+    
     //plotDirectCompLookup(ALL_A_STRING, "results/complookup-alla.txt", 10000, 499, preproc_limit);
     /*
     Algorithm* cacheAlgs[] = {

### 2011-07-11 doc/report.tex
m

@@
 
-\begin{table}[ht]
+\begin{table}[tp]
 \centering
@@
 
+\begin{table}[tp]
+\centering
+\begin{tabular}{l|r|r|r|r|r|r|r}
+\hline\hline
+% File & \proc{DirectComp} & \fprint[2]\textless$\sqrt n$\textgreater ac & \fprint[3]\textless$n^{2/3}$, $n^{1/3}$\textgreater ac & \fprint[\log n]ac & RMQ<n;1> \\ [0.5ex] \hline
+File & $n$ & $\sigma$ & DC & FP$_2$ & FP$_3$ & FP$_{\log n}$ & RMQ \\ [0.5ex] \hline
+book1 & $0.7\cdot 2^{20}$ & 82 & 8.1 & 11.4 & 10.6 & 12.0 & 218.0 \\ \hline
+kennedy.xls & $1.0\cdot 2^{20}$ & 256 & 11.9 & 16.0 & 16.1 & 18.6 & 114.4 \\ \hline
+E.coli & $4.4\cdot 2^{20}$& 4 & 12.7 & 16.5 & 16.6 & 19.2 & 320.0 \\ \hline
+bible.txt & $3.9\cdot 2^{20}$ & 63 & 8.5 & 11.3 & 10.5 & 12.6 & 284.0 \\ \hline
+world192.txt & $2.3\cdot 2^{20}$ & 93 & 7.9 & 10.5 & 9.8 & 12.7 & 291.7 \\ \hline
+\end{tabular}
+\caption{Query times in nano seconds for \proc{DirectComp} (DC), \fprintk\ (FP$_k$) and \proc{LcpRmq} (RMQ) on the five largest files from the Canterbury corpus.}\label{tab:text-fprint-files}
+\end{table}
+
 \pref{fig:test-fingerprint} shows our experimental results on average case strings with a small average LCE value, worst case strings with a large average LCE value, and strings with a medium average LCE value.
 
-On average case strings, our new \fprintk\ algorithm is approximately 20\% slower than \proc{DirectComp}, and it is more than five times faster than \proc{LcpRmq}.
+On average case strings, our new \fprintk\ algorithm is approximately 20\% slower than \proc{DirectComp}, and it is between than 5 and 25 times faster than \proc{LcpRmq}.
+We see the same results on some real world strings in \pref{tab:text-fprint-files}.
 
@@
 For longer strings the $k=3$ variant performs the best up to strings of length $250.000$, at which point the $k=\logceil$ variant becomes the best.
-This indicates that for shorter strings, it is best to use fewer levels, and when the input size increases, the \fprintk\ variants with better asymptotic query times have better worst case times in practice.
+This indicates that for shorter strings, using fewer levels is better, and when the input size increases, the \fprintk\ variants with better asymptotic query times have better worst case times in practice.
 
@@
 
-\fixme{Results: Add some results on real world files, in a table like Navarro. STATUS: Pizza\&Chili-teksterne er for store. Brug nogle andre?}
-\fixme{Results: Investigate if we can explain the jump in Fingerprint-log}
+% \fixme{Results: Investigate if we can explain the jump in Fingerprint-log}
 % \paragraph{\fprint[\log n] cache limit.}
@@
 
-\fixme{Results: Conclusions on Experimental Results}
+Of the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, we have seen that \proc{DirectComp} is best on average case strings, while \proc{LcpRmq} is best on worst case strings.
+
+We have found that our new \proc{DirectCompLookup} algorithm can improve the worst case query time performance of \proc{DirectComp}, while it does not degrade the average case query time to the level of \proc{LcpRmq}.
+However it is not not as interesting as our new \fprintk\ algorithm, because it uses too much space.
+
+Our new \fprintk\ algorithm is able to achieve a balance between worst case and average case query times.
+It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have found cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}.
+\fprintk\ gives a good time space tradeoff, and it uses less space than \proc{LcpRmq} when $k$ is small.
+The perfprmance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
 
@@
 
-Skal vi antage at $i < j$ eller ej, eller skal begge dele testes?
-
-Skal vi antage at strengen ender i et specielt tegn eller ej, eller skal begge dele testes?
-
-Hvor meget skal et tegn fylde? Altid 8 bit som i Tinta-artiklen eller varieres?
-
-Er valg af random-generator vigtig?
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Structure}
-
-To test the practical performance of the algorithms, they are implemented in C++.
-The implementation is split into three parts.
-\pref{tab:test-variables} summarizes where the different variables for testing are implemented.
-
-\begin{table}[ht]
+To test practical performance, we have implemented the algorithms in C++.
+The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified.
+We have made it compile on a 32-bit Windows archictecture, and generalized it to allow more algorithms and tests to be implemented.
+\pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.
+
+\begin{table}[tp]
 \centering
@@
 character pattern & uniformly at random & string \\ \hline
-algorithm & Perfect Hashing & algorithm \\
+algorithm & \fprintk & algorithm \\
 alg. parameter & $t = 2\sqrt n$ & algorithm \\ \hline
@@
 
-("Cache set count is not a power of two.")
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Also tell about c++ bug hunting in tinta, cachegrind code}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Also tell about attempting by-LCE-value plot and why it fails}
+In our implementation, we have made the following assumptions about the LCE problem:
+\begin{itemize}
+\item A character in a string is always eight bits.
+\item A string always ends with a special null character, which does not occur anywehere else in the string.
+This does not affect our results, since we do not measure practical preprocessing time.
+If an algorithm has faster query times when the string ends in a special character, it could just add that character to the end of the string in its preprocessing.
+\item We make no assumptions on the order of $i$ and $j$.
+In some applications of the LCE problem, for example approximate string matching, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times.
+We do not measure this case.
+\item Wehenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
+\end{itemize}
+
+\fixme{Results: Test Code: tell about c++ bug hunting in tinta, cachegrind code "Cache set count is not a power of two."}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsubsection{Query time by LCE value}
+
+\fixme{Results: Test Code: tell about attempting by-LCE-value plot and why it fails}
 

### 2011-07-12 doc/report.tex
m

@@
 
-\listoffixmes
-
 \newpage
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-Blah blah blah\fixme{Introduction}
+\section{Introduction\label{sec:intro}}
+
+The Longest Common Extension problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
+Optimal solutions exists for the problem, with constant query time and linear space.
+These teoretically optimal solutions are however not the best in practice, since they have large constant factors for both time and space usage.
+In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance.
+This algorithm is ideal when only average case performance is relevant.
+In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.
+
+We have found two new algortihms.
+They both have better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms.
+The best of these two uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.
+
+In \pref{sec:algorithms} we theorectically describe and analyze the LCE algorithms introduced here in \pref{sec:intro}, and in \pref{sec:results} we analyze results from practical C++ implementations of these algorithms.
+
+We also look at how to apply the LCE problem to Ziv-Lempel compressed strings, such that we can answer LCE queries fast while only using space, which is a function of the compressed size of the string.
+In \pref{sec:lz-compress} we describe LZ compression, and how we can convert the simple linear time LCE algorithm to work on compressed strings with space linear to the size of the compressed string, worst case query time linear to the size of the uncompressed string, and average case query time $O(\log\log N)$.
+In \pref{sec:tree-lce} we look at how we can use the tree structure of LZ compression to get better query times on compressed strings.
 
@@
 Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is a lot faster than \proc{DirectComp} on a worst case input.
-Compared to \proc{LcpRmq}, our solution has a significantly better performance on an average case input, but its worst case performance is not as good as \proc{LcpRmq}.
+Compared to \proc{LcpRmq}, our solution has a significantly better performance on an average case input, but its worst case performance is not as good as that of \proc{LcpRmq}.
 The space usage for \proc{LcpRmq} and \fprintk\ are approximately the same when $k=6$.
@@
 
-\fixme{Introduction: Our Table Lookup Algorithm}
+We also present another new LCE algorithm, which we call \proc{DirectCompLookup}.
+
+\begin{theorem}
+For a string $s$ of length $n$, the \proc{DirectCompLookup} algorithm can solve the LCE problem in $O(t)$ worst case query time and $O(1)$ average case query time using $O(n^2/t)$ space and $O(n^2)$ preprocessing time, where $t$ is a parameter $1 \leq t \leq n$.
+\end{theorem}
+
+The algorithm stores LCE values for some of the $n^2$ possible query inputs on a given string.
+It uses \proc{DirectComp} to move towards one of these stored values, and stops when it reaches one.
+The \proc{DirectCompLookup} algorithm is very simple, but the tradoff between worst case query time and space is not very favorable, and preprocessing the data structure takes $O(n^2)$ time unless we use another LCE algorithm in the preprocessing.
 
@@
 The suffix of $s$ starting at index $i$ is written $\suff_i = s[i\twodots n]$.
+
+\paragraph{Tree notation.}
+The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
+The depth of a node $u$ is one greather than the depth of $u$'s parent, and the depth of the root is zero.
 
@@
 
-\paragraph{NCA.} The ancestors of a node $u$ is $u$ itself and any node, which is a parent of an ancestor of $u$.
-The common ancestors of two nodes $u$ and $v$ is the intersection between the ancestors of $u$ and the ancestors of $v$.
+\paragraph{NCA.} The common ancestors of two nodes $u$ and $v$ is the intersection between the ancestors of $u$ and the ancestors of $v$.
 The Nearest Common Ancestor of two nodes $u$ and $v$ is the common ancestor of $u$ and $v$ of greatest depth.
@@
 
-\paragraph{RMQ.} The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, or $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
+\paragraph{RMQ.} The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, i.e.\ $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
 A Range Minimum Query on a static array of $n$ elements can be answered in $O(1)$ time with $O(n)$ space and preprocessing time~\cite{jf-rmq}.
@@
 \paragraph{Predecessor.}
-\fixme{Preliminaries: Predecessor}
+Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natual number $x$ is an object with maximal key less than or equal to $x$.
+We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.
 
 \paragraph{Level Ancestor.}
-\fixme{Preliminaries: Level Ancestor}
-
-\paragraph{Trees.}
-\fixme{Preliminaries: Trees}
-depth root = 0
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{LCE Algorithms}
+The Level Ancestor at level $\ell$ of node $u$ is the ancestor of $u$, whoes depth is $\ell$.
+We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{LCE Algorithms\label{sec:algorithms}}
 This section describes existing and new algorithms for solving the LCE problem.
@@
 \proc{DirectComp} solves the LCE problem without any preprocessing in $O(|\LCE(i,j)|)$ query time.
-For a query $\LCE(i,j)$, the algorithm compares $s[i+v]$ with $s[j+v]$ in a loop, for $v$ starting at zero and incrementing, until the two characters differ, or the end of the string is reached.
+For a query $\LCE(i,j)$, the algorithm compares $s[i+v]$ to $s[j+v]$ in a loop, for $v$ starting at zero and incrementing, until the two characters differ, or the end of the string is reached, at which point the LCE value is $v$.
 \proc{DirectComp} does not need to check if the end of the string is reached, i.e.
@@
 The path label from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes.
-Since two child edges of the same node never have the same first character on their edge labels, the path label from the root to the $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but longest common prefix of $\suff_i$ and $\suff_j$.
+Since two child edges of the same node never have the same first character on their edge labels, the path label from the root to $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but longest common prefix of $\suff_i$ and $\suff_j$.
 We can therefore find the LCE value as $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$, where $\mathcal{T}$ is the suffix tree for $s$, and $D[v]$ is the label depth of node $v$ in the tree.
@@
 
-The input to the $LCE$ problem is two numbers $1 \leq i < j \leq n$ \footnote{We can assume that $i< j$ without loss of generality, since $\LCE(i,j)=\LCE(j,i)$}, so there are $n*(n-1)/2 = O(n^2)$ different input possibilities in total.
+The input to the $LCE$ problem is two numbers $1 \leq i < j \leq n$ \footnote{We can assume that $i< j$ without loss of generality, since $\LCE(i,j)=\LCE(j,i)$}, so there are $n\cdot(n-1)/2 = O(n^2)$ different input possibilities in total.
 Because of this limited space of possible query inputs, we can tabulate a number of LCE values, and use those to speed up queries.
@@
 \proc{DirectComp} and \proc{DirectLookup} preprocesses and stores no results and all results respectively.
-By only storing some results, we can use \proc{DirectComp} until we reach a stored result, and then retrieve this result using \proc{DirectLookup}.
+If we only store some results, we can use \proc{DirectComp} until we reach a stored result, and then retrieve this result using \proc{DirectLookup}.
 If we store every $t$'th row, as shown in \pref{fig:resultspace-t-row}, the space usage will be $O(n^2/t)$ and the worst case query time will be $O(t)$.
@@
 Algorithm & Space \\ [0.5ex] \hline
-\proc{DirectComp} & $1$ words \\ \hline
+\proc{DirectComp} & $0$ words \\ \hline
 \proc{DirectLookup} & $2n^2$ words \footnote{The factor two comes from using shift instead of multiplication, which may double the size depending on $n$.} \\ \hline
@@
 \end{minipage}
-\caption{Space requirements for the LCE algorithms, where $n$ is the number of characters, one character is byte and one word is four bytes.}\label{tab:algorithms-space}
+\caption{Space requirements for the LCE algorithms, where $n$ is the number of characters, a character is one byte and a word is four bytes.}\label{tab:algorithms-space}
 \end{table}
@@
 
-We implemented different variants \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other.
+We implemented different variants of \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other.
 We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}.
@@
 \item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$.
-The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater describe the exact size of fingerprinted substrings at each level.
+The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
 \item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater bc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_1$, ..., $u_{k-1}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
 \item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
@@
 \item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations are used.
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
 \end{description}
@@
 We want to compare our new algortithms against the algortihm which is the best in practice of the asymptotically best $O(1)$ algorithms.
-The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at that one.
+The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at \proc{LcpRmq}.
 We have looked at a number of RMQ algorithms to find out which works best for the LCE problem.
@@
 \pref{fig:plot-rmq-directcomp} also shows \proc{DirectComp+LcpRmq}, where we first try \proc{DirectComp}, and if we have not found the LCE value after one iteration, we use \proc{LcpRmq} instead.
+We see that \proc{DirectComp+LcpRmq} is worse than \proc{DirectComp} on average case strings.
 We could increase the cutoff from one iteration to something bigger, which might improve average case performance.
-That would however decrease worst case performance.
+That would however increase worst case time.
 None of the two plots we measure represents a good worst case for \proc{DirectComp+LcpRmq} compared to the other algorithms we test.
-The worst case for \proc{DirectComp+LcpRmq} relative to other the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithm still work on LCE values as small as possible.
+The worst case for \proc{DirectComp+LcpRmq} relative to other the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithms still work on LCE values as small as possible.
 As described in \pref{sec:dc-lcprmq}, \proc{DirectComp+LcpRmq} also uses a lot of space.
@@
 
-Figure \pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs $t$ together with \proc{LcpRmq}.
+\pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs together with \proc{LcpRmq}.
 In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra ckeck to find out if it has reached a position for which a result is stored.
@@
 We see no improvement on worst case strings, and we see increased query times on average case strings.
-We conclude that this attempt at cache optimization does not improve practical query times, which could be expected, since it does not improve the asymptotic I/O bund either.
+We conclude that this attempt at cache optimization does not improve practical query times, which could be expected, since it does not improve the asymptotic I/O bound either.
 
@@
 We have found that our new \proc{DirectCompLookup} algorithm can improve the worst case query time performance of \proc{DirectComp}, while it does not degrade the average case query time to the level of \proc{LcpRmq}.
-However it is not not as interesting as our new \fprintk\ algorithm, because it uses too much space.
+However it is not as interesting as our new \fprintk\ algorithm, because it uses too much space.
 
@@
 \item We make no assumptions on the order of $i$ and $j$.
-In some applications of the LCE problem, for example approximate string matching, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times.
+In some applications of the LCE problem, for example approximate string searching, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times.
 We do not measure this case.
@@
 
-\fixme{Results: Test Code: tell about c++ bug hunting in tinta, cachegrind code "Cache set count is not a power of two."}
+% TODO \fixme{Results: Test Code: tell about c++ bug hunting in tinta, cachegrind code "Cache set count is not a power of two."}
 
@@
 
-\fixme{Results: Test Code: tell about attempting by-LCE-value plot and why it fails}
-
-Plot results using LCE value as x-axis: Results looks wrong, why? Likely because of queries in a given pattern, v=max(i,j)
-
+\begin{figure}[tp]
+    \begin{center}
+        %\includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/value-article-rand4.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/value-article-alla.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/value-article-repeat10-10.plt}
+    \end{center}
+    \caption{\label{fig:plot-by-value}Query times of some of our tested algorithms as a function of the returned LCE value.}
+\end{figure}
+
+As an alternative way of testing our alorithms, we have looked at how long a query takes in relation to the LCE value it returns, since most of our algorithms have query times, which depend on the returned LCE value.
+\pref{fig:plot-by-value} shows query times for a fixed string of 1000 characters.
+We can see how the query times of \fprint[2] and \fprint[3] increases as the LCE value increases, and drops when the LCE value becomes long enough to make one more step at the higher level with fingerprints of longer substrings.
+We also see that the query time of \fprint[\log n] increases whenever the LCE value becomes long enough to reach a higher level of the data structure.
+
+The plots in \pref{fig:plot-by-value} contains a lot of noise compared to our other plots.
+The number of measurements we make on these plots is many times bigger than the number of measurements we make on other plots.
+Therefore each measurement is an average over only 10000 queries instead of a million queries, which gives less accurate results.
+
+We need to measure a sum of many queries to get an accurate timing, since each individual query is too fast to measure.
+While measureing a sequence of randomly generated queries is a very artificial test, which is unlikely to occur in practice, we think that it is even more artificial to measure a sequence of queries for which we already know that the LCE value is the same each time.
+On our worst case string $s=aaaaaaa...$, the queries we measure in each point in \pref{fig:plot-by-value} have the pattern $\max(i,j)=v$, which may give very different performance characteristics compared to random queries due to cache effects.
+We cannot meaningfully measure performance by LCE value on average case strings, as these strings do not contain large LCE values.
+The second plot in \pref{fig:plot-by-value} is on a string, which attempts to avoid a strong pattern like $\max(i,j)=v$ while it still has large LCE values.
+
+Because of the amount of noise in the plots and the strong pattern of $\max(i,j)=v$, we have not focused on this way of measuring query time performance in our analysis.
+
+\begin{comment}
 \begin{algorithm}
@@
 Where $\proc{Plot}(f,x,y)$ adds point $(x,y)$ to series $f$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Ziv-Lempel compression}
+\end{comment}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Ziv-Lempel Compression\label{sec:lz-compress}}
 
@@
 Our original goal of looking into compression was to see if using the techniques used in compression could improve the worst case query time of \proc{DirectComp} to something better than $O(N)$.
-Our idea is that whenever you have a long LCE value, you have a repetition within the string, and repetitions would typically mean good compression.
+Our idea is that whenever you have a long LCE value, you have a repetition within the string, and repetitions would typically give good compression.
 If the compressed string can be searched over similarly to how \proc{DirectComp} searches the uncompressed string, it might be possible to establish a better worst case query time.
 
-We have found one case where compressed \proc{DirectComp} is faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time.
+We have found one case where we can make compressed \proc{DirectComp} faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time.
 We explain this in terms of the reversed compressed \proc{DirectComp}, as it is simpler, but it also works without the string reversal.
@@
 If $x_i = x_j$ we know that we can add $\textit{depth}[x_i]$ to the LCE value, where \textit{depth} is the depth of a node in the tree of compression elements.
-\proc{DirectComp} can then compare $\textit{depth}[x_i]$ characters in constant time.
+Compressed \proc{DirectComp} can then compare $\textit{depth}[x_i]$ characters in constant time.
 
@@
 This is because we cannot be sure to hit the condition allowing us to compare many characters in constant time, even for long LCE values.
-As an example, $LCE(1,2)$ on a string where all $N$ characters are the same, takes $\Omega(N)$ time.
+As an example, $LCE(1,2)$ on a string where all $N$ characters are the same takes $\Omega(N)$ time.
 For such a string, the tree representing the compressed string is always a path, and $x_i$ and $x_j$ will always be either a node and its parent or the root and some other node.
-They will therefore never end up being the same node at the same time, as shown in the example in \pref{fig:lz-no-jumps}.
+They will therefore never end up being the same node at the same time, as shown in \pref{fig:lz-no-jumps}.
 
@@
 
-\fixme{Compression: But can we cache some results as we go even if pointers are not the same? No, that is Theta(n**2) = Omega(N) space. But do we really need it all?}
+% \fixme{Compression: But can we cache some results as we go even if pointers are not the same? No, that is Theta(n**2) = Omega(N) space. But do we really need it all?}
 
@@
 We have looked into the cache performance of compressed \proc{DirectComp}.
-If the string does not compress at all, i.e.
-the compressed string is $z=(0,a)(0,b)(0,c)...$, \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case.
+If the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case.
 When the string is LZ-compressable, the I/O pattern becomes much worse.
-After reading compression element $x_1$ with $\zreference(x_1) = x_2 > 0$, the code will read from position $x_2$, where all we know is that $1 \leq x_2 < x_1$.
-This gives $O(N)$ I/O's in worst case.
-
-We could lay out the tree of compression elements in memory as a B-tree\fixme{Compression: hvad hedder dette layout?
-B-tree??} to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the number of I/O's.\fixme{Compression: Er det klart?}
+If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be next to each other in memory.
+This can happen many times, which gives $O(N)$ I/O's in worst case.
+
+We could lay out the tree of compression elements in memory as a B-tree, where we store groups of adjacant nodes next to eachother in memory, to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the worst case number of I/O's, as each node visited may be in a new group.
 
@@
 
+\bibitem{predecessor} Dan E. Willard. Log-logarithmic worst-case range queries are possible in space $\Theta$(N). Information Processing Letters, Volume 17, Issue 2, 24 August 1983, pages 81-84.
+
+\bibitem{level-ancestor} Michael A. Bendera, Martín Farach-Colton. The Level Ancestor Problem simplified. Theoretical Computer Science 321, 2004, pages 5-12.
+
+\bibitem{approx-search} Gad M. Landau and Uzi Vishkin. Introducing efficient parallelism into approximate string matching and a new serial algorithm. 18th ACM STOC, pages 220–230, 1986.
+
 \end{thebibliography}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{THE END}
-
-\subsection{Tinta tests}
+\section*{THE END}
+
+\subsection*{Tinta tests}
 The two tests of LCE query performance used in the Tinta paper are shown in \pref{alg:tinta-testlceonebyone} and \pref{alg:tinta-randomtestlce}.
@@
 
-\subsection{Dropped}
+\subsection*{Dropped}
 \begin{enumerate}
@@
 \item CODE: Non-random strings? Non-random queries? Different t-values for Hash-3L?
-\item DOC: Explain why Karp-Miller-Rosenberg is $O(n\log n)$.
+\item DOC: Why is Karp-Miller-Rosenberg $O(n\log n)$?.
 \end{enumerate}
+
+\listoffixmes
 

### 2011-07-12 src/plot2.sh

@@ -41,20 +41,20 @@ ymc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Medium LCE-size strings'
 style="lines"
 
 # x-axis = LCE value
-xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:10]"
+xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:10]\nset title 'Average case strings, alphabet = 4'"
 datafile="results/value-rand4.txt"
 plotfile="results/value-article-rand4.plt"
-oneplot "DirectComp," "Fingerprint_2<sqrt n>wc," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}wc," "Fingerprint_{log n}ac," "RMQ<n;1>," "DirectComp<1>RMQ<n;1>,"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
 
-xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:1000]"
+xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:1000]\nset title 'Worst case strings'"
 datafile="results/value-alla.txt"
 plotfile="results/value-article-alla.plt"
-oneplot "DirectComp," "Fingerprint_2<sqrt n>wc," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}wc," "Fingerprint_{log n}ac," "RMQ<n;1>," "DirectComp<1>RMQ<n;1>,"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
 
-xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:1000]"
+xtra="unset logscale x\nset yrange [1e-8:1e-7]\nset xrange [0:1000]\nset title 'String of 10 characters repeated 100 times'"
 datafile="results/value-repeat10-10.txt"
 plotfile="results/value-article-repeat10-10.plt"
[…]

### 2011-07-13 doc/report.tex
m

@@
 \newcommand{\LCP}{\textit{LCP}}
+\newcommand{\LA}{\textit{LA}}
 \newcommand{\suff}{\textit{suff}}
@@
 
+\hyphenation{Di-rect-Comp Di-rect-Look-up Di-rect-Comp-Look-up Fin-ger-print}
+
 \tableofcontents
@@
 \paragraph{Level Ancestor.}
-The Level Ancestor at level $\ell$ of node $u$ is the ancestor of $u$, whoes depth is $\ell$.
+The Level Ancestor at level $\ell$ of node $u$, $\LA(\ell,u)$, is the ancestor of $u$, whoes depth is $\ell$.
 We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.
@@
 \begin{theorem}
+\label{thm:lz-tree}
 Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$.
@@
 
-\begin{verbatim}
-* LCE on trees
-  * Definition
-  * Heavy paths
-  * Hashing
-  * Ladder attempt
-\end{verbatim}
-
-TREE-LCE: Vi kan finde LCE på et træ i $O(\log n)$ tid ved at lave tung/let-vej og lave streng-LCE på de tunge veje.
-
-TREE-LCE: Vi kan lave hashing på et træ, ved at hver knude gemmer hash-værdien for de sidste t led af stien fra roden til knuden.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Definition}
+To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space. In this section we analyze ways to achieve this.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Definition\label{sec:tree-def}}
 
 Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $p$ to child node $c$ is labeled with a character $\zlabel(c)$.
-Let $x_i$ and $x_j$ be two nodes in T.
-Let $y_i$ be a decendent of $x_i$, and let $y_j$ be a decendent of $x_j$.
-Let $s_i$ be the string obtained from characters on the path from $x_i$ to $y_i$, and let $s_j$ be the string obtained from characters on the path from $x_j$ to $y_j$.
-Then $\LCE_T(x_i, y_i, x_j, y_j)$ is the length of the longest common prefix of $s_i$ and $s_j$.
+Let $u_i$, $w_i$, $u_j$ and $w_j$ be nodes in T, such that $w_i$ is a decendent of $u_i$, and $w_j$ is a decendent of $u_j$.
+Then $\LCE_T(u_i, w_i, u_j, w_j)$ is the length of the longest common prefix of $s_i$ and $s_j$, where $s_i$ is the string obtained from characters on the path from $u_i$ to $w_i$, and $s_j$ is the string obtained from characters on the path from $u_j$ to $w_j$.
 
@@
 
+We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor.
+If $u_i$ has depth $d_i$ and $u_j$ has depth $d_j$, then first compare $\textit{label}(\LA(d_i+1,w_i))$ to $\textit{label}(\LA(d_j+1,w_j))$, then compare $\textit{label}(\LA(d_i+2,w_i))$ to $\textit{label}(\LA(d_j+2,w_j))$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
+
+Each Level Ancestor query takes constant time, and we do $2|\LCE(i,j)| + 2$ of them, giving us $O(|\LCE(i,j)|)$ query time.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Gem fingerprint for hver knude for forfæder sti. Samme princip som hash-LCE for strenge
+\newcommand{\tfprint}[1][k]{\ensuremath{\proc{TreeFingerprint}_{#1}}}
+\newcommand{\tfprintk}{\tfprint[k]}
+
+We can adapt our \fprintk\ algorithm on strings to work on trees, which we call \tfprintk.
+If we store the fingerprint of $s[i\twodots i+t_\ell]$ in $H_\ell[i+t_\ell]$ instead of $H_\ell[i]$, then the \fprintk\ query directly adapts to \tfprintk.
+If we wanted to use the same preprocessing as we use for \fprintk\ on strings, we would have to adapt the suffix array to apply to trees, and find a way to construct it in $\sortt$ time, which we have not looked at.
+Instead we can preprocess the fingerprints of the tree with Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}.
+
+\begin{theorem}
+\tfprintk, where $k$ is a parameter $1\leq k\leq\logceil$, can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(k\cdot n^{1/k})$ worst case query time, $O(1)$ average case query time, $O(k\cdot n)$ space, and $O(n\log n)$ preprocessing time.
+\end{theorem}
+
+\subsubsection{Data Structure}
+
+We define a natural number $F_t[u]$ to be the fingerprint for node $u$ of length $t$.
+For two nodes $u$ and $w$ in a tree $T$, their fingerprints $F_t[u]$ and $F_t[w]$ of the same length $t$ are the same if and only if the last $t$ characters on the path from the root of $T$ to $u$ is the same as the last $t$ characters on the path from the root of $T$ to $w$.
+
+For a given parameter $k$, where $1\leq k \leq \logceil$, we define the \tfprintk\ data structure as follows: For each number $\ell$ between $0$ and $k-1$, let $t_\ell=\Theta(n^{\ell/k})$, and let $t_0 = 1$.
+For each number $\ell$ between $0$ and $k-1$ and each node $u$ in $T$, whoes depth is at least $t_\ell$, let $H_\ell[u] = F_{t_\ell}[u]$
+
+\begin{lemma}
+The \tfprintk\ data structure takes $O(n)$ space.
+\end{lemma}
+\begin{proof}
+The data structure contains $k$ numbers $t_\ell$ and $k$ tables of fingerprints $H_\ell$, and each table contains fingerprints for at most all of the $n$ nodes in the tree.
+\end{proof}
+
+\subsubsection{Query}
+
+The quey for \tfprintk\ is nearly the same as the query for \fprintk\ on strings, except we replace $H_\ell[i+v]$ and $H_\ell[j+v]$ with $H_\ell[\LA(d(u_i)+v+t_\ell,w_i)]$ and $H_\ell[\LA(d(u_j)+v+t_\ell,w_j)]$, where $d(u)$ is the depth of node $u$.
+We also stop the algorithm, when $d(u_i)+v+t_\ell$ is greater than the depth of $w_i$, or $d(u_j)+v+t_\ell$ is greater than the depth of $w_j$.
+
+\begin{lemma}
+The \tfprintk\ query finds the correct tree LCE value in $O(\log n)$ time.
+\end{lemma}
+\begin{proof}
+We have changed nothing from the \fprintk\ query on strings, which negatively affects the query time, so the query time is the same.
+
+% The combination of Level Ancestor and the definition that a fingerprint relates to a substring from a given node to an ancestor that node, ensures that the fingerprints we choose to compare are the fingerprints, which represents the strings on the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$.
+
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.7\textwidth]{tree-fingerprint.pdf}
+    \end{center}
+    \caption{\label{fig:tree-query}Finging the right fingerprint to look at for a \tfprintk\ query.}
+\end{figure}
+
+The fingerprints we choose to compare are the fingerprints, which represents the strings on the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$.
+Let $x_i$ and $y_i$ be two nodes on the path from $u_i$ to $w_i$, such that the distance between $u_i$ and $x_i$ is $v$, the distance between $x_i$ and $y_i$ is $t_\ell$ and $y_i$ is a decendent of $x_i$, like in \pref{fig:tree-query}.
+At any given iteration at level $\ell$, the query algorithm has already looked at the string on the path from $u_i$ to $x_i$, and we now need to look at the string on the path from $x_i$ to $y_i$.
+From our definition of fingerprints in a tree, this string has fingerprint $H_\ell[y_i]$.
+We have $y_i = \LA(d(u_i)+v+t_\ell,w_i)$, and the LCE query is therefore correct when we use $H_\ell[\LA(d(u_i)+v+t_\ell,w_i)]$ (and similar for $j$ instead of $i$).
+\end{proof}
+
+\subsubsection{Preprocessing}
+
+We use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to preprocess the \tfprintk\ data structure. When $t_\ell$ is $t_{\ell-1} < t_\ell \leq 2t_{\ell-1}$, we preprocess level $\ell$ like this:
+\begin{enumerate}
+\item Make pairs of fingerprints $H_\ell[u] = (H_{\ell-1}[LA(d(u)-t_\ell+t_{\ell-1},u)], H_{\ell-1}[u])$ for all $u$ where $d(u) \geq t_\ell$.
+\item Sort the pairs lexicographically.
+\item Replace each pair with a new fingerprint, by going throug them in lexichographical order, while updating the fingerprint whenever the pair changes.
+\end{enumerate}
+
+Repeat the steps above for each level, starting with $\ell = 1$ and ending with $\ell = k-1$.
+When $k < \logceil$, the condition that $t_\ell \leq 2t_{\ell-1}$ is not satisfied.
+In this case we need to add extra levels in the preprocessing, which we will only store temporarely.
+We therefore always need to preprocess $O(\log n)$ levels, even if we only store some of them.
+
+\begin{lemma}
+The KMR algorithm for preprocessing \tfprintk\ gives the correct data structure.
+\end{lemma}
+\begin{proof}
+Two strings $S$ and $T$ of length $a$ are the same if and only if we can split each of them into two smaller substrings $S = S_1\cdot S_2$ and $T=T_1\cdot T_2$ such that $S_1=T_1$ and $S_2=T_2$, where "$\cdot$" is string concatenation.
+Therefore we can assign fingerprints $F_a[S]=F_b[S_1]\cdot F_c[S_2]$ and $F_a[T]=F_b[T_1]\cdot F_c[T_2]$ in step one of the preprocessing algorithm.
+The same also holds when the two smaller substrings overlap.
+
+The fingerprints from level $\ell-1$ are natural numbers, less than $n$.
+Step one of the preprocessing algorithm makes each fingerprint at level $\ell$ a natural number less than $n^2$.
+Step two and three of the preprocessing algorithm reduceses that universe of fingerprints back to $n$ without changing the equality of the fingerprints.
+\end{proof}
+
+\begin{lemma}
+The KMR algorithm for preprocessing \tfprintk\ takes $O(n\log n)$ time.
+\end{lemma}
+\begin{proof}
+Step one and three use linear time.
+In step two we can use radix sort, which uses linear time, because our universe of elements to sort is $n^2$.
+The total preprocessing time becomes $O(n\log n)$ for preprocessing $\log n$ levels in $O(n)$ time each.
+\end{proof}
+
+HOW TO PREPROCESS?
+
+SA: ved ikke
+
+KMR: Ja!
 
@@
 \subsection{Using Constant Time String LCE in Heavy Paths}
+
+Vi kan finde LCE på et træ i $O(\log n)$ tid ved at lave tung/let-vej og lave streng-LCE på de tunge veje.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{thebibliography}{9}

### 2011-07-14 doc/report.tex
m

@@
 \usepackage{textcomp}
+\usepackage{enumerate}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Walking the Paths with Level Ancestor}
+\subsection{Walking the Paths with Level Ancestor\label{sec:tree-dc}}
 
 We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor.
-If $u_i$ has depth $d_i$ and $u_j$ has depth $d_j$, then first compare $\textit{label}(\LA(d_i+1,w_i))$ to $\textit{label}(\LA(d_j+1,w_j))$, then compare $\textit{label}(\LA(d_i+2,w_i))$ to $\textit{label}(\LA(d_j+2,w_j))$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
+First compare $\textit{label}(\LA(d(u_i)+1,w_i))$ to $\textit{label}(\LA(d(u_j)+1,w_j))$, where $d(u)$ is the depth of node $u$, then compare $\textit{label}(\LA(d(u_i)+2,w_i))$ to $\textit{label}(\LA(d(u_j)+2,w_j))$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
 
@@
     \begin{center}
-        \includegraphics[width=0.7\textwidth]{tree-fingerprint.pdf}
+        \includegraphics[width=0.7\textwidth,page=1]{tree-fingerprint.pdf}
     \end{center}
@@
 
-HOW TO PREPROCESS?
-
-SA: ved ikke
-
-KMR: Ja!
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Vi kan finde LCE på et træ i $O(\log n)$ tid ved at lave tung/let-vej og lave streng-LCE på de tunge veje.
-
-Vi kan lave Træ-LCE i O(n) plads, O(log n) tid ved at lave streng-LCE på tunge stier + Level-Ancestor.
-
-\begin{verbatim}
-    Current nodes are i_c and j_c, and end nodes a i_d and j_d.
-    while i_c != i_d && j_c != j_d
-        Let m_i be the number of initial heavy edges on the path from i_c to i_d before any light edge.
+\paragraph{Heavy Path Decomposition}
+The weight of a node $u$ is the size of the subtree rooted at $u$.
+For a given node $u$, its child of greatest weight is marked heavy, and all other children are marked light.
+The root is always marked light.
+An edge going from a node to a heavy child node is heavy, and an edge going from a node to a light child node is light.
+A path of only heavy edges is a heavy path, and the combined length of all heavy paths in a tree of $n$ nodes is less than $n$.
+The markings of heavy and light edges and nodes in a tree is a Heavy Path Decomposition.
+Every simple path in a tree $T$ of size $n$ contains at most $O(\log n)$ light edges.
+We can construct a Heavy Path Decomposition in WHAT??\fixme{heavy tree time} time.
+%% end of preliminary
+
+To answer LCE queries on a tree, we can create a heavy path decomposition\fixme{define heavy path decomposition} of the tree and create a string of the labels of all the heavy paths concatenated.
+We can then use constant time string LCE on this string.
+We call this algorithm \proc{HeavyTreeLce}.
+
+\begin{theorem}
+\proc{HeavyTreeLce} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ worst case query time, $O(1)$ average case query time, and $O(n)$ space.
+\fixme{preproc time}
+\end{theorem}
+
+\subsubsection{Data Structure}
+
+Let $T$ be a tree with a heavy path decomposition.
+A path label for a path in $T$ is a concatenation of all edge labels on the path starting at smallest depth and ending at greatest depth.
+The string $s$ is a concatenation of path labels of all heavy paths in $T$.
+For a node $u$, $\textit{end}(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ us the depth of $u$, and if $u$ is heavy, $\textit{index}(u)$ is the index in $s$, which stores $\zlabel(u)$.
+The tree must be preprocessed for constant time NCA and LA, and the string $s$ must be preprocessed for LCE.
+
+\begin{lemma}
+The \proc{HeavyTreeLce} data structure takes $O(n)$ space.
+\end{lemma}
+\begin{proof}
+The string $s$ contains less than $n$ characters, since the combined length of all heavy paths in a tree of $n$ nodes is less than $n$.
+The space usage for NCA, LA, LCE, and the extra information about each node (\textit{end}, \textit{index}, \textit{d}, and heavy markings) is all $O(n)$
+\end{proof}
+
+\begin{lemma}
+We can preprocess \proc{HeavyTreeLce} data structure in $\sortt$ time.
+\end{lemma}
+\begin{proof}
+Preprocessing LCE using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time.
+If we walk through the tree and for each leaf append the heavy path ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
+The rest of the data structures can be preprocessed in $O(n)$ time.
+\end{proof}
+
+\subsubsection{Query}
+
+When we directly walk the paths as described in \pref{sec:tree-dc}, we start with $v=0$ and increment $v$ by one as long as $\zlabel(\LA(d(u_i)+v+1,w_i))=\zlabel(\LA(d(u_j)+v+1,w_j))$, and we have not yet reached either $w_i$ or $w_j$.
+
+In a \proc{HeavyTreeLce} query $\LCE_T(u_i,w_i,u_j,w_j)$ we also start with $v=0$ and do the following repeatedly as long as we have not reached either $w_i$ or $w_j$:
+\begin{enumerate}[~~~a.~]
+  \item When either $\LA(d(u_i)+v+1,w_i)$ or $\LA(d(u_j)+v+1,w_j)$ is a light node:
+  \begin{enumerate}[~1.~]
+    \item Increment $v$ by one if $\zlabel(\LA(d(u_i)+v+1,w_i))=\zlabel(\LA(d(u_j)+v+1,w_j))$, otherwise stop and return $v$.
+  \end{enumerate}
+  \item When both $\LA(d(u_i)+v+1,w_i)$ and $\LA(d(u_j)+v+1,w_j)$ are heavy nodes:
+  \begin{enumerate}[~1.~]
+    \item Find $v'=\LCE_s(\LA(d(u_i)+v+1,w_i),\LA(d(u_j)+v+1,w_j))$.
+    \item Find $m_i = d(\NCA_T(\textit{end}(\LA(d(u_i)+v,w_i)), w_i))-(d(u_i)+v)$.
+    \item Find $m_j = d(\NCA_T(\textit{end}(\LA(d(u_j)+v,w_j)), w_j))-(d(u_j)+v)$.
+    \item Stop and return $v+v'$, if $v'<\min(m_i,m_j)$.
+    \item otherwise increment $v$ by $m$.
+  \end{enumerate}
+\end{enumerate}
+
+When we resch either $\LA(d(u_i)+v,w_i)=w_i$ or $\LA(d(u_j)+v,w_j)=w_j$, we stop and return $v$.
+
+\begin{lemma}
+The \proc{HeavyTreeLce} query finds the correct LCE value.
+\end{lemma}
+\begin{proof}
+The query algorithm maintains the invariant $v\leq\LCE_T(u_i,w_i,u_j,w_j)$.
+Part $a$ simply checks an extra character and increments $v$ if it finds a match.
+In part $b$, $m_i$ is the number of heavy edges at the beginning of the path from $\LA(d(u_i)+v,w_i)$ to $w_i$, and similar for $m_j$, like illustrated in \pref{fig:tree-heavy-query}.
+If $v'<\min(m_i,m_j)$ it means that the LCE query found the longest common prefix to be within both of the two paths we are searching, and we have therefore found the LCE value and stop.
+If $v'\geq\min(m_i,m_j)$, we had a matching prefix which was as long or longer than the part of the heavy path we are interested in, and we therefore need to continue looking outside that heavy path to find the LCE value.
+
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.7\textwidth,page=2]{tree-fingerprint.pdf}
+    \end{center}
+    \caption{\label{fig:tree-heavy-query}How we find $m_i$ and $m_j$ in a \proc{HeavyTreeLce} query.}
+\end{figure}
+\end{proof}
+
+\begin{lemma}
+The \proc{HeavyTreeLce} query takes $O(\log n)$ time.
+\end{lemma}
+\begin{proof}
+Since there is at most $O(\log n)$ light edges in any simple path in $T$, the two paths we look at contains at most $O(\log n)$ light edges, and part $a$ of the algoritmh is chosen at most $O(\log n)$ times.
+Part $b$ is never chosen two times without part $a$ being chosen in between, so part $b$ is also chosen at most $O(\log n)$ times.
+Both part $a$ and $b$ take $O(1)$ time.
+In step $b.1$ we can use either \proc{SuffixNca} or \proc{LcpRmq} to achieve $O(1)$ time.
+\end{proof}
+
+\begin{comment}
+LCE(u_i,w_i,u_j,w_j)
+    v_i = u_i
+    v_j = u_j
+    v = 0
+    while v_i != w_i && v_j != w_j
+        Let m_i be the number of initial heavy edges on the path from v_i to w_i before any light edge.
         Let m_j be ...
-        m_i and m_j can be calculated in constant time.
+        //m_i and m_j can be calculated in constant time.
         Let m = min(m_i, m_j)
@@
         else
-            Do heavy path LCE starting at i_c and j_c, and let v be the LCE value.
+            v=LCE_s(index(v_i),index(v_j))
             If v >= m
@@
     O(log n) time, one step for each light edge encountered on each of the two paths.
-\end{verbatim}
+\end{comment}
 

### 2011-07-15 doc/report.tex
m

@@
 The depth of a node $u$ is one greather than the depth of $u$'s parent, and the depth of the root is zero.
+A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.
 
@@
 We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.
+
+\paragraph{Heavy Path Decomposition}
+The weight of a node $u$ is the size of the subtree rooted at $u$.
+For a given node $u$, a child of greatest weight is marked heavy, and all other children are marked light.
+The root is always marked light.
+An edge going from a node to a heavy child node is heavy, and an edge going from a node to a light child node is light.
+A path of only heavy edges is a heavy path, and the combined length of all heavy paths in a tree of $n$ nodes is less than $n$.
+The markings of heavy and light edges and nodes in a tree is a Heavy Path Decomposition.
+Every simple path in a tree of size $n$ contains at most $O(\log n)$ light edges, and we can construct a Heavy Path Decomposition in $O(n)$ time~\cite{nca}.
 
@@
 \begin{theorem}
-\tfprintk, where $k$ is a parameter $1\leq k\leq\logceil$, can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(k\cdot n^{1/k})$ worst case query time, $O(1)$ average case query time, $O(k\cdot n)$ space, and $O(n\log n)$ preprocessing time.
+\tfprintk, where $k$ is a parameter $1\leq k\leq\logceil$, can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(k\cdot n^{1/k})$ query time, $O(k\cdot n)$ space, and $O(n\log n)$ preprocessing time.
 \end{theorem}
@@
 
-\paragraph{Heavy Path Decomposition}
-The weight of a node $u$ is the size of the subtree rooted at $u$.
-For a given node $u$, its child of greatest weight is marked heavy, and all other children are marked light.
-The root is always marked light.
-An edge going from a node to a heavy child node is heavy, and an edge going from a node to a light child node is light.
-A path of only heavy edges is a heavy path, and the combined length of all heavy paths in a tree of $n$ nodes is less than $n$.
-The markings of heavy and light edges and nodes in a tree is a Heavy Path Decomposition.
-Every simple path in a tree $T$ of size $n$ contains at most $O(\log n)$ light edges.
-We can construct a Heavy Path Decomposition in WHAT??\fixme{heavy tree time} time.
-%% end of preliminary
-
-To answer LCE queries on a tree, we can create a heavy path decomposition\fixme{define heavy path decomposition} of the tree and create a string of the labels of all the heavy paths concatenated.
+To answer LCE queries on a tree, we can create a heavy path decomposition of the tree and create a string of the labels of all the heavy paths concatenated.
 We can then use constant time string LCE on this string.
-We call this algorithm \proc{HeavyTreeLce}.
+We call this algorithm \proc{HeavyPathLce}.
 
 \begin{theorem}
-\proc{HeavyTreeLce} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ worst case query time, $O(1)$ average case query time, and $O(n)$ space.
-\fixme{preproc time}
+\proc{HeavyPathLce} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
 \end{theorem}
@@
 \begin{lemma}
-The \proc{HeavyTreeLce} data structure takes $O(n)$ space.
+The \proc{HeavyPathLce} data structure takes $O(n)$ space.
 \end{lemma}
@@
 \begin{lemma}
-We can preprocess \proc{HeavyTreeLce} data structure in $\sortt$ time.
+We can preprocess \proc{HeavyPathLce} data structure in $O(\sortt)$ time.
 \end{lemma}
 \begin{proof}
-Preprocessing LCE using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time.
+Preprocessing LCE on $s$ using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time.
 If we walk through the tree and for each leaf append the heavy path ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
@@
 
-In a \proc{HeavyTreeLce} query $\LCE_T(u_i,w_i,u_j,w_j)$ we also start with $v=0$ and do the following repeatedly as long as we have not reached either $w_i$ or $w_j$:
+In a \proc{HeavyPathLce} query $\LCE_T(u_i,w_i,u_j,w_j)$ we also start with $v=0$ and do the following repeatedly as long as we have not reached either $w_i$ or $w_j$:
 \begin{enumerate}[~~~a.~]
@@
 \begin{lemma}
-The \proc{HeavyTreeLce} query finds the correct LCE value.
+The \proc{HeavyPathLce} query finds the correct LCE value.
 \end{lemma}
@@
     \end{center}
-    \caption{\label{fig:tree-heavy-query}How we find $m_i$ and $m_j$ in a \proc{HeavyTreeLce} query.}
+    \caption{\label{fig:tree-heavy-query}How we find $m_i$ and $m_j$ in a \proc{HeavyPathLce} query.}
 \end{figure}
@@
 \begin{lemma}
-The \proc{HeavyTreeLce} query takes $O(\log n)$ time.
+The \proc{HeavyPathLce} query takes $O(\log n)$ time.
 \end{lemma}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Directly Mapping Other Algorithms}
-
-\begin{verbatim}
-Directly apply SuffixNca or LcpRmq. Not possible.
-A suffix tree would take too much space.
-Each leaf would have d suffixes, where d is the deapth of the leaf.
-That is more than n.
-(Only leafs, since for internal nodes we can choose a decendant leaf and cut off the excess later)
-The input is bigger, on strings we had i (n possibilities) and j (n possibilities), and on trees we have (i_c,i_d) (>n possibilities) and (j_c,j_d) (n> possibilities).
-So directly using suffix trees or SA-LCP-RMQ would be >n space.
-\end{verbatim}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Level Ancestor way?}
-
-\begin{verbatim}
-Kan vi lige som i Level Ancestor smække to O(log n) sammen til en O(1)?
-
-Vi kan lave træ-LCE i O(hn) plads, O(1) tid ved at lave strenge for hver sti til et blad. Hver knude kender en sti den ligger på, og vi finder to stier for slut-knuderne og laver streng-LCE på dem.
-\end{verbatim}
+\subsection{Directly Mapping \proc{SuffixNca} and \proc{LcpRmq}}
+
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.25\textwidth]{quadratic-tree.pdf}
+    \end{center}
+    \caption{\label{fig:quadratic-tree}A tree with $\Omega(n^2)$ ways to choose a leaf and an ancestor of that leaf.}
+\end{figure}
+
+We have looked at applying \proc{SuffixNca} and \proc{LcpRmq} directly to trees like we did with \proc{DirectComp} and \fprintk, such that we can get $O(1)$ query time.
+
+\begin{theorem}
+\proc{TreeLcpRmq} and \proc{TreeSuffixNca} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(1)$ query time and $O(n^2)$ space.
+\end{theorem}
+\begin{proof}
+\proc{LcpRmq} relies on accessing two suffixes in SA, $i$ and $j$.
+In \proc{TreeLcpRmq} the two suffixes are pairs $(u_i,w_i)$ and $(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways.
+We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the tree original LCE query in $O(1)$ time, but with this modification a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each pair.
+The SA and LCP arrays would therefore be of length $O(n^2)$.
+When we look at \proc{TreeSuffixNca}, the same argument holds for the nessesary number of leafs in the suffix tree.
+\end{proof}
+
+Because of the quadratic space requirement, this result is not interesting when we use LCE on a tree to answer LCE queries on a compressed string in \pref{sec:lz-other}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Looking at Level Ancestor}
+
+The constant time solution for the Level Ancestor problem by Bendera~et~al.~\cite{level-ancestor} combines two $O(\log n)$ time solutions to get constant query times in linear space.
+We have looked at this algorithm to see if we can use similar techniques to achieve constant time LCE on trees in linear space.
+The two solutions used by the Level Ancestors algorithm are Jump Pointers and Ladder Decomposition.
+For the query $v=\LA(\ell,u)$, first use jump pointers to jump from $u$ half way to $v$ in constant time, and then use ladder decomposition to jump from the half way node to $v$.
+
+\proc{LadderLce}, whic we describe in the next section, finds the LCE value of the first half of the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
+We have not found a constant time solution to find the LCE value of the second half of these paths, in the samy way jump pointers is used to jump from $u$ node to the half way node node in the Level Ancestor algorithm.
+We have thus not found a constant time linear space solution for LCE on trees.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Using Constant Time String LCE in a Ladder Decomposition}
+
+%% BEGIN PRELIMINARY
+\fixme{preliminary}
+\paragraph{Ladder Decomposition}
+To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth.
+We then remove the nodes in ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
+Node $u$'s ladder is the ladder which contains $u$ at this point.
+Finally, we double the length of each ladder by extending it towards the root of the tree.
+If we reach the root before we have doubled the length of the ladder, we stop at the root.
+Each node may thus be in multiple ladders.
+
+For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path.
+If a node $u$ is half way down a path, $u$'s ladder contains all ancestors of $u$, which are in that path.
+The combined size of the ladders is at most $2n$.~\cite{level-ancestor}
+%% END PRELIMINARY
+
+We can use the ladder decomposition in the same way we use the heavy tree decomposition to get an algorithm for LCE on trees, which we call \proc{LadderLce}, and which has the same properties as \proc{HeavyPathLce}.
+
+\begin{theorem}
+\proc{LadderLce} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
+\end{theorem}
+
+\subsubsection{Data Structure}
+
+The input tree $T$ is preprocessed with a ladder decomposition, and for each ladder, the labels of nodes in the ladder forms a substring of a string $s$.
+The tree $T$ is preprocessed for constant time LA and NCA queries, and $s$ is preprocessed for constant time LCE queries.
+For a node $u$, $\textit{index}(u)$ is an index in $s$, where $\zlabel(u)$ is stored for $u$'s ladder.
+
+\fixme{INSERT FIGURE OF DS}
+\begin{lemma}
+The \proc{LadderLce} data structure takes $O(n)$ space.
+\end{lemma}
+\begin{proof}
+The string $s$ contains at most $2n$ characters. The rest of the data structures takes $O(n)$ space.
+\end{proof}
+
+\begin{lemma}
+We can preprocess the \proc{LadderLce} data structure in $O(\sortt)$ time.
+\end{lemma}
+\begin{proof}
+Constructing $s$ from the ladders takes $O(n)$ time, and preprocessing it for constant time LCE takes $O(\sortt)$ time. The rest of the data structures have $O(n)$ preprocessing times.
+\end{proof}
+
+\subsubsection{Query}
+
+The \proc{LadderLce} query is:
+\begin{enumerate}
+\item Start with $v_i = u_i$ and $v_j = u_j$
+\item Find $m_i = d(v_i')-d(v_i)$ and $i = \textit{index}(v_i') - m_i$, where $v_i'$ is the node half way from $v_i$ to $w_i$.
+\item Find $m_j$ and $j$ in the same way.
+\item Find $v' = LCE_s(i,j)$
+\item If $v' < \min(n_i,m_j)$ then stop and return $v'$ plus the distance between $u_i$ and $v_i$. Otherwise set $v_i = v_i'$ and $v_j = v_j'$ and continue from step two.
+\item Stop when $v_i = w_i$ or $v_j = w_j$.
+\end{enumerate}
+
+\begin{lemma}
+The \proc{LadderLce} query finds the correct LCE value.
+\end{lemma}
+\begin{proof}
+The algorithm maintains the invariant that the labels on the path from $u_i$ to $v_i$ is the same as the labels on the path from $u_j$ to $v_j$.
+
+In step two, $i$ refres to a position in $v_i'$'s ladder, and therefore $s[i]$ is the label of $v_i$.
+\end{proof}
+
+We split the tree into ladders and do constant time string LCE on each, just like \proc{HeavyPathLce} does. Since we need to look at at most $O(\log n)$ ladders
+ For any path in the tree, we need to look at at most $O(\log n)$ different ladders, which gives \proc{LadderLce} a query time of $O(\log n)$.
+
+\fixme{ARGH}
+
+The main ladder of the node at the middle of the path also contains the node at the beginning of the path.
+
+The ladder decomposition has an extra property, which heavy path decomposition does not have: If $\LCE(u_i,w_i,u_j,w_j)$ is less than half the length of the path from $u_i$ to $w_i$ and less than half the length of the path from $u_j$ to $w_j$, the LCE query takes constant time, since it only needs to look at one ladder.
 
@@
 
-\bibitem{nca} S. Alstrup, C. Gavoille, H. Kaplan, and T. Rauhe. Nearest Common Ancestors: A Survey and a New Algorithm for a Distributed Environment. Theory of Comput. Sys., 2004.
+\bibitem{nca} D. Harel, R. E. Tarjan. Fast Algorithms for Finding Nearest Common Ancestors. SIAM J. Comput., 1984.
+% S. Alstrup, C. Gavoille, H. Kaplan, and T. Rauhe. Nearest Common Ancestors: A Survey and a New Algorithm for a Distributed Environment. Theory of Comput. Sys., 2004.
 
@@
 
+\begin{comment}
 \subsection*{Tinta tests}
@@
 \end{algorithm}
+\end{comment}
 
@@
 \item CODE: Non-random strings? Non-random queries? Different t-values for Hash-3L?
-\item DOC: Why is Karp-Miller-Rosenberg $O(n\log n)$?.
 \end{enumerate}

### 2011-07-17 doc/report.tex
m

@@
 \begin{proof}
-In \pref{sec:tree-lce} we describe how to perform LCE queries on a tree of $n$ nodes in $O(\log n)$ query time and $O(n)$ space.
+In \pref{sec:tree-lce} we describe how to perform LCE queries on a static tree of $n$ nodes in $O(\log n)$ query time and $O(n)$ space.
 We can first find the positions of $i$ and $j$ within $z$ in $O(\log\log N)$ time, and then we can repeat the $O(\log n)$ tree LCE query \textit{\#phrases} times to find the LCE value.
@@
 Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $p$ to child node $c$ is labeled with a character $\zlabel(c)$.
-Let $u_i$, $w_i$, $u_j$ and $w_j$ be nodes in T, such that $w_i$ is a decendent of $u_i$, and $w_j$ is a decendent of $u_j$.
-Then $\LCE_T(u_i, w_i, u_j, w_j)$ is the length of the longest common prefix of $s_i$ and $s_j$, where $s_i$ is the string obtained from characters on the path from $u_i$ to $w_i$, and $s_j$ is the string obtained from characters on the path from $u_j$ to $w_j$.
+Let $p_i$ be a path in $T$ from a node $u_i$ to a node $w_i$, such that $w_i$ is a decendent of $u_i$, and
+let $p_j$ be a path in $T$ from a node $u_j$ to a node $w_j$, such that $w_j$ is a decendent of $u_j$.
+Then $\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $s_i$ and $s_j$, where
+$s_i$ is the concatenation of the edge labels of $p_i$, and
+$s_j$ is the concatenation of the edge labels of $p_j$.
 
@@
 
+To simplify our description of our tree LCE algorithms, we define $p_i(v)$ to be the node on $p_i$ with distance $v$ to $u_i$, and similarly for $p_j(v)$.
+We can calculate $p_i$ in constant time as $p_i(v)=\LA(d(u_i)+v,w_i)$, where $d(u_i)$ is the depth of $u_i$, and $\LA$ is a Level Ancestor query.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor.
-First compare $\textit{label}(\LA(d(u_i)+1,w_i))$ to $\textit{label}(\LA(d(u_j)+1,w_j))$, where $d(u)$ is the depth of node $u$, then compare $\textit{label}(\LA(d(u_i)+2,w_i))$ to $\textit{label}(\LA(d(u_j)+2,w_j))$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
-
-Each Level Ancestor query takes constant time, and we do $2|\LCE(i,j)| + 2$ of them, giving us $O(|\LCE(i,j)|)$ query time.
+We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\textit{label}(p_i(1))$ to $\textit{label}(p_j(1))$, then compare $\textit{label}(p_i(2))$ to $\textit{label}(p_j(2))$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
+
+\begin{theorem}
+\proc{TreeDirectComp} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(|\LCE(i,j)|)$ query time, and $O(n)$ space and preprocessing time.
+\end{theorem}
+\begin{proof}
+Each Level Ancestor query takes constant time, and we do $2|\LCE(i,j)| + 2$ of them, giving us $O(|\LCE(i,j)|)$ query time. The space and preprocessing requirement of Level Ancestor is $O(n)$.
+\end{proof}
 
@@
 \begin{theorem}
-\tfprintk, where $k$ is a parameter $1\leq k\leq\logceil$, can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(k\cdot n^{1/k})$ query time, $O(k\cdot n)$ space, and $O(n\log n)$ preprocessing time.
+\tfprintk, where $k$ is a parameter $1\leq k\leq\logceil$, can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(k\cdot n^{1/k})$ query time, $O(k\cdot n)$ space, and $O(n\log n)$ preprocessing time.
 \end{theorem}
@@
 \begin{lemma}
-The \tfprintk\ data structure takes $O(n)$ space.
+The \tfprintk\ data structure takes $O(k\cdot n)$ space.
 \end{lemma}
@@
 
-The quey for \tfprintk\ is nearly the same as the query for \fprintk\ on strings, except we replace $H_\ell[i+v]$ and $H_\ell[j+v]$ with $H_\ell[\LA(d(u_i)+v+t_\ell,w_i)]$ and $H_\ell[\LA(d(u_j)+v+t_\ell,w_j)]$, where $d(u)$ is the depth of node $u$.
-We also stop the algorithm, when $d(u_i)+v+t_\ell$ is greater than the depth of $w_i$, or $d(u_j)+v+t_\ell$ is greater than the depth of $w_j$.
+The quey for \tfprintk\ is nearly the same as the query for \fprintk\ on strings, except we replace $H_\ell[i+v]$ and $H_\ell[j+v]$ with $H_\ell[p_i(v+t_\ell)]$ and $H_\ell[p_j(v+t_\ell)]$.
+We also stop the algorithm after we reach $p_i(v+t_\ell)=w_i$ or $p_j(v+t_\ell)=w_j$.
 
 \begin{lemma}
-The \tfprintk\ query finds the correct tree LCE value in $O(\log n)$ time.
+The \tfprintk\ query finds the correct tree LCE value in $O(k\cdot n^{1/k})$ time.
 \end{lemma}
@@
 
-The fingerprints we choose to compare are the fingerprints, which represents the strings on the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$.
-Let $x_i$ and $y_i$ be two nodes on the path from $u_i$ to $w_i$, such that the distance between $u_i$ and $x_i$ is $v$, the distance between $x_i$ and $y_i$ is $t_\ell$ and $y_i$ is a decendent of $x_i$, like in \pref{fig:tree-query}.
-At any given iteration at level $\ell$, the query algorithm has already looked at the string on the path from $u_i$ to $x_i$, and we now need to look at the string on the path from $x_i$ to $y_i$.
-From our definition of fingerprints in a tree, this string has fingerprint $H_\ell[y_i]$.
-We have $y_i = \LA(d(u_i)+v+t_\ell,w_i)$, and the LCE query is therefore correct when we use $H_\ell[\LA(d(u_i)+v+t_\ell,w_i)]$ (and similar for $j$ instead of $i$).
+The fingerprints we choose to compare are the fingerprints, which represents substrings on $p_i$ and $p_j$.
+At any given iteration at level $\ell$, the query algorithm has already looked at the string on the path from $u_i$ to $p_i(v)$, and we now need to look at the string on the path from $p_i(v)$ to $p_i(v+t_\ell)$, as shown in \pref{fig:tree-query}.
+From our definition of fingerprints in a tree, this string has fingerprint $H_\ell[p_i(v+t_\ell)]$.
+The LCE query is therefore correct when we use $H_\ell[p_i(v+t_\ell)]$ (and similar for $j$ instead of $i$).
 \end{proof}
@@
 \begin{theorem}
-\proc{HeavyPathLce} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
+\proc{HeavyPathLce} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
 \end{theorem}
@@
 
-Let $T$ be a tree with a heavy path decomposition.
-A path label for a path in $T$ is a concatenation of all edge labels on the path starting at smallest depth and ending at greatest depth.
-The string $s$ is a concatenation of path labels of all heavy paths in $T$.
-For a node $u$, $\textit{end}(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ us the depth of $u$, and if $u$ is heavy, $\textit{index}(u)$ is the index in $s$, which stores $\zlabel(u)$.
-The tree must be preprocessed for constant time NCA and LA, and the string $s$ must be preprocessed for LCE.
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.5\textwidth,page=4]{tree-fingerprint.pdf}
+    \end{center}
+    \caption{\label{fig:tree-heavy-ds}A heavy path decomposition of a tree and a corresponding \proc{HeavyPathLce} string $s$.}
+\end{figure}
+
+The \proc{HeavyPathLce} data structure consists of the following:
+\begin{itemize}
+\item An input tree $T$ with a heavy path decomposition.
+\item A string $s$, which is a concatenation of all heavy path labels of $T$, where the label of a heavy path is the concatenation of the label on a light edge, and the labels on all edges on the heavy path following that light edge.
+\pref{fig:tree-heavy-ds} shows an example of $s$.
+\item For a node $u$, $\textit{end}(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\textit{index}(u)$ is the index in $s$, which stores $\zlabel(u)$.
+\item Data structurs for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
+\end{itemize}
 
@@
 \begin{proof}
-The string $s$ contains less than $n$ characters, since the combined length of all heavy paths in a tree of $n$ nodes is less than $n$.
-The space usage for NCA, LA, LCE, and the extra information about each node (\textit{end}, \textit{index}, \textit{d}, and heavy markings) is all $O(n)$
+The string $s$ contains $n-1$ characters, one for each edge in $T$.
+The space usage for NCA, LA, LCE, and the extra information about each node (\textit{end}, \textit{index}, and \textit{d}) is all $O(n)$.
 \end{proof}
@@
 \begin{lemma}
-We can preprocess \proc{HeavyPathLce} data structure in $O(\sortt)$ time.
+We can preprocess the \proc{HeavyPathLce} data structure in $O(\sortt)$ time.
 \end{lemma}
@@
 
-When we directly walk the paths as described in \pref{sec:tree-dc}, we start with $v=0$ and increment $v$ by one as long as $\zlabel(\LA(d(u_i)+v+1,w_i))=\zlabel(\LA(d(u_j)+v+1,w_j))$, and we have not yet reached either $w_i$ or $w_j$.
-
-In a \proc{HeavyPathLce} query $\LCE_T(u_i,w_i,u_j,w_j)$ we also start with $v=0$ and do the following repeatedly as long as we have not reached either $w_i$ or $w_j$:
-\begin{enumerate}[~~~a.~]
-  \item When either $\LA(d(u_i)+v+1,w_i)$ or $\LA(d(u_j)+v+1,w_j)$ is a light node:
-  \begin{enumerate}[~1.~]
-    \item Increment $v$ by one if $\zlabel(\LA(d(u_i)+v+1,w_i))=\zlabel(\LA(d(u_j)+v+1,w_j))$, otherwise stop and return $v$.
+In a \proc{HeavyPathLce} query $\LCE_T(p_i,p_j)$ we start with $v=0$ and do the following repeatedly:
+\begin{enumerate}
+\item Stop when we resch either $p_i(v)=w_i$ or $p_j(v)=w_j$, and return $v$.
+\item Find $v'=\LCE_s(\textit{index}(p_i(v+1)),\textit{index}(p_j(v+1)))$.
+\item Find $m_i = d(\NCA_T(\textit{end}(p_i(v+1)), w_i))-(d(u_i)+v)$.
+\item Find $m_j = d(\NCA_T(\textit{end}(p_j(v+1)), w_j))-(d(u_j)+v)$.
+\item Stop if $v'<\min(m_i,m_j)$ and return $v+v'$.
+\item Otherwise increment $v$ by $m$.
 \end{enumerate}
-  \item When both $\LA(d(u_i)+v+1,w_i)$ and $\LA(d(u_j)+v+1,w_j)$ are heavy nodes:
-  \begin{enumerate}[~1.~]
-    \item Find $v'=\LCE_s(\LA(d(u_i)+v+1,w_i),\LA(d(u_j)+v+1,w_j))$.
-    \item Find $m_i = d(\NCA_T(\textit{end}(\LA(d(u_i)+v,w_i)), w_i))-(d(u_i)+v)$.
-    \item Find $m_j = d(\NCA_T(\textit{end}(\LA(d(u_j)+v,w_j)), w_j))-(d(u_j)+v)$.
-    \item Stop and return $v+v'$, if $v'<\min(m_i,m_j)$.
-    \item otherwise increment $v$ by $m$.
-  \end{enumerate}
-\end{enumerate}
-
-When we resch either $\LA(d(u_i)+v,w_i)=w_i$ or $\LA(d(u_j)+v,w_j)=w_j$, we stop and return $v$.
 
@@
 \begin{proof}
-The query algorithm maintains the invariant $v\leq\LCE_T(u_i,w_i,u_j,w_j)$.
-Part $a$ simply checks an extra character and increments $v$ if it finds a match.
-In part $b$, $m_i$ is the number of heavy edges at the beginning of the path from $\LA(d(u_i)+v,w_i)$ to $w_i$, and similar for $m_j$, like illustrated in \pref{fig:tree-heavy-query}.
-If $v'<\min(m_i,m_j)$ it means that the LCE query found the longest common prefix to be within both of the two paths we are searching, and we have therefore found the LCE value and stop.
-If $v'\geq\min(m_i,m_j)$, we had a matching prefix which was as long or longer than the part of the heavy path we are interested in, and we therefore need to continue looking outside that heavy path to find the LCE value.
+The query algorithm maintains the invariant $v\leq\LCE_T(p_i,p_j)$.
+In step three, $m_i$ is the number of edges on a single heavy path at the beginning of the path from $p_i(v)$ to $w_i$, and similar for $m_j$ in step four, like illustrated in \pref{fig:tree-heavy-query}.
+The labels of these edges are stored concecutively in $s$.
+If $v'<\min(m_i,m_j)$, the LCE query found the longest common prefix to be within both of the two paths $p_i$ and $p_j$ we are searching, and we have therefore found the LCE value and stop.
+If $v'\geq\min(m_i,m_j)$, we had a matching prefix which was as long or longer than the part of the heavy path we are interested in, and we therefore need to continue looking at the next heavy path to find the LCE value.
 
@@
 \end{proof}
+
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.25\textwidth,page=3]{tree-fingerprint.pdf}
+    \end{center}
+    \caption{\label{fig:tree-heavy-query2}A \proc{HeavyPathLce} query with at most five iterations. Heavy edges on $p_i$ and $p_j$ are marked bold.}
+\end{figure}
 
@@
 \begin{proof}
-Since there is at most $O(\log n)$ light edges in any simple path in $T$, the two paths we look at contains at most $O(\log n)$ light edges, and part $a$ of the algoritmh is chosen at most $O(\log n)$ times.
-Part $b$ is never chosen two times without part $a$ being chosen in between, so part $b$ is also chosen at most $O(\log n)$ times.
-Both part $a$ and $b$ take $O(1)$ time.
-In step $b.1$ we can use either \proc{SuffixNca} or \proc{LcpRmq} to achieve $O(1)$ time.
+Since there is at most $O(\log n)$ light edges in any simple path in $T$, the two paths $p_i$ and $p_j$ contains at most $O(\log n)$ light edges in total.
+Since and each iteration of the query takes constant time, and we reach reach a new light edge in each iteration, as shown in \pref{fig:tree-heavy-query2}, the query time is $O(\log n)$.
 \end{proof}
@@
 \begin{theorem}
-\proc{TreeLcpRmq} and \proc{TreeSuffixNca} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(1)$ query time and $O(n^2)$ space.
+\proc{TreeLcpRmq} and \proc{TreeSuffixNca} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(1)$ query time and $O(n^2)$ space.
 \end{theorem}
@@
 \proc{LcpRmq} relies on accessing two suffixes in SA, $i$ and $j$.
-In \proc{TreeLcpRmq} the two suffixes are pairs $(u_i,w_i)$ and $(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways.
+In \proc{TreeLcpRmq} the two suffixes are pairs $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways.
 We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the tree original LCE query in $O(1)$ time, but with this modification a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each pair.
@@
 
-\proc{LadderLce}, whic we describe in the next section, finds the LCE value of the first half of the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
+\proc{LadderLce}, whic we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
 We have not found a constant time solution to find the LCE value of the second half of these paths, in the samy way jump pointers is used to jump from $u$ node to the half way node node in the Level Ancestor algorithm.
@@
 To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth.
-We then remove the nodes in ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
+We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
 Node $u$'s ladder is the ladder which contains $u$ at this point.
@@
 %% END PRELIMINARY
+\subsection*{END PRELIMINARY}
 
@@
 \begin{theorem}
-\proc{LadderLce} can preprocess a tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
+\proc{LadderLce} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
 \end{theorem}
@@
 
-The input tree $T$ is preprocessed with a ladder decomposition, and for each ladder, the labels of nodes in the ladder forms a substring of a string $s$.
-The tree $T$ is preprocessed for constant time LA and NCA queries, and $s$ is preprocessed for constant time LCE queries.
-For a node $u$, $\textit{index}(u)$ is an index in $s$, where $\zlabel(u)$ is stored for $u$'s ladder.
-
-\fixme{INSERT FIGURE OF DS}
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=0.4\textwidth,page=1]{ladder.pdf}
+        \includegraphics[width=0.4\textwidth,page=2]{ladder.pdf}
+    \end{center}
+    \caption{\label{fig:ladder-ds} Ladders of a tree before and after their length is doubled.
+A \proc{LadderLce} string $s$ for this ladder decomposition is $s=abc|ade|f|ghijkl|hno|m$, where the splits between ladders are marked with a vertical line.}
+\end{figure}
+
+The \proc{LadderLce} data structure consists of the following:
+\begin{itemize}
+\item An input tree $T$ with a ladder decomposition.
+\item A strings $s$, which is a concatenation of ladder labels for each ladder in $T$, where a ladder label is the concatenation of labels of edges between nodes in the ladder.
+\pref{fig:ladder-ds} shows an example of $s$.
+\item For a node $u$, $\textit{index}(u)$ is an index in $s$ within the ladder label of $u$'s ladder, where $\zlabel(u)$ is stored.
+\item Data structures for constant time LA on $T$ and constant time LCE on $s$.
+\end{itemize}
+
 \begin{lemma}
@@
 
-The \proc{LadderLce} query is:
+The \proc{LadderLce} query starts with $v=0$ and repeats the following:
 \begin{enumerate}
-\item Start with $v_i = u_i$ and $v_j = u_j$
-\item Find $m_i = d(v_i')-d(v_i)$ and $i = \textit{index}(v_i') - m_i$, where $v_i'$ is the node half way from $v_i$ to $w_i$.
-\item Find $m_j$ and $j$ in the same way.
+\item Stop when $d_i(v) = w_i$ or $d_j(v) = w_j$ and return $v$.
+\item Find $m_i = d(v_i')-d(p_i(v))$ and $i = \textit{index}(v_i') - m_i + 1$, where $v_i'$ is the node half way from $p_i(v)$ to $w_i$.
+\item Find $m_j = d(v_j')-d(p_j(v))$ and $j = \textit{index}(v_j') - m_j + 1$, where $v_j'$ is the node half way from $p_j(v)$ to $w_j$.
 \item Find $v' = LCE_s(i,j)$
-\item If $v' < \min(n_i,m_j)$ then stop and return $v'$ plus the distance between $u_i$ and $v_i$. Otherwise set $v_i = v_i'$ and $v_j = v_j'$ and continue from step two.
-\item Stop when $v_i = w_i$ or $v_j = w_j$.
+\item Stop if $v' < \min(m_i,m_j)$ and return $v+v'$.
+\item Otherwise add $\min(m_i,m_j)$ to $v$.
 \end{enumerate}
@@
 \begin{proof}
-The algorithm maintains the invariant that the labels on the path from $u_i$ to $v_i$ is the same as the labels on the path from $u_j$ to $v_j$.
-
-In step two, $i$ refres to a position in $v_i'$'s ladder, and therefore $s[i]$ is the label of $v_i$.
-\end{proof}
-
-We split the tree into ladders and do constant time string LCE on each, just like \proc{HeavyPathLce} does. Since we need to look at at most $O(\log n)$ ladders
- For any path in the tree, we need to look at at most $O(\log n)$ different ladders, which gives \proc{LadderLce} a query time of $O(\log n)$.
-
-\fixme{ARGH}
-
-The main ladder of the node at the middle of the path also contains the node at the beginning of the path.
-
-The ladder decomposition has an extra property, which heavy path decomposition does not have: If $\LCE(u_i,w_i,u_j,w_j)$ is less than half the length of the path from $u_i$ to $w_i$ and less than half the length of the path from $u_j$ to $w_j$, the LCE query takes constant time, since it only needs to look at one ladder.
+The algorithm maintains the invariant that $v\leq\LCE(p_i,p_j)$. Which means that we know that $\zlabel(p_i(v))=\zlabel(p_j(v))$, but we do not know if $\zlabel(p_i(v+1))=\zlabel(p_j(v+1))$.
+
+In step two, $i$ refers to a position in $v_i'$'s ladder, and $s[i]$ is the label of $p_i(v+1)$.
+Similarly $s[j]$ is the label of $p_j(v+1)$.
+The LCE query thus determines the LCE value starting from $p_i(v+1)$ and $p_j(v+1)$ until we reach either $v_i'$ or $v_j'$, whichever comes first, which is distance $\min(m_i,m_j)$ from $d_i(v)$ and $d_j(v)$.
+When $v' < \min(m_i,m_j)$, we have found the final LCE value within the current pair of ladders and we can stop, otherwise we need to continue looking at other ladders.
+\end{proof}
+
+\begin{lemma}
+The \proc{LadderLce} query takes $O(\log n)$ time.
+\end{lemma}
+\begin{proof}
+Each iteration is constant time, and in each iteration we halve the distance between $d_i(v)$ and $w_i$ or between $d_j(v)$ and $w_j$.
+\end{proof}
+
+\proc{LadderLce} is very similar to \proc{HeavyPathLce}, however it has an extra property, which \proc{HeavyPathLce} does not have: If $\LCE(p_i,p_j)$ is less than half the length of $p_i$ and less than half the length of $p_j$, the LCE query takes constant time, since it only needs to look at one ladder.
 

### 2011-07-17 doc/report.tex
m

@@
 Optimal solutions exists for the problem, with constant query time and linear space.
-These teoretically optimal solutions are however not the best in practice, since they have large constant factors for both time and space usage.
+These theoretically optimal solutions are however not the best in practice, since they have large constant factors for both time and space usage.
 In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance.
@@
 
-We have found two new algortihms.
+We have found two new algorithms.
 They both have better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms.
@@
 
-In \pref{sec:algorithms} we theorectically describe and analyze the LCE algorithms introduced here in \pref{sec:intro}, and in \pref{sec:results} we analyze results from practical C++ implementations of these algorithms.
+In \pref{sec:algorithms} we theoretically describe and analyze the LCE algorithms introduced here in \pref{sec:intro}, and in \pref{sec:results} we analyze results from practical C++ implementations of these algorithms.
 
@@
 It uses \proc{DirectComp} to move towards one of these stored values, and stops when it reaches one.
-The \proc{DirectCompLookup} algorithm is very simple, but the tradoff between worst case query time and space is not very favorable, and preprocessing the data structure takes $O(n^2)$ time unless we use another LCE algorithm in the preprocessing.
+The \proc{DirectCompLookup} algorithm is very simple, but the tradeoff between worst case query time and space is not very favorable, and preprocessing the data structure takes $O(n^2)$ time unless we use another LCE algorithm in the preprocessing.
 
@@
 The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
-The depth of a node $u$ is one greather than the depth of $u$'s parent, and the depth of the root is zero.
+The depth of a node $u$ is one greater than the depth of $u$'s parent, and the depth of the root is zero.
 A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.
@@
 There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes.
-Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be reprecented by two indexes of $s$.
+Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be represented by two indexes of $s$.
 The suffix tree therefore takes $O(n)$ space.
@@
 \paragraph{Predecessor.}
-Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natual number $x$ is an object with maximal key less than or equal to $x$.
+Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key less than or equal to $x$.
 We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.
@@
 \paragraph{Level Ancestor.}
-The Level Ancestor at level $\ell$ of node $u$, $\LA(\ell,u)$, is the ancestor of $u$, whoes depth is $\ell$.
+The Level Ancestor at level $\ell$ of node $u$, $\LA(\ell,u)$, is the ancestor of $u$, whose depth is $\ell$.
 We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.
@@
 Every simple path in a tree of size $n$ contains at most $O(\log n)$ light edges, and we can construct a Heavy Path Decomposition in $O(n)$ time~\cite{nca}.
+
+\paragraph{Ladder Decomposition}
+To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth.
+We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
+Node $u$'s ladder is the ladder which contains $u$ at this point.
+Finally, we double the number of nodes in each ladder by extending it towards the root of the tree.
+If we reach the root before we have doubled the number of nodes, we stop at the root.
+Each node may thus be in multiple ladders.
+
+For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path.
+If a node $u$ has height $h$, $u$'s ladder contains at least the last $h$ ancestors of $u$.
+The combined size of the ladders is at most $2n$.~\cite{level-ancestor}
 
@@
 
-For a string of length $n$ with aplhabet size $\sigma$, we can preprocess \proc{SuffixNca} in $O(\sortt)$ time by building the suffix tree, preprocessing it for NCA and building the array of label depths.
+For a string of length $n$ with alphabet size $\sigma$, we can preprocess \proc{SuffixNca} in $O(\sortt)$ time by building the suffix tree, preprocessing it for NCA and building the array of label depths.
 The query time is $O(1)$, since looking up $L_i$ and $L_j$, doing an NCA query, and looking up the label depth in $D$ all takes constant time.
@@
 The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
-\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct seach through a $O(\log n)$ length block.
+\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct search through a $O(\log n)$ length block.
 \end{description}
@@
 The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$.
-However if we structure our tables of fingerprints differently, we can improve ths number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
+However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
 Instead of storing the fingerprint of $s[i\twodots i+t_\ell]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$.
@@
 
-For larger values of $k$, we can attempt to optimize memory access accros different levels.
+For larger values of $k$, we can attempt to optimize memory access accross different levels.
 When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$.
@@
 We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other.
-This will not improve the asymptotic I/O bound, but could impove practical query times.
+This will not improve the asymptotic I/O bound, but could improve practical query times.
 
@@
 The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater bc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater wc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
 \item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
@@
 
-We want to compare our new algortithms against the algortihm which is the best in practice of the asymptotically best $O(1)$ algorithms.
+We want to compare our new algorithms against the algorithm which is the best in practice of the asymptotically best $O(1)$ algorithms.
 The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at \proc{LcpRmq}.
@@
 \pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs together with \proc{LcpRmq}.
-In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra ckeck to find out if it has reached a position for which a result is stored.
+In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra check to find out if it has reached a position for which a result is stored.
 For short strings, \proc{DirectLookup} is faster than \proc{DirectCompLookup}, since it does not need to check if the value it needs is stored.
@@
 
-On worst case strings, the \fprintk\ algorithms are signicicantly better than \proc{DirectComp} and somewhat worse than \proc{LcpRmq}.
+On worst case strings, the \fprintk\ algorithms are significantly better than \proc{DirectComp} and somewhat worse than \proc{LcpRmq}.
 Up until $n=30.000$ the three measured \fprintk\ algorithms have nearly the same query times.
@@
 
-On strings with medium aveage LCE values, we see that our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}
-
-We conclude that our new \fprintk\ algorithm acieves a tradeoff between worst case times and average case times, which is better than the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, yet it is not strictly better than the existing algorithms on all inputs.
+On strings with medium average LCE values, we see that our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}
+
+We conclude that our new \fprintk\ algorithm achieves a tradeoff between worst case times and average case times, which is better than the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, yet it is not strictly better than the existing algorithms on all inputs.
 \fprintk\ is therefore a good choice in cases where both average case and worst case performance is important.
@@
 
-In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the aveage case optimization.
-Both variations of \fprint[\logceil] show query times, whic are $O(\log n)$ as expected.
+In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the average case optimization.
+Both variations of \fprint[\logceil] show query times, which are $O(\log n)$ as expected.
 On average case strings, the query time of \fprint[\logceil] improves significantly when using average case optimization, and both variants show their expected $O(1)$ and $O(\log n)$ average case query times.
@@
 
-We conclude that using average case optimization yields a query time improvement on average case strings, which is significantly greater than the query time degration on worst case strings, and that \fprintk\ therefore should use the average case optimization.
+We conclude that using average case optimization yields a query time improvement on average case strings, which is significantly greater than the query time increase on worst case strings, and that \fprintk\ therefore should use the average case optimization.
 
@@
 
-On average case strings the cache optimization does not change the query times, while on worst case strings and strings with medium size LCE values, cache optimization gives a noticable improvement for large inputs.
+On average case strings the cache optimization does not change the query times, while on worst case strings and strings with medium size LCE values, cache optimization gives a noticeable improvement for large inputs.
 The cache optimized \fprint[3] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.
@@
 \fprintk\ gives a good time space tradeoff, and it uses less space than \proc{LcpRmq} when $k$ is small.
-The perfprmance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
+The performance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
 
@@
 The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified.
-We have made it compile on a 32-bit Windows archictecture, and generalized it to allow more algorithms and tests to be implemented.
+We have made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented.
 \pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.
@@
 \end{tabular}
-\caption{List of performance veriables tested.}\label{tab:test-variables}
+\caption{List of performance variables tested.}\label{tab:test-variables}
 \end{table}
@@
 \item A character in a string is always eight bits.
-\item A string always ends with a special null character, which does not occur anywehere else in the string.
+\item A string always ends with a special null character, which does not occur anywhere else in the string.
 This does not affect our results, since we do not measure practical preprocessing time.
@@
 We do not measure this case.
-\item Wehenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
+\item Whenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
 \end{itemize}
@@
 
-As an alternative way of testing our alorithms, we have looked at how long a query takes in relation to the LCE value it returns, since most of our algorithms have query times, which depend on the returned LCE value.
+As an alternative way of testing our algorithms, we have looked at how long a query takes in relation to the LCE value it returns, since most of our algorithms have query times, which depend on the returned LCE value.
 \pref{fig:plot-by-value} shows query times for a fixed string of 1000 characters.
@@
 We need to measure a sum of many queries to get an accurate timing, since each individual query is too fast to measure.
-While measureing a sequence of randomly generated queries is a very artificial test, which is unlikely to occur in practice, we think that it is even more artificial to measure a sequence of queries for which we already know that the LCE value is the same each time.
+While measuring a sequence of randomly generated queries is a very artificial test, which is unlikely to occur in practice, we think that it is even more artificial to measure a sequence of queries for which we already know that the LCE value is the same each time.
 On our worst case string $s=aaaaaaa...$, the queries we measure in each point in \pref{fig:plot-by-value} have the pattern $\max(i,j)=v$, which may give very different performance characteristics compared to random queries due to cache effects.
@@
 \end{align*}
-where "$\cdot$" denotes string contatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.
+where "$\cdot$" denotes string concatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.
 
@@
 \end{enumerate}
-The zompressed string $z$ is now the tree except the special root node $z_0$.
+The compressed string $z$ is now the tree except the special root node $z_0$.
 
@@
 
-We can adapt \proc{DirectComp} to work on a compressed string. To work directly on a compressed string, we first need to have ramdom access to the characters of the original string.
+We can adapt \proc{DirectComp} to work on a compressed string. To work directly on a compressed string, we first need to have random access to the characters of the original string.
 
@@
 The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree.
-The LCE value is the length of a substring occuring at two positions $i$ and $j$ in the input string.
+The LCE value is the length of a substring, which occurs at two positions $i$ and $j$ in the input string.
 This substring spans a number of phrases, where each phrase is a path in the tree of compression elements.
@@
 
-We could lay out the tree of compression elements in memory as a B-tree, where we store groups of adjacant nodes next to eachother in memory, to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the worst case number of I/O's, as each node visited may be in a new group.
+We could lay out the tree of compression elements in memory as a B-tree, where we store groups of adjacent nodes next to each other in memory, to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the worst case number of I/O's, as each node visited may be in a new group.
 
@@
 Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $p$ to child node $c$ is labeled with a character $\zlabel(c)$.
-Let $p_i$ be a path in $T$ from a node $u_i$ to a node $w_i$, such that $w_i$ is a decendent of $u_i$, and
-let $p_j$ be a path in $T$ from a node $u_j$ to a node $w_j$, such that $w_j$ is a decendent of $u_j$.
+Let $p_i$ be a path in $T$ from a node $u_i$ to a node $w_i$, such that $w_i$ is a descendant of $u_i$, and
+let $p_j$ be a path in $T$ from a node $u_j$ to a node $w_j$, such that $w_j$ is a descendant of $u_j$.
 Then $\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $s_i$ and $s_j$, where
@@
 
-This definition of LCE on a tree is chosen to match the tree of LZ compression elements and the nessecary LCE queries on it described in \pref{sec:lz-other}.
+This definition of LCE on a tree is chosen to match the tree of LZ compression elements and the necessary LCE queries on it described in \pref{sec:lz-other}.
 
@@
 For a given parameter $k$, where $1\leq k \leq \logceil$, we define the \tfprintk\ data structure as follows: For each number $\ell$ between $0$ and $k-1$, let $t_\ell=\Theta(n^{\ell/k})$, and let $t_0 = 1$.
-For each number $\ell$ between $0$ and $k-1$ and each node $u$ in $T$, whoes depth is at least $t_\ell$, let $H_\ell[u] = F_{t_\ell}[u]$
+For each number $\ell$ between $0$ and $k-1$ and each node $u$ in $T$, whose depth is at least $t_\ell$, let $H_\ell[u] = F_{t_\ell}[u]$
 
@@
     \end{center}
-    \caption{\label{fig:tree-query}Finging the right fingerprint to look at for a \tfprintk\ query.}
+    \caption{\label{fig:tree-query}Finding the right fingerprint to look at for a \tfprintk\ query.}
 \end{figure}
@@
 \item Sort the pairs lexicographically.
-\item Replace each pair with a new fingerprint, by going throug them in lexichographical order, while updating the fingerprint whenever the pair changes.
+\item Replace each pair with a new fingerprint, by going through them in lexicographical order, while updating the fingerprint whenever the pair changes.
 \end{enumerate}
@@
 When $k < \logceil$, the condition that $t_\ell \leq 2t_{\ell-1}$ is not satisfied.
-In this case we need to add extra levels in the preprocessing, which we will only store temporarely.
+In this case we need to add extra levels in the preprocessing, which we will only store temporarily.
 We therefore always need to preprocess $O(\log n)$ levels, even if we only store some of them.
@@
 Step one of the preprocessing algorithm makes each fingerprint at level $\ell$ a natural number less than $n^2$.
-Step two and three of the preprocessing algorithm reduceses that universe of fingerprints back to $n$ without changing the equality of the fingerprints.
+Step two and three of the preprocessing algorithm reduces that universe of fingerprints back to $n$ without changing the equality of the fingerprints.
 \end{proof}
@@
 \item For a node $u$, $\textit{end}(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\textit{index}(u)$ is the index in $s$, which stores $\zlabel(u)$.
-\item Data structurs for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
+\item Data structures for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
 \end{itemize}
@@
 \begin{enumerate}
-\item Stop when we resch either $p_i(v)=w_i$ or $p_j(v)=w_j$, and return $v$.
+\item Stop when we reach either $p_i(v)=w_i$ or $p_j(v)=w_j$, and return $v$.
 \item Find $v'=\LCE_s(\textit{index}(p_i(v+1)),\textit{index}(p_j(v+1)))$.
@@
 \item Stop if $v'<\min(m_i,m_j)$ and return $v+v'$.
-\item Otherwise increment $v$ by $m$.
+\item Otherwise increment $v$ by $\min(m_i,m_j)$.
 \end{enumerate}
@@
 In step three, $m_i$ is the number of edges on a single heavy path at the beginning of the path from $p_i(v)$ to $w_i$, and similar for $m_j$ in step four, like illustrated in \pref{fig:tree-heavy-query}.
-The labels of these edges are stored concecutively in $s$.
+The labels of these edges are stored consecutively in $s$.
 If $v'<\min(m_i,m_j)$, the LCE query found the longest common prefix to be within both of the two paths $p_i$ and $p_j$ we are searching, and we have therefore found the LCE value and stop.
@@
 The SA and LCP arrays would therefore be of length $O(n^2)$.
-When we look at \proc{TreeSuffixNca}, the same argument holds for the nessesary number of leafs in the suffix tree.
+When we look at \proc{TreeSuffixNca}, the same argument holds for the necessary number of leafs in the suffix tree.
 \end{proof}
@@
 \proc{LadderLce}, whic we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
-We have not found a constant time solution to find the LCE value of the second half of these paths, in the samy way jump pointers is used to jump from $u$ node to the half way node node in the Level Ancestor algorithm.
+We have not found a constant time solution to find the LCE value of the second half of these paths, in the same way jump pointers is used to jump from $u$ node to the half way node node in the Level Ancestor algorithm.
 We have thus not found a constant time linear space solution for LCE on trees.
@@
 \subsection{Using Constant Time String LCE in a Ladder Decomposition}
-
-%% BEGIN PRELIMINARY
-\fixme{preliminary}
-\paragraph{Ladder Decomposition}
-To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth.
-We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
-Node $u$'s ladder is the ladder which contains $u$ at this point.
-Finally, we double the length of each ladder by extending it towards the root of the tree.
-If we reach the root before we have doubled the length of the ladder, we stop at the root.
-Each node may thus be in multiple ladders.
-
-For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path.
-If a node $u$ is half way down a path, $u$'s ladder contains all ancestors of $u$, which are in that path.
-The combined size of the ladders is at most $2n$.~\cite{level-ancestor}
-%% END PRELIMINARY
-\subsection*{END PRELIMINARY}
 
@@
 
-In step two, $i$ refers to a position in $v_i'$'s ladder, and $s[i]$ is the label of $p_i(v+1)$.
-Similarly $s[j]$ is the label of $p_j(v+1)$.
+In step two, $i$ refers to a position in $v_i'$'s ladder, because the height of $v_i'$ is at least $m_i$.
+Therefore $s[i\twodots i+m_i-1]$ is the label of the path from $p_i(v)$ to $v_i'$, and similarly $s[j\twodots m_i-1]$ is the label of the path from $p_j(v)$ to $v_j'$.
 The LCE query thus determines the LCE value starting from $p_i(v+1)$ and $p_j(v+1)$ until we reach either $v_i'$ or $v_j'$, whichever comes first, which is distance $\min(m_i,m_j)$ from $d_i(v)$ and $d_j(v)$.

### 2011-07-18 doc/report.tex
m

@@
 At each iteration, it then updates $(x_i,y_i)$ and $(x_j,y_j)$ to find the previous two positions.
-If $x_i = x_j$ we know that we can add $\textit{depth}[x_i]$ to the LCE value, where \textit{depth} is the depth of a node in the tree of compression elements.
-Compressed \proc{DirectComp} can then compare $\textit{depth}[x_i]$ characters in constant time.
+If $x_i = x_j$ we know that we can add $d(x_i)$ to the LCE value, where $d(x_i)$ is the depth of node $x_i$ in the tree of compression elements.
+Compressed \proc{DirectComp} can then compare $d(x_i)$ characters in constant time.
 
@@
 
+\newcommand{\zlength}{\textit{length}}
+\newcommand{\zindex}{\textit{index}}
+\newcommand{\zend}{\textit{end}}
+
 To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space. In this section we analyze ways to achieve this.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Definition\label{sec:tree-def}}
-
-Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $p$ to child node $c$ is labeled with a character $\zlabel(c)$.
-Let $p_i$ be a path in $T$ from a node $u_i$ to a node $w_i$, such that $w_i$ is a descendant of $u_i$, and
-let $p_j$ be a path in $T$ from a node $u_j$ to a node $w_j$, such that $w_j$ is a descendant of $u_j$.
-Then $\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $s_i$ and $s_j$, where
-$s_i$ is the concatenation of the edge labels of $p_i$, and
-$s_j$ is the concatenation of the edge labels of $p_j$.
+\subsection{Definitions\label{sec:tree-def}}
+
+\paragraph{Trees}
+Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $u$ to child node $w$ is labeled with a character $\zlabel(w)$.
+
+\paragraph{Paths}
+A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on path $p=(u,w)$:
+\begin{itemize}
+\item The path has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
+\item The node $p[a]$ is a node in $p$, such that the distance between $u$ and $p[a]$ is $a$.
+The start and end nodes of the path are $p[0]=u$ and $p[e]=w$.
+We can calculate $p[a]$ in constant time as $p[a]=\LA(d(u)+a,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
+\item The subpath $p[a\twodots b]=(p[a],p[b])$ is the path, which starts at node $p[a]$ and ends at node $p[b]$. The entire path is $p = p[0,e]$.
+\item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
+\end{itemize}
+
+\paragraph{LCE on trees}
+Given two paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$ of lengths $e_i$ and $e_j$ in a tree $T$, the Longest Common Extension $v=\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $\zlabel(p_i)$ and $\zlabel(p_j)$.
+
+In other words $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and either $v=e_i$, $v=e_j$, or $\zlabel(p_i[v+1]) \neq \zlabel(p_j[v+1])$.
 
@@
 
-To simplify our description of our tree LCE algorithms, we define $p_i(v)$ to be the node on $p_i$ with distance $v$ to $u_i$, and similarly for $p_j(v)$.
-We can calculate $p_i$ in constant time as $p_i(v)=\LA(d(u_i)+v,w_i)$, where $d(u_i)$ is the depth of $u_i$, and $\LA$ is a Level Ancestor query.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\textit{label}(p_i(1))$ to $\textit{label}(p_j(1))$, then compare $\textit{label}(p_i(2))$ to $\textit{label}(p_j(2))$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
+We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\zlabel(p_i[1])$ to $\zlabel(p_j[1])$, then compare $\zlabel(p_i[2])$ to $\zlabel(p_j[2])$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
 
@@
 
-The quey for \tfprintk\ is nearly the same as the query for \fprintk\ on strings, except we replace $H_\ell[i+v]$ and $H_\ell[j+v]$ with $H_\ell[p_i(v+t_\ell)]$ and $H_\ell[p_j(v+t_\ell)]$.
-We also stop the algorithm after we reach $p_i(v+t_\ell)=w_i$ or $p_j(v+t_\ell)=w_j$.
+The query for \tfprintk\ is nearly the same as the query for \fprintk\ on strings, except we replace $H_\ell[i+v]$ and $H_\ell[j+v]$ with $H_\ell[p_i[v+t_\ell]]$ and $H_\ell[p_j[v+t_\ell]]$.
+We also decrement the level $\ell$ whenever we reach $v+t_\ell=e_i$ or $v+t_\ell=e_j$, since we no longer have a special character \$ to tell when we have reached the end.
 
@@
 The fingerprints we choose to compare are the fingerprints, which represents substrings on $p_i$ and $p_j$.
-At any given iteration at level $\ell$, the query algorithm has already looked at the string on the path from $u_i$ to $p_i(v)$, and we now need to look at the string on the path from $p_i(v)$ to $p_i(v+t_\ell)$, as shown in \pref{fig:tree-query}.
-From our definition of fingerprints in a tree, this string has fingerprint $H_\ell[p_i(v+t_\ell)]$.
-The LCE query is therefore correct when we use $H_\ell[p_i(v+t_\ell)]$ (and similar for $j$ instead of $i$).
+At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$, as shown in \pref{fig:tree-query}.
+From our definition of fingerprints in a tree, these two labels have fingerprints $H_\ell[p_i[v+t_\ell]]$ and  $H_\ell[p_j[v+t_\ell]]$, which is what our LCE query uses.
 \end{proof}
@@
 \item An input tree $T$ with a heavy path decomposition.
-\item A string $s$, which is a concatenation of all heavy path labels of $T$, where the label of a heavy path is the concatenation of the label on a light edge, and the labels on all edges on the heavy path following that light edge.
+\item A string $s$, which is a concatenation of strings of all light edges of $T$, where the string of a light edge is the label of the light edge concatenated with the label of the heavy path, which follows that light edge.
 \pref{fig:tree-heavy-ds} shows an example of $s$.
-\item For a node $u$, $\textit{end}(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\textit{index}(u)$ is the index in $s$, which stores $\zlabel(u)$.
+\item For a node $u$, $\zend(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\zindex(u)$ is the index in $s$, which stores $\zlabel(u)$.
 \item Data structures for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
@@
 \begin{proof}
-The string $s$ contains $n-1$ characters, one for each edge in $T$.
-The space usage for NCA, LA, LCE, and the extra information about each node (\textit{end}, \textit{index}, and \textit{d}) is all $O(n)$.
+The string $s$ contains $n-1$ characters, one for each edge in $T$. The space usage for NCA, LA, LCE, and the extra information about each node (\zend, \zindex, and $d$) is all $O(n)$.
 \end{proof}
@@
 Preprocessing LCE on $s$ using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time.
-If we walk through the tree and for each leaf append the heavy path ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
+If we walk through the tree and for each leaf append the light edge string ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
 The rest of the data structures can be preprocessed in $O(n)$ time.
@@
 \begin{enumerate}
-\item Stop when we reach either $p_i(v)=w_i$ or $p_j(v)=w_j$, and return $v$.
-\item Find $v'=\LCE_s(\textit{index}(p_i(v+1)),\textit{index}(p_j(v+1)))$.
-\item Find $m_i = d(\NCA_T(\textit{end}(p_i(v+1)), w_i))-(d(u_i)+v)$.
-\item Find $m_j = d(\NCA_T(\textit{end}(p_j(v+1)), w_j))-(d(u_j)+v)$.
-\item Stop if $v'<\min(m_i,m_j)$ and return $v+v'$.
-\item Otherwise increment $v$ by $\min(m_i,m_j)$.
+\item Stop when we reach either $v=e_i$ or $v=e_j$, and return $v$.
+\item Find $m_i = \zlength(d_i[v], \NCA_T(\zend(p_i[v+1]), w_i))$ and $i=\zindex(p_i[v+1])$.
+\item Find $m_j = \zlength(d_j[v], \NCA_T(\zend(p_j[v+1]), w_j))$ and $j=\zindex(p_j[v+1])$.
+\item Find $v'=\min\{\LCE_s(i,j),m_i,m_j\}$.
+\item Increment $v$ by $v'$.
+\item Stop if $v'<\min\{m_i,m_j\}$ and return $v$.
 \end{enumerate}
@@
 \begin{proof}
-The query algorithm maintains the invariant $v\leq\LCE_T(p_i,p_j)$.
-In step three, $m_i$ is the number of edges on a single heavy path at the beginning of the path from $p_i(v)$ to $w_i$, and similar for $m_j$ in step four, like illustrated in \pref{fig:tree-heavy-query}.
-The labels of these edges are stored consecutively in $s$.
-If $v'<\min(m_i,m_j)$, the LCE query found the longest common prefix to be within both of the two paths $p_i$ and $p_j$ we are searching, and we have therefore found the LCE value and stop.
-If $v'\geq\min(m_i,m_j)$, we had a matching prefix which was as long or longer than the part of the heavy path we are interested in, and we therefore need to continue looking at the next heavy path to find the LCE value.
+In step two, $m_i$ is the number of edges on a single heavy path at the beginning of $p_i[v\twodots e_i]$, like illustrated in \pref{fig:tree-heavy-query}.
+The labels of these edges are stored consecutively in $s$, such that $\zlabel(p_i[v\twodots v+m_i])=s[i\twodots i + m_i-1]$.
+The same holds for $m_j$ in step three.
+
+The algorithm keeps the invariant $\zlabel(p_i[0\twodots v]) = \zlabel(p_j[0\twodots v])$.
+At each iteration we find $v'$, which is the longest common prefix of $\zlabel(p_i[v\twodots v+m_i])$ and $\zlabel(p_j[v\twodots v+m_j])$.
+This gives $\zlabel(p_i[0\twodots v+v']) = \zlabel(p_j[0\twodots v+v'])$, and we therefore keep the invariant when we add $v'$ to $v$ in step five.
+
+If $v'<\min\{m_i,m_j\}$ in step six, we also know that $\zlabel(p_i[v+1]) \neq \zlabel(p_j[v+1])$, and therefore that $v=\LCE(p_i,p_j)$.
 
@@
 Since there is at most $O(\log n)$ light edges in any simple path in $T$, the two paths $p_i$ and $p_j$ contains at most $O(\log n)$ light edges in total.
-Since and each iteration of the query takes constant time, and we reach reach a new light edge in each iteration, as shown in \pref{fig:tree-heavy-query2}, the query time is $O(\log n)$.
+Since and each iteration of the query takes constant time, and we reach reach a new light edge of either $p_i$ or $p_j$ in each iteration, as shown in \pref{fig:tree-heavy-query2}, the query time is $O(\log n)$.
 \end{proof}
@@
 \proc{LcpRmq} relies on accessing two suffixes in SA, $i$ and $j$.
-In \proc{TreeLcpRmq} the two suffixes are pairs $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways.
-We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the tree original LCE query in $O(1)$ time, but with this modification a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each pair.
+In \proc{TreeLcpRmq} we change the two suffixes to paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways.
+We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the original LCE query in $O(1)$ time, but even with this modification, a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each path.
 The SA and LCP arrays would therefore be of length $O(n^2)$.
@@
 \item An input tree $T$ with a ladder decomposition.
-\item A strings $s$, which is a concatenation of ladder labels for each ladder in $T$, where a ladder label is the concatenation of labels of edges between nodes in the ladder.
+\item A strings $s$, which is a concatenation of ladder labels for each ladder in $T$, where a ladder label for is $\zlabel(p)$, where $p$ is the path from the first node in the ladder to the last node in the ladder.
 \pref{fig:ladder-ds} shows an example of $s$.
-\item For a node $u$, $\textit{index}(u)$ is an index in $s$ within the ladder label of $u$'s ladder, where $\zlabel(u)$ is stored.
+\item For a node $u$, $\zindex(u)$ is an index in $s$ within the ladder label of $u$'s ladder, where $\zlabel(u)$ is stored.
 \item Data structures for constant time LA on $T$ and constant time LCE on $s$.
@@
 \begin{proof}
-The string $s$ contains at most $2n$ characters. The rest of the data structures takes $O(n)$ space.
+The string $s$ contains at most $2n$ characters. The rest of the data structures take $O(n)$ space.
 \end{proof}
@@
 \begin{enumerate}
-\item Stop when $d_i(v) = w_i$ or $d_j(v) = w_j$ and return $v$.
-\item Find $m_i = d(v_i')-d(p_i(v))$ and $i = \textit{index}(v_i') - m_i + 1$, where $v_i'$ is the node half way from $p_i(v)$ to $w_i$.
-\item Find $m_j = d(v_j')-d(p_j(v))$ and $j = \textit{index}(v_j') - m_j + 1$, where $v_j'$ is the node half way from $p_j(v)$ to $w_j$.
-\item Find $v' = LCE_s(i,j)$
-\item Stop if $v' < \min(m_i,m_j)$ and return $v+v'$.
-\item Otherwise add $\min(m_i,m_j)$ to $v$.
+\item Stop when $v = e_i$ or $v = e_j$ and return $v$.
+\item Find $m_i = \zlength(p_i[v\twodots e_i])/2$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
+\item Find $m_j = \zlength(p_j[v\twodots e_j])/2$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
+\item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
+\item Increment $v$ by $v'$.
+\item Stop if $v' < \min\{m_i,m_j\}$ and return $v$.
 \end{enumerate}
@@
 \begin{proof}
-The algorithm maintains the invariant that $v\leq\LCE(p_i,p_j)$. Which means that we know that $\zlabel(p_i(v))=\zlabel(p_j(v))$, but we do not know if $\zlabel(p_i(v+1))=\zlabel(p_j(v+1))$.
-
-In step two, $i$ refers to a position in $v_i'$'s ladder, because the height of $v_i'$ is at least $m_i$.
-Therefore $s[i\twodots i+m_i-1]$ is the label of the path from $p_i(v)$ to $v_i'$, and similarly $s[j\twodots m_i-1]$ is the label of the path from $p_j(v)$ to $v_j'$.
-The LCE query thus determines the LCE value starting from $p_i(v+1)$ and $p_j(v+1)$ until we reach either $v_i'$ or $v_j'$, whichever comes first, which is distance $\min(m_i,m_j)$ from $d_i(v)$ and $d_j(v)$.
-When $v' < \min(m_i,m_j)$, we have found the final LCE value within the current pair of ladders and we can stop, otherwise we need to continue looking at other ladders.
+In step two, $i$ refers to a position in $p_i[v+m_i]$'s ladder, because the height of $p_i[v+m_i]$ is at least $m_i$.
+Therefore $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$, and similarly $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.
+
+By the same argument as we use for \proc{HeavyPathLce}, the algorithm maintains the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and gives the correct LCE value.
 \end{proof}
@@
 \begin{proof}
-Each iteration is constant time, and in each iteration we halve the distance between $d_i(v)$ and $w_i$ or between $d_j(v)$ and $w_j$.
+Each iteration takes constant time, and in each iteration we halve either $\zlength(d_i[v\twodots e_i])$ or $\zlength(d_j[v\twodots e_j])$.
 \end{proof}

### 2011-07-18 doc/report.tex
m

@@
 
-Given a string $s$, the fingerprint $F_t[i]$ is a natural number identifying the substring $s[i\twodots i+t]$ among all $t$-length substrings of $s$.
-We assign fingerprints such that for any $i$, $j$ and $t$, $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t] = s[j\twodots j+t]$.
+Given a string $s$, the fingerprint $F_t[i]$ is a natural number identifying the substring $s[i\twodots i+t-1]$ among all $t$-length substrings of $s$.
+We assign fingerprints such that for any $i$, $j$ and $t$, $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t-1] = s[j\twodots j+t-1]$.
 In other words, if two substrings of $s$ have the same length, they have the same fingerprints if and only if the substrings themselves are the same.
 
-At the end of a string when $i+t>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.
+At the end of a string when $i+t-1>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.
 
@@
 
-To perform an LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
+To perform a LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
 \begin{samepage}
@@
 \item Assign an arbitrary fingerprint to the first substring.
-\item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell]$ is equal to the substring examined in the previous iteration of the loop, give the current substring the same fingerprint as the previous substring, otherwise give the current substring a new unused fingerprint.
+\item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell-1]$ is equal to the substring examined in the previous iteration of the loop, give the current substring the same fingerprint as the previous substring, otherwise give the current substring a new unused fingerprint.
 The two substrings are equal when $\LCE[i] \geq t_\ell$.
@@
 Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$.
-We can preprocess $z$ using $O(n)$ space to answer LCE queries $v=\LCE(i,j)$ in $O(\log\log N + \log n \cdot \textit{\#phrases})$ time, where \textit{\#phrases} is the number of phrases of $z$ used to store $s[i\twodots i+v]$ and $s[j\twodots j+v]$.
+We can preprocess $z$ using $O(n)$ space to answer LCE queries $v=\LCE(i,j)$ in $O(\log\log N + \log n \cdot \textit{\#phrases})$ time, where \textit{\#phrases} is the number of phrases of $z$ used to store $s[i\twodots i+v-1]$ and $s[j\twodots j+v-1]$.
 \end{theorem}
@@
 \newcommand{\zlength}{\textit{length}}
+\newcommand{\tfprint}[1][k]{\ensuremath{\proc{TreeFingerprint}_{#1}}}
+\newcommand{\tfprintk}{\tfprint[k]}
 \newcommand{\zindex}{\textit{index}}
@@
 
-To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space. In this section we analyze ways to achieve this.
+To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space.
+In this section we analyze ways to achieve this.
+We first define LCE on trees, and then describe different solutions we have found to the problem.
+\pref{tab:tree-lce-algs} lists the different solutions we have found.
+
+\begin{table}[tp]
+\centering
+\begin{tabular}{l|l|l|l}
+\hline\hline
+Algorithm & Space & Query time & Preprocessing \\ [0.5ex] \hline
+\proc{TreeDirectComp} & $O(n)$ & $O(n)$ & $O(n)$ \\ \hline
+\tfprintk, $1\leq k\leq\logceil$ & $O(kn)$ & $O(k\cdot n^{1/k})$ & $O(n\log n)$ \\ \hline
+\proc{HeavyPathLce} & $O(n)$ & $O(\log n)$ & $O(\sortt)$ \\ \hline
+\proc{TreeLcpRmq} & $O(n^2)$ & $O(1)$ & - \\ \hline
+\proc{TreeSuffixNca} & $O(n^2)$ & $O(1)$ & - \\ \hline
+\proc{LadderLce} & $O(n)$ & $O(\log n)$ & $O(\sortt)$ \\ \hline
+\end{tabular}
+\caption{Algorithms for LCE on trees with their space requirements, query times and preprocessing times.}\label{tab:tree-lce-algs}
+\end{table}
 
@@
 \paragraph{Paths}
-A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on path $p=(u,w)$:
+A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on paths:
 \begin{itemize}
-\item The path has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
-\item The node $p[a]$ is a node in $p$, such that the distance between $u$ and $p[a]$ is $a$.
+\item The path $p=(u,w)$ has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
+\item The node $p[x]$ is a node in $p=(u,w)$, such that the distance between $u$ and $p[x]$ is $x$.
 The start and end nodes of the path are $p[0]=u$ and $p[e]=w$.
-We can calculate $p[a]$ in constant time as $p[a]=\LA(d(u)+a,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
-\item The subpath $p[a\twodots b]=(p[a],p[b])$ is the path, which starts at node $p[a]$ and ends at node $p[b]$. The entire path is $p = p[0,e]$.
+We can calculate $p[x]$ in constant time as $p[x]=\LA(d(u)+x,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
+\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0,e]$.
 \item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
@@
 \paragraph{LCE on trees}
-Given two paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$ of lengths $e_i$ and $e_j$ in a tree $T$, the Longest Common Extension $v=\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $\zlabel(p_i)$ and $\zlabel(p_j)$.
+Given two paths $p_i$ and $p_j$ of lengths $e_i$ and $e_j$ in a tree $T$, the Longest Common Extension $v=\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $\zlabel(p_i)$ and $\zlabel(p_j)$.
 
@@
 \subsection{Using Fingerprinting}
-
-\newcommand{\tfprint}[1][k]{\ensuremath{\proc{TreeFingerprint}_{#1}}}
-\newcommand{\tfprintk}{\tfprint[k]}
 
@@
 
+\fixme{Write out the query in full}
+
 \begin{lemma}
@@
 
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.7\textwidth,page=1]{tree-fingerprint.pdf}
-    \end{center}
-    \caption{\label{fig:tree-query}Finding the right fingerprint to look at for a \tfprintk\ query.}
-\end{figure}
-
-The fingerprints we choose to compare are the fingerprints, which represents substrings on $p_i$ and $p_j$.
-At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$, as shown in \pref{fig:tree-query}.
+\fixme{Improve this paragraph}
+At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$.
 From our definition of fingerprints in a tree, these two labels have fingerprints $H_\ell[p_i[v+t_\ell]]$ and $H_\ell[p_j[v+t_\ell]]$, which is what our LCE query uses.
@@
 \item Stop when we reach either $v=e_i$ or $v=e_j$, and return $v$.
-\item Find $m_i = \zlength(d_i[v], \NCA_T(\zend(p_i[v+1]), w_i))$ and $i=\zindex(p_i[v+1])$.
-\item Find $m_j = \zlength(d_j[v], \NCA_T(\zend(p_j[v+1]), w_j))$ and $j=\zindex(p_j[v+1])$.
+\item Find $m_i = \zlength(p_i[v], \NCA_T(\zend(p_i[v+1]), p_i[e_i]))$ \\ and $i=\zindex(p_i[v+1])$.
+\item Find $m_j = \zlength(p_j[v], \NCA_T(\zend(p_j[v+1]), p_j[e_j]))$ \\ and $j=\zindex(p_j[v+1])$.
 \item Find $v'=\min\{\LCE_s(i,j),m_i,m_j\}$.
@@
 \item Find $m_i = \zlength(p_i[v\twodots e_i])/2$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
+\fixme{afrunding}
 \item Find $m_j = \zlength(p_j[v\twodots e_j])/2$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.

### 2011-07-18 doc/report.tex
m

@@
     \caption{\label{fig:ladder-ds} Ladders of a tree before and after their length is doubled.
-A \proc{LadderLce} string $s$ for this ladder decomposition is $s=abc|ade|f|ghijkl|hno|m$, where the splits between ladders are marked with a vertical line.}
+A \proc{LadderLce} string $s$ for this ladder decomposition is $s=abc|ade|df|ghijkl|ghno|jm$, where the splits between ladders are marked with a vertical line.}
 \end{figure}
@@
 \item An input tree $T$ with a ladder decomposition.
-\item A strings $s$, which is a concatenation of ladder labels for each ladder in $T$, where a ladder label for is $\zlabel(p)$, where $p$ is the path from the first node in the ladder to the last node in the ladder.
+\item A strings $s$, which is a concatenation of ladder labels of each ladder in $T$, where the label of a ladder is the concatenation of labels of each node in the ladder, ordered ascending by the depth of the nodes.
 \pref{fig:ladder-ds} shows an example of $s$.
@@
 \item Stop when $v = e_i$ or $v = e_j$ and return $v$.
-\item Find $m_i = \zlength(p_i[v\twodots e_i])/2$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
+\item Find $m_i = \lceil\zlength(p_i[v\twodots e_i])/2\rceil$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
 \fixme{afrunding}
-\item Find $m_j = \zlength(p_j[v\twodots e_j])/2$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
+\item Find $m_j = \lceil\zlength(p_j[v\twodots e_j])/2\rceil$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
 \item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
@@
 \begin{proof}
-In step two, $i$ refers to a position in $p_i[v+m_i]$'s ladder, because the height of $p_i[v+m_i]$ is at least $m_i$.
+In step two, $i$ refers to a position in $p_i[v+m_i]$'s ladder, because the height of $p_i[v+m_i]$ is at least $m_i-1$.
 Therefore $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$, and similarly $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.

### 2011-07-19 doc/report.tex
m

@@
 
-\title{Algorithms for Longest Common Extensions}
-
-\author{Jesper Kristensen, s062397}
+\title{{\large Master Thesis}\\{\normalsize on}\\Algorithms for Longest Common Extensions}
+
+\author{Jesper Kristensen\\s062397\\DTU Informatics\\Technical University of Denmark}
+
+\date{August 1, 2011}
 
@@
 
-\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers of $\sigma$ distinct values is written as $\sortt$.
+\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of $\sigma$ is written as $\sortt$.
 
@@
 For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path.
-If a node $u$ has height $h$, $u$'s ladder contains at least the last $h$ ancestors of $u$.
+If a node $u$ has height $h$, $u$'s ladder contains all the ancestors of $u$ whose distance to $u$ is at most $h+1$.
 The combined size of the ladders is at most $2n$.~\cite{level-ancestor}
@@
 \item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
-\item When $H_\ell[i+v] \neq H_\ell[j+v]$, decrement $\ell$ by one and repeat step 2.
-\item Stop when $\ell = 0$ and $H_\ell[i+v] \neq H_\ell[j+v]$.
+\item Stop and return $v$ when $\ell = 0$, otherwise decrement $\ell$ by one and go to step two.
 \end{enumerate}
 
-The LCE value is now $v = \LCE(i,j)$. An example is shown in \pref{fig:fingerprint-query}.
+An example of a query is shown in \pref{fig:fingerprint-query}.
 \end{samepage}
@@
 
-For larger values of $k$, we can attempt to optimize memory access accross different levels.
+For larger values of $k$, we can attempt to optimize memory access across different levels.
 When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$.
@@
 In this section we explore the LCE problem on strings compressed using LZ78 compression.
-We look at solutions, which require space proportional to the compressed size of a string, and are able to answer LCE queries fast.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{LZ78 definition\label{sec:lz-definition}}
+We look at solutions, which require space relative to the compressed size of a string, and we try to find solutions, which can answer LCE queries fast.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{LZ78 definition}
 A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$.
@@
 \begin{cases}
-    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference \neq 0\\
+    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference_z(x) \neq 0\\
     \zlabel_z(x) & \textrm{otherwise}
@@
 
+\subsubsection{Compressing and Decompressing\label{sec:lz-comp-alg}}
+
 We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions.
@@
 \begin{enumerate}
-\item Let the tree be single special root node $z_0$, let $i = 1$ and $c=z_0$.
-\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create the child node $u$, set $\zlabel(u)=s[i]$ and set $c=z_0$.
-\item Increment $i$ by one and repeat from step 2 until the entire string is consumed when $i = n$.
+\item Let the tree be a single special root node $z_0$, let $i = 1$ and $c=z_0$.
+\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create a new node $u$ as a child of $c$, set $\zlabel(u)=s[i]$, and set $c=z_0$.
+\item Increment $i$ by one and repeat from step two until the entire string is consumed when $i = N$.
 \end{enumerate}
@@
 \begin{proof}
-To find the character $s[i]$, we can first use a Predecessor query to find out which compression element $s[i]$ is in.
-We can then use a Level Ancestor query to find the correct character from the found compression element.
+To find the character $s[i]$, we can first use a Predecessor query to find out which phrase $s[i]$ is in. We can then use a Level Ancestor query to find the correct character from the found phrase.
 
@@
 The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$.
-In subsequent iterations of \proc{DirectComp} we know that we always need the same compression element or the following compression element as we used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the compression element.
+In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase.
 The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
@@
 
-A simpler way of doing LCE on compressed a string is to reverse the string and then run \proc{DirectComp} in the reverse direction on the reversed string, which gives the same result since reversing a string twice gives the original string.
-Where \proc{DirectComp} compares $s[i+v]$ against $s[j+v]$ in each iteration of its loop, reversed \proc{DirectComp} compares $s[N+1-i-v]$ against $s[N+1-j-v]$ in each iteration.
+A simpler way of doing LCE on a compressed string is to reverse the string and then run \proc{DirectComp} in the reverse direction on the reversed string, which gives the same result since reversing a string twice gives the original string.
+Where \proc{DirectComp} compares $s[i+v]$ against $s[j+v]$ in each iteration of its loop, reversed \proc{DirectComp} compares $s^R[N+1-i-v]$ against $s^R[N+1-j-v]$ in each iteration.
 This is simpler than using \proc{DirectComp} directly on the compressed string, because Level Ancestor queries are no longer needed at each iteration, since we can use the reference of the current compression element as follows, where the position of each of $i$ and $j$ is stored as a pair $(x,y)$:
@@
 
-Reversing the string can be done in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively.
-To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-definition} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.
+We can reverse the string in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively.
+To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.
 
@@
 
-We have found one case where we can make compressed \proc{DirectComp} faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time.
-We explain this in terms of the reversed compressed \proc{DirectComp}, as it is simpler, but it also works without the string reversal.
+We have found one case where we can make the reversed compressed \proc{DirectComp} faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time.
 Reversed compressed \proc{DirectComp} starts with two positions $(x_i,y_i)$ and $(x_j,y_j)$ corresponding to the two positions $i$ and $j$ in $\LCE(i,j)$.
@@
 If $x_i = x_j$ we know that we can add $d(x_i)$ to the LCE value, where $d(x_i)$ is the depth of node $x_i$ in the tree of compression elements.
-Compressed \proc{DirectComp} can then compare $d(x_i)$ characters in constant time.
+Reversed compressed \proc{DirectComp} can then compare $d(x_i)$ characters in constant time.
 
@@
 
-We can look at the other LCE algorithms we have examined in addition to \proc{DirectComp}, and see if we can apply them to LZ-compressed strings.
+We can look at the other LCE algorithms we have examined in addition to \proc{DirectComp}, to see if we can apply them to LZ-compressed strings.
 We have not found a way to apply \proc{SuffixNca} or \proc{LcpRmq}, as the string we want to find LCE values on still has N suffixes, and we have not found a way to relate the compression rate of the NCA structure or suffix array to the compression rate of the string.
@@
 Fingerprinting does not compress well either.
-We could LC-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each elvel, the compression ration becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.
+We could LZ-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.
 
@@
 We have looked into the cache performance of compressed \proc{DirectComp}.
-If the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case.
-When the string is LZ-compressable, the I/O pattern becomes much worse.
+If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case.
+When the string has a better LZ-compression ratio, the I/O pattern becomes much worse.
 If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be next to each other in memory.
@@
 We can calculate $p[x]$ in constant time as $p[x]=\LA(d(u)+x,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
-\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0,e]$.
+\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0\twodots e]$.
 \item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
@@
 
-We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\zlabel(p_i[1])$ to $\zlabel(p_j[1])$, then compare $\zlabel(p_i[2])$ to $\zlabel(p_j[2])$ and so on until the two characters differ or either $w_i$ or $w_j$ is reached.
+We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\zlabel(p_i[1])$ to $\zlabel(p_j[1])$, then compare $\zlabel(p_i[2])$ to $\zlabel(p_j[2])$ and so on until the two characters differ or either $p_i[e_i]$ or $p_j[e_j]$ is reached.
 
@@
 We can adapt our \fprintk\ algorithm on strings to work on trees, which we call \tfprintk.
-If we store the fingerprint of $s[i\twodots i+t_\ell]$ in $H_\ell[i+t_\ell]$ instead of $H_\ell[i]$, then the \fprintk\ query directly adapts to \tfprintk.
+If we store the fingerprint of $s[i\twodots i+t_\ell-1]$ in $H_\ell[i+t_\ell]$ instead of $H_\ell[i]$, then the \fprintk\ query directly adapts to \tfprintk.
 If we wanted to use the same preprocessing as we use for \fprintk\ on strings, we would have to adapt the suffix array to apply to trees, and find a way to construct it in $\sortt$ time, which we have not looked at.
@@
 
-\fixme{Write out the query in full}
+In the \tfprintk\ query, start with $v=0$ and $\ell=0$, then do the following steps:
+\begin{samepage}
+\begin{enumerate}
+\item As long as $v+t_\ell\leq e_i\lor v+t_\ell\leq e_j$ and $H_\ell[p_i[v+t_\ell]] = H_\ell[p_j[v+t_\ell]]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless $\ell = k-1$.
+\item As long as  $v+t_\ell\leq e_i\lor v+t_\ell\leq e_j$ and $H_\ell[p_i[v+t_\ell]] = H_\ell[p_j[v+t_\ell]]$, increment $v$ by $t_\ell$ and repeat this step.
+\item Stop and return $v$ when $\ell = 0$, otherwise decrement $\ell$ by one and go to step two.
+\end{enumerate}
+\end{samepage}
 
 \begin{lemma}
-The \tfprintk\ query finds the correct tree LCE value in $O(k\cdot n^{1/k})$ time.
+The \tfprintk\ query finds the correct tree LCE value.
 \end{lemma}
 \begin{proof}
+The query algorithm keeps the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$.
+At any given step at level $\ell$, we check if $H_\ell[p_i[v+t_\ell]]=H_\ell[p_j[v+t_\ell]]$, which by the definition of our data structure is equivalent to $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$.
+If the equality holds, we know that $\zlabel(p_i[0\twodots v+t_\ell])=\zlabel(p_j[0\twodots v+t_\ell])$, and we maintain our invariant when we increment $v$ by $t_\ell$.
+
+We stop when $v+1>e_i$, $v+1>e_j$, or $H_0[p_i[v+1]] \neq H_0[p_j[v+1]]$, at which point $v=\LCE(p_i,p_j)$.
+\end{proof}
+
+\begin{lemma}
+The \tfprintk\ query takes $O(k\cdot n^{1/k})$ time.
+\end{lemma}
+\begin{proof}
 We have changed nothing from the \fprintk\ query on strings, which negatively affects the query time, so the query time is the same.
-
-% The combination of Level Ancestor and the definition that a fingerprint relates to a substring from a given node to an ancestor that node, ensures that the fingerprints we choose to compare are the fingerprints, which represents the strings on the paths from $u_i$ to $w_i$ and from $u_j$ to $w_j$.
-
-\fixme{Improve this paragraph}
-At any given iteration at level $\ell$, we know that $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and we now check if $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$.
-From our definition of fingerprints in a tree, these two labels have fingerprints $H_\ell[p_i[v+t_\ell]]$ and $H_\ell[p_j[v+t_\ell]]$, which is what our LCE query uses.
 \end{proof}
@@
 \begin{enumerate}
-\item Make pairs of fingerprints $H_\ell[u] = (H_{\ell-1}[LA(d(u)-t_\ell+t_{\ell-1},u)], H_{\ell-1}[u])$ for all $u$ where $d(u) \geq t_\ell$.
+\item Make pairs of fingerprints $H_\ell[u] = (H_{\ell-1}[LA(d(u)-t_\ell+t_{\ell-1},u)], H_{\ell-1}[u])$ for all nodes $u$ in $T$, where $d(u) \geq t_\ell$.
 \item Sort the pairs lexicographically.
@@
 \item An input tree $T$ with a heavy path decomposition.
-\item A string $s$, which is a concatenation of strings of all light edges of $T$, where the string of a light edge is the label of the light edge concatenated with the label of the heavy path, which follows that light edge.
+\item A string $s$, which is a concatenation of strings of all heavy paths of $T$, where the string of a heavy path is the label of the light edge leading to the heavy path concatenated with the label of the heavy path.
 \pref{fig:tree-heavy-ds} shows an example of $s$.
@@
 Preprocessing LCE on $s$ using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time.
-If we walk through the tree and for each leaf append the light edge string ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
+If we walk through the tree and for each leaf append the heavy path string ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
 The rest of the data structures can be preprocessed in $O(n)$ time.
@@
 We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the original LCE query in $O(1)$ time, but even with this modification, a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each path.
-The SA and LCP arrays would therefore be of length $O(n^2)$.
+The SA and LCP arrays in \proc{TreeLcpRmq} would therefore be of length $O(n^2)$.
 When we look at \proc{TreeSuffixNca}, the same argument holds for the necessary number of leafs in the suffix tree.
@@
 
-\proc{LadderLce}, whic we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
-We have not found a constant time solution to find the LCE value of the second half of these paths, in the same way jump pointers is used to jump from $u$ node to the half way node node in the Level Ancestor algorithm.
+\proc{LadderLce}, which we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
+We have not found a constant time solution to find the LCE value of the second half of these paths, in the same way jump pointers is used to jump from node $u$ to the half way node in the Level Ancestor algorithm.
 We have thus not found a constant time linear space solution for LCE on trees.
@@
 \item An input tree $T$ with a ladder decomposition.
-\item A strings $s$, which is a concatenation of ladder labels of each ladder in $T$, where the label of a ladder is the concatenation of labels of each node in the ladder, ordered ascending by the depth of the nodes.
+\item A strings $s$, which is a concatenation of ladder labels of each ladder in $T$, where the label of a ladder is the concatenation of labels on each edge leading to a node in the ladder, ordered ascending by the depth of the nodes.
 \pref{fig:ladder-ds} shows an example of $s$.
@@
 \item Find $m_i = \lceil\zlength(p_i[v\twodots e_i])/2\rceil$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
-\fixme{afrunding}
 \item Find $m_j = \lceil\zlength(p_j[v\twodots e_j])/2\rceil$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
@@
 \begin{proof}
-In step two, $i$ refers to a position in $p_i[v+m_i]$'s ladder, because the height of $p_i[v+m_i]$ is at least $m_i-1$.
-Therefore $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$, and similarly $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.
-
-By the same argument as we use for \proc{HeavyPathLce}, the algorithm maintains the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and gives the correct LCE value.
+In step two, $\zlength(p_i[v\twodots v+m_i]) +1 \leq \zlength(p_i[v+m_i\twodots e_i])\leq h(p_i[v+m_i])$, where $h$ is the height of a node.
+For a node $u$ of height $h$, the ladder decomposition guarantees that $u$'s ancestor with distance $h+1$ to $u$ is in $u$'s ladder.
+Therefore, $p_i[v+1]$ is in $p_i[v+m_i]$'s ladder, and $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$.
+Similarly in step three, $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.
+
+By the same argument we use for \proc{HeavyPathLce}, the algorithm maintains the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and gives the correct LCE value.
 \end{proof}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{THE END}
+%\section*{THE END}
 
@@
 
+\begin{comment}
 \subsection*{Dropped}
@@
 
-\begin{comment}
 \section{What is the time for KMR?}

### 2011-07-19 doc/report.tex
m

@@
 
-We also look at how to apply the LCE problem to Ziv-Lempel compressed strings, such that we can answer LCE queries fast while only using space, which is a function of the compressed size of the string.
-In \pref{sec:lz-compress} we describe LZ compression, and how we can convert the simple linear time LCE algorithm to work on compressed strings with space linear to the size of the compressed string, worst case query time linear to the size of the uncompressed string, and average case query time $O(\log\log N)$.
+We also look at how to apply the LCE problem to Ziv-Lempel compressed strings, such that we can answer LCE queries fast while only using space relative to the compressed size of the string.
+In \pref{sec:lz-compress} we describe LZ compression, and how we can convert the simple linear time LCE algorithm to work on compressed strings.
+It uses space linear to the size of the compressed string, and it has worst case query time linear to the size of the uncompressed string, and average case query time $O(\log\log N)$.
 In \pref{sec:tree-lce} we look at how we can use the tree structure of LZ compression to get better query times on compressed strings.

### 2011-07-20 doc/report.tex
m

@@
+% TODOs
+% skarpere plots
+% hav plot for $r=\sqrt n$ med til forsvar
+
 \documentclass[a4]{article}
@@
 We call this algorithm \proc{DirectComp}.
-For a query $\LCE(i,j)$, the algorithm compares $s[i]$ with $s[j]$, then $s[i+1]$ with $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
+For a query $\LCE(i,j)$, the algorithm compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
 The worst case query time is $O(n)$ on a string of length $n$, and the average case query time is $O(1)$, because the average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}.
@@
 The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA.
-The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e.
+The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering.
+I.e.
 $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
-% An example is shown in \pref{fig:sa+lcp+min}.
 
@@
 \begin{corollary}
-\fprint[2] has $O(n)$ space and $O(\sqrt n)$ query time. It has two levels, where one uses a table of fingerprints and the other uses the original string.
+\fprint[2] uses $O(n)$ space and $O(\sqrt n)$ query time. It has two levels, where one uses a table of fingerprints and the other uses the original string.
 \end{corollary}
@@
 \begin{corollary}
-\fprint[\logceil] has $O(n\cdot\log n)$ space and $O(\log n)$ query time.
+\fprint[\logceil] uses $O(n\cdot\log n)$ space and $O(\log n)$ query time.
 The data structure is equivalent to the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, and a query only needs to do one comparison at each level.
@@
 
-Existing state of the art solutions are either good in worst case, while poor in average case (\proc{LcpRmq}), or good in average case while poor in worst case (\proc{DirectComp}).
+In practice, existing state of the art solutions are either good in worst case, while poor in average case (\proc{LcpRmq}), or good in average case while poor in worst case (\proc{DirectComp}).
 Our \fprintk\ solution targets a worst case vs.
@@
 \pref{fig:wc-avg} illustrates the relative query time performances of these solutions.
-Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is a lot faster than \proc{DirectComp} on a worst case input.
+Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is significantly faster than \proc{DirectComp} on a worst case input.
 Compared to \proc{LcpRmq}, our solution has a significantly better performance on an average case input, but its worst case performance is not as good as that of \proc{LcpRmq}.
@@
 The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
-The depth of a node $u$ is one greater than the depth of $u$'s parent, and the depth of the root is zero.
 A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.
-
-\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of $\sigma$ is written as $\sortt$.
+The length of a path is the number of edges between nodes in the path.
+The depth of a node $u$ is the length of the path from the root to $u$, and the height of $u$ is the length of the longest path from $u$ to a descendant of $u$.
+E.g.
+the depth of the root and the height of a leaf is both zero.
+
+\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of size $\sigma$ is written as $\sortt$.
 
@@
 There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes.
-Each edge uses $O(1)$ space, as the label is a substring of $s$ and it can therefore be represented by two indexes of $s$.
+Each edge uses $O(1)$ space, since the label is a substring of $s$ and it can therefore be represented by two indexes of $s$.
 The suffix tree therefore takes $O(n)$ space.
@@
 
-\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet $\sigma$, the Suffix Array, \SA, is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$.
+\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet size $\sigma$, the Suffix Array, \SA, is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$.
 The lexicographically smallest suffix is $\suff_{\SA[1]}$, the lexicographically largest suffix is $\suff_{\SA[n]}$, and the lexicographically $i$'th smallest suffix is $\suff_{\SA[i]}$.
@@
 \paragraph{Predecessor.}
-Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key less than or equal to $x$.
+Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key, whose key is at most $x$.
 We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.
@@
 The weight of a node $u$ is the size of the subtree rooted at $u$.
-For a given node $u$, a child of greatest weight is marked heavy, and all other children are marked light.
+For each node $u$, a child of greatest weight is marked heavy, and all other children are marked light.
 The root is always marked light.
@@
 \paragraph{Ladder Decomposition}
-To make a ladder decomposition of a tree, we first create a ladder of nodes from the root of the tree to the leaf of greatest depth.
+To make a ladder decomposition of a tree, we first create a ladder of nodes on the path from the root of the tree to the leaf of greatest depth.
 We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
@@
 In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$).
-The path label from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes.
-Since two child edges of the same node never have the same first character on their edge labels, the path label from the root to $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but longest common prefix of $\suff_i$ and $\suff_j$.
+The path label on the path from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes.
+Since two child edges of the same node never have the same first character on their edge labels, the path label on the path from the root to $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but the longest common prefix of $\suff_i$ and $\suff_j$.
 We can therefore find the LCE value as $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$, where $\mathcal{T}$ is the suffix tree for $s$, and $D[v]$ is the label depth of node $v$ in the tree.
@@
 The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA.
-The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e.
+The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering.
+I.e.
 $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
@@
 The array is split into blocks of size $O(\log n)$.
-The top level is an array of length $O(n/\log n)$, where each element contains the minimum value one block.
+The top level is an array of length $O(n/\log n)$, where each element contains the minimum value of one block.
 The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
@@
 \begin{proof}
-Step 1 takes $O(k)$ time.
-In step 2 and 3, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$).
+Step one takes $O(k)$ time.
+In step two and three, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$).
 Since we can check $t_\ell = \Theta(n^{\ell/k})$ characters in constant time at level $\ell$, the algorithm uses $O(n^{(\ell+1)/k})/\Theta(n^{\ell/k}) = O(n^{1/k})$ time at that level.
@@
 
-At each step except step 3, the algorithm increments $v$.
-Step 3 is executed the same number of times as step 1, in which $v$ is incremented.
+At each step except step three, the algorithm increments $v$.
+Step three is executed the same number of times as step one, in which $v$ is incremented.
 The query time is therefore linear to the number of times $v$ is incremented, and it is thereby $O(v)$.
@@
 
-We could have left out step 1 of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$, which would change the query pattern as shown in \pref{fig:fingerprint-updown}.
-This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, but it would increase our average case query time to $O(k)$.
-Our experiments have shown that whenever $k$ is small, keeping step 1 improves average case query time, while it does not have a measurable effect on worst case query times.
-When $k = \logceil$, keeping step 1 can double the worst case query time, while it can make the average case query time 40 times faster for an input string of ten million characters.
+We could have left out step one of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$, which would change the query pattern as shown in \pref{fig:fingerprint-updown}.
+This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
+\ifarticle
+Our experiments have shown that whenever $k$ is small, keeping step one improves average case query time, while it does not have a measurable effect on worst case query times.
+When $k = \logceil$, keeping step one can double the worst case query time, while it can make the average case query time 40 times faster for an input string of ten million characters.
 We want to optimize our LCE query time for the average case where the LCE value is small, so our results in \pref{sec:results} does not include the variant of \fprintk\ with $O(k)$ average case query time.
-
-In step 1 of our query algorithm we perform one comparison at each level.
-We could instead do up to $O(n^{1/k})$ comparisons at each level without affecting asymptotic times.
+\fi % article
+
+\begin{figure}[tp]
+    \begin{center}
+        \includegraphics[width=1\textwidth,page=1]{fingerprint-updown.pdf}
+    \end{center}
+    \caption{\label{fig:fingerprint-updown}Two ways of querying the fingerprint levels, one starting from level $\ell=k-1$ and one starting at level $\ell=0$.}
+\end{figure}
+
+In step one of our query algorithm we perform one comparison at each level. We could instead do up to $O(n^{1/k})$ comparisons at each level without affecting asymptotic times.
+\ifarticle
 Our experiments have shown that always doing one comparison at each level is the best in practice.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=1\textwidth,page=1]{fingerprint-updown.pdf}
-    \end{center}
-    \caption{\label{fig:fingerprint-updown}Two ways of querying the fingerprint levels, one starting from level $\ell=k-1$ and one starting at level $\ell=0$.}
-\end{figure}
+\fi % article
 
@@
 However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
-Instead of storing the fingerprint of $s[i\twodots i+t_\ell]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$.
+Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$.
 This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.
@@
 \item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater wc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step 1 of the query in \pref{sec:fingerprint-query}.
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step one of the query in \pref{sec:fingerprint-query}.
 \item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
@@
 \item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache-horizontal-}\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache}-\textit{dir}-\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{dir} describes which of our two cache optimized variants is used, and \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
 \end{description}
@@
 \item For each tested algorithm, preprocess the given string, and run a query for each of the million pairs. Measure the time it takes to run all the queries combined.
-\item Double the value of $n$ and repeat from step 1.
+\item Double the value of $n$ and repeat from step one.
 \end{enumerate}
@@
 None of the two plots we measure represents a good worst case for \proc{DirectComp+LcpRmq} compared to the other algorithms we test.
-The worst case for \proc{DirectComp+LcpRmq} relative to other the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithms still work on LCE values as small as possible.
+The worst case for \proc{DirectComp+LcpRmq} relative to the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithms still work on LCE values as small as possible.
 As described in \pref{sec:dc-lcprmq}, \proc{DirectComp+LcpRmq} also uses a lot of space.
@@
 
-In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] degrade a small amount when using the average case optimization.
-Both variations of \fprint[\logceil] show query times, which are $O(\log n)$ as expected.
-On average case strings, the query time of \fprint[\logceil] improves significantly when using average case optimization, and both variants show their expected $O(1)$ and $O(\log n)$ average case query times.
+In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] doubles when we use the average case optimization.
+On average case strings, the query time of \fprint[\logceil] is up to 40 times faster when using average case optimization.
+All four cases in \pref{fig:plot-fingerprint-avglog} show their expected $O(1)$ and $O(\log n)$ query times.
+In \pref{fig:plot-fingerprint-avg3l}, we see the same results for \fprint[3], as we saw for \fprint[\logceil], but at a smaller scale.
+The negative effect of doing average case optimization is barely measurable on worst case strings, while the improvement on average case strings is up to 40\%.
 
@@
 
-In \pref{fig:plot-fingerprint-avg3l} we see that doing one comparison at each level on the way up is better than doing $n^{1/k}$ comparisons.
+In \pref{fig:plot-fingerprint-avg3l} we see that doing one comparison at each level on the way up in \fprint[3] is better than doing $n^{1/k}$ comparisons.
 This is most visible on worst case strings, where the algorithm doing one comparison at each level uses less time to reach level $\ell=k-1$, which it will reach anyway.
 On average case strings, the code specialized in handling only one comparison at each level is simpler.
-On the question of doing average case optimization or not, we see the same results for \fprint[3], as we saw for \fprint[\logceil], but at a smaller scale.
 
@@
     \end{center}
-    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without cache optimization.}
+    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without intra-level cache optimization.}
 \end{figure}
@@
     \end{center}
-    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without cache optimization.}
+    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without inter-level cache optimization.}
 \end{figure}
@@
 The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified.
-We have made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented.
+We have added a working build system, made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented.
 \pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.
@@
 algorithm & \fprintk & algorithm \\
-alg. parameter & $t = 2\sqrt n$ & algorithm \\ \hline
+alg. parameters & $t_1 = 2\sqrt n$ & algorithm \\ \hline
 string length & exponentially increasing from $1$ to $100.000$ & test \\
@@
 \item We make no assumptions on the order of $i$ and $j$.
-In some applications of the LCE problem, for example approximate string searching, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times.
+In some applications of the LCE problem, for example approximate string searching~\cite{approx-search}, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times.
 We do not measure this case.

### 2011-07-20 doc/report.tex
m

@@
 The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$.
-In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase.
+In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase.
 The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
@@
 If we store $t$ positions equally spaced across the string $s$, we would get $O(n+t)$ space and $O(N/t+|\LCE(i,j)|)$ query time.
-To keep the compressed space, we can set $t=n$, which gives $O(N/n+|\LCE(i,j)|) = O(\sqrt{N}+|\LCE(i,j)|)$ query time.
+To keep the space within $O(n)$, we can set $t=n$, which gives $O(N/n+|\LCE(i,j)|) = O(\sqrt{N}+|\LCE(i,j)|)$ query time.
 
@@
 We can reverse the string in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively.
-To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n)$ space for its tree of compression elements.
+To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n_2)$ space for its tree of compression elements.
 
@@
 Fingerprinting does not compress well either.
-We could LZ-compress each level of fingerprints just like the original string is compressed, but as the substrings fingerprinted becomes longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.
+We could LZ-compress each level of fingerprints just like the original string is compressed, but as the fingerprinted substrings become longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.
 
@@
 The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree.
-The LCE value is the length of a substring, which occurs at two positions $i$ and $j$ in the input string.
-This substring spans a number of phrases, where each phrase is a path in the tree of compression elements.
-If we can make fast LCE queries on such a tree, we can make fast LCE queries on highly compressable strings.
+The LCE value is the length of two equivalent substrings, which occur at positions $i$ and $j$ in the input string.
+These substrings span a number of phrases.
+If we can make fast LCE queries on a tree of compression elements, we can make fast LCE queries on highly compressable strings.
 
@@
 We have looked into the cache performance of compressed \proc{DirectComp}.
-If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's in worst case.
+If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's.
 When the string has a better LZ-compression ratio, the I/O pattern becomes much worse.
-If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be next to each other in memory.
+If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be in the same block in memory.
 This can happen many times, which gives $O(N)$ I/O's in worst case.

### 2011-07-20 doc/report.tex
m

@@
 % skarpere plots
+% sørg for at figurer ikke sidder underligt
+% Prøv at formindske for store mellemrum mellem ord ved at lave påtvungne linjeskift
 % hav plot for $r=\sqrt n$ med til forsvar
@@
 The algorithm keeps the invariant $\zlabel(p_i[0\twodots v]) = \zlabel(p_j[0\twodots v])$.
-At each iteration we find $v'$, which is the longest common prefix of $\zlabel(p_i[v\twodots v+m_i])$ and $\zlabel(p_j[v\twodots v+m_j])$.
+At each iteration we find $v'$, which is the length of the longest common prefix of $\zlabel(p_i[v\twodots v+m_i])$ and $\zlabel(p_j[v\twodots v+m_j])$.
 This gives $\zlabel(p_i[0\twodots v+v']) = \zlabel(p_j[0\twodots v+v'])$, and we therefore keep the invariant when we add $v'$ to $v$ in step five.
@@
     \end{center}
-    \caption{\label{fig:ladder-ds} Ladders of a tree before and after their length is doubled.
+    \caption{\label{fig:ladder-ds} Ladders of a tree before and after their size is doubled.
 A \proc{LadderLce} string $s$ for this ladder decomposition is $s=abc|ade|df|ghijkl|ghno|jm$, where the splits between ladders are marked with a vertical line.}
@@
 \begin{proof}
-In step two, $\zlength(p_i[v\twodots v+m_i]) +1 \leq \zlength(p_i[v+m_i\twodots e_i])\leq h(p_i[v+m_i])$, where $h$ is the height of a node.
+In step two, $\zlength(p_i[v\twodots v+m_i]) +1 \leq \zlength(p_i[v+m_i\twodots e_i])\leq h(p_i[v+m_i])$, where $h(u)$ is the height of node $u$.
 For a node $u$ of height $h$, the ladder decomposition guarantees that $u$'s ancestor with distance $h+1$ to $u$ is in $u$'s ladder.
 Therefore, $p_i[v+1]$ is in $p_i[v+m_i]$'s ladder, and $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$.
-Similarly in step three, $s[j\twodots j+m_i-1] = \zlabel(p_j[v\twodots v+m_j])$.
+Similarly in step three, $s[j\twodots j+m_i-1]\\ = \zlabel(p_j[v\twodots v+m_j])$.
 

### 2011-07-20 doc/report.tex
m

@@
 Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space.
-But like for the LCE problem, the asymptotically best RMQ solution may not be the best RMQ solution in practice.
+But like for the LCE problem, the asymp-\\totically best RMQ solution may not be the best RMQ solution in practice.
 We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
@@
 We then preprocess each of the $k$ levels in $O(n)$ time, since we loop through $n$ substrings, and comparing neighboring substrings takes constant time when we use the LCE array.
-The total preprocessing time becomes $O(\sortt + k\cdot n)$.
+The total preprocessing time becomes $O(\sortt\\ + k\cdot n)$.
 \end{proof}
@@
 
-The amount if I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$.
+The amount of I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$.
 However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
@@
 At each iteration we find $v'$, which is the length of the longest common prefix of $\zlabel(p_i[v\twodots v+m_i])$ and $\zlabel(p_j[v\twodots v+m_j])$.
-This gives $\zlabel(p_i[0\twodots v+v']) = \zlabel(p_j[0\twodots v+v'])$, and we therefore keep the invariant when we add $v'$ to $v$ in step five.
+This gives $\zlabel(p_i[0\twodots v+v']) = \\\zlabel(p_j[0\twodots v+v'])$, and we therefore keep the invariant when we add $v'$ to $v$ in step five.
 
@@
 \begin{proof}
-Constructing $s$ from the ladders takes $O(n)$ time, and preprocessing it for constant time LCE takes $O(\sortt)$ time. The rest of the data structures have $O(n)$ preprocessing times.
+Constructing $s$ from the ladders takes $O(n)$ time, and preprocessing it for constant time LCE takes $O(\sortt)$ time. The rest of the data structures have\\ $O(n)$ preprocessing times.
 \end{proof}

### 2011-07-20 doc/report.tex
m

@@
 % Prøv at formindske for store mellemrum mellem ord ved at lave påtvungne linjeskift
+
 % hav plot for $r=\sqrt n$ med til forsvar
@@
 
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.5\textwidth,page=1]{resultspace.pdf}
+\begin{figure}[p]
+    \begin{center}
+        \includegraphics[width=0.42\textwidth,page=1]{resultspace.pdf}
     \end{center}
@@
 
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.5\textwidth,page=2]{resultspace.pdf}
+\begin{figure}[p]
+    \begin{center}
+        \includegraphics[width=0.42\textwidth,page=2]{resultspace.pdf}
     \end{center}
     \caption{\label{fig:resultspace-directcomp}$LCE(3,7)$ using \proc{DirectComp}.}
+\end{figure}
+
+\begin{figure}[p]
+    \begin{center}
+        \includegraphics[width=0.42\textwidth,page=3]{resultspace.pdf}
+    \end{center}
+    \caption{\label{fig:resultspace-t-row}$LCE(3,7)$ with distance $t=4$ between each stored row.}
 \end{figure}
@@
 We call this algorithm \proc{DirectCompLookup}.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.5\textwidth,page=3]{resultspace.pdf}
-    \end{center}
-    \caption{\label{fig:resultspace-t-row}$LCE(3,7)$ with distance $t=4$ between each stored row.}
-\end{figure}
 
@@
 
+\begin{samepage}
 \begin{lemma}
@@
 \end{lemma}
-
 \begin{proof}
@@
 \end{proof}
+\end{samepage}
 
@@
 
-\begin{figure}[tp]
+\begin{figure}[p]
     \begin{center}
@@
 \subsection{\proc{DirectCompLookup} Results}
+
+\begin{figure}[p]
+    \begin{center}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-alla.plt}
+    \end{center}
+    \caption{\label{fig:plot-complookup}Query times of our new DirectCompLookup algorithm versus the existing \proc{DirectComp} and \proc{LcpRmq} algorithms.}
+\end{figure}
 
@@
 The only way we have found to improve this preprocessing time is to use one of the other algorithms we have covered to preprocess the data structure of \proc{DirectCompLookup}.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-complookup}Query times of our new DirectCompLookup algorithm versus the existing \proc{DirectComp} and \proc{LcpRmq} algorithms.}
-\end{figure}
 
@@
     \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without intra-level cache optimization.}
+\end{figure}
+
+\begin{figure}[p]
+    \begin{center}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-alla.plt}
+    \end{center}
+    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without inter-level cache optimization.}
 \end{figure}
@@
 However, due to late timing of this discovery, the cache optimized variant of \fprintk\ is not the main focus of our analysis.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without inter-level cache optimization.}
-\end{figure}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 \section{Ziv-Lempel Compression\label{sec:lz-compress}}
@@
 
+\begin{samepage}
 \subsubsection{Data Structure}
@@
 \end{itemize}
+\end{samepage}
 
@@
 
-We have looked at applying \proc{SuffixNca} and \proc{LcpRmq} directly to trees like we did with \proc{DirectComp} and \fprintk, such that we can get $O(1)$ query time.
+We have looked at applying \proc{SuffixNca} and \proc{LcpRmq} directly to trees like we did with \proc{DirectComp} and \fprintk, to get $O(1)$ query time.
 

### 2011-08-13 presentation/presentation.tex
m

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\setbeamertemplate{navigation symbols}{}
\usetheme{Boadilla}
\title{Algorithms for Longest Common Extensions}
\author{Jesper Kristensen}
\institute[DTU Informatics]{DTU Informatics\\Technical University of Denmark}
\date{August 31, 2011}
\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}{Contents}
\tableofcontents
\end{frame}

\section{Introduction}
\begin{frame}{The LCE Problem the DirectComp algorithm}
    \begin{block}{Input}
        \begin{itemize}
            \item $s=ababba$
            \item $(i,j)=(2,3)$
        \end{itemize}
    \end{block}
    \begin{block}{The DirectComp algorithm}
        TODO
    \end{block}
    \begin{block}{Result}
        $LCE_s(2,3)=3$
    \end{block}
\end{frame}

\end{document}

### 2011-08-13 presentation/compile.sh

@@ -1,0 +1,1 @@
+pdflatex -shell-escape presentation.tex

### 2011-08-16 presentation/presentation.tex
m

@@
 \usepackage[utf8]{inputenc}
+\usepackage{clrscode}
+\usepackage{multicol}
 \setbeamertemplate{navigation symbols}{}
@@
 \begin{document}
+
+\newcommand{\sortt}{\textit{sort}(n,\sigma)}
+\newcommand{\LCE}{\textit{LCE}}
+\newcommand{\NCA}{\textit{NCA}}
+\newcommand{\RMQ}{\textit{RMQ}}
+\newcommand{\SA}{\textit{SA}}
+\newcommand{\SAinv}{\textit{SA}^{-1}} % math
+\newcommand{\SAi}{SA$^{-1}$} % no math
+\newcommand{\LCP}{\textit{LCP}}
+\newcommand{\LA}{\textit{LA}}
+\newcommand{\suff}{\textit{suff}}
+\newcommand{\logceil}{\lceil\log n\rceil}
+\newcommand{\fprint}[1][k]{\ensuremath{\proc{Fingerprint}_{#1}}}
+\newcommand{\fprintk}{\fprint[k]}
+\newcommand{\RMQpq}[2]{RMQ\textless$#1$, $#2$\textgreater}
+\newcommand{\RMQn}{\RMQpq{1}{n}}
+\newcommand{\RMQq}{\RMQpq{n}{1}}
+\newcommand{\RMQlog}{\RMQpq{n}{\log n}}
 
@@
 \section{Introduction}
-\begin{frame}{The LCE Problem the DirectComp algorithm}
+\subsection{The LCE Problem}
+\begin{frame}{The LCE Problem and the \proc{DirectComp} algorithm}
     \begin{block}{Input}
         \begin{itemize}
-            \item $s=ababba$
-            \item $(i,j)=(2,3)$
+            \item $s=abbababba$
+            \item $(i,j)=(4,6)$
         \end{itemize}
     \end{block}
-    \begin{block}{The DirectComp algorithm}
-        TODO
+    \begin{block}{The \proc{DirectComp} algorithm}
+        %\setlength{\unitlength}{1cm}
+        \definecolor{matchcolor}{rgb}{.6,1,.6}
+        \definecolor{mismatchcolor}{rgb}{1,.6,.6}
+        \begin{picture}(300,50)
+            \put(0, 30){\makebox(20, 20){$s=$}}
+            \put(20, 30){\framebox(20, 20){a}}
+            \put(40, 30){\framebox(20, 20){b}}
+            \put(60, 30){\framebox(20, 20){b}}
+            \put(80, 30){\framebox(20, 20){a}}
+            \put(100, 30){\framebox(20, 20){b}}
+            \put(120, 30){\framebox(20, 20){a}}
+            \put(140, 30){\framebox(20, 20){b}}
+            \put(160, 30){\framebox(20, 20){b}}
+            \put(180, 30){\framebox(20, 20){a}}
+            \uncover<2-2>{
+                \put(200, 30){\makebox(100, 20){1 match}}
+                \put(80,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
+                \put(90, 15){\vector(0,1){13}}
+                \put(80, 0){\makebox(20, 15){4}}
+                \put(120,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
+                \put(130, 15){\vector(0,1){13}}
+                \put(120, 0){\makebox(20, 15){6}}
+            }
+            \uncover<3->{
+                \put(200, 30){\makebox(100, 20){2 matches}}
+            }
+            \uncover<3-3>{
+                \put(100,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
+                \put(110, 15){\vector(0,1){13}}
+                \put(100, 0){\makebox(20, 15){5}}
+                \put(140,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
+                \put(150, 15){\vector(0,1){13}}
+                \put(140, 0){\makebox(20, 15){7}}
+            }
+            \uncover<4-4>{
+                \put(120,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){a}}}}
+                \put(130, 15){\vector(0,1){13}}
+                \put(120, 0){\makebox(20, 15){6}}
+                \put(160,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){b}}}}
+                \put(170, 15){\vector(0,1){13}}
+                \put(160, 0){\makebox(20, 15){8}}
+            }
+        \end{picture}
     \end{block}
+    \uncover<5->{
         \begin{block}{Result}
-        $LCE_s(2,3)=3$
+            $LCE_s(4,6)=2$
         \end{block}
+    }
+\end{frame}
+
+\subsection{The \proc{DirectComp} algorithm}
+\begin{frame}{Existing Algorithm: \proc{DirectComp}}
+    \begin{tabular}{r l}
+        Preprocessing & $O(1)$ \\
+        Space & $O(1)$ \\
+        Query & $O(|\LCE(i,j)|)=O(n)$ \\
+        Average query & $O(1)$ \\
+    \end{tabular}
+
+    \vspace{1cm}
+    For a string length $n$ and alphabet size $\sigma$, the average LCE value over all $n^\sigma$ strings and $n^2$ query pairs is $O(1)$.
+\end{frame}
+
+\section{Existing Results}
+\subsection{The \proc{SuffixNca} Algorithm}
+\begin{frame}{Existing Algorithm: \proc{SuffixNca}}
+    $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$
+\end{frame}
+
+\subsection{The \proc{LcpRmq} Algorithm}
+\begin{frame}{Existing Algorithm: \proc{LcpRmq}}
+    \begin{multicols}{2}{
+        \begin{align*}
+            s&=abbababba\\
+            s[2\twodots n]&=bbababba\\
+            s[3\twodots n]&=bababba\\
+            \LCE_s(2,3)&=1
+        \end{align*}
+        \newpage
+        \begin{center}
+            \uncover<2->{
+                \includegraphics[width=0.26\textwidth,page=1]{../doc/sa+lcp+min.pdf}
+            }
+        \end{center}
+    }
+    \end{multicols}
+    \uncover<3->{
+        \begin{center}
+            $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$\\
+            where $\SAinv[i] < \SAinv[j]$\\
+            \vspace{1em}
+            \begin{tabular}{r l}
+                Preprocessing & $O(\sortt)$ \\
+                Space & $O(n)$ \\
+                Query & $O(1)$ \\
+                Average query & $O(1)$ \\
+            \end{tabular}
+        \end{center}
+    }
+\end{frame}
+
+\section{The \fprintk\ Algorithm}
+\subsection{Data Structure}
+\begin{frame}{The \fprintk\ Algorithm: Data Structure}
+    \begin{itemize}
+        \item For a string $s[1\twodots n]$, the $t$-length fingerprints $F_t[1\twodots n]$ are natural numbers, such that $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t+1] = s[j\twodots j+t+1]$.
+        \item For each of $k$ levels, $\ell = 0\twodots k-1$:
+        \begin{itemize}
+            \item $t_\ell = \Theta(n^{\ell/k}), t_0=1$
+            \item $H_\ell = F_{t_\ell}$
+        \end{itemize}
+    \end{itemize}
+    \begin{center}
+        \includegraphics[width=0.98\textwidth,page=1]{../doc/fingerprint.pdf}
+    \end{center}
+    \begin{tabular}{r l}
+        Space & $O(k\cdot n)$\\
+    \end{tabular}
+\end{frame}
+
+\subsection{Query}
+\begin{frame}{The \fprintk Algorithm: Query}
+    \begin{enumerate}
+        \item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
+        \item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
+        \item Stop and return $v$ when $\ell = 0$, otherwise decrement $\ell$ by one and go to step two.
+    \end{enumerate}
+    \begin{center}
+        \includegraphics[width=0.8\textwidth,page=2]{../doc/fingerprint.pdf}\\
+         $\LCE(3,12)$
+    \end{center}
+    \begin{tabular}{r l}
+        Query & $O(k\cdot n^{1/k})$ \\
+        Average query & $O(1)$ \\
+    \end{tabular}
 \end{frame}

### 2011-08-16 presentation/presentation.tex
m

@@
 \usepackage[utf8]{inputenc}
+\usepackage{verbatim}
 \usepackage{clrscode}
@@
 \setbeamertemplate{navigation symbols}{}
-\usetheme{Boadilla}
+%\usetheme{Boadilla}
+\setbeamertemplate{footline}[frame number]
 \title{Algorithms for Longest Common Extensions}
@@
 \section{Existing Results}
-\subsection{The \proc{SuffixNca} Algorithm}
-\begin{frame}{Existing Algorithm: \proc{SuffixNca}}
-    $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$
-\end{frame}
+%\subsection{The \proc{SuffixNca} Algorithm}
+%\begin{frame}{Existing Algorithm: \proc{SuffixNca}}
+%    $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$
+%\end{frame}
 
@@
         \begin{center}
-            \uncover<2->{
+            %\uncover<2->{
             \includegraphics[width=0.26\textwidth,page=1]{../doc/sa+lcp+min.pdf}
-            }
+            %}
         \end{center}
@@
     \end{multicols}
-    \uncover<3->{
+    %\uncover<3->{
     \begin{center}
@@
     \end{center}
+    %}
+\end{frame}
+
+\subsection{Practical results}
+\begin{frame}{Existing Algorithms: Practical Results}
+    Query times of
+    \textcolor{red}{\proc{DirectComp}} and
+    \textcolor{green}{\proc{LcpRmq}}
+    by string length
+    \begin{multicols}{2}{
+        \begin{center}
+            Average case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-slide-rand10.plt}
+            \newpage
+            Worst case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-slide-alla.plt}
+        \end{center}
     }
+    \end{multicols}
+
 \end{frame}
@@
     \begin{itemize}
-        \item For a string $s[1\twodots n]$, the $t$-length fingerprints $F_t[1\twodots n]$ are natural numbers, such that $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t+1] = s[j\twodots j+t+1]$.
-        \item For each of $k$ levels, $\ell = 0\twodots k-1$:
+        \item For a string $s[1\twodots n]$, the $t$-length fingerprints $F_t[1\twodots n]$ are natural numbers, such that $F_t[i] = F_t[j]$ if and only if $s[i\twodots i+t-1] = s[j\twodots j+t-1]$.
+        \item $k$ levels, $1\leq k\leq\logceil$
+        \item For each level, $\ell = 0\twodots k-1$:
         \begin{itemize}
@@
 \subsection{Query}
-\begin{frame}{The \fprintk Algorithm: Query}
+\begin{frame}{The \fprintk\ Algorithm: Query}
     \begin{enumerate}
@@
 
+\subsection{Preprocessing}
+\begin{frame}{The \fprintk\ Algorithm: Preprocessing}
+    \begin{itemize}
+        \item For each level $\ell$
+        \begin{itemize}
+            \item For each $t_\ell$-length substring in lexicographically sorted order
+            \begin{itemize}
+                \item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell-1]$ is equal to the previous substring, give it the same fingerprint as the previous substring, otherwise give it a new unused fingerprint.
+The two substrings are equal when $\LCE[i] \geq t_\ell$.
+            \end{itemize}
+        \end{itemize}
+    \end{itemize}
+
+    \begin{multicols}{2}{
+        \begin{center}
+        \hfill\\
+            $s=$ abbababba\\\hfill\\
+            For $t_\ell = 3$:\\
+            $H_\ell =$ [3, 8, 6, 2, 6, 3, 8, 5, 1]
+        \end{center}
+        \newpage
+        \begin{center}
+            \includegraphics[width=0.2\textwidth,page=1]{../doc/fingerprint-preproc.pdf}
+        \end{center}
+    }
+    \end{multicols}
+
+    \begin{tabular}{r l}
+        Preprocessing & $O(k\cdot n + \sortt)$ \\
+    \end{tabular}
+
+\end{frame}
+
+\begin{frame}{The \fprintk\ Algorithm}
+    \begin{tabular}{r l}
+        Preprocessing & $O(k\cdot n + \sortt)$ \\
+        Space & $O(k\cdot n)$\\
+        Query & $O(k\cdot n^{1/k})$ \\
+        Average query & $O(1)$ \\
+    \end{tabular}
+
+    \hfill\\$k=1$:
+
+    ~~Same as \proc{DirectComp}
+
+%    \hfill\\$k=2$:
+
+%    \begin{tabular}{r l}
+%        Preprocessing & $O(\sortt)$ \\
+%        Space & $O(n)$\\
+%        Query & $O(\sqrt n)$ \\
+%        Average query & $O(1)$ \\
+%    \end{tabular}
+
+    \hfill\\$k=\logceil$:
+
+    \begin{tabular}{r l}
+        Preprocessing & $O(n\log n)$ \\
+        Space & $O(n\log n)$\\
+        Query & $O(\log n)$ \\
+        Average query & $O(1)$ \\
+    \end{tabular}
+\end{frame}
+
+\subsection{Practical Results}
+\begin{frame}{The \fprintk\ Algorithm: Practical Results}
+    Query times of
+    \textcolor{red}{\proc{DirectComp}},
+    \textcolor{green}{\fprint[2]},
+    \textcolor{blue}{\fprint[3]},
+    \textcolor{magenta}{\fprint[\logceil]} and
+    \textcolor{cyan}{\proc{LcpRmq}}
+    by string length
+    \begin{multicols}{2}{
+        \begin{center}
+            Average case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-rand10.plt}
+            \newpage
+            Worst case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-alla.plt}
+        \end{center}
+    }
+    \end{multicols}
+\end {frame}
+
+\begin{frame}{The \fprintk\ Algorithm: Practical Results}
+    Query times of
+    \textcolor{red}{\proc{DirectComp}},
+    \textcolor{green}{\fprint[2]},
+    \textcolor{blue}{\fprint[3]},
+    \textcolor{magenta}{\fprint[\logceil]} and
+    \textcolor{cyan}{\proc{LcpRmq}}
+    by string length
+    \begin{center}
+        Medium LCE size\\
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-repeat-pow.plt}
+    \end{center}
+\end {frame}
+
 \end{document}

### 2011-08-16 src/plot2.sh

@@ -99,6 +99,23 @@ datafile="results/length-repeat-sqrt.txt"
 plotfile="results/length-article-repeat-sqrt.plt"
 oneplot "DirectComp," "Fingerprint_2<sqrt n>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
 
+#Overall Fingerprint Slides
+slides="\nunset title\nunset xlabel\nunset ylabel\nunset key"
+xtra="$yavg$slides"
+datafile="results/length-rand10.txt"
+plotfile="results/length-slides-rand10.plt"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
+
+xtra="$ywc$slides"
+datafile="results/length-alla.txt"
+plotfile="results/length-slides-alla.plt"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
+
+xtra="$ymc$slides"
+datafile="results/length-repeat-pow.txt"
+plotfile="results/length-slides-repeat-pow.plt"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
+
 # Not used
 #xtra=$yavg
 #datafile="results/length-rand10.txt"
@@ -198,7 +215,7 @@ datafile="results/length-alla.txt"
 plotfile="results/length-rmq-alla.plt"
 oneplot "RMQ<1;n>_{virtual}," "RMQ<n;1>_{virtual}," "RMQ<n;2log n>_{virtual}," "RMQ<n;4log n>_{virtual}," #"RMQ<succinct;?>_{virtual}," "RMQ<optimal>_{virtual},"
 
-# RMQ vs. DirectComp
+# RMQ vs. DirectComp vs. combination
 xtra=$yavg
 datafile="results/length-rand10.txt"
 plotfile="results/length-rmq-directcomp-rand10.plt"
@@ -208,6 +225,17 @@ xtra=$ywc
 datafile="results/length-alla.txt"
[…]

### 2011-08-17 article/article.tex
m

--- 2011-07-20 doc/report.tex
@@
 
-\title{{\large Master Thesis}\\{\normalsize on}\\Algorithms for Longest Common Extensions}
-
-\author{Jesper Kristensen\\s062397\\DTU Informatics\\Technical University of Denmark}
+\title{Algorithms for Longest Common Extensions}
+
+\author{Jesper Kristensen\\DTU Informatics\\Technical University of Denmark}
 
@@
 
-\reporttrue
+\articletrue
 
@@
 
-\vspace{1cm}
+%\vspace{1cm}
 % stuff on toc page
 
-\newpage
+%\newpage
 
@@
 
-We have found two new algorithms.
-They both have better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms.
-The best of these two uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.
+We look at a new algorithm, which has better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms.
+This algorithm uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.
+
+\ifreport
 
@@
 
+\fi % report
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 The longest common prefix of the two suffixes will therefore be the path label from the root to the nearest common ancestor of $L_i$ and $L_j$.
-This gives $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.
+I.e.
+$\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Existing Practical Results}
+% \subsubsection{Existing Practical Results}
 
@@
 
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.5\textwidth,page=1]{wc-avg.pdf}
-    \end{center}
-    \caption{\label{fig:wc-avg}Graphical comparison of query times for different LCE algorithms.}
-\end{figure}
+\pref{tab:article-algorithms} shows an overview of asymptotic bounds of the different LCE algorithms.
+
+\begin{table}[tp]
+\centering
+% \begin{tabular}{l|p{160px}|p{160px}}
+\begin{tabular}{l|l|l|l}
+\hline\hline
+Algorithm & Space & Query time & Preprocessing \\ [0.5ex] \hline
+\proc{SuffixNca} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
+\proc{LcpRmq} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
+\proc{DirectComp} & $O(1)$ & $O(n)$ & $O(1)$ \\ \hline
+\fprintk & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
+~~~~$k=\logceil$ & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline
+\end{tabular}
+\caption{LCE algorithms with their space requirements, worst case query times and preprocessing times. Average case query times are $O(1)$ for all shown algorithms.}\label{tab:article-algorithms}
+\end{table}
+
+%\begin{figure}[tp]
+%    \begin{center}
+%        \includegraphics[width=0.5\textwidth,page=1]{wc-avg.pdf}
+%    \end{center}
+%    \caption{\label{fig:wc-avg}Graphical comparison of query times for different LCE algorithms.}
+%\end{figure}
 
@@
 average case query time tradeoff between these two extremes.
-\pref{fig:wc-avg} illustrates the relative query time performances of these solutions.
 Our solution is almost as fast as \proc{DirectComp} on an average case input, and it is significantly faster than \proc{DirectComp} on a worst case input.
@@
 
+\ifreport
 \paragraph{Tree notation.}
@@
 the depth of the root and the height of a leaf is both zero.
+\fi % report
 
@@
 
-\paragraph{NCA.} The common ancestors of two nodes $u$ and $v$ is the intersection between the ancestors of $u$ and the ancestors of $v$.
-The Nearest Common Ancestor of two nodes $u$ and $v$ is the common ancestor of $u$ and $v$ of greatest depth.
+\paragraph{NCA.} The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
+The Nearest Common Ancestor of two nodes $u$ and $v$ is the node of greatest depth, which is an ancestor of both $u$ and $v$.
 An NCA query can be answered in $O(1)$ time with $O(n)$ space and preprocessing time in a static tree with $n$ nodes~\cite{nca}.
@@
     \begin{center}
-        \includegraphics[width=0.98\textwidth,page=1]{fingerprint.pdf}
+        \includegraphics[width=0.98\textwidth,page=1]{../doc/fingerprint.pdf}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=0.98\textwidth,page=2]{fingerprint.pdf}
+        \includegraphics[width=0.98\textwidth,page=2]{../doc/fingerprint.pdf}
     \end{center}
@@
 
-We could have left out step one of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$, which would change the query pattern as shown in \pref{fig:fingerprint-updown}.
+We could have left out step one of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$.
 This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
@@
 
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=1\textwidth,page=1]{fingerprint-updown.pdf}
-    \end{center}
-    \caption{\label{fig:fingerprint-updown}Two ways of querying the fingerprint levels, one starting from level $\ell=k-1$ and one starting at level $\ell=0$.}
-\end{figure}
+%\begin{figure}[tp]
+%    \begin{center}
+%        \includegraphics[width=1\textwidth,page=1]{fingerprint-updown.pdf}
+%    \end{center}
+%    \caption{\label{fig:fingerprint-updown}Two ways of querying the fingerprint levels, one starting from level $\ell=k-1$ and one starting at level $\ell=0$.}
+%\end{figure}
 
@@
     \begin{center}
-        \includegraphics[width=0.2\textwidth,page=1]{fingerprint-preproc.pdf}
+        \includegraphics[width=0.2\textwidth,page=1]{../doc/fingerprint-preproc.pdf}
     \end{center}
@@
 
-\ifreport
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\paragraph{Horizontal, intra-level optimization}
+%\paragraph{Horizontal, intra-level optimization}
 
@@
 When $k=\logceil$, this cache optimization gives no difference in the amount of I/O's.
+
+\ifreport
 
@@
 \item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$.
+\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is the \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$.
 The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater describe the exact size of fingerprinted substrings at each level.
@@
 
-\begin{figure}[p]
+\begin{figure}[tp]
     \begin{center}
@@
 
+\ifreport
 \begin{figure}[p]
@@
 \end{figure}
-
-\paragraph{Horizontal, intra-level optimization}
-
-\pref{fig:plot-fingerprint-cache-horiz} shows results of the intra-level cache optimization described in \pref{sec:fingerprint-cache}.
-We have implemented two intra-level cache optimized variants.
+\fi % report
+
+%\paragraph{Horizontal, intra-level optimization}
+
+\pref{fig:plot-fingerprint-cache-horiz} shows results of the cache optimization described in \pref{sec:fingerprint-cache}.
+We have implemented two cache optimized variants.
 One as described in \pref{sec:fingerprint-cache}, and one where multiplication and division is replaced with shift operations.
@@
 
+\ifreport
+
 \paragraph{Vertical, inter-level optimization}
@@
 
+\ifreport
 Of the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, we have seen that \proc{DirectComp} is best on average case strings, while \proc{LcpRmq} is best on worst case strings.
@@
 However it is not as interesting as our new \fprintk\ algorithm, because it uses too much space.
-
-Our new \fprintk\ algorithm is able to achieve a balance between worst case and average case query times.
+\fi % report
+
+The \fprintk\ algorithm is able to achieve a balance between worst case and average case query times.
 It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have found cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}.
@@
 The performance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
+
+\ifreport
 

### 2011-08-17 article/compile.sh

@@ -1,0 +1,1 @@
+pdflatex -shell-escape article.tex

### 2011-08-18 src/algFingerprint3AQCacheHorizMult.cpp
m

@@ -1,0 +1,143 @@
+#include "algFingerprint3AQCacheHorizMult.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+Fingerprint3AQCacheHorizMult::Fingerprint3AQCacheHorizMult(NamedFunc* f1, NamedFunc* f2) {
+    s = NULL;
+    n = 0;
+    H1 = NULL;
+    H2 = NULL;
+    t1 = 0;
+    t2 = 0;
+    this->f1 = f1;
+    this->f2 = f2;
+    ndt1 = 0;
+    ndt2 = 0;
+}
+
+// H[(i mod t)*ndt + i div t]
+#define HH1(i) ((i)%t1)*ndt1 + (i)/t1
+#define HH2(i) ((i)%t2)*ndt2 + (i)/t2
+
+Fingerprint3AQCacheHorizMult::~Fingerprint3AQCacheHorizMult() {
+}
+
+bool Fingerprint3AQCacheHorizMult::preproc(TestString* ts) {
+    s = ts->s;
+    n = ts->n;
+
+    t1 = f1->f(n);
+    t2 = f2->f(n);
+    if (t1 < 1)
+        t1 = 1;
+    if (t2 < t1)
+        t2 = t1;
+
+    // ndt = ceil((n+1)/t)
+    ndt1 = (n+1+t1)/t1;
+    ndt2 = (n+1+t2)/t2;
+
+    uint32 max1 = (t1 - 1) * ndt1 + n / t1;
+    uint32 max2 = (t2 - 1) * ndt2 + n / t2;
+
+    H1 = new (std::nothrow) uint32[max1+1];
+    H2 = new (std::nothrow) uint32[max2+1];
+    if (!H1 || !H2) {
+        if (H1) {
+            delete[] H1;
+            H1 = NULL;
+        }
+        if (H2) {
+            delete[] H2;
+            H2 = NULL;
+        }
+        s = NULL;
+        n = 0;
+        t1 = 0;
+        t2 = 0;
+        ndt1 = 0;
+        ndt2 = 0;
+        return false;
+    }
+
+    uint32 h1;
+    uint32 h2;
+    // loop through each suffix s->sa[i] in sorted order.
+    for (uint32 i = 0; i < n; i++) {
+        // if suffix s->sa[i] does not have a prefix of length t in common with its predecessor, then update the current hash
+        if (ts->lcp[i] < t1)
+            h1 = i;
+        if (ts->lcp[i] < t2)
[…]

### 2011-08-18 src/algFingerprint3AQCacheHorizMult.h

@@ -1,0 +1,26 @@
+#ifndef _ALGFINGERPRINT3AQCACHEHORIZMULT_H_
+#define _ALGFINGERPRINT3AQCACHEHORIZMULT_H_
+
+#include "algorithm.h"
+#include "bittricks.h"
+
+class Fingerprint3AQCacheHorizMult : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint3AQCacheHorizMult(NamedFunc*, NamedFunc*);
+    ~Fingerprint3AQCacheHorizMult();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32 *H1, *H2;
+    uint32 t1, t2, ndt1, ndt2;
+    NamedFunc *f1, *f2;
+};
+
+#endif

### 2011-08-18 src/algFingerprint3AQCacheHorizShift.cpp

@@ -1,0 +1,161 @@
+#include "algFingerprint3AQCacheHorizShift.h"
+#include <stdlib.h>
+#include <stdio.h>
+#include <new>
+
+Fingerprint3AQCacheHorizShift::Fingerprint3AQCacheHorizShift(NamedFunc* f1, NamedFunc* f2) {
+    s = NULL;
+    n = 0;
+    H1 = NULL;
+    H2 = NULL;
+    t1 = 0;
+    t2 = 0;
+    this->f1 = f1;
+    this->f2 = f2;
+    S_CLEAR(t1)
+    S_CLEAR(t2)
+    S_CLEAR(ndt1);
+    S_CLEAR(ndt2);
+}
+
+// H[(i mod t)*ndt + i div t]
+#define HH1(i) S_MUL(S_MOD(i, t1), ndt1) + S_DIV(i, t1)
+#define HH2(i) S_MUL(S_MOD(i, t2), ndt2) + S_DIV(i, t2)
+
+Fingerprint3AQCacheHorizShift::~Fingerprint3AQCacheHorizShift() {
+}
+
+bool Fingerprint3AQCacheHorizShift::preproc(TestString* ts) {
+    s = ts->s;
+    n = ts->n;
+
+    t1 = f1->f(n);
+    t2 = f2->f(n);
+    if (t1 < 1)
+        t1 = 1;
+    if (t2 < t1)
+        t2 = t1;
+    S_PREP_MOD(t1)
+    S_PREP_MOD(t2)
+
+    // ndt = ceil((n+1)/t)
+    uint32 ndt1 = (n+1+t1)/t1;
+    uint32 ndt2 = (n+1+t2)/t2;
+    S_PREP(ndt1)
+    S_PREP(ndt2)
+
+    uint32 max1 = S_MUL(t1 - 1, ndt1) + S_DIV(n, t1);
+    uint32 max2 = S_MUL(t2 - 1, ndt2) + S_DIV(n, t2);
+
+    H1 = new (std::nothrow) uint32[max1+1];
+    H2 = new (std::nothrow) uint32[max2+1];
+    if (!H1 || !H2) {
+        if (H1) {
+            delete[] H1;
+            H1 = NULL;
+        }
+        if (H2) {
+            delete[] H2;
+            H2 = NULL;
+        }
+        s = NULL;
+        n = 0;
+        t1 = 0;
+        t2 = 0;
+        S_CLEAR(t1)
[…]

### 2011-08-18 src/algFingerprint3AQCacheHorizShift.h

@@ -1,0 +1,30 @@
+#ifndef _ALGFINGERPRINT3AQCACHEHORIZSHIFT_H_
+#define _ALGFINGERPRINT3AQCACHEHORIZSHIFT_H_
+
+#include "algorithm.h"
+#include "bittricks.h"
+
+class Fingerprint3AQCacheHorizShift : public Algorithm {
+public:
+    virtual bool preproc(TestString*);
+    virtual uint32 query(uint32, uint32);
+    virtual void cleanup();
+    virtual void getName(char*, uint32);
+    virtual uint64 spaceUsage(uint32);
+
+    Fingerprint3AQCacheHorizShift(NamedFunc*, NamedFunc*);
+    ~Fingerprint3AQCacheHorizShift();
+
+protected:
+    uint8* s;
+    uint32 n;
+    uint32 *H1, *H2;
+    uint32 t1, t2;
+    S_DECL_MOD(t1)
+    S_DECL_MOD(t2)
+    S_DECL(ndt1)
+    S_DECL(ndt2)
+    NamedFunc *f1, *f2;
+};
+
+#endif

### 2011-08-18 src/algHash2LCache.cpp

@@ -62,8 +62,9 @@ uint32 Fingerprint2AQCacheHorizShift::query(uint32 i, uint32 j) {
     if (i == j)
         return n - i;
     if (s[i + v] == s[j + v]) {
-        uint32 ihh = HH(i);
-        uint32 jhh = HH(j);
+        v++;
+        uint32 ihh = HH(i + v);
+        uint32 jhh = HH(j + v);
         while (H[ihh] == H[jhh]) {
             v += t;
             ihh++;

### 2011-08-18 src/algHash2LCacheMul.cpp

@@ -57,8 +57,9 @@ uint32 Fingerprint2AQCacheHorizMult::query(uint32 i, uint32 j) {
     if (i == j)
         return n - i;
     if (s[i + v] == s[j + v]) {
-        uint32 ihh = HH(i);
-        uint32 jhh = HH(j);
+        v++;
+        uint32 ihh = HH(i + v);
+        uint32 jhh = HH(j + v);
         while (H[ihh] == H[jhh]) {
             v += t;
             ihh++;

### 2011-08-18 src/main.cpp

@@ -23,6 +23,8 @@
 #ifndef _MSC_VER
 #include "algRmq.h"
 #endif
+#include "algFingerprint3AQCacheHorizShift.h"
+#include "algFingerprint3AQCacheHorizMult.h"
 
 #include "correctness.h"
 #include "functionalTest.h"
@@ -57,6 +59,32 @@ uint32 log2_nf(uint32 n) { return (uint32) log2((double)n); } NamedFunc log2_n =
 uint32 ndlog2nf(uint32 n) { return (uint32) (((double)n)/log2((double)n)); } NamedFunc ndlog2n = {ndlog2nf, "n/log n"};
 
 int main(int argc, char* argv[]) {
+    double preproc_limit = 6;
+    double query_limit = 1e-6;
+    uint32 max_n = 10000000;
+
+    Algorithm* slidealgs[] = {
+        new AlgorithmDirectComp,
+        new Fingerprint2AQCacheHorizShift(&sqrt_n),
+        new Fingerprint2AQCacheHorizMult(&sqrt_n),
+        new Fingerprint3AQCacheHorizShift(&pow23, &pow13),
+        new Fingerprint3AQCacheHorizMult(&pow23, &pow13),
+        new FingerprintLA(),
+        new AlgorithmRmq1,
+        NULL
+    };
+    correctness(slidealgs, RANDOM10_STRING);
+    correctness(slidealgs, RANDOM2_STRING);
+    correctness(slidealgs, ALL_A_STRING);
+    correctness(slidealgs, REPEAT_POW_STRING);
+    graphIncreasingLength(slidealgs, RANDOM10_STRING, "results/length-slides-rand10.txt", preproc_limit, query_limit, max_n);
+    graphIncreasingLength(slidealgs, RANDOM2_STRING, "results/length-slides-rand2.txt", preproc_limit, query_limit, max_n);
+    graphIncreasingLength(slidealgs, ALL_A_STRING, "results/length-slides-alla.txt", preproc_limit, 1e-5, max_n);
+    graphIncreasingLength(slidealgs, REPEAT_POW_STRING, "results/length-slides-repeat-pow.txt", preproc_limit, 1e-5, max_n);
+    printf("\nThe slides end\n");
+}
+
+int notmain(int argc, char* argv[]) {
 #if defined(CACHE_ALLA)
     cacheTest(2000000, ALL_A_STRING);
 #elif defined(CACHE_RAND10)

### 2011-08-18 src/main.cpp
m

@@ -65,8 +65,10 @@ int main(int argc, char* argv[]) {
 
     Algorithm* slidealgs[] = {
         new AlgorithmDirectComp,
+        new Fingerprint2AQ(&sqrt_n),
         new Fingerprint2AQCacheHorizShift(&sqrt_n),
         new Fingerprint2AQCacheHorizMult(&sqrt_n),
+        new Fingerprint3AQ(&pow23, &pow13),
         new Fingerprint3AQCacheHorizShift(&pow23, &pow13),
         new Fingerprint3AQCacheHorizMult(&pow23, &pow13),
         new FingerprintLA(),

### 2011-08-18 presentation/presentation.tex
m

@@
         Average query & $O(1)$ \\
+        Query I/Os & $O\Big(\frac{|\LCE(i,j)|}{B}\Big)$ \\
     \end{tabular}
@@
             Average query & $O(1)$ \\
+            Query I/Os & $O(1)$ \\
         \end{tabular}
@@
 
+\subsection{I/O}
+\begin{frame}{The \fprintk\ Algorithm: I/O}
+    \begin{itemize}
+        \item Original:
+        \begin{itemize}
+            \item Data structure: $H_\ell[i] = F_{t_\ell}[i]$
+            \item Size: $|H_\ell| = n$
+            \item I/Os: $O\big(k\cdot n^{1/k}\big)$
+        \end{itemize}
+        \item Cache optimized:
+        %Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at 
+        \begin{itemize}
+            \item Data structure: \\~~$H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1] = F_{t_\ell}[i]$
+            \item Size: $|H_\ell| = n+t_\ell$
+            \item I/Os: $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$
+            \begin{itemize}
+                \item Best when $k$ is small $\implies$ $n^{1/k}$ is large.
+            \end{itemize}
+        \end{itemize}
+     \end{itemize}
+     %$n+t_\ell-1/t_\ell+2$
+\end{frame}
+
 \begin{frame}{The \fprintk\ Algorithm}
@@
         Average query & $O(1)$ \\
+        Query I/Os & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
     \end{tabular}
@@
 
-    \hfill\\$k=\logceil$:
+    \begin{multicols}{2}{
+        $k=2$:
+
+        \begin{tabular}{r l}
+            Preprocessing & $O(\sortt)$ \\
+            Space & $O(n)$\\
+            Query & $O(\sqrt n)$ \\
+            Average query & $O(1)$ \\
+            Query I/Os & $O\Big(\frac{\sqrt n}{B}\Big)$ \\
+        \end{tabular}
+        \newpage
+        $k=\logceil$:
 
@@
             Average query & $O(1)$ \\
+            Query I/Os & $O(\log n)$ \\
         \end{tabular}
+    }
+    \end{multicols}
 \end{frame}
@@
     \textcolor{red}{\proc{DirectComp}},
-    \textcolor{green}{\fprint[2]},
-    \textcolor{blue}{\fprint[3]},
-    \textcolor{magenta}{\fprint[\logceil]} and
+    \textcolor{green}{\fprint[2] (cache opt.)},
+    \textcolor{blue}{\fprint[3] (not cache opt.)},
+    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
     \textcolor{cyan}{\proc{LcpRmq}}
@@
             Average case
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-rand10.plt}
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
             \newpage
             Worst case
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-alla.plt}
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-alla.plt}
         \end{center}
@@
     \textcolor{red}{\proc{DirectComp}},
-    \textcolor{green}{\fprint[2]},
-    \textcolor{blue}{\fprint[3]},
-    \textcolor{magenta}{\fprint[\logceil]} and
+    \textcolor{green}{\fprint[2] (cache opt.)},
+    \textcolor{blue}{\fprint[3] (not cache opt.)},
+    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
     \textcolor{cyan}{\proc{LcpRmq}}
     by string length
+    \begin{multicols}{2}{
         \begin{center}
             Medium LCE size\\
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-repeat-pow.plt}
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-repeat-pow.plt}
+            \newpage
+            Cache stress\\
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand2.plt}
         \end{center}
+    }
+    \end{multicols}
+\end {frame}
+
+\subsection{Practical I/O}
+\begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
+    Query times of \fprint[2]
+    \textcolor{red}{without cache optimization}
+    and with cache optimization using
+    \textcolor{green}{shift operations} vs.\
+    \textcolor{blue}{multiplication and division}
+    %by string length
+    \begin{multicols}{2}{
+        \begin{center}
+            Average case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-fp2-rand10.plt}
+            \newpage
+            Worst case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-fp2-alla.plt}
+        \end{center}
+    }
+    \end{multicols}
+\end {frame}
+
+\subsection{Practical I/O}
+\begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
+    Query times of \fprint[3]
+    \textcolor{red}{without cache optimization}
+    and with cache optimization using
+    \textcolor{green}{shift operations} vs.\
+    \textcolor{blue}{multiplication and division}
+    %by string length
+    \begin{multicols}{2}{
+        \begin{center}
+            Average case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-fp3-rand10.plt}
+            \newpage
+            Worst case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-fp3-alla.plt}
+        \end{center}
+    }
+    \end{multicols}
 \end {frame}

### 2011-08-18 src/plot-slides.sh

@@ -1,0 +1,86 @@
+
+oneplot()
+{
+    cat plotheader.txt > $plotfile
+    echo "set out '$plotfile.eps'" >> $plotfile
+    echo "set xrange [10:]" >> $plotfile
+    title=$(head -1 $datafile | sed -e 's/,/\n/g' | head -1)
+    # echo "set title '$title'" >> $plotfile
+    echo -e "$xtra" >> $plotfile
+    echo "plot \\" >> $plotfile
+    first=1
+    for algo in "$@"
+    do
+        if test $first = 0
+        then
+            echo ", \\" >> $plotfile
+        fi
+        first=0
+        algoidx=$(head -1 $datafile | sed -e 's/,/,\n/g' | grep -nF "$algo" -m1 | cut -d':' -f1)
+        echo -n "  '$datafile' using 1:$algoidx with $style" >> $plotfile
+    done
+    echo "Plotting $plotfile";
+    gnuplot $plotfile; ps2pdf -dEPSCrop $plotfile.eps $plotfile.pdf
+}
+
+listdata()
+{
+    head -1 $datafile | sed -e 's/,/,\n/g'
+    echo ""
+}
+
+yavg="set logscale y\nset yrange [7e-9:1e-6]\nset title 'Average case strings'"
+ywc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Worst case strings'"
+ymc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Medium LCE-size strings'"
+
+style="linespoints"
+
+#Overall Cache Fingerprint Slides
+
+slides="\nunset title\nunset xlabel\nunset ylabel\nunset key"
+xtra="$yavg$slides"
+datafile="results/length-slides-rand10.txt"
+plotfile="results/length-slides-cache-rand10.plt"
+oneplot "DirectComp," "Fingerprint_2<sqrt n>ac_{cache-horizontal-mult}," "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_{log n}ac," "RMQ<n;1>,"
+
+xtra="$yavg$slides"
+datafile="results/length-slides-rand2.txt"
+plotfile="results/length-slides-cache-rand2.plt"
[…]

### 2011-08-19 article/article.tex
m

@@
-% TODOs
-% skarpere plots
-% sørg for at figurer ikke sidder underligt
-% Prøv at formindske for store mellemrum mellem ord ved at lave påtvungne linjeskift
+%- Efter "..constant query time and linear space" skal der være referencer.
+%  * Til hvem? Navarri refererer til konstanttids NCA og RMQ men ikke til en kilde der beskriver hvordan de bruges til at lave LCE. Johannes Fischer skriver bare "It is well-known".
+%- Er navnene på algoritmerne de samme som i de oprindelige artikler? I
+%  så fald, så drop "we call". Der skal selvfølgelig også være
+%  referencer i dette afsnit lige i starten af definitionen på hver
+%  algoritme/datastruktur. Skriv gerne navnene på forfatterne: "Navarro
+%  et al[?] gave an algorithm..." eller lign. 
+%  * DirectComp har jeg ikke selv navngivet. SuffixNca og LcpRmq navnene har jeg selv fundet på. De har ikke noget navn i de tidligere artikler.
 
@@
 
-\tableofcontents
+%\tableofcontents
 
@@
 
-The Longest Common Extension problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
-Optimal solutions exists for the problem, with constant query time and linear space.
-These theoretically optimal solutions are however not the best in practice, since they have large constant factors for both time and space usage.
-In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance.
+The \emph{longest common extension} (LCE) problem is to preprocess a string in order to allow for a large number of LCE querie, such that the queries are efficient.
+The LCE value, $\LCE_s(i,j)$, is the length of the longest common prefix of the pair of suffixes starting at index $i$ and $j$ in the string $s$.
+The LCE problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
+Solutions with linear space with constant query time exists for the problem~[TODO cite who].
+These theoretically good solutions are however not the best in practice, since they have large constant factors for both time and space usage.
+In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance~\cite{ilie-navarro-tinta}.
 This algorithm is ideal when only average case performance is relevant.
 In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.
-
-We look at a new algorithm, which has better than linear worst case query times, and achieves significantly better average case practical query times compared to the theoretically best algorithms.
-This algorithm uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.
+Such a situation could be use of approximate string searching in a firewall, which should not allow an attacker to significantly degrade its performance by sending it carefully crafted packages, while the average case of scanning legitimate data must still be highly performant.
+
+In this paper we give a new algorithm, which is a generalization of the simple linear time algorithm, which is almost as simple, yet it has significantly better worst case query times.
+This algorithm also achieves significantly better average case practical query times compared to the theoretically best algorithms.
+The algorithm uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The LCE Problem}
-
-Given a string $s$ of length $n$ and a pair of indexes $1\leq i\leq n$ and $1\leq j\leq n$, the Longest Common Extension of $i$ and $j$ on $s$, $\LCE_s(i,j)$, is the length of the longest common prefix of $\suff_i$ and $\suff_j$.
+%\subsection{The LCE Problem}
+
+%Given a string $s$ of length $n$ and a pair of indexes $1\leq i\leq n$ and $1\leq j\leq n$, the \emph{longest common extension} of $i$ and $j$ on $s$, $\LCE_s(i,j)$, is the length of the longest common prefix of $\suff_i$ and $\suff_j$.
 If for example $s = abbababba$, $i = 4$ and $j=6$, then $\suff_4 = ababba$ and $\suff_6 = abba$.
@@
 
-The LCE problem is to preprocess a string in order to allow for a large number of LCE queries on different pairs $(i,j)$, such that the queries are efficient.
-
-Given a string length $n$ and an alphabet size $\sigma$, we define average case query time as the average of query times over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Existing Results\label{sec:existing-results}}
-
-One way of solving the LCE problem without any preprocessing uses $O(|\LCE(i,j)|)$ query time.
-We call this algorithm \proc{DirectComp}.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Previous Results\label{sec:existing-results}}
+
+Ilie~et~al.~\cite{ilie-navarro-tinta} gave and algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and $O(|\LCE(i,j)|)$ query time.
 For a query $\LCE(i,j)$, the algorithm compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
-The worst case query time is $O(n)$ on a string of length $n$, and the average case query time is $O(1)$, because the average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}.
-
-The LCE problem can be solved optimally with $O(1)$ worst case query time, using $O(n)$ space and $O(\sortt)$ preprocessing time. Two different ways of doing this exists.
-
-One method, which we call \proc{SuffixNca}, uses constant time Nearest Common Ancestor queries on a suffix tree.
+The worst case query time is $O(n)$ on a string of length $n$.
+Given a string length $n$ and an alphabet size $\sigma$, we define \emph{average case query time} as the average of query times over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
+The average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}, which gives \proc{DirectComp} average case query time $O(1)$.
+The LCE problem can be solved with $O(1)$ worst case query time, using $O(n)$ space and $O(\sortt)$ preprocessing time. Two different ways of doing this exists.
+One method, \proc{SuffixNca}, uses constant time nearest common ancestor (NCA) queries on a suffix tree.
 The LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$.
@@
 $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.
-
-The other method, which we call \proc{LcpRmq}, uses constant time Range Minimum Queries on a Longest Common Prefix array.
-The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA.
+The other method, \proc{LcpRmq}, uses constant time range minimum queries (RMQ) on a longest common prefix (LCP) array.
+The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in the suffix array (SA).
 The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering.
@@
 $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \subsubsection{Existing Practical Results}
-
 Ilie~et~al.~\cite{ilie-navarro-tinta} have looked at a number of real world texts as well as texts of randomly generated characters, and found that all the texts they examined each has an average LCE of at most one, over all $n^2$ input pairs.
 Therefore it is interesting to have LCE algorithms, which perform good on average when the LCE value is small.
-
 Both \proc{SuffixNca} and \proc{LcpRmq} have the same asymptotic space and times.
@@
 
-We present a new LCE algorithm based on string fingerprinting.
-We call our algorithm \fprintk, where $k$ is a parameter $1\leq k\leq\logceil$, which describes the number of levels used\footnote{All logarithms are base two.}.
+We present a new LCE algorithm, \fprintk, where $k$ is a parameter $1\leq k\leq\logceil$, which describes the number of levels used\footnote{All logarithms are base two.}.
+This algorithm is based on string fingerprinting.
 
@@
 To preprocess the $O(k\cdot n)$ fingerprints used by our algorithm, we can use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, which takes $O(n\log n)$ time.
-For $k=o(\log n)$, we can speed up preprocessing to $O(\sortt + k\cdot n)$ by using the Suffix Array and Longest Common Prefix array.
+For $k=o(\log n)$, we can speed up preprocessing to $O(\sortt + k\cdot n)$ by using the SA and LCP arrays.
 
@@
 \proc{DirectComp} & $O(1)$ & $O(n)$ & $O(1)$ \\ \hline
-\fprintk & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
-~~~~$k=\logceil$ & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline
+\fprintk * & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
+~~~~$k=\logceil$ * & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline
 \end{tabular}
-\caption{LCE algorithms with their space requirements, worst case query times and preprocessing times. Average case query times are $O(1)$ for all shown algorithms.}\label{tab:article-algorithms}
+\caption{LCE algorithms with their space requirements, worst case query times and preprocessing times.
+Average case query times are $O(1)$ for all shown algorithms.
+Rows marked with * show the new algorithm we present.}\label{tab:article-algorithms}
 \end{table}
@@
 
-\ifreport
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Our Table Lookup Algorithm}
-
-We also present another new LCE algorithm, which we call \proc{DirectCompLookup}.
-
-\begin{theorem}
-For a string $s$ of length $n$, the \proc{DirectCompLookup} algorithm can solve the LCE problem in $O(t)$ worst case query time and $O(1)$ average case query time using $O(n^2/t)$ space and $O(n^2)$ preprocessing time, where $t$ is a parameter $1 \leq t \leq n$.
-\end{theorem}
-
-The algorithm stores LCE values for some of the $n^2$ possible query inputs on a given string.
-It uses \proc{DirectComp} to move towards one of these stored values, and stops when it reaches one.
-The \proc{DirectCompLookup} algorithm is very simple, but the tradeoff between worst case query time and space is not very favorable, and preprocessing the data structure takes $O(n^2)$ time unless we use another LCE algorithm in the preprocessing.
-
-\fi % report
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Overview}
+In \pref{sec:fprintk} we present the \fprintk\ algorithm and analyze its theoretical complexity, and in \pref{sec:results} we present our findings from tests of a practical implementation of \fprintk, which we compare against the existing algorithms.
 
@@
 
-\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet size $\sigma$, the Suffix Array, \SA, is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$.
+\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet size $\sigma$, the \emph{suffix array} (\SA) is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$.
 The lexicographically smallest suffix is $\suff_{\SA[1]}$, the lexicographically largest suffix is $\suff_{\SA[n]}$, and the lexicographically $i$'th smallest suffix is $\suff_{\SA[i]}$.
-The inverse Suffix Array, \SAi, describes where a given suffix is in the lexicographical order.
+The \emph{inverse suffix array} (\SAi) describes where a given suffix is in the lexicographical order.
 Suffix $\suff_i$ is the lexicographically $\SAinv[i]$'th smallest suffix.
 
-The Longest Common Prefix Array, LCP, describes the length of longest common prefixes of neighboring suffixes in SA.
-$\LCP[i]$ is the length of the longest common prefix of $\suff_{\SA[i-1]}$ and $\suff_{\SA[i]}$ for $2 \leq i \leq n$.
+The \emph{longest common prefix array} (LCP array) describes the length of longest common prefixes of neighboring suffixes in SA.
+The length of the longest common prefix of $\suff_{\SA[i-1]}$ and $\suff_{\SA[i]}$ is $\LCP[i]$, for $2 \leq i \leq n$.
 The first element $\LCP[1]$ is always zero.
 
-Building the SA, \SAi and LCP arrays takes $O(\sortt)$ time~\cite{sort-complexity}.
+Building the SA, \SAi\ and LCP arrays takes $O(\sortt)$ time~\cite{sort-complexity}.
 
 \paragraph{NCA.} The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
-The Nearest Common Ancestor of two nodes $u$ and $v$ is the node of greatest depth, which is an ancestor of both $u$ and $v$.
+The \emph{nearest common ancestor} (NCA) of two nodes $u$ and $v$ is the node of greatest depth, which is an ancestor of both $u$ and $v$.
 An NCA query can be answered in $O(1)$ time with $O(n)$ space and preprocessing time in a static tree with $n$ nodes~\cite{nca}.
@@
 \paragraph{RMQ.} The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, i.e.\ $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
-A Range Minimum Query on a static array of $n$ elements can be answered in $O(1)$ time with $O(n)$ space and preprocessing time~\cite{jf-rmq}.
-
-\ifreport
-
-\paragraph{I/O's and Cache-Oblivious Model.}
+A \emph{range minimum query} (RMQ) on a static array of $n$ elements can be answered in $O(1)$ time with $O(n)$ space and preprocessing time~\cite{jf-rmq}.
+
+\paragraph{I/O and Cache-Oblivious Model.}
 The I/O model describes the number of memory blocks an algorithm moves between two layers of a layered memory architecture, where the size of the internal memory layer is $M$ words, and data is moved between internal and external memory in blocks of $B$ words.
-In the cache-oblivious model, the algorithm has no knowledge of the values of $M$ and $B$.
-
-\paragraph{Predecessor.}
-Given a set of $n$ objects, where the key of each object is a natural number between one and $U$, the predecessor of a natural number $x$ is an object with maximal key, whose key is at most $x$.
-We can solve the static Predecessor problem in $O(\log\log U)$ time using $O(n)$ space~\cite{predecessor}.
-
-\paragraph{Level Ancestor.}
-The Level Ancestor at level $\ell$ of node $u$, $\LA(\ell,u)$, is the ancestor of $u$, whose depth is $\ell$.
-We can preprocess a static tree in $O(n)$ time and space to answer Level Ancestor queries in $O(1)$ time~\cite{level-ancestor}.
-
-\paragraph{Heavy Path Decomposition}
-The weight of a node $u$ is the size of the subtree rooted at $u$.
-For each node $u$, a child of greatest weight is marked heavy, and all other children are marked light.
-The root is always marked light.
-An edge going from a node to a heavy child node is heavy, and an edge going from a node to a light child node is light.
-A path of only heavy edges is a heavy path, and the combined length of all heavy paths in a tree of $n$ nodes is less than $n$.
-The markings of heavy and light edges and nodes in a tree is a Heavy Path Decomposition.
-Every simple path in a tree of size $n$ contains at most $O(\log n)$ light edges, and we can construct a Heavy Path Decomposition in $O(n)$ time~\cite{nca}.
-
-\paragraph{Ladder Decomposition}
-To make a ladder decomposition of a tree, we first create a ladder of nodes on the path from the root of the tree to the leaf of greatest depth.
-We then remove the nodes in the ladder from the tree and repeat this operation recursively for all remaining subtrees, until all nodes are part of a ladder.
-Node $u$'s ladder is the ladder which contains $u$ at this point.
-Finally, we double the number of nodes in each ladder by extending it towards the root of the tree.
-If we reach the root before we have doubled the number of nodes, we stop at the root.
-Each node may thus be in multiple ladders.
-
-For any path in a tree of $n$ nodes, we can find $O(\log n)$ ladders, which contain all the nodes in the path.
-If a node $u$ has height $h$, $u$'s ladder contains all the ancestors of $u$ whose distance to $u$ is at most $h+1$.
-The combined size of the ladders is at most $2n$.~\cite{level-ancestor}
+In the \emph{cache-oblivious model}, the algorithm has no knowledge of the values of $M$ and $B$.
+
+\ifreport
 
@@
 
-Different algorithms exists for solving the RMQ problem, and we can solve it in optimal $O(1)$ time and $O(n)$ space.
+Different algorithms exists for solving the RMQ problem, and we can solve it in $O(1)$ time and $O(n)$ space.
 But like for the LCE problem, the asymp-\\totically best RMQ solution may not be the best RMQ solution in practice.
@@
 \ifarticle
-\section{The \fprintk\ Algorithm}
+\section{The \fprintk\ Algorithm\label{sec:fprintk}}
 \fi % article
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Definition of a Fingerprint}
+\subsection{Data Structure\label{sec:fingerprint-ds}}
 
@@
 At the end of a string when $i+t-1>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Data Structure\label{sec:fingerprint-ds}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Query\label{sec:fingerprint-query}}
+\subsection{Query\label{sec:fingerprint-query}}
 
@@
 \begin{lemma}
-The \fprintk\ query algorithm given above finds the correct LCE value.
+The \fprintk\ query algorithm is correct.
 \end{lemma}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Average Case Optimization\label{sec:improved-avg}}
-
-We could have left out step one of the query algorithm in \pref{sec:fingerprint-query} and started with $\ell=k-1$.
+\paragraph{Average Case Optimization\label{sec:improved-avg}}
+
+We could have left out step one of the query algorithm and started with $\ell=k-1$.
 This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Preprocessing}
+\subsection{Preprocessing}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Cache Optimization\label{sec:fingerprint-cache}}
+\subsection{Cache Optimization\label{sec:fingerprint-cache}}
 
@@
 
-The amount of I/O's used by \fprintk\ is $O(k\cdot n^{1/k})$.
-However if we structure our tables of fingerprints differently, we can improve the number of I/O's to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
+The amount of I/O used by \fprintk\ is $O(k\cdot n^{1/k})$.
+However if we structure our tables of fingerprints differently, we can improve the number of I/O operations to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
 Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$.
-This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O's at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.
+This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.
 
@@
 We achieve the greatest I/O improvement when $k$ is small.
-When $k=\logceil$, this cache optimization gives no difference in the amount of I/O's.
+When $k=\logceil$, this cache optimization gives no difference in the amount of I/O.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Cache Optimization}
+\subsection{Cache Optimization}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Conclusions on Experimental Results}
+\section{Conclusions}
+
+We have presented the \fprintk\ algorithm, where $k$ is a parameter $1 \leq k \leq \logceil$, which for a string $s$ of length $n$ and alphabet size $\sigma$, can solve the LCE problem in $O(k\cdot n^{1/k})$ worst case query time and $O(1)$ average case query time using $O(k\cdot n)$ space and $O(\sortt + k\cdot n)$ preprocessing time.
 
@@
 
-The \fprintk\ algorithm is able to achieve a balance between worst case and average case query times.
+The \fprintk\ algorithm is able to achieve a balance between practical worst case and average case query times.
 It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have found cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}.
 \fprintk\ gives a good time space tradeoff, and it uses less space than \proc{LcpRmq} when $k$ is small.
-The performance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
+%The performance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-\section{Ziv-Lempel Compression\label{sec:lz-compress}}
-
-\newcommand{\zreference}{\textit{reference}}
-\newcommand{\zlabel}{\textit{label}}
-\newcommand{\zphrase}{\textit{phrase}}
-
-Sometimes it can be useful to store a string in a compressed format and query the compressed string directly without using extra space and time to first decompress the string.
-In this section we explore the LCE problem on strings compressed using LZ78 compression.
-We look at solutions, which require space relative to the compressed size of a string, and we try to find solutions, which can answer LCE queries fast.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{LZ78 definition}
-A string $s$ of length $N$ has a corresponding \textit{compressed string} $z$ of compressed length $n$, where $z$ is a list of \textit{compression elements}, $z = z_1, z_2, ..., z_n$.
-Each compression element $z_x$ is a pair $(r_x, \alpha_x)$, where the reference, $\zreference_z(x)=r_x$, is an index of an earlier compression element in the compressed string or zero, i.e.
-$0 \leq \zreference_z(x) < x$, and the label, $\zlabel_z(x)=\alpha_z$, is a character from the original uncompressed string $s$.
-The \zphrase\ of a compression element is defined as:
-\begin{align*}
-\zphrase_z(x) =
-\begin{cases}
-    \zphrase_z(\zreference_z(x)) \cdot \zlabel_z(x) & \textrm{if}~ \zreference_z(x) \neq 0\\
-    \zlabel_z(x) & \textrm{otherwise}
-\end{cases}
-\end{align*}
-where "$\cdot$" denotes string concatenation. From a compressed string $z = z_1,...,z_n$, we can get the original string $s=\zphrase_z(1)\cdot...\cdot\zphrase_z(n)$.
-
-\subsubsection{Compressing and Decompressing\label{sec:lz-comp-alg}}
-
-We can easily get the decompressed string $s$ from the compressed string $z$ using the definitions.
-To get the compressed string $z$ from the original string $s$ we can look at the compression elements as a tree, where the reference of each compression element corresponds to the parent pointer of each node in the tree.
-We consume $s$ from left to right while we build the tree:
-\begin{enumerate}
-\item Let the tree be a single special root node $z_0$, let $i = 1$ and $c=z_0$.
-\item If $c$ has a child $u$, where $\zlabel(u)=s[i]$, set $c=u$. Otherwise create a new node $u$ as a child of $c$, set $\zlabel(u)=s[i]$, and set $c=z_0$.
-\item Increment $i$ by one and repeat from step two until the entire string is consumed when $i = N$.
-\end{enumerate}
-The compressed string $z$ is now the tree except the special root node $z_0$.
-
-The size of the compressed string is at most $n=O(N)$ and at least $n=\Omega(\sqrt{N})$.
-% N= 1+2+3+4+...+n=(n^2+n)/2
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Compressed \proc{DirectComp}\label{sec:lz-iterate-forward}}
-
-We can adapt \proc{DirectComp} to work on a compressed string. To work directly on a compressed string, we first need to have random access to the characters of the original string.
-
-\begin{lemma}
-\label{lem:lz-random-access}
-Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$. For any $1 \leq i \leq N$, we can find $s[i]$ in $O(\log\log N)$ query time using $O(n)$ space.
-\end{lemma}
-
-\begin{proof}
-To find the character $s[i]$, we can first use a Predecessor query to find out which phrase $s[i]$ is in. We can then use a Level Ancestor query to find the correct character from the found phrase.
-
-% We can find the $i$'th character of $s$ as $s[i] = \zphrase(\proc{Pred}(i))[i - \textit{start}[\proc{Pred}(i)]]$ where $\proc{Pred}(i)$ is the index of the compression element containing $s[i]$, and $\textit{start}[x]$ is the position in $s$ of the first character of $\zphrase(x)$, i.e.
-$s[\textit{start}[x]] = \zphrase(x)[1]$.
-
-% We can find \proc{Pred}(i) with a predecessor query over predecessor datastructure of the $n$ compression elemets, where each compression element $x$ is added using $\textit{start}[x] as its key, and the universe has size N.
-When we look at the compression elements as a tree, we have $\zlabel(x)[i]=\proc{Level-Ancestor}(i,x)$.
-We therefore have random access to che characters of a compressed string using one Predecessor query and one Level Ancestor query.
-
-The query time for a random access query is $O(\log\log N)$, since we use $O(\log\log N)$ time on the Predecessor query and $O(1)$ time on the Level-Ancestor query.
-The space needed to support random access queries is $O(n)$, since both the predecessor data structure and the level ancestor data structure use $O(n)$ space.
-\end{proof}
-
-\begin{theorem}
-Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$.
-Compressed \proc{DirectComp} can find $\LCE(i,j)$ in $O(\log\log N + |\LCE(i,j)|)$ query time using $O(n)$ space.
-\end{theorem}
-
-\begin{proof}
-Whenever \proc{DirectComp} reads a character of the string $s$, we instead use the random access described in \pref{lem:lz-random-access} to read the character from the compressed string.
-The predecessor query is only needed in the first comparison of the two indexes $i$ and $j$ in $\LCE(i,j)$.
-In subsequent iterations of \proc{DirectComp} we always need the same phrase or the phrase following the one we used in the previous iteration of the loop, and we therefore do not need the predecessor query to find the phrase.
-The total query time for \proc{DirectComp} on LZ-compressed strings becomes $O(\log\log N + |\LCE(i,j)|)$, which is $O(N)$ in worst case and $O(\log \log N)$ on average.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Reversed Compressed \proc{DirectComp}}
-
-A simpler way of doing LCE on a compressed string is to reverse the string and then run \proc{DirectComp} in the reverse direction on the reversed string, which gives the same result since reversing a string twice gives the original string.
-Where \proc{DirectComp} compares $s[i+v]$ against $s[j+v]$ in each iteration of its loop, reversed \proc{DirectComp} compares $s^R[N+1-i-v]$ against $s^R[N+1-j-v]$ in each iteration.
-This is simpler than using \proc{DirectComp} directly on the compressed string, because Level Ancestor queries are no longer needed at each iteration, since we can use the reference of the current compression element as follows, where the position of each of $i$ and $j$ is stored as a pair $(x,y)$:
-\begin{align*}
-\proc{PreviousPosition}(x, y)=&
-\begin{cases}
-(\zreference(x), y) & \textrm{if} ~ \zreference(x) \neq 0 \\
-(y-1, y-1) & \textrm{if} ~ y > 1\\
-\textit{End of string} & \textrm{otherwise}
-\end{cases}\\
-\proc{Character}(x, y)=&~\zlabel(x)\\
-\proc{Position}(i)=&~(x,y)~\textrm{where}\\
-&~~y=\proc{Predecessor}(i)~\textrm{and}\\
-&~~x=\proc{Level-Ancestor}(i-\proc{Start-Of}(y), y)
-\end{align*}
-We then only need one Predecessor and one Level Ancestor query to initialize each of the two pairs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Random Access Initialization}
-
-We can try to completely avoid Predecessor and Level Ancestor when initializing the pair $(x,y)$.
-We could try to store all $N$ pairs, but that would require $O(N)$ space and defeat the purpose of compression.
-We could also just walk from the end of the string to the desired position, but that would bring the average case query time up to $O(N)$.
-Similarly to how \proc{DirectCompLookup} works, we can store some of these positions.
-If we store $t$ positions equally spaced across the string $s$, we would get $O(n+t)$ space and $O(N/t+|\LCE(i,j)|)$ query time.
-To keep the space within $O(n)$, we can set $t=n$, which gives $O(N/n+|\LCE(i,j)|) = O(\sqrt{N}+|\LCE(i,j)|)$ query time.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Length of the Reversed String}
-
-We can reverse the string in $O(N)$ time and $O(n_1 + n_2)$ space, where $n_1$ and $n_2$ are the compressed lengths of the original and reversed string respectively.
-To achieve this, we have already shown how we can read a string from right to left one character at a time in $O(1)$ space, and the compression algorithm described in \pref{sec:lz-comp-alg} reads its input string one character at a time from left to right and uses $O(n_2)$ space for its tree of compression elements.
-
-But the question is how big $n_2$ is relative to $n_1$.
-From knowing how much compression is possible with LZ-compression, we can see that $n_2=O(n_1^{~2})$.
-It might be possible to improve that bound, but for the string $z=(0,a)$ $(1,b)$ $(2,c)$ $...$ $(n-1,z)$, we have experimentally determined that reversing $z$ gives $n_2=0.89n_1^{~1.51}$, so we cannot get an optimal bound of $n_2=O(n_1)$.
-This is a disadvantage of doing LCE on the reversed string compared to doing it on the original compressed string using Level Ancestor queries.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using Compression to Speed Up \proc{DirectComp}}
-
-Our original goal of looking into compression was to see if using the techniques used in compression could improve the worst case query time of \proc{DirectComp} to something better than $O(N)$.
-Our idea is that whenever you have a long LCE value, you have a repetition within the string, and repetitions would typically give good compression.
-If the compressed string can be searched over similarly to how \proc{DirectComp} searches the uncompressed string, it might be possible to establish a better worst case query time.
-
-We have found one case where we can make the reversed compressed \proc{DirectComp} faster than \proc{DirectComp} on uncompressed strings, but our finding does not improve worst case query time.
-Reversed compressed \proc{DirectComp} starts with two positions $(x_i,y_i)$ and $(x_j,y_j)$ corresponding to the two positions $i$ and $j$ in $\LCE(i,j)$.
-At each iteration, it then updates $(x_i,y_i)$ and $(x_j,y_j)$ to find the previous two positions.
-If $x_i = x_j$ we know that we can add $d(x_i)$ to the LCE value, where $d(x_i)$ is the depth of node $x_i$ in the tree of compression elements.
-Reversed compressed \proc{DirectComp} can then compare $d(x_i)$ characters in constant time.
-
-Through this technique can improve query time in some cases, it does not improve the worst case time of $\Theta(N)$.
-This is because we cannot be sure to hit the condition allowing us to compare many characters in constant time, even for long LCE values.
-As an example, $LCE(1,2)$ on a string where all $N$ characters are the same takes $\Omega(N)$ time.
-For such a string, the tree representing the compressed string is always a path, and $x_i$ and $x_j$ will always be either a node and its parent or the root and some other node.
-They will therefore never end up being the same node at the same time, as shown in \pref{fig:lz-no-jumps}.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.9\textwidth]{lz.pdf}
-    \end{center}
-    \caption{\label{fig:lz-no-jumps}Reversed compressed \proc{DirectComp} for $LCE(1,2)$ on the string $s=aaaaaaaaaa\$$. None of the steps can be skipped, since $x_i$ and $x_j$ are never the same.}
-\end{figure}
-
-% \fixme{Compression: But can we cache some results as we go even if pointers are not the same? No, that is Theta(n**2) = Omega(N) space. But do we really need it all?}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Compression with Other Algorithms\label{sec:lz-other}}
-
-We can look at the other LCE algorithms we have examined in addition to \proc{DirectComp}, to see if we can apply them to LZ-compressed strings.
-We have not found a way to apply \proc{SuffixNca} or \proc{LcpRmq}, as the string we want to find LCE values on still has N suffixes, and we have not found a way to relate the compression rate of the NCA structure or suffix array to the compression rate of the string.
-
-Fingerprinting does not compress well either.
-We could LZ-compress each level of fingerprints just like the original string is compressed, but as the fingerprinted substrings become longer at each level, the compression ratio becomes worse, and at the top level with $t_\ell = \Omega(N)$, we need $\Omega(N)$ space to store $H_\ell$, since all fingerprints would be unique or have very few copies.
-
-Since we have not been able to find improvements to the $O(N)$ worst case query time in $O(n)$ space using existing algorithms, we have looked into new algorithms.
-The compression elements of a compressed string $z$ forms a tree, and each phrase in $z$ is a path in the tree.
-The LCE value is the length of two equivalent substrings, which occur at positions $i$ and $j$ in the input string.
-These substrings span a number of phrases.
-If we can make fast LCE queries on a tree of compression elements, we can make fast LCE queries on highly compressable strings.
-
-\begin{theorem}
-\label{thm:lz-tree}
-Let $s$ be a string of length $N$ and let $z$ of length $n$ be the LZ-compression of $s$.
-We can preprocess $z$ using $O(n)$ space to answer LCE queries $v=\LCE(i,j)$ in $O(\log\log N + \log n \cdot \textit{\#phrases})$ time, where \textit{\#phrases} is the number of phrases of $z$ used to store $s[i\twodots i+v-1]$ and $s[j\twodots j+v-1]$.
-\end{theorem}
-
-\begin{proof}
-In \pref{sec:tree-lce} we describe how to perform LCE queries on a static tree of $n$ nodes in $O(\log n)$ query time and $O(n)$ space.
-We can first find the positions of $i$ and $j$ within $z$ in $O(\log\log N)$ time, and then we can repeat the $O(\log n)$ tree LCE query \textit{\#phrases} times to find the LCE value.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Cache Performance}
-
-We have looked into the cache performance of compressed \proc{DirectComp}.
-If we ignore the predecessor and level ancestor queries, and the string does not compress at all, i.e.\ the compressed string is $z=(0,a)(0,b)(0,c)...$, compressed \proc{DirectComp} reads from the string in the best way possible using $O(N/B)$ I/O's.
-When the string has a better LZ-compression ratio, the I/O pattern becomes much worse.
-If $\zreference(x_1) = x_2 > 0$, the algorithm will read from position $x_1$ and then $x_2$, and these two may not be in the same block in memory.
-This can happen many times, which gives $O(N)$ I/O's in worst case.
-
-We could lay out the tree of compression elements in memory as a B-tree, where we store groups of adjacent nodes next to each other in memory, to optimize the number of I/O's within each phrase, but each node may have more than $B$ children, and in this case it would not improve the worst case number of I/O's, as each node visited may be in a new group.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{LCE on Trees\label{sec:tree-lce}}
-
-\newcommand{\zlength}{\textit{length}}
-\newcommand{\tfprint}[1][k]{\ensuremath{\proc{TreeFingerprint}_{#1}}}
-\newcommand{\tfprintk}{\tfprint[k]}
-\newcommand{\zindex}{\textit{index}}
-\newcommand{\zend}{\textit{end}}
-
-To achieve the result of \pref{thm:lz-tree}, we need to do LCE on a tree in $O(\log n)$ query time and $O(n)$ space.
-In this section we analyze ways to achieve this.
-We first define LCE on trees, and then describe different solutions we have found to the problem.
-\pref{tab:tree-lce-algs} lists the different solutions we have found.
-
-\begin{table}[tp]
-\centering
-\begin{tabular}{l|l|l|l}
-\hline\hline
-Algorithm & Space & Query time & Preprocessing \\ [0.5ex] \hline
-\proc{TreeDirectComp} & $O(n)$ & $O(n)$ & $O(n)$ \\ \hline
-\tfprintk, $1\leq k\leq\logceil$ & $O(kn)$ & $O(k\cdot n^{1/k})$ & $O(n\log n)$ \\ \hline
-\proc{HeavyPathLce} & $O(n)$ & $O(\log n)$ & $O(\sortt)$ \\ \hline
-\proc{TreeLcpRmq} & $O(n^2)$ & $O(1)$ & - \\ \hline
-\proc{TreeSuffixNca} & $O(n^2)$ & $O(1)$ & - \\ \hline
-\proc{LadderLce} & $O(n)$ & $O(\log n)$ & $O(\sortt)$ \\ \hline
-\end{tabular}
-\caption{Algorithms for LCE on trees with their space requirements, query times and preprocessing times.}\label{tab:tree-lce-algs}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Definitions\label{sec:tree-def}}
-
-\paragraph{Trees}
-Let $T$ be a rooted tree with $n$ nodes where each edge from parent node $u$ to child node $w$ is labeled with a character $\zlabel(w)$.
-
-\paragraph{Paths}
-A path $p=(u,w)$ in a tree is a path, which starts at node $u$ and ends at node $w$, where $u$ is an ancestor of $w$. We use the following notation on paths:
-\begin{itemize}
-\item The path $p=(u,w)$ has length $e=\zlength(p)=d(w)-d(u)$, i.e.\ $p$ has $e$ edges and $e+1$ nodes.
-\item The node $p[x]$ is a node in $p=(u,w)$, such that the distance between $u$ and $p[x]$ is $x$.
-The start and end nodes of the path are $p[0]=u$ and $p[e]=w$.
-We can calculate $p[x]$ in constant time as $p[x]=\LA(d(u)+x,w)$, where $d(u)$ is the depth of $u$, and $\LA$ is a Level Ancestor query.
-\item The subpath $p[x\twodots y]=(p[x],p[y])$ of a path $p$ is the path, which starts at node $p[x]$ and ends at node $p[y]$. The entire path is $p = p[0\twodots e]$.
-\item The label of a path $\zlabel(p)=\zlabel(p[1])\cdot...\cdot\zlabel(p[e])$ is the concatenation of the labels on all edges between nodes in the path.
-\end{itemize}
-
-\paragraph{LCE on trees}
-Given two paths $p_i$ and $p_j$ of lengths $e_i$ and $e_j$ in a tree $T$, the Longest Common Extension $v=\LCE_T(p_i, p_j)$ is the length of the longest common prefix of $\zlabel(p_i)$ and $\zlabel(p_j)$.
-
-In other words $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and either $v=e_i$, $v=e_j$, or $\zlabel(p_i[v+1]) \neq \zlabel(p_j[v+1])$.
-
-This definition of LCE on a tree is chosen to match the tree of LZ compression elements and the necessary LCE queries on it described in \pref{sec:lz-other}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Walking the Paths with Level Ancestor\label{sec:tree-dc}}
-
-We can find $\LCE(i,j)$ in $O(|\LCE(i,j)|)$ query time using Level Ancestor, in the same way \proc{DirectComp} finds LCE on strings: First compare $\zlabel(p_i[1])$ to $\zlabel(p_j[1])$, then compare $\zlabel(p_i[2])$ to $\zlabel(p_j[2])$ and so on until the two characters differ or either $p_i[e_i]$ or $p_j[e_j]$ is reached.
-
-\begin{theorem}
-\proc{TreeDirectComp} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(|\LCE(i,j)|)$ query time, and $O(n)$ space and preprocessing time.
-\end{theorem}
-\begin{proof}
-Each Level Ancestor query takes constant time, and we do $2|\LCE(i,j)| + 2$ of them, giving us $O(|\LCE(i,j)|)$ query time. The space and preprocessing requirement of Level Ancestor is $O(n)$.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using Fingerprinting}
-
-We can adapt our \fprintk\ algorithm on strings to work on trees, which we call \tfprintk.
-If we store the fingerprint of $s[i\twodots i+t_\ell-1]$ in $H_\ell[i+t_\ell]$ instead of $H_\ell[i]$, then the \fprintk\ query directly adapts to \tfprintk.
-If we wanted to use the same preprocessing as we use for \fprintk\ on strings, we would have to adapt the suffix array to apply to trees, and find a way to construct it in $\sortt$ time, which we have not looked at.
-Instead we can preprocess the fingerprints of the tree with Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}.
-
-\begin{theorem}
-\tfprintk, where $k$ is a parameter $1\leq k\leq\logceil$, can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(k\cdot n^{1/k})$ query time, $O(k\cdot n)$ space, and $O(n\log n)$ preprocessing time.
-\end{theorem}
-
-\subsubsection{Data Structure}
-
-We define a natural number $F_t[u]$ to be the fingerprint for node $u$ of length $t$.
-For two nodes $u$ and $w$ in a tree $T$, their fingerprints $F_t[u]$ and $F_t[w]$ of the same length $t$ are the same if and only if the last $t$ characters on the path from the root of $T$ to $u$ is the same as the last $t$ characters on the path from the root of $T$ to $w$.
-
-For a given parameter $k$, where $1\leq k \leq \logceil$, we define the \tfprintk\ data structure as follows: For each number $\ell$ between $0$ and $k-1$, let $t_\ell=\Theta(n^{\ell/k})$, and let $t_0 = 1$.
-For each number $\ell$ between $0$ and $k-1$ and each node $u$ in $T$, whose depth is at least $t_\ell$, let $H_\ell[u] = F_{t_\ell}[u]$
-
-\begin{lemma}
-The \tfprintk\ data structure takes $O(k\cdot n)$ space.
-\end{lemma}
-\begin{proof}
-The data structure contains $k$ numbers $t_\ell$ and $k$ tables of fingerprints $H_\ell$, and each table contains fingerprints for at most all of the $n$ nodes in the tree.
-\end{proof}
-
-\subsubsection{Query}
-
-The query for \tfprintk\ is nearly the same as the query for \fprintk\ on strings, except we replace $H_\ell[i+v]$ and $H_\ell[j+v]$ with $H_\ell[p_i[v+t_\ell]]$ and $H_\ell[p_j[v+t_\ell]]$.
-We also decrement the level $\ell$ whenever we reach $v+t_\ell=e_i$ or $v+t_\ell=e_j$, since we no longer have a special character \$ to tell when we have reached the end.
-
-In the \tfprintk\ query, start with $v=0$ and $\ell=0$, then do the following steps:
-\begin{samepage}
-\begin{enumerate}
-\item As long as $v+t_\ell\leq e_i\lor v+t_\ell\leq e_j$ and $H_\ell[p_i[v+t_\ell]] = H_\ell[p_j[v+t_\ell]]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless $\ell = k-1$.
-\item As long as  $v+t_\ell\leq e_i\lor v+t_\ell\leq e_j$ and $H_\ell[p_i[v+t_\ell]] = H_\ell[p_j[v+t_\ell]]$, increment $v$ by $t_\ell$ and repeat this step.
-\item Stop and return $v$ when $\ell = 0$, otherwise decrement $\ell$ by one and go to step two.
-\end{enumerate}
-\end{samepage}
-
-\begin{lemma}
-The \tfprintk\ query finds the correct tree LCE value.
-\end{lemma}
-\begin{proof}
-The query algorithm keeps the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$.
-At any given step at level $\ell$, we check if $H_\ell[p_i[v+t_\ell]]=H_\ell[p_j[v+t_\ell]]$, which by the definition of our data structure is equivalent to $\zlabel(p_i[v\twodots v+t_\ell])=\zlabel(p_j[v\twodots v+t_\ell])$.
-If the equality holds, we know that $\zlabel(p_i[0\twodots v+t_\ell])=\zlabel(p_j[0\twodots v+t_\ell])$, and we maintain our invariant when we increment $v$ by $t_\ell$.
-
-We stop when $v+1>e_i$, $v+1>e_j$, or $H_0[p_i[v+1]] \neq H_0[p_j[v+1]]$, at which point $v=\LCE(p_i,p_j)$.
-\end{proof}
-
-\begin{lemma}
-The \tfprintk\ query takes $O(k\cdot n^{1/k})$ time.
-\end{lemma}
-\begin{proof}
-We have changed nothing from the \fprintk\ query on strings, which negatively affects the query time, so the query time is the same.
-\end{proof}
-
-\subsubsection{Preprocessing}
-
-We use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to preprocess the \tfprintk\ data structure. When $t_\ell$ is $t_{\ell-1} < t_\ell \leq 2t_{\ell-1}$, we preprocess level $\ell$ like this:
-\begin{enumerate}
-\item Make pairs of fingerprints $H_\ell[u] = (H_{\ell-1}[LA(d(u)-t_\ell+t_{\ell-1},u)], H_{\ell-1}[u])$ for all nodes $u$ in $T$, where $d(u) \geq t_\ell$.
-\item Sort the pairs lexicographically.
-\item Replace each pair with a new fingerprint, by going through them in lexicographical order, while updating the fingerprint whenever the pair changes.
-\end{enumerate}
-
-Repeat the steps above for each level, starting with $\ell = 1$ and ending with $\ell = k-1$.
-When $k < \logceil$, the condition that $t_\ell \leq 2t_{\ell-1}$ is not satisfied.
-In this case we need to add extra levels in the preprocessing, which we will only store temporarily.
-We therefore always need to preprocess $O(\log n)$ levels, even if we only store some of them.
-
-\begin{lemma}
-The KMR algorithm for preprocessing \tfprintk\ gives the correct data structure.
-\end{lemma}
-\begin{proof}
-Two strings $S$ and $T$ of length $a$ are the same if and only if we can split each of them into two smaller substrings $S = S_1\cdot S_2$ and $T=T_1\cdot T_2$ such that $S_1=T_1$ and $S_2=T_2$, where "$\cdot$" is string concatenation.
-Therefore we can assign fingerprints $F_a[S]=F_b[S_1]\cdot F_c[S_2]$ and $F_a[T]=F_b[T_1]\cdot F_c[T_2]$ in step one of the preprocessing algorithm.
-The same also holds when the two smaller substrings overlap.
-
-The fingerprints from level $\ell-1$ are natural numbers, less than $n$.
-Step one of the preprocessing algorithm makes each fingerprint at level $\ell$ a natural number less than $n^2$.
-Step two and three of the preprocessing algorithm reduces that universe of fingerprints back to $n$ without changing the equality of the fingerprints.
-\end{proof}
-
-\begin{lemma}
-The KMR algorithm for preprocessing \tfprintk\ takes $O(n\log n)$ time.
-\end{lemma}
-\begin{proof}
-Step one and three use linear time.
-In step two we can use radix sort, which uses linear time, because our universe of elements to sort is $n^2$.
-The total preprocessing time becomes $O(n\log n)$ for preprocessing $\log n$ levels in $O(n)$ time each.
-\end{proof}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using Constant Time String LCE in Heavy Paths}
-
-To answer LCE queries on a tree, we can create a heavy path decomposition of the tree and create a string of the labels of all the heavy paths concatenated.
-We can then use constant time string LCE on this string.
-We call this algorithm \proc{HeavyPathLce}.
-
-\begin{theorem}
-\proc{HeavyPathLce} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
-\end{theorem}
-
-\begin{samepage}
-\subsubsection{Data Structure}
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.5\textwidth,page=4]{tree-fingerprint.pdf}
-    \end{center}
-    \caption{\label{fig:tree-heavy-ds}A heavy path decomposition of a tree and a corresponding \proc{HeavyPathLce} string $s$.}
-\end{figure}
-
-The \proc{HeavyPathLce} data structure consists of the following:
-\begin{itemize}
-\item An input tree $T$ with a heavy path decomposition.
-\item A string $s$, which is a concatenation of strings of all heavy paths of $T$, where the string of a heavy path is the label of the light edge leading to the heavy path concatenated with the label of the heavy path.
-\pref{fig:tree-heavy-ds} shows an example of $s$.
-\item For a node $u$, $\zend(u)$ points to the node of greatest depth in $u$'s heavy path, $d(u)$ is the depth of $u$, and $\zindex(u)$ is the index in $s$, which stores $\zlabel(u)$.
-\item Data structures for constant time NCA on $T$, constant time LA on $T$, and constant time LCE on $s$.
-\end{itemize}
-\end{samepage}
-
-\begin{lemma}
-The \proc{HeavyPathLce} data structure takes $O(n)$ space.
-\end{lemma}
-\begin{proof}
-The string $s$ contains $n-1$ characters, one for each edge in $T$. The space usage for NCA, LA, LCE, and the extra information about each node (\zend, \zindex, and $d$) is all $O(n)$.
-\end{proof}
-
-\begin{lemma}
-We can preprocess the \proc{HeavyPathLce} data structure in $O(\sortt)$ time.
-\end{lemma}
-\begin{proof}
-Preprocessing LCE on $s$ using \proc{SuffixNca} or \proc{LcpRmq} takes $O(\sortt)$ time.
-If we walk through the tree and for each leaf append the heavy path string ending in that leaf to $s$, we can construct $s$ in $O(n)$ time.
-The rest of the data structures can be preprocessed in $O(n)$ time.
-\end{proof}
-
-\subsubsection{Query}
-
-In a \proc{HeavyPathLce} query $\LCE_T(p_i,p_j)$ we start with $v=0$ and do the following repeatedly:
-\begin{enumerate}
-\item Stop when we reach either $v=e_i$ or $v=e_j$, and return $v$.
-\item Find $m_i = \zlength(p_i[v], \NCA_T(\zend(p_i[v+1]), p_i[e_i]))$ \\ and $i=\zindex(p_i[v+1])$.
-\item Find $m_j = \zlength(p_j[v], \NCA_T(\zend(p_j[v+1]), p_j[e_j]))$ \\ and $j=\zindex(p_j[v+1])$.
-\item Find $v'=\min\{\LCE_s(i,j),m_i,m_j\}$.
-\item Increment $v$ by $v'$.
-\item Stop if $v'<\min\{m_i,m_j\}$ and return $v$.
-\end{enumerate}
-
-\begin{lemma}
-The \proc{HeavyPathLce} query finds the correct LCE value.
-\end{lemma}
-\begin{proof}
-In step two, $m_i$ is the number of edges on a single heavy path at the beginning of $p_i[v\twodots e_i]$, like illustrated in \pref{fig:tree-heavy-query}.
-The labels of these edges are stored consecutively in $s$, such that $\zlabel(p_i[v\twodots v+m_i])=s[i\twodots i + m_i-1]$.
-The same holds for $m_j$ in step three.
-
-The algorithm keeps the invariant $\zlabel(p_i[0\twodots v]) = \zlabel(p_j[0\twodots v])$.
-At each iteration we find $v'$, which is the length of the longest common prefix of $\zlabel(p_i[v\twodots v+m_i])$ and $\zlabel(p_j[v\twodots v+m_j])$.
-This gives $\zlabel(p_i[0\twodots v+v']) = \\\zlabel(p_j[0\twodots v+v'])$, and we therefore keep the invariant when we add $v'$ to $v$ in step five.
-
-If $v'<\min\{m_i,m_j\}$ in step six, we also know that $\zlabel(p_i[v+1]) \neq \zlabel(p_j[v+1])$, and therefore that $v=\LCE(p_i,p_j)$.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.7\textwidth,page=2]{tree-fingerprint.pdf}
-    \end{center}
-    \caption{\label{fig:tree-heavy-query}How we find $m_i$ and $m_j$ in a \proc{HeavyPathLce} query.}
-\end{figure}
-\end{proof}
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.25\textwidth,page=3]{tree-fingerprint.pdf}
-    \end{center}
-    \caption{\label{fig:tree-heavy-query2}A \proc{HeavyPathLce} query with at most five iterations. Heavy edges on $p_i$ and $p_j$ are marked bold.}
-\end{figure}
-
-\begin{lemma}
-The \proc{HeavyPathLce} query takes $O(\log n)$ time.
-\end{lemma}
-\begin{proof}
-Since there is at most $O(\log n)$ light edges in any simple path in $T$, the two paths $p_i$ and $p_j$ contains at most $O(\log n)$ light edges in total.
-Since and each iteration of the query takes constant time, and we reach reach a new light edge of either $p_i$ or $p_j$ in each iteration, as shown in \pref{fig:tree-heavy-query2}, the query time is $O(\log n)$.
-\end{proof}
-
-\begin{comment}
-LCE(u_i,w_i,u_j,w_j)
-    v_i = u_i
-    v_j = u_j
-    v = 0
-    while v_i != w_i && v_j != w_j
-        Let m_i be the number of initial heavy edges on the path from v_i to w_i before any light edge.
-        Let m_j be ...
-        //m_i and m_j can be calculated in constant time.
-        Let m = min(m_i, m_j)
-        if m = 0
-            if the two next characters are the same
-                advance one character
-            else
-                stop
-        else
-            v=LCE_s(index(v_i),index(v_j))
-            If v >= m
-                advance m characters using Level Ancestor
-            else
-                advance v characters
-                stop
-
-    O(log n) time, one step for each light edge encountered on each of the two paths.
-\end{comment}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Directly Mapping \proc{SuffixNca} and \proc{LcpRmq}}
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.25\textwidth]{quadratic-tree.pdf}
-    \end{center}
-    \caption{\label{fig:quadratic-tree}A tree with $\Omega(n^2)$ ways to choose a leaf and an ancestor of that leaf.}
-\end{figure}
-
-We have looked at applying \proc{SuffixNca} and \proc{LcpRmq} directly to trees like we did with \proc{DirectComp} and \fprintk, to get $O(1)$ query time.
-
-\begin{theorem}
-\proc{TreeLcpRmq} and \proc{TreeSuffixNca} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(1)$ query time and $O(n^2)$ space.
-\end{theorem}
-\begin{proof}
-\proc{LcpRmq} relies on accessing two suffixes in SA, $i$ and $j$.
-In \proc{TreeLcpRmq} we change the two suffixes to paths $p_i=(u_i,w_i)$ and $p_j=(u_j,w_j)$, and each of these can be chosen in $O(n^2)$ ways.
-We can restrict $w_i$ and $w_j$ to be leafs and still be able to answer the original LCE query in $O(1)$ time, but even with this modification, a tree like the one shown in \pref{fig:quadratic-tree} still has $\Omega(n^2)$ ways to choose each path.
-The SA and LCP arrays in \proc{TreeLcpRmq} would therefore be of length $O(n^2)$.
-When we look at \proc{TreeSuffixNca}, the same argument holds for the necessary number of leafs in the suffix tree.
-\end{proof}
-
-Because of the quadratic space requirement, this result is not interesting when we use LCE on a tree to answer LCE queries on a compressed string in \pref{sec:lz-other}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Looking at Level Ancestor}
-
-The constant time solution for the Level Ancestor problem by Bendera~et~al.~\cite{level-ancestor} combines two $O(\log n)$ time solutions to get constant query times in linear space.
-We have looked at this algorithm to see if we can use similar techniques to achieve constant time LCE on trees in linear space.
-The two solutions used by the Level Ancestors algorithm are Jump Pointers and Ladder Decomposition.
-For the query $v=\LA(\ell,u)$, first use jump pointers to jump from $u$ half way to $v$ in constant time, and then use ladder decomposition to jump from the half way node to $v$.
-
-\proc{LadderLce}, which we describe in the next section, finds the LCE value of the first half of $p_i$ and $p_j$ in constant time, similar to how the ladder decomposition is used to jump from the half way node to $v$ in the Level Ancestor algorithm.
-We have not found a constant time solution to find the LCE value of the second half of these paths, in the same way jump pointers is used to jump from node $u$ to the half way node in the Level Ancestor algorithm.
-We have thus not found a constant time linear space solution for LCE on trees.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using Constant Time String LCE in a Ladder Decomposition}
-
-We can use the ladder decomposition in the same way we use the heavy tree decomposition to get an algorithm for LCE on trees, which we call \proc{LadderLce}, and which has the same properties as \proc{HeavyPathLce}.
-
-\begin{theorem}
-\proc{LadderLce} can preprocess a static tree of $n$ nodes to answer LCE queries as defined in \pref{sec:tree-def} using $O(\log n)$ query time, $O(n)$ space, and $O(\sortt)$ preprocessing time.
-\end{theorem}
-
-\subsubsection{Data Structure}
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.4\textwidth,page=1]{ladder.pdf}
-        \includegraphics[width=0.4\textwidth,page=2]{ladder.pdf}
-    \end{center}
-    \caption{\label{fig:ladder-ds} Ladders of a tree before and after their size is doubled.
-A \proc{LadderLce} string $s$ for this ladder decomposition is $s=abc|ade|df|ghijkl|ghno|jm$, where the splits between ladders are marked with a vertical line.}
-\end{figure}
-
-The \proc{LadderLce} data structure consists of the following:
-\begin{itemize}
-\item An input tree $T$ with a ladder decomposition.
-\item A strings $s$, which is a concatenation of ladder labels of each ladder in $T$, where the label of a ladder is the concatenation of labels on each edge leading to a node in the ladder, ordered ascending by the depth of the nodes.
-\pref{fig:ladder-ds} shows an example of $s$.
-\item For a node $u$, $\zindex(u)$ is an index in $s$ within the ladder label of $u$'s ladder, where $\zlabel(u)$ is stored.
-\item Data structures for constant time LA on $T$ and constant time LCE on $s$.
-\end{itemize}
-
-\begin{lemma}
-The \proc{LadderLce} data structure takes $O(n)$ space.
-\end{lemma}
-\begin{proof}
-The string $s$ contains at most $2n$ characters. The rest of the data structures take $O(n)$ space.
-\end{proof}
-
-\begin{lemma}
-We can preprocess the \proc{LadderLce} data structure in $O(\sortt)$ time.
-\end{lemma}
-\begin{proof}
-Constructing $s$ from the ladders takes $O(n)$ time, and preprocessing it for constant time LCE takes $O(\sortt)$ time. The rest of the data structures have\\ $O(n)$ preprocessing times.
-\end{proof}
-
-\subsubsection{Query}
-
-The \proc{LadderLce} query starts with $v=0$ and repeats the following:
-\begin{enumerate}
-\item Stop when $v = e_i$ or $v = e_j$ and return $v$.
-\item Find $m_i = \lceil\zlength(p_i[v\twodots e_i])/2\rceil$ and $i = \zindex(p_i[v+m_i]) - m_i + 1$.
-\item Find $m_j = \lceil\zlength(p_j[v\twodots e_j])/2\rceil$ and $j = \zindex(p_j[v+m_j]) - m_j + 1$.
-\item Find $v' = \min\{LCE_s(i,j),m_i,m_j\}$
-\item Increment $v$ by $v'$.
-\item Stop if $v' < \min\{m_i,m_j\}$ and return $v$.
-\end{enumerate}
-
-\begin{lemma}
-The \proc{LadderLce} query finds the correct LCE value.
-\end{lemma}
-\begin{proof}
-In step two, $\zlength(p_i[v\twodots v+m_i]) +1 \leq \zlength(p_i[v+m_i\twodots e_i])\leq h(p_i[v+m_i])$, where $h(u)$ is the height of node $u$.
-For a node $u$ of height $h$, the ladder decomposition guarantees that $u$'s ancestor with distance $h+1$ to $u$ is in $u$'s ladder.
-Therefore, $p_i[v+1]$ is in $p_i[v+m_i]$'s ladder, and $s[i\twodots i+m_i-1] = \zlabel(p_i[v\twodots v+m_i])$.
-Similarly in step three, $s[j\twodots j+m_i-1]\\ = \zlabel(p_j[v\twodots v+m_j])$.
-
-By the same argument we use for \proc{HeavyPathLce}, the algorithm maintains the invariant $\zlabel(p_i[0\twodots v])=\zlabel(p_j[0\twodots v])$, and gives the correct LCE value.
-\end{proof}
-
-\begin{lemma}
-The \proc{LadderLce} query takes $O(\log n)$ time.
-\end{lemma}
-\begin{proof}
-Each iteration takes constant time, and in each iteration we halve either $\zlength(d_i[v\twodots e_i])$ or $\zlength(d_j[v\twodots e_j])$.
-\end{proof}
-
-\proc{LadderLce} is very similar to \proc{HeavyPathLce}, however it has an extra property, which \proc{HeavyPathLce} does not have: If $\LCE(p_i,p_j)$ is less than half the length of $p_i$ and less than half the length of $p_j$, the LCE query takes constant time, since it only needs to look at one ladder.
-
 \fi % report
@@
 
-\bibitem{predecessor} Dan E. Willard. Log-logarithmic worst-case range queries are possible in space $\Theta$(N). Information Processing Letters, Volume 17, Issue 2, 24 August 1983, pages 81-84.
-
-\bibitem{level-ancestor} Michael A. Bendera, Martín Farach-Colton. The Level Ancestor Problem simplified. Theoretical Computer Science 321, 2004, pages 5-12.
+%\bibitem{predecessor} Dan E. Willard. Log-logarithmic worst-case range queries are possible in space $\Theta$(N). Information Processing Letters, Volume 17, Issue 2, 24 August 1983, pages 81-84.
+
+%\bibitem{level-ancestor} Michael A. Bendera, Martín Farach-Colton. The Level Ancestor Problem simplified. Theoretical Computer Science 321, 2004, pages 5-12.
 

### 2011-08-19 presentation/presentation.tex
m

@@
         Average query & $O(1)$ \\
-        Query I/Os & $O\Big(\frac{|\LCE(i,j)|}{B}\Big)$ \\
+        Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big)$ \\
     \end{tabular}
@@
             Average query & $O(1)$ \\
-            Query I/Os & $O(1)$ \\
+            Query I/O & $O(1)$ \\
         \end{tabular}
@@
             \item Size: $|H_\ell| = n$
-            \item I/Os: $O\big(k\cdot n^{1/k}\big)$
+            \item I/O: $O\big(k\cdot n^{1/k}\big)$
         \end{itemize}
@@
             \item Size: $|H_\ell| = n+t_\ell$
-            \item I/Os: $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$
+            \item I/O: $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$
             \begin{itemize}
@@
         Average query & $O(1)$ \\
-        Query I/Os & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
+        Query I/O & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
     \end{tabular}
@@
             Average query & $O(1)$ \\
-            Query I/Os & $O\Big(\frac{\sqrt n}{B}\Big)$ \\
+            Query I/O & $O\Big(\frac{\sqrt n}{B}\Big)$ \\
         \end{tabular}
@@
             Average query & $O(1)$ \\
-            Query I/Os & $O(\log n)$ \\
+            Query I/O & $O(\log n)$ \\
         \end{tabular}
@@
 
-\subsection{Practical I/O}
+%\subsection{Practical I/O}
 \begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
@@
     \end{multicols}
-\end {frame}
+\end{frame}
 
-\subsection{Practical I/O}
 \begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
@@
     \end{multicols}
-\end {frame}
+\end{frame}
+
+\section{LCE on trees}
+\subsection{Compression}
+\begin{frame}{LCE on Compressed Strings}
+    Goal
+    \begin{itemize}
+        \item Allow LCE queries without decompressing the string
+        \item Using Ziv-Lempel compression (LZ)
+    \end{itemize}
+    How
+    \begin{itemize}
+        \item LZ compression represents the string as a tree
+        \item An LCE query on a LZ compressed string is a number of LCE queries on a tree
+    \end{itemize}
+\end{frame}
+
+\begin{frame}{LCE on Trees}
+    \begin{itemize}
+        \item Trees:
+        \begin{itemize}
+            \item One character on each edge
+            \item LCE is the length of the longest common prefix of two strings along two paths
+        \end{itemize}
+        \item $p_i=bbc$ and $p_j=bbba$ gives $\LCE(p_i,p_j)=2$
+    \end{itemize}
+    \begin{center}
+        \includegraphics[width=0.4\textwidth,page=1]{tree-lce.pdf}
+    \end{center}
+\end{frame}
 

### 2011-08-19 presentation/presentation.tex
m

@@
     \begin{center}
-        \includegraphics[width=0.98\textwidth,page=1]{../doc/fingerprint.pdf}
+        \includegraphics[width=1\textwidth,page=1]{../doc/fingerprint.pdf}
     \end{center}
@@
     \begin{center}
-        \includegraphics[width=0.8\textwidth,page=2]{../doc/fingerprint.pdf}\\
-         $\LCE(3,12)$
+        \includegraphics[width=1\textwidth,page=2]{../doc/fingerprint.pdf}\\
+         $\LCE(3,12)=9$
     \end{center}
@@
 
+\subsection{Constant Time String LCE on Heavy Paths}
+\begin{frame}{Constant Time String LCE on Heavy Paths}
+    \begin{multicols}{2}{
+        \begin{itemize}
+            \item Data structure:
+            \begin{itemize}
+                \item Construct a heavy tree decomposition
+                \item For each heavy path, store characters as a substring
+            \end{itemize}
+            \item Query:
+            \begin{itemize}
+                \item Use constant time string LCE on each heavy path
+            \end{itemize}
+        \end{itemize}
+        \begin{center}
+            \includegraphics[width=0.44\textwidth,page=4]{../doc/tree-fingerprint.pdf}
+        \end{center}
+        \begin{itemize}
+            \item How to find the indexes $(i,j)$ in the string:
+            \begin{itemize}
+                \item Store an index at each node
+            \end{itemize}
+            \item How to know then the heavy path splits from the queried path:
+            \begin{itemize}
+                \item Store a pointer to the end of the heavy path at each node
+                \item Find NCA of the end of the heavy path and the end of the queried path
+            \end{itemize}
+        \end{itemize}
+        \begin{tabular}{r l}
+            Preprocessing & $O(\sortt)$ \\
+            Space & $O(n)$\\
+            Query & $O(\log n)$ \\
+        \end{tabular}
+    }
+    \end{multicols}
+\end{frame}
+
 \end{document}

### 2011-08-20 presentation/presentation.tex
m

@@
         \end{center}
+        \begin{tabular}{r l}
+            Preprocessing & $O(\sortt)$ \\
+            Space & $O(n)$\\
+            Query & $O(\log n)$ \\
+        \end{tabular}
+    }
+    \end{multicols}
+\end{frame}
+
+\begin{frame}{Constant Time String LCE on Heavy Paths}
+    \begin{columns}[t]
+        \begin{column}{0.7\textwidth}
             \begin{itemize}
@@
                 \end{itemize}
+                \item How to find a node on the queried path:
+                \begin{itemize}
+                    \item Use level ancestor
                 \end{itemize}
+            \end{itemize}
+        \end{column}
+        \begin{column}{0.3\textwidth}
+        \hfill\\
+            \includegraphics[width=1\textwidth,page=3]{../doc/tree-fingerprint.pdf}
+        \end{column}
+    \end{columns}
+\end{frame}
+
+\section{Summary}
+\begin{frame}{Summary}
+    \begin{tabular}{r l l l}
+        & \proc{Direct-} & \proc{LcpRmq} / & \\
+        & \proc{Comp} & \proc{SuffixNca} & \fprintk \\
+        Preprocessing & $O(1)$ & $O(\sortt)$ & $O(k\cdot n+\sortt)$ \\
+        Space & $O(1)$ & $O(n)$ & $O(k\cdot n)$\\
+        Query & $O(n)$ & $O(1)$ & $O(k\cdot n^{1/k})$ \\
+        Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
+        Query I/O & $O(n/B)$ & $O(1)$ & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
+    \end{tabular}
+    \begin{itemize}
+        \item In practice, the \fprintk\ algorithm is...
+        \begin{itemize}
+            \item ...almost as good as \proc{DirectComp} and significantly better than \proc{LcpRmq} in average case
+            \item ...significantly better than \proc{DirectComp} but worse than \proc{LcpRmq} in worst case
+        \end{itemize}
+        \item We can generalize LCE to use on trees:
         \begin{tabular}{r l}
@@
         \end{tabular}
-    }
-    \end{multicols}
-\end{frame}
+    \end{itemize}
+\end {frame}
 

### 2011-08-20 presentation/presentation.tex
m

@@
             \put(180, 30){\framebox(20, 20){a}}
-            \uncover<2-2>{
-                \put(200, 30){\makebox(100, 20){1 match}}
-                \put(80,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
+            \uncover<1-2>{
                 \put(90, 15){\vector(0,1){13}}
                 \put(80, 0){\makebox(20, 15){4}}
-                \put(120,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
                 \put(130, 15){\vector(0,1){13}}
@@
             }
-            \uncover<3->{
-                \put(200, 30){\makebox(100, 20){2 matches}}
+            \uncover<2-2>{
+                \put(80,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
+                \put(120,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
             }
-            \uncover<3-3>{
-                \put(100,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
+            \uncover<2-3>{
+                \put(200, 30){\makebox(100, 20){1 match}}
+            }
+            \uncover<3-4>{
                 \put(110, 15){\vector(0,1){13}}
                 \put(100, 0){\makebox(20, 15){5}}
-                \put(140,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
                 \put(150, 15){\vector(0,1){13}}
@@
             \uncover<4-4>{
-                \put(120,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){a}}}}
+                \put(100,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
+                \put(140,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
+            }
+            \uncover<4->{
+                \put(200, 30){\makebox(100, 20){2 matches}}
+            }
+            \uncover<5-6>{
                 \put(130, 15){\vector(0,1){13}}
                 \put(120, 0){\makebox(20, 15){6}}
-                \put(160,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){b}}}}
                 \put(170, 15){\vector(0,1){13}}
@@
             }
+            \uncover<6-6>{
+                \put(120,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){a}}}}
+                \put(160,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){b}}}}
+            }
         \end{picture}
     \end{block}
-    \uncover<5->{
+    \uncover<7->{
         \begin{block}{Result}
             $LCE_s(4,6)=2$
+        \end{block}
+        \begin{block}{LCE problem}
+            Efficiently perform multiple queries $(i,j)$ on a static string $s$
         \end{block}
@@
         Average query & $O(1)$ \\
-        Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big)$ \\
+        Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big) = O\big(\frac{n}{B}\big)$ \\
     \end{tabular}
@@
      %$n+t_\ell-1/t_\ell+2$
+    \begin{center}
+        \includegraphics[width=0.7\textwidth,page=2]{../doc/fingerprint.pdf}\\
+    \end{center}
 \end{frame}
@@
 \subsection{Practical Results}
-\begin{frame}{The \fprintk\ Algorithm: Practical Results}
+\begin{frame}{Practical Results}
     Query times of
@@
 
-\begin{frame}{The \fprintk\ Algorithm: Practical Results}
+\begin{frame}{Practical Results}
     Query times of
@@
     by string length
-    \begin{multicols}{2}{
     \begin{center}
@@
         \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-repeat-pow.plt}
-            \newpage
-            Cache stress\\
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand2.plt}
     \end{center}
-    }
-    \end{multicols}
 \end {frame}
+
+\begin{frame}{Cache Optimization, Practical Results}
+    Is I/O optimization good in practice?
+    \begin{itemize}
+        \item Pro: better cache efficiency
+            \begin{itemize}
+                \item Best for small $k$, no change for $k=\logceil$
+            \end{itemize}
+        \item Con: Calculating memory addresses is more complicated
+            \begin{itemize}
+                \item $((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1$ vs.\ $i$
+            \end{itemize}
+    \end{itemize}
+\end{frame}
 
@@
 
+\begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
+    Query times of
+    \textcolor{red}{\proc{DirectComp}},
+    \textcolor{green}{\fprint[2] (cache opt.)},
+    \textcolor{blue}{\fprint[3] (not cache opt.)},
+    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
+    \textcolor{cyan}{\proc{LcpRmq}}
+    by string length
+    \begin{multicols}{2}{
+        \begin{center}
+            Average case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
+            \newpage
+            Cache stress\\
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand2.plt}
+        \end{center}
+    }
+    \end{multicols}
+\end {frame}
+
+\begin{comment}
 \section{LCE on trees}
@@
 \end{frame}
+\end{comment}
 
@@
         Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
-        Query I/O & $O(n/B)$ & $O(1)$ & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
+        Query I/O & $O\big(\frac{n}{B}\big)$ & $O(1)$ & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
     \end{tabular}
+    \vspace{1em}
     \begin{itemize}
@@
         \end{itemize}
-        \item We can generalize LCE to use on trees:
-        \begin{tabular}{r l}
-            Preprocessing & $O(\sortt)$ \\
-            Space & $O(n)$\\
-            Query & $O(\log n)$ \\
-        \end{tabular}
+        \item Cache optimization of \fprintk\ improves query times at $k=2$ and worsens query times at $k\geq 3$
+        %\item We can generalize LCE to use on trees:
+        %\begin{tabular}{r l}
+        %    Preprocessing & $O(\sortt)$ \\
+        %    Space & $O(n)$\\
+        %    Query & $O(\log n)$ \\
+        %\end{tabular}
     \end{itemize}

### 2011-08-22 article/article.tex
m

@@
 %\newpage
+
+\begin{abstract}
+The \emph{longest common extension} (LCE) problem is to preprocess a string in order to allow for a large number of LCE querie, such that the queries are efficient.
+The LCE value, $\LCE_s(i,j)$, is the length of the longest common prefix of the pair of suffixes starting at index $i$ and $j$ in the string $s$.
+The LCE problem can be solved in linear space with constant query time and a preprocessing of sorting complexity.
+There are two known approaches achieving these bounds, which use nearest common ancestors and range minimum queries respectively.
+However, in practice a much simpler approach with linear query time and no preprocessing achieves significantly better average case performance.
+We show a new algorithm, \fprintk\, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time.
+Though this solution is asymptotically strictly worse than the asymptotically best previously known algorithms, it outperforms them in practice in average case and is almost as fast as the simple linear time algorithm.
+On worst case input, this new algorithm is significantly faster in practice compared to the simple linear time algorithm.
+\end{abstract}
 

### 2011-08-23 article/article.tex
m

@@
-%- Efter "..constant query time and linear space" skal der være referencer.
-%  * Til hvem? Navarri refererer til konstanttids NCA og RMQ men ikke til en kilde der beskriver hvordan de bruges til at lave LCE. Johannes Fischer skriver bare "It is well-known".
-%- Er navnene på algoritmerne de samme som i de oprindelige artikler? I
-%  så fald, så drop "we call". Der skal selvfølgelig også være
-%  referencer i dette afsnit lige i starten af definitionen på hver
-%  algoritme/datastruktur. Skriv gerne navnene på forfatterne: "Navarro
-%  et al[?] gave an algorithm..." eller lign. 
-%  * DirectComp har jeg ikke selv navngivet. SuffixNca og LcpRmq navnene har jeg selv fundet på. De har ikke noget navn i de tidligere artikler.
-
 % hav plot for $r=\sqrt n$ med til forsvar
@@
 
-\author{Jesper Kristensen\\DTU Informatics\\Technical University of Denmark}
+\author{Philip Bille\\\texttt{phbi@imm.dtu.dk}\and Inte Li Gørtz\\\texttt{ilg@imm.dtu.dk}\and Jesper Kristensen\\\texttt{mail@jesperkristensen.dk}}
 
-\date{August 1, 2011}
+\date{Technical University of Denmark\\\hfill\\August 1, 2011}
 
@@
 There are two known approaches achieving these bounds, which use nearest common ancestors and range minimum queries respectively.
-However, in practice a much simpler approach with linear query time and no preprocessing achieves significantly better average case performance.
+However, in practice a much simpler approach with linear query time, no extra space and no preprocessing achieves significantly better average case performance.
 We show a new algorithm, \fprintk\, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time.
@@
 On worst case input, this new algorithm is significantly faster in practice compared to the simple linear time algorithm.
+We also look at cache performance of the new algorithm, and we show that optimizing for cache only improves practical query time for $k=2$.
 \end{abstract}
@@
 
-The \emph{longest common extension} (LCE) problem is to preprocess a string in order to allow for a large number of LCE querie, such that the queries are efficient.
+The \emph{longest common extension} (LCE) problem is to preprocess a string in order to allow for a large number of LCE queries, such that the queries are efficient.
 The LCE value, $\LCE_s(i,j)$, is the length of the longest common prefix of the pair of suffixes starting at index $i$ and $j$ in the string $s$.
 The LCE problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
-Solutions with linear space with constant query time exists for the problem~[TODO cite who].
+Solutions with linear space and constant query time exists for the problem~\cite{nca,jf-rmq}.
 These theoretically good solutions are however not the best in practice, since they have large constant factors for both time and space usage.
-In average cases, a much simpler solution with worst case linear time, average case constant time, and no preprocessing has significantly better practical performance~\cite{ilie-navarro-tinta}.
-This algorithm is ideal when only average case performance is relevant.
+A much simpler solution exists with average case constant time and no space required other than storing the input string.
+It has has significantly better practical performance for average case input as well as for average case queries on some real world strings, when compared to the asymptotically best known algorithms~\cite{ilie-navarro-tinta}.
+However this algorithm has linear worst case query time, and is thus only ideal when worst case performance is irrelevant.
 In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.
-Such a situation could be use of approximate string searching in a firewall, which should not allow an attacker to significantly degrade its performance by sending it carefully crafted packages, while the average case of scanning legitimate data must still be highly performant.
+Such a situation could be use of approximate string searching in a firewall.
+The firewall should not allow an attacker to significantly degrade its performance by sending it carefully crafted packages.
+At the same time it must scan legitimate data fast.
 
-In this paper we give a new algorithm, which is a generalization of the simple linear time algorithm, which is almost as simple, yet it has significantly better worst case query times.
-This algorithm also achieves significantly better average case practical query times compared to the theoretically best algorithms.
-The algorithm uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$.
-
-\ifreport
-
-In \pref{sec:algorithms} we theoretically describe and analyze the LCE algorithms introduced here in \pref{sec:intro}, and in \pref{sec:results} we analyze results from practical C++ implementations of these algorithms.
-
-We also look at how to apply the LCE problem to Ziv-Lempel compressed strings, such that we can answer LCE queries fast while only using space relative to the compressed size of the string.
-In \pref{sec:lz-compress} we describe LZ compression, and how we can convert the simple linear time LCE algorithm to work on compressed strings.
-It uses space linear to the size of the compressed string, and it has worst case query time linear to the size of the uncompressed string, and average case query time $O(\log\log N)$.
-In \pref{sec:tree-lce} we look at how we can use the tree structure of LZ compression to get better query times on compressed strings.
-
-\fi % report
+In this paper we give a new algorithm, which is a generalization of the simple linear time algorithm.
+The new algorithm is almost as simple, yet it has significantly better worst case query time.
+This algorithm also achieves significantly better average case practical query time compared to the asymptotically best known algorithms.
+The algorithm uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$, on a string of length $n$.
 
@@
 
-Ilie~et~al.~\cite{ilie-navarro-tinta} gave and algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and $O(|\LCE(i,j)|)$ query time.
+Ilie~et~al.~\cite{ilie-navarro-tinta} gave an algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and $O(|\LCE(i,j)|)$ query time.
 For a query $\LCE(i,j)$, the algorithm compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
@@
 The LCE problem can be solved with $O(1)$ worst case query time, using $O(n)$ space and $O(\sortt)$ preprocessing time. Two different ways of doing this exists.
-One method, \proc{SuffixNca}, uses constant time nearest common ancestor (NCA) queries on a suffix tree.
+One method, \proc{SuffixNca}, uses constant time nearest common ancestor (NCA) queries~\cite{nca} on a suffix tree.
 The LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$.
@@
 $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.
-The other method, \proc{LcpRmq}, uses constant time range minimum queries (RMQ) on a longest common prefix (LCP) array.
+The other method, \proc{LcpRmq}, uses constant time range minimum queries (RMQ)~\cite{jf-rmq} on a longest common prefix (LCP) array.
 The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in the suffix array (SA).
@@
 
-\ifreport
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Our Fingerprinting Algorithm}
-\fi % report
-\ifarticle
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Our Results}
-\fi % article
 
-We present a new LCE algorithm, \fprintk, where $k$ is a parameter $1\leq k\leq\logceil$, which describes the number of levels used\footnote{All logarithms are base two.}.
-This algorithm is based on string fingerprinting.
+We present a new LCE algorithm, \fprintk, based on multiple levels of string fingerprinting.
+The algorithm has a parameter $k$ in the range $1\leq k\leq\logceil$, which describes the number of levels used\footnote{All logarithms are base two.}.
 
@@
 
+For $k=2$ we can improve practical \fprintk\ query time even further by optimizing it for cache efficiency.
+However for $k>2$ the query time performance degrades, as the added overhead outweighs the improved cache efficiency.
+
 Our algorithm is fairly simple.
@@
 
-\paragraph{String notation.} Let $s$ be a string of length $n$.
+% \paragraph{String notation.}
+Let $s$ be a string of length $n$.
 Then $s[i]$ is the $i$'th character of $s$, and $s[i\twodots j]$ is a substring of $s$ containing characters $s[i]$ to $s[j]$, both inclusive.
@@
 The suffix of $s$ starting at index $i$ is written $\suff_i = s[i\twodots n]$.
+%\paragraph{Sorting complexity.}
+The time it takes to sort $n$ numbers within an universe of size $\sigma$ is written as $\sortt$.
 
-\ifreport
-\paragraph{Tree notation.}
-The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
-A path in a tree begins at node $u$ and ends at node $v$, where $u$ is an ancestor of $v$.
-The length of a path is the number of edges between nodes in the path.
-The depth of a node $u$ is the length of the path from the root to $u$, and the height of $u$ is the length of the longest path from $u$ to a descendant of $u$.
-E.g.
-the depth of the root and the height of a leaf is both zero.
-\fi % report
-
-\paragraph{Sorting complexity.} The time it takes to sort $n$ numbers within an universe of size $\sigma$ is written as $\sortt$.
-
-\paragraph{Suffix tree.} A suffix tree $\mathcal{T}$ encodes all suffixes of a string $s$ of length $n$ with alphabet $\sigma$.
+%\paragraph{Suffix tree.}
+A \emph{suffix tree} $\mathcal{T}$ encodes all suffixes of a string $s$ of length $n$ with alphabet $\sigma$.
 The tree has $n$ leaves named $L_1$ to $L_n$, one for each suffix of $s$.
@@
 
-\ifreport
-
-There are $n$ leaves in a suffix tree, and since each internal node has at least two children, the tree has less than $2n$ nodes.
-Each edge uses $O(1)$ space, since the label is a substring of $s$ and it can therefore be represented by two indexes of $s$.
-The suffix tree therefore takes $O(n)$ space.
-
-\fi % report
-
-\paragraph{SA and LCP.} For a string $s$ of length $n$ with alphabet size $\sigma$, the \emph{suffix array} (\SA) is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$.
+%\paragraph{SA and LCP.}
+For a string $s$ of length $n$ with alphabet size $\sigma$, the \emph{suffix array} (\SA) is an array of length $n$, which encodes the lexicographical ordering of all suffixes of $s$.
 The lexicographically smallest suffix is $\suff_{\SA[1]}$, the lexicographically largest suffix is $\suff_{\SA[n]}$, and the lexicographically $i$'th smallest suffix is $\suff_{\SA[i]}$.
@@
 The first element $\LCP[1]$ is always zero.
-
 Building the SA, \SAi\ and LCP arrays takes $O(\sortt)$ time~\cite{sort-complexity}.
 
-\paragraph{NCA.} The ancestors of a node $u$ is $u$ itself as well as any node, which is a parent of an ancestor of $u$.
+%\paragraph{NCA.}
 The \emph{nearest common ancestor} (NCA) of two nodes $u$ and $v$ is the node of greatest depth, which is an ancestor of both $u$ and $v$.
+The ancestors of a node $u$ includes $u$ itself.
 An NCA query can be answered in $O(1)$ time with $O(n)$ space and preprocessing time in a static tree with $n$ nodes~\cite{nca}.
 
-\paragraph{RMQ.} The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, i.e.\ $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
+%\paragraph{RMQ.}
+The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, i.e.\ $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
 A \emph{range minimum query} (RMQ) on a static array of $n$ elements can be answered in $O(1)$ time with $O(n)$ space and preprocessing time~\cite{jf-rmq}.
 
-\paragraph{I/O and Cache-Oblivious Model.}
-The I/O model describes the number of memory blocks an algorithm moves between two layers of a layered memory architecture, where the size of the internal memory layer is $M$ words, and data is moved between internal and external memory in blocks of $B$ words.
+%\paragraph{I/O and Cache-Oblivious Model.}
+The \emph{I/O model} describes the number of memory blocks an algorithm moves between two layers of a layered memory architecture, where the size of the internal memory layer is $M$ words, and data is moved between internal and external memory in blocks of $B$ words.
 In the \emph{cache-oblivious model}, the algorithm has no knowledge of the values of $M$ and $B$.
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{LCE Algorithms\label{sec:algorithms}}
-This section describes existing and new algorithms for solving the LCE problem.
-It contains a theoretical analysis of their asymptotic query times and space requirements, shown in summary in \pref{tab:algorithms}.
-\pref{sec:results} describes the results from experiments on practical implementations of these algorithms.
-
-\begin{table}[tp]
-\centering
-% \begin{tabular}{l|p{160px}|p{160px}}
-\begin{tabular}{l|l|l|l}
-\hline\hline
-Algorithm & Space & Query time & Preprocessing \\ [0.5ex] \hline
-\proc{SuffixNca} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
-\proc{LcpRmq} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
-\proc{DirectComp} & $O(1)$ & $O(n)$ & $O(1)$ \\ \hline
-\proc{DirectCompLookup}, $1 \leq t \leq n$ & $O(n^2/t)$ & $O(t)$ & $O(n^2)$ \\
-$t=1$ & $O(n^2)$ & $O(1)$ & $O(n^2)$ \\ \hline
-\fprintk, $1 \leq k \leq \logceil$ & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
-$k=\logceil$ & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline
-\end{tabular}
-\caption{LCE algorithms with their space requirements, worst case query times and preprocessing times. Average case query times are $O(1)$ for all shown algorithms.}\label{tab:algorithms}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Existing Algorithms}
-
-In this section we look at the existing LCE algorithms described in \pref{sec:existing-results} with a bit more details.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{\proc{DirectComp}}
-
-\proc{DirectComp} solves the LCE problem without any preprocessing in $O(|\LCE(i,j)|)$ query time.
-For a query $\LCE(i,j)$, the algorithm compares $s[i+v]$ to $s[j+v]$ in a loop, for $v$ starting at zero and incrementing, until the two characters differ, or the end of the string is reached, at which point the LCE value is $v$.
-\proc{DirectComp} does not need to check if the end of the string is reached, i.e.
-$i+v>n$ or $j+v>n$.
-Instead we can preprocess the string by adding a special character \$ to the end of the string, which does not occur anywhere else in the string.
-When the algorithm compares $s[n] = \$$ to any other character, it will stop, as the characters differ.
-This can give better practical performance compared to checking $i+v>n$ and $j+v>n$ in each iteration of the algorithm.
-The query time is $O(|\LCE(i,j)|)$, which in worst case is $O(n)$ on a string of length $n$, and in average case is $O(1)$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsubsection{\proc{SuffixNca}: NCA on a Suffix Tree}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{\proc{LcpRmq}: RMQs on a LCP Array}
-
-The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in SA.
-The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering.
-I.e.
-$\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
-An example is shown in \pref{fig:sa+lcp+min}.
-Preprocessing time is $O(\sortt)$, space requirement is $O(n)$ and query time is $O(1)$.
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.2\textwidth,page=1]{sa+lcp+min.pdf}
-    \end{center}
-    \caption{\label{fig:sa+lcp+min}For $s=abbababba$, $\LCE(2,3)$ is the longest common prefix of $bababba$ and $bbababba$.
-There are two other suffixes between these in the lexicographical ordering, and the longest common prefixes between each pair of neighbors are 3, 1 and 3.
-Therefore $\LCE(2,3)= \min ~ \{3, 1, 3\} = 1$.}
-\end{figure}
-
-Different algorithms exists for solving the RMQ problem, and we can solve it in $O(1)$ time and $O(n)$ space.
-But like for the LCE problem, the asymp-\\totically best RMQ solution may not be the best RMQ solution in practice.
-We examine the following RMQ solutions, where \RMQpq{p}{q} has $O(p)$ space and preprocessing time, and $O(q)$ query time:
-\begin{samepage}
-\begin{description}
-\item[\RMQn] walks through the array directly in $O(n)$ time and does not use any preprocessing (except for transforming LCE into RMQ).
-\item[\RMQq] preprocesses the array in $O(n)$ space in order to do queries in $O(1)$ time.
-This is a twolevel data structure as shown in \pref{fig:sa+lcp+min-twolevel}.
-The array is split into blocks of size $O(\log n)$.
-The top level is an array of length $O(n/\log n)$, where each element contains the minimum value of one block.
-The top level uses a \RMQpq{n\log n}{1}-solution, and the lower level precomputes all answers for each block, which is done in $O(n)$ space, since there are only a limited number of different block types of size $O(\log n)$.
-\item[\RMQlog] is the same as \RMQq\ except that the lower level is replaced with a direct search through a $O(\log n)$ length block.
-\end{description}
-\end{samepage}
-
-\begin{figure}[tp]
-    \begin{center}
-        \includegraphics[width=0.8\textwidth,page=2]{sa+lcp+min.pdf}
-    \end{center}
-    \caption{\label{fig:sa+lcp+min-twolevel}Twolevel Range Minimum Query. The range minimum is calculated as the minimum of one query in the top level and two queries in the lower level.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Combining \proc{DirectComp} and \proc{LcpRmq}\label{sec:dc-lcprmq}}
-
-Ilie~et~al.~\cite{ilie-navarro-tinta} suggests combining \proc{DirectComp} and \RMQn, by using \RMQn\ whenever the two suffixes we want to compare are close to each other in SA, and otherwise use \proc{DirectComp}.
-However that still gives $O(n)$ worst case query time.
-For example in the string $s=aaaaa...$, the query $\LCE(1,n/2)$ would require $\Omega(n)$ time for both \proc{DirectComp} and \RMQn.
-
-We can instead combine \proc{DirectComp} and \RMQq, which we call \proc{DirectComp+LcpRmq}.
-\proc{DirectComp+LcpRmq} first uses \proc{DirectComp} until a specific cutoff and then switches over to \RMQq\ instead.
-This could have both the good average case query time of \proc{DirectComp} and the good worst case query time of \proc{LcpRmq}.
-However it still keeps the large space requirements of \proc{LcpRmq}, and it is not simple to implement like \proc{DirectComp}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The \proc{DirectCompLookup} Algorithm}
-
-The input to the $LCE$ problem is two numbers $1 \leq i < j \leq n$ \footnote{We can assume that $i< j$ without loss of generality, since $\LCE(i,j)=\LCE(j,i)$}, so there are $n\cdot(n-1)/2 = O(n^2)$ different input possibilities in total.
-Because of this limited space of possible query inputs, we can tabulate a number of LCE values, and use those to speed up queries.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{\proc{DirectLookup}}
-
-We can store results for all $O(n^2)$ possible queries in $O(n^2)$ space, and a query will then be a single table lookup in $O(1)$ time.
-We call this algorithm \proc{DirectLookup}.
-A result space table for \proc{DirectLookup} is shown in \pref{fig:resultspace-empty}, and \pref{fig:resultspace-directcomp} shows how \proc{DirectComp} moves through that table.
-
-We can preprocess the table in in $O(n^2)$ time, since each value can be calculated in constant time like follows, if $LCE(i+1,j+1)$ is always calculated before $LCE(i,j)$ for any $i$ and $j$:
-\begin{align*}
-LCE(i,j) &=
-\begin{cases}
-    0 & \textrm{if} ~ s[i] \neq s[j] \\
-    1 & \textrm{if} ~ s[i] = s[j] \wedge (i = n \lor j = n) \\
-    LCE(i+1, j+1) + 1 & \textrm{otherwise} \\
-\end{cases}
-\end{align*}
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.42\textwidth,page=1]{resultspace.pdf}
-    \end{center}
-    \caption{\label{fig:resultspace-empty}The result space for $LCE_s(i,j)$ for $1 \leq i < j \leq n$ where $s = abacbbacbacacabc$.}
-\end{figure}
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.42\textwidth,page=2]{resultspace.pdf}
-    \end{center}
-    \caption{\label{fig:resultspace-directcomp}$LCE(3,7)$ using \proc{DirectComp}.}
-\end{figure}
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.42\textwidth,page=3]{resultspace.pdf}
-    \end{center}
-    \caption{\label{fig:resultspace-t-row}$LCE(3,7)$ with distance $t=4$ between each stored row.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{\proc{DirectComp} and \proc{DirectLookup} Combined}
-
-\proc{DirectComp} and \proc{DirectLookup} preprocesses and stores no results and all results respectively.
-If we only store some results, we can use \proc{DirectComp} until we reach a stored result, and then retrieve this result using \proc{DirectLookup}.
-If we store every $t$'th row, as shown in \pref{fig:resultspace-t-row}, the space usage will be $O(n^2/t)$ and the worst case query time will be $O(t)$.
-We call this algorithm \proc{DirectCompLookup}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{The \fprintk\ Algorithm}
-
 \fi % report
 
-\ifarticle
 \section{The \fprintk\ Algorithm\label{sec:fprintk}}
-\fi % article
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Cache Optimization\label{sec:fingerprint-cache}}
-
-%\paragraph{Horizontal, intra-level optimization}
-
-The amount of I/O used by \fprintk\ is $O(k\cdot n^{1/k})$.
-However if we structure our tables of fingerprints differently, we can improve the number of I/O operations to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
-Instead of storing the fingerprint of $s[i\twodots i+t_\ell-1]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$.
-This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.
-
-The size of each fingerprint table will grow from $|H_\ell| = n$ to $|H_\ell| = n+t_\ell$, because the rounding operations may introduce one-element gaps in the table after every $n/t_\ell$ elements.
-We achieve the greatest I/O improvement when $k$ is small.
-When $k=\logceil$, this cache optimization gives no difference in the amount of I/O.
-
-\ifreport
-
-\paragraph{Vertical, inter-level optimization}
-
-For larger values of $k$, we can attempt to optimize memory access across different levels.
-When we move down through the levels for $k=\logceil$, we can continue in two directions after reading $H_\ell[i]$: $H_{\ell-1}[i]$ or $H_{\ell-1}[i+t_\ell]$.
-Neither of these positions are next to $H_\ell[i]$ in memory.
-We can optimize for the case where we reach $H_{\ell-1}[i]$ by flipping the direction in which we store the fingerprints, such that all fingerprints $H_\ell[i]$ for all $\ell$ and a specific $i$ are stored next to each other.
-This will not improve the asymptotic I/O bound, but could improve practical query times.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Space Usage}
-
-\pref{tab:algorithms-space} shows the space requirement of each of our analyzed algorithms. The size of the input string of $n/4$ words is not included, and we ignore constant terms.
-
-\begin{table}[tp]
-\centering
-\begin{minipage}{0.7\textwidth}
-\centering
-\begin{tabular}{l|l}
-\hline\hline
-Algorithm & Space \\ [0.5ex] \hline
-\proc{DirectComp} & $0$ words \\ \hline
-\proc{DirectLookup} & $2n^2$ words \footnote{The factor two comes from using shift instead of multiplication, which may double the size depending on $n$.} \\ \hline
-\proc{DirectCompLookup} & $2n^2/t+n$ words \\ \hline
-\proc{LcpRmq} & ca.
-$5.6n$ words \footnote{Space use for \proc{LcpRmq} is $2n + 2nb + (bs \cdot (bs+1) / 2 + 1)\cdot C_{bs} + (nb+1)\cdot \lfloor\log nb\rfloor \approx 5.6n$, where $bs = \lceil\log n / 4\rceil$, $nb = (n-1)/bs+1$ and $C_i$ is the $i$'t Catalan Number.} \\ \hline
-\fprint[2] & $n$ words \\ \hline
-\fprint[3] & $2n$ words \\ \hline
-\fprint[\logceil] & $n\cdot\logceil$ words \\ \hline
-\end{tabular}
-\end{minipage}
-\caption{Space requirements for the LCE algorithms, where $n$ is the number of characters, a character is one byte and a word is four bytes.}\label{tab:algorithms-space}
-\end{table}
-
-\fi % report
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\ifarticle
-
 We implemented different variants of the \fprintk\ algorithm in C++ and compared them with optimized versions of the \proc{DirectComp} and \proc{LcpRmq} algorithms.
@@
 \end{description}
-
-\fi % article
-
-\ifreport
-
-We implemented different variants of \proc{DirectComp}, \proc{LcpRmq}, \proc{DirectCompLookup} and \fprintk\ in C++ and compared them to each other.
-We have not implemented \proc{SuffixNca}, since others have shown that \proc{LcpRmq} is better than \proc{SuffixNca} in practice~\cite{ilie-navarro-tinta}.
-The algorithms we compared are the following:
-\begin{description}
-\item[\proc{DirectComp}] is the simple \proc{DirectComp} algorithm with no preprocessing and worst case $O(n)$ query time.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is our new \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$.
-The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater wc] is a variant of our new \fprintk\ algorithm, which starts at level $\ell=k-1$ instead of $\ell=0$, as discussed in \pref{sec:improved-avg}.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac\textless$u_0$, ..., $u_{k-2}$\textgreater] is a variant of our new \fprintk\ algorithm, which stays at level $\ell$ until $v\geq u_\ell$, as discussed in \pref{sec:improved-avg}, instead of doing one comparison at each level in step one of the query in \pref{sec:fingerprint-query}.
-\item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
-\item[\proc{RMQ}\textless$p$, $q$\textgreater$_{\textrm{virtual}}$] is the \proc{LcpRmq} algorithm using different RMQ algorithms.
-They all contain an additional virtual method call to make implementation simpler, but their results cannot be directly compared to other algorithms, which do not have this extra virtual method call.
-\item[\proc{DirectComp}\textless$c$\textgreater\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{DirectComp+LcpRmq} algorithm, where the cutoff from \proc{DirectComp} to \proc{LcpRmq} is after $c$ iterations.
-\item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac$_{\textrm{cache}-\textit{dir}-\textit{op}}$] is a cache optimized variant of our \fprintk\ algorithm, where \textit{dir} describes which of our two cache optimized variants is used, and \textit{op} describes which of the shift or multiplication arithmetic operations it uses.
-\end{description}
-
-\fi % report
 
@@
 \item[Medium LCE value strings] These strings have an average LCE value over all $n^2$ query pairs of $n/2r$, where $r=0.73n^{0.42}$.
+This value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where the each of the values of $n$ were chosen to give the best result for \fprintk\ on a plot with the given constant value of $r$.
 We use results on these strings as an indication of query times somewhere between the average case and worst case.
@@
 
+Each measurement we make is an average of query times over a million random query pairs $(i, j)$.
+For a given string length and string type (worst case strings vs.\ average case strings), we use the same string and the same million query pairs on all tested algorithms.
+
+\begin{comment}
 For each kind of strings, we tested the algorithms using the following pattern:
@@
 \end{enumerate}
+\end{comment}
 
-\ifreport
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\proc{LcpRmq} Results\label{sec:test-rmq}}
-
-We want to compare our new algorithms against the algorithm which is the best in practice of the asymptotically best $O(1)$ algorithms.
-The \proc{LcpRmq} algorithm is better than \proc{SuffixNca}~\cite{ilie-navarro-tinta}, so we only look at \proc{LcpRmq}.
-We have looked at a number of RMQ algorithms to find out which works best for the LCE problem.
-The results are shown in \pref{fig:plot-rmq}.
-\RMQq\ is the best in practice, and we therefore use that when we test against other LCE algorithms.
-This result is different from that of Ilie~et~al.~\cite{ilie-navarro-tinta}, where \RMQpq{n}{4\log n} is the best.
-This could be due to hardware or compiler differences.
-
-The performance of the implementations shown in \pref{fig:plot-rmq} cannot be compared with those shown on other figures, because they all contain an additional, slow virtual method call, which gives them an unfair disadvantage.
-We have optimized the best variant of \proc{LcpRmq} to avoid the virtual method call, when we compare it against other algorithms.
-We compare this optimized version to \proc{DirectComp} in \pref{fig:plot-rmq-directcomp}, which shows that \proc{DirectComp} is significantly better in average case, though both have constant asymptotic average case query times.
-
-The RMQ algorithms show a significant jump in query times around $n=1.000.000$ on the plot with average case strings, but not on the plot with worst case strings.
-We have run the tests in Cachegrind, and found that the number of instructions executed and the number of data reads and writes are exactly the same for both average case strings and worst case strings.
-The cache miss rate for \RMQq\ on average case strings is 14\% and 9\% for the L1 and L2 caches, and on worst case strings the miss rate is 17\% and 13\%, which is the opposite of what could explain the jump we see in the plot.
-
-% size = 3,5*2^30 = 3758096384
-% assoc = (3,5*2^30)/(4*2^10) = 917504
-% line = 4*2^10 = 4096
-% --LL=3758096384,917504,4096
-
-% size = 1*2^30 = 1073741824
-% assoc = (1*2^30)/(4*2^10) = 262144
-% line = 4*2^10 = 4096
-% --LL=1073741824,262144,4096
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-rmq}Query times of \proc{LcpRmq} using different variants of RMQ.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{\proc{DirectComp+LcpRmq} Results}
-
-\pref{fig:plot-rmq-directcomp} also shows \proc{DirectComp+LcpRmq}, where we first try \proc{DirectComp}, and if we have not found the LCE value after one iteration, we use \proc{LcpRmq} instead.
-We see that \proc{DirectComp+LcpRmq} is worse than \proc{DirectComp} on average case strings.
-We could increase the cutoff from one iteration to something bigger, which might improve average case performance.
-That would however increase worst case time.
-None of the two plots we measure represents a good worst case for \proc{DirectComp+LcpRmq} compared to the other algorithms we test.
-The worst case for \proc{DirectComp+LcpRmq} relative to the other algorithms would be when the average LCE value is slightly larger than the cutoff, such that \proc{DirectComp+LcpRmq} reaches \proc{LcpRmq} in most queries, while other algorithms still work on LCE values as small as possible.
-As described in \pref{sec:dc-lcprmq}, \proc{DirectComp+LcpRmq} also uses a lot of space.
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-rmq-directcomp}Query times of constant time \proc{LcpRmq}, \proc{DirectComp} and \proc{DirectComp+LcpRmq}.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\proc{DirectCompLookup} Results}
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-complookup-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-complookup}Query times of our new DirectCompLookup algorithm versus the existing \proc{DirectComp} and \proc{LcpRmq} algorithms.}
-\end{figure}
-
-\pref{fig:plot-complookup} shows \proc{DirectCompLookup} with different time/space-tradeoffs together with \proc{LcpRmq}.
-In average case, \proc{DirectCompLookup} is slower than \proc{DirectComp}, most likely because it needs to do an extra check to find out if it has reached a position for which a result is stored.
-For short strings, \proc{DirectLookup} is faster than \proc{DirectCompLookup}, since it does not need to check if the value it needs is stored.
-However, it quickly grows to be as slow as \proc{LcpRmq}, while \proc{DirectCompLookup} does not experience this slowdown.
-We explain this with that \proc{DirectLookup} reads from its $O(n^2)$ table of results in every query, whereas \proc{DirectCompLookup} only reads from the $O(n)$ string in average case, and thus has a significantly smaller working set.
-
-In the worst case, the \proc{DirectCompLookup} variants which use more space are faster in practice, just like in theory.
-If we are willing to use $O(n\log n)$ space, \proc{DirectCompLookup}\textless$n/\log n$\textgreater\ will only halve the worst case query time compared to \proc{DirectComp}.
-This makes \proc{DirectCompLookup} better than \proc{DirectComp}, but still not a very good candidate when worst case query time is important.
-
-\proc{DirectCompLookup} uses $O(n^2)$ preprocessing time no matter what parameter $t$ we choose, which is significantly slower than any of the other algorithms we have tested, and it is also the reason why we have only tested this algorithm up until $n=100.000$.
-The only way we have found to improve this preprocessing time is to use one of the other algorithms we have covered to preprocess the data structure of \proc{DirectCompLookup}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{\fprintk\ Results}
-
-\fi % report
-
-\ifarticle
 \subsection{Results}
-\fi % article
 
@@
     \begin{center}
-        \includegraphics[width=0.7\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-rand10.plt}
-        \includegraphics[width=0.7\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-alla.plt}
-        \includegraphics[width=0.7\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-repeat-pow.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-alla.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-repeat-pow.plt}
     \end{center}
@@
     \end{center}
-    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without intra-level cache optimization.}
+    \caption{\label{fig:plot-fingerprint-cache-horiz}Query times of \fprint[2] with and without cache optimization.}
 \end{figure}
 
-\ifreport
-\begin{figure}[p]
+\begin{figure}[tp]
     \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-log-cache-alla.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint3-cache-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint3-cache-alla.plt}
     \end{center}
-    \caption{\label{fig:plot-fingerprint-cache-vert}Query times of \fprint[2] with and without inter-level cache optimization.}
+    \caption{\label{fig:plot-fingerprint3-cache}Query times of \fprint[3] with and without cache optimization.}
 \end{figure}
-\fi % report
 
@@
 
-\pref{fig:plot-fingerprint-cache-horiz} shows results of the cache optimization described in \pref{sec:fingerprint-cache}.
+The amount of I/O used by \fprintk\ is $O(k\cdot n^{1/k})$.
+However if we structure our tables of fingerprints differently, we can improve the number of I/O operations to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
+Instead of storing $F_{t_\ell}[i]$ at $H_\ell[i]$, we can store it at $H_\ell[((i-1)\mod t_\ell)\cdot\lceil n/t_\ell\rceil+\lfloor (i-1)/t_\ell\rfloor+1]$.
+This will group all used fingerprints at level $\ell$ next to each other in memory, such that the amount of I/O at each level is reduced from $O(n^{1/k})$ to $O(n^{1/k}/B)$.
+
+The size of each fingerprint table will grow from $|H_\ell| = n$ to $|H_\ell| = n+t_\ell$, because the rounding operations may introduce one-element gaps in the table after every $n/t_\ell$ elements.
+We achieve the greatest I/O improvement when $k$ is small.
+When $k=\logceil$, this cache optimization gives no difference in the amount of I/O.
+
 We have implemented two cache optimized variants.
-One as described in \pref{sec:fingerprint-cache}, and one where multiplication and division is replaced with shift operations.
+One as described above, and one where multiplication, division and modulo is replaced with shift operations.
 To use shift operations, $t_\ell$ and $\lceil n/t_\ell\rceil$ must both be powers of two.
@@
 
+\pref{fig:plot-fingerprint-cache-horiz} shows our measurements for \fprint[2].
 On average case strings the cache optimization does not change the query times, while on worst case strings and strings with medium size LCE values, cache optimization gives a noticeable improvement for large inputs.
-The cache optimized \fprint[3] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.
+The cache optimized \fprint[2] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.
 
 The last plot on \pref{fig:plot-fingerprint-cache-horiz} shows a variant of average case where the alphabet size is changed to two.
-This plot attempts to show the worst case for cache optimized \fprint[3].
+This plot attempts to show the worst case for cache optimized \fprint[2].
 LCE values in this plot are large enough to ensure that $H_1$ is used often, which should make the extra complexity of calculating indexes into $H_1$ visible.
 At the same time the LCE values are small enough to ensure, that the cache optimization has no effect.
-In this plot we see that the cache optimized variant of \fprint[3] has only slightly worse query time compared to the variant, which is not cache optimized.
+In this plot we see that the cache optimized variant of \fprint[2] has only slightly worse query time compared to the variant, which is not cache optimized.
 
-We conclude that this cache optimization does not visibly affect average case query times, while it improves worst case query times.
-However, due to late timing of this discovery, the cache optimized variant of \fprintk\ is not the main focus of our analysis.
-
-\ifreport
-
-\paragraph{Vertical, inter-level optimization}
-
-\pref{fig:plot-fingerprint-cache-vert} shows results of our inter-level cache optimization.
-We see no improvement on worst case strings, and we see increased query times on average case strings.
-We conclude that this attempt at cache optimization does not improve practical query times, which could be expected, since it does not improve the asymptotic I/O bound either.
-
-\fi % report
+\pref{fig:plot-fingerprint3-cache} shows our measurements for \fprint[3].
+Unlike for \fprint[2], we see a degraded performance on worst case strings compared to the not cache optimized variant.
+Our results thus show that cache optimization of \fprintk\ only gives a performance improvement for $k=2$.
 
@@
 
-% TODO \fixme{Results: Test Code: tell about c++ bug hunting in tinta, cachegrind code "Cache set count is not a power of two."}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Query time by LCE value}
-
-\begin{figure}[tp]
-    \begin{center}
-        %\includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/value-article-rand4.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/value-article-alla.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/value-article-repeat10-10.plt}
-    \end{center}
-    \caption{\label{fig:plot-by-value}Query times of some of our tested algorithms as a function of the returned LCE value.}
-\end{figure}
-
-As an alternative way of testing our algorithms, we have looked at how long a query takes in relation to the LCE value it returns, since most of our algorithms have query times, which depend on the returned LCE value.
-\pref{fig:plot-by-value} shows query times for a fixed string of 1000 characters.
-We can see how the query times of \fprint[2] and \fprint[3] increases as the LCE value increases, and drops when the LCE value becomes long enough to make one more step at the higher level with fingerprints of longer substrings.
-We also see that the query time of \fprint[\log n] increases whenever the LCE value becomes long enough to reach a higher level of the data structure.
-
-The plots in \pref{fig:plot-by-value} contains a lot of noise compared to our other plots.
-The number of measurements we make on these plots is many times bigger than the number of measurements we make on other plots.
-Therefore each measurement is an average over only 10000 queries instead of a million queries, which gives less accurate results.
-
-We need to measure a sum of many queries to get an accurate timing, since each individual query is too fast to measure.
-While measuring a sequence of randomly generated queries is a very artificial test, which is unlikely to occur in practice, we think that it is even more artificial to measure a sequence of queries for which we already know that the LCE value is the same each time.
-On our worst case string $s=aaaaaaa...$, the queries we measure in each point in \pref{fig:plot-by-value} have the pattern $\max(i,j)=v$, which may give very different performance characteristics compared to random queries due to cache effects.
-We cannot meaningfully measure performance by LCE value on average case strings, as these strings do not contain large LCE values.
-The second plot in \pref{fig:plot-by-value} is on a string, which attempts to avoid a strong pattern like $\max(i,j)=v$ while it still has large LCE values.
-
-Because of the amount of noise in the plots and the strong pattern of $\max(i,j)=v$, we have not focused on this way of measuring query time performance in our analysis.
-
-\begin{comment}
-\begin{algorithm}
-\caption{Algorithm for plotting query time as a function of string length\label{alg:graph-length}}
-\begin{codebox}
-\li \For $\id{type} \in \{\textrm{string of random characters with}~\sigma=10,~\textrm{string with}~\sigma = 1\}$
-    \Indentmore
-\li     \For $n = 1$ \To $1e7$ \Comment Double $n$ at each iteration
-        \Indentmore
-\li         $s = \proc{MakeString}_{\id{type}}(n)$
-\li         $\id{data} =$ a sequence of a million pairs $(i,j)$ where
-            \Indentmore
-\zi             $i\in[1\twodots n]$ and $j\in[1\twodots n]$ are chosen
-\zi             uniformly at random.
-            \End
-\li         \For each $\id{alg} \in \id{Algorithms}$
-            \Indentmore
-\li             $\id{alg}.\proc{preprocess}(s)$
-\li             start timing
-\li             \For each $(i,j) \in \id{data}$
-                \Indentmore
-\li                 $\id{alg}.\proc{query}(i,j)$
-                \End
-\li             end timing
-\li             \proc{Plot}$_{\id{type}}$($alg$, $n$, timing)
-\li             $\id{alg}.\proc{cleanup}()$
-            \End
-        \End
-    \End
-\end{codebox}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{Algorithm for plotting query time as a function of LCE value\label{alg:graph-value}}
-\begin{codebox}
-\li $s = \proc{MakeString}(1000)$
-\li \id{data} = an array where $\id{data}[v] \subseteq \{(i,j)~|~\LCE(i,j)=v\} \land |\id{data}[v]| =10.000$
-\li \For each $\id{alg} \in \id{Algorithms}$
-    \Indentmore
-\li     $\id{alg}.\proc{preprocess}(s)$
-\li     \For $v = 0$ \To $n$
-        \Indentmore
-\li         \Repeat 3 times
-% \li     $A =$ generate 1000000 random pairs $(i,j)$ where $i<j$
-\li             start timing
-\li             \For each $(i,j) \in data[v]$
-                \Indentmore
-\li                 $\id{alg}.\proc{query}(i,j)$
-                \End
-\li             end timing
-            \End
-\li         \proc{Plot}($alg$, $v$, median of the three timings)
-        \End
-\li     $\id{alg}.\proc{cleanup}()$
-    \End
-\end{codebox}
-\end{algorithm}
-
-Where $\proc{Plot}(f,x,y)$ adds point $(x,y)$ to series $f$.
-\end{comment}
-
 \fi % report
@@
 \ifreport
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{THE END}
-
-\begin{comment}
-\subsection*{Tinta tests}
-The two tests of LCE query performance used in the Tinta paper are shown in \pref{alg:tinta-testlceonebyone} and \pref{alg:tinta-randomtestlce}.
-
-% http://www.cs.dartmouth.edu/~thc/clrscode/clrscode.sty
-
-\begin{algorithm}
-\caption{Test algorithm for \texttt{test\_lce\_one\_by\_one}\label{alg:tinta-testlceonebyone}}
-\begin{codebox}
-\li \For each $LCE \in Algorithms$
-    \Indentmore
-\li     \For $k = 2$ \To $n$
-        \Indentmore
-\li         $A =\{(i,j) ~|~ j-i=k\}$
-\li         \Repeat 3 times
-\li             shuffle $A$
-\li             start timing
-\li             \For each $(i,j) \in A$
-                \Indentmore
-\li                 $LCE(i,j)$
-                \End
-\li             end timing
-            \End
-\li         $avg =$ average of the 3 timings
-\li         log(each of the 3 timings)
-\li         print($avg/length(A)$)
-        \End
-    \End
-\end{codebox}
-\end{algorithm}
-
-\begin{algorithm}
-\caption{Test algorithm for \texttt{random\_test\_lce}\label{alg:tinta-randomtestlce}}
-\begin{codebox}
-\li \Repeat 3 times
-\li     $A =$ generate 1000000 random pairs $(i,j)$ where $i<j$
-\li     \For each $\proc{LCE} \in \id{Algorithms}$
-        \Indentmore
-\li         start timing
-\li         \For each $(i,j) \in A$
-            \Indentmore
-\li             $LCE(i,j)$
-            \End
-\li         end timing
-        \End
-    \End
-\li \For each $LCE \in Algorithms$
-    \Indentmore
-\li     log(each of the 3 timings)
-    \End
-\li \For each $LCE \in Algorithms$
-    \Indentmore
-\li     print($($ average of the 3 timings $)/length(A)$)
-    \End
-\end{codebox}
-\end{algorithm}
-\end{comment}
-
 \ifarticle
@@
 \fi % article
-
-\begin{comment}
-\subsection*{Dropped}
-\begin{enumerate}
-\item DOC: Test on another machine.
-\item CODE: Improve cache-vertical and make shift variant.
-\item CODE: Non-random strings? Non-random queries? Different t-values for Hash-3L?
-\end{enumerate}
-
-\listoffixmes
-
-\section{What is the time for KMR?}
-\subsection{Original KMR}
-"Rapid identification of repeated patterns in strings, trees and arrays"
-says $O(n\cdot(\log n)^2)$.
-
-http://portal.acm.org.globalproxy.cvt.dk/citation.cfm?id=804905
-
-\subsection{RMR in $O(n \log n)$}
-\begin{itemize}
-\item
-"A Theoretical and Experimental Study on the Construction of Suffix Arrays in External Memory"
-page 10, 3.3 The Doubling Algorithm.
-
-http://www.springerlink.com.globalproxy.cvt.dk/content/h1tejj3bamjj5qrg/
-\item
-"Compressing and indexing labeled trees, with applications"
-
-http://portal.acm.org.globalproxy.cvt.dk/citation.cfm?id=1613680
-\end{itemize}
-
-KMR uses $O(sort(n,n^2)\cdot\log n) = O(n\log n)$ time.
-
-At a given level:
-\begin{enumerate}
-\item create 2-touples of fingerprints from the previous level.
-\item sort these touples lexicographically. (radix sort takes $O(kn)$ for $\sigma=n^k$, here $\sigma=n^2$)
-\item scan through the sorted touples to assign new fingerprints.
-\end{enumerate}
-\end{comment}
-
 \fi % report

### 2011-08-23 src/plot-article.sh

@@ -1,0 +1,54 @@
+
+oneplot()
+{
+    cat plotheader.txt > $plotfile
+    echo "set out '$plotfile.eps'" >> $plotfile
+    echo "set xrange [10:]" >> $plotfile
+    title=$(head -1 $datafile | sed -e 's/,/\n/g' | head -1)
+    # echo "set title '$title'" >> $plotfile
+    echo -e "$xtra" >> $plotfile
+    echo "plot \\" >> $plotfile
+    first=1
+    for algo in "$@"
+    do
+        if test $first = 0
+        then
+            echo ", \\" >> $plotfile
+        fi
+        first=0
+        algoidx=$(head -1 $datafile | sed -e 's/,/,\n/g' | grep -nF "$algo" -m1 | cut -d':' -f1)
+        echo -n "  '$datafile' using 1:$algoidx with $style" >> $plotfile
+    done
+    echo "Plotting $plotfile";
+    gnuplot $plotfile; ps2pdf -dEPSCrop $plotfile.eps $plotfile.pdf
+}
+
+listdata()
+{
+    head -1 $datafile | sed -e 's/,/,\n/g'
+    echo ""
+}
+
+yavg="set logscale y\nset yrange [7e-9:1e-6]\nset title 'Average case strings'"
+ywc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Worst case strings'"
+ymc="set logscale y\nset yrange [7e-9:1e-5]\nset title 'Medium LCE-size strings'"
+
+style="linespoints"
+
+#Overall Cache Fingerprint Slides
+
+# fp3 cache
+
+xtra=$yavg
+datafile="results/length-slides-rand10.txt"
+plotfile="results/length-fingerprint3-cache-rand10.plt"
+oneplot "Fingerprint_3<n^{2/3};n^{1/3}>ac," "Fingerprint_3<n^{2/3};n^{1/3}>ac_{cache-horizontal-shift}," "Fingerprint_3<n^{2/3};n^{1/3}>ac_{cache-horizontal-mult},"
+
+xtra=$ywc
+datafile="results/length-slides-alla.txt"
+plotfile="results/length-fingerprint3-cache-alla.plt"
[…]

### 2011-08-23 article/article.tex
m

@@
 \begin{abstract}
-The \emph{longest common extension} (LCE) problem is to preprocess a string in order to allow for a large number of LCE querie, such that the queries are efficient.
+The \emph{longest common extension} (LCE) problem is to preprocess a string in order to allow for a large number of LCE queries, such that the queries are efficient.
 The LCE value, $\LCE_s(i,j)$, is the length of the longest common prefix of the pair of suffixes starting at index $i$ and $j$ in the string $s$.

### 2011-08-25 article/article.tex
m

@@
 
-\title{Algorithms for Longest Common Extensions}
+\title{Longest Common Extension via Fingerprinting}
 
@@
 The LCE problem can be solved in linear space with constant query time and a preprocessing of sorting complexity.
-There are two known approaches achieving these bounds, which use nearest common ancestors and range minimum queries respectively.
+There are two known approaches achieving these bounds, which use nearest common ancestors and range minimum queries, respectively.
 However, in practice a much simpler approach with linear query time, no extra space and no preprocessing achieves significantly better average case performance.
-We show a new algorithm, \fprintk\, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time.
+We show a new algorithm, \fprintk\, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time, where $\sortt$ is the time it takes to sort $n$ numbers from $\sigma$.
 Though this solution is asymptotically strictly worse than the asymptotically best previously known algorithms, it outperforms them in practice in average case and is almost as fast as the simple linear time algorithm.
 On worst case input, this new algorithm is significantly faster in practice compared to the simple linear time algorithm.
-We also look at cache performance of the new algorithm, and we show that optimizing for cache only improves practical query time for $k=2$.
+We also look at cache performance of the new algorithm, and we show that for $k=2$, cache optimization can improve practical query time.
 \end{abstract}
@@
 The LCE value, $\LCE_s(i,j)$, is the length of the longest common prefix of the pair of suffixes starting at index $i$ and $j$ in the string $s$.
-The LCE problem can be used in many algorithms for solving other algorithmic problems, e.g.\ the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
-Solutions with linear space and constant query time exists for the problem~\cite{nca,jf-rmq}.
-These theoretically good solutions are however not the best in practice, since they have large constant factors for both time and space usage.
-A much simpler solution exists with average case constant time and no space required other than storing the input string.
-It has has significantly better practical performance for average case input as well as for average case queries on some real world strings, when compared to the asymptotically best known algorithms~\cite{ilie-navarro-tinta}.
+The LCE problem can be used in many algorithms for solving other algorithmic problems, e.g., the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
+Solutions with linear space, constant query time, and $O(\sortt)$ preprocessing time exists for the problem~\cite{nca,jf-rmq}.
+Here $\sortt$ is the time it takes to sort $n$ numbers from an alphabet of size $\sigma$.
+For $\sigma=O(n^c)$, where $c$ is a constant, we have $\sortt=O(n)$.
+These theoretically good solutions are however not the best in practice, since they have large constant factors for both query time and space usage.
+A much simpler solution exists with average case constant time and no space or preprocessing required other than storing the input string.
+It has significantly better practical performance for average case input as well as for average case queries on some real world strings, when compared to the asymptotically best known algorithms~\cite{ilie-navarro-tinta}.
 However this algorithm has linear worst case query time, and is thus only ideal when worst case performance is irrelevant.
@@
 This algorithm also achieves significantly better average case practical query time compared to the asymptotically best known algorithms.
-The algorithm uses string fingerprinting to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$, on a string of length $n$.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{The LCE Problem}
-
-%Given a string $s$ of length $n$ and a pair of indexes $1\leq i\leq n$ and $1\leq j\leq n$, the \emph{longest common extension} of $i$ and $j$ on $s$, $\LCE_s(i,j)$, is the length of the longest common prefix of $\suff_i$ and $\suff_j$.
-If for example $s = abbababba$, $i = 4$ and $j=6$, then $\suff_4 = ababba$ and $\suff_6 = abba$.
-The longest common prefix of these two suffixes is $ab$, which has length $2$, and therefore we have $\LCE_s(4,6) = 2$.
+The algorithm uses string fingerprinting inspired by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$, on a string of length $n$.
+The preprocessing time for this algorithm is $O(kn+\sortt)$.
+We look at a cache optimization for the new fingerprinting based algorithm, which improves query I/O for small values of $k$, and we show that for $k=2$, this cache optimization improves practical query times.
 
@@
 The algorithm has a parameter $k$ in the range $1\leq k\leq\logceil$, which describes the number of levels used\footnote{All logarithms are base two.}.
+The performance of the algorithm is summarized by the following theorem:
 
@@
 
-The exact worst case performance of our solution depends on the amount of space you are willing to use.
+By choosing $k$ we introduce the following tradeoffs:
 
@@
 For $k=2$ we can improve practical \fprintk\ query time even further by optimizing it for cache efficiency.
-However for $k>2$ the query time performance degrades, as the added overhead outweighs the improved cache efficiency.
+However for $k>2$, this cache optimization degrades practical query time performance, as the added overhead outweighs the improved cache efficiency.
 
@@
 %\paragraph{Sorting complexity.}
-The time it takes to sort $n$ numbers within an universe of size $\sigma$ is written as $\sortt$.
+%The time it takes to sort $n$ numbers within an universe of size $\sigma$ is written as $\sortt$.
 
@@
 \subsection{Query\label{sec:fingerprint-query}}
+The \fprintk\ query speeds up \proc{DirectComp} by comparing fingerprints of substrings of the input string instead of individual characters.
+The central part of the query algorithm moves down through the levels comparing fingerprints of progressively shorter substrings.
+This gives speed at the higher levels and then precision at the lower levels.
+To achieve constant average case query time, the algorithm starts at the lowest level and moves up through the levels before doing the central part.
 
@@
 \begin{enumerate}
-\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless and $\ell = k-1$.
+\item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$, increment $\ell$ by one, and repeat this step unless $\ell = k-1$.
 \item As long as $H_\ell[i+v] = H_\ell[j+v]$, increment $v$ by $t_\ell$ and repeat this step.
@@
 
-\begin{lemma}
+\begin{lemma}\label{lem:query-correct}
 The \fprintk\ query algorithm is correct.
@@
 \begin{proof}
-Step one takes $O(k)$ time.
+First we consider the worst case: Step one takes $O(k)$ time.
 In step two and three, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$).
@@
 
-At each step except step three, the algorithm increments $v$.
+Next we consider the average case: At each step except step three, the algorithm increments $v$.
 Step three is executed the same number of times as step one, in which $v$ is incremented.
-The query time is therefore linear to the number of times $v$ is incremented, and it is thereby $O(v)$.
-The query time is thus $O(1)$ in average case, where $v=O(1)$.
+The query time is therefore linear in the number of times $v$ is incremented, and it is thereby $O(v)$.
+From the proof of \pref{lem:query-correct} we have $v=\LCE(i,j)$, and in average case $\LCE(i,j)=O(1)$~\cite{ilie-navarro-tinta}.
+The query time is thus $O(1)$ in average case.
 \end{proof}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{Average Case Optimization\label{sec:improved-avg}}
+\paragraph{Average Case vs.\ Worst Case\label{sec:improved-avg}}
 
+The query algorithm is optimized for average case.
+Without this, the average case query time would be $O(k)$, while the asymptotic worst case time would not be affected.
+The practical worst case query time could however be negatively affected.
+Our experiments have shown that the improvement in practical average case query time far outweighs any degraded worst case performance, and for small values of $k$, changes in worst case performance is barely measurable.
+%Our experiments have shown that for small values of $k$, optimizing for the average case significantly improves average case query time, while any negative effect on worst case query time is barely measurable.
+For large values of $k$ the improvement in average case far outweighs the slightly degraded practical worst case.
+
+\begin{comment}
 We could have left out step one of the query algorithm and started with $\ell=k-1$.
@@
 \fi % article
+\end{comment}
 
@@
 \item[Medium LCE value strings] These strings have an average LCE value over all $n^2$ query pairs of $n/2r$, where $r=0.73n^{0.42}$.
-This value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where the each of the values of $n$ were chosen to give the best result for \fprintk\ on a plot with the given constant value of $r$.
-We use results on these strings as an indication of query times somewhere between the average case and worst case.
+We use results on these strings to show that \fprint\ can be faster than both \proc{DirectComp} and \proc{LcpRmq} on some inputs between the average case and worst case.
+The value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
 We construct these strings by repeating a substring of $r$ characters, where each character is unique.
@@
 
-On strings with medium average LCE values, we see that our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}
+On the plot of strings with medium average LCE values, we see a case where our \fprintk\ algorithms are faster than both \proc{DirectComp} and \proc{LcpRmq}.
 
@@
 We achieve the greatest I/O improvement when $k$ is small.
-When $k=\logceil$, this cache optimization gives no difference in the amount of I/O.
+When $k=\logceil$, this cache optimization gives no asymptotic difference in the amount of I/O.
 
@@
 The last plot on \pref{fig:plot-fingerprint-cache-horiz} shows a variant of average case where the alphabet size is changed to two.
-This plot attempts to show the worst case for cache optimized \fprint[2].
+This plot shows a bad case for cache optimized \fprint[2].
 LCE values in this plot are large enough to ensure that $H_1$ is used often, which should make the extra complexity of calculating indexes into $H_1$ visible.
 At the same time the LCE values are small enough to ensure, that the cache optimization has no effect.
-In this plot we see that the cache optimized variant of \fprint[2] has only slightly worse query time compared to the variant, which is not cache optimized.
+In this bad case plot we see that the cache optimized variant of \fprint[2] has only slightly worse query time compared to the variant, which is not cache optimized.
 
@@
 Unlike for \fprint[2], we see a degraded performance on worst case strings compared to the not cache optimized variant.
-Our results thus show that cache optimization of \fprintk\ only gives a performance improvement for $k=2$.
+Our results thus show that this cache optimization of \fprintk\ only gives a performance improvement for $k=2$.
 
@@
 
-\ifreport
-Of the existing best \proc{DirectComp} and \proc{LcpRmq} algorithms, we have seen that \proc{DirectComp} is best on average case strings, while \proc{LcpRmq} is best on worst case strings.
-
-We have found that our new \proc{DirectCompLookup} algorithm can improve the worst case query time performance of \proc{DirectComp}, while it does not degrade the average case query time to the level of \proc{LcpRmq}.
-However it is not as interesting as our new \fprintk\ algorithm, because it uses too much space.
-\fi % report
-
 The \fprintk\ algorithm is able to achieve a balance between practical worst case and average case query times.
-It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have found cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}.
+It has almost as good average case query times as \proc{DirectComp}, its worst case query times are significantly better than those of \proc{DirectComp}, and we have shown cases between average and worst case where \fprintk\ is better than both \proc{DirectComp} and \proc{LcpRmq}.
 \fprintk\ gives a good time space tradeoff, and it uses less space than \proc{LcpRmq} when $k$ is small.
 %The performance of \fprintk\ can be tweaked on many parameters, and we have found that optimizing for average case queries and for memory caches improve practical query time performance.
-
-\ifreport
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Test Code}
-
-To test practical performance, we have implemented the algorithms in C++.
-The implementation is based on that of Ilie~et~al.~\cite{ilie-navarro-tinta}, which we have heavily modified.
-We have added a working build system, made it compile on a 32-bit Windows architecture, and generalized it to allow more algorithms and tests to be implemented.
-\pref{tab:test-variables} summarizes the structure of the code and the parameters we can vary.
-
-\begin{table}[tp]
-\centering
-\begin{tabular}{l l l}
-Variable & Example & Component \\ \hline
-alphabet size & $\sigma = 10$ & string \\
-character pattern & uniformly at random & string \\ \hline
-algorithm & \fprintk & algorithm \\
-alg. parameters & $t_1 = 2\sqrt n$ & algorithm \\ \hline
-string length & exponentially increasing from $1$ to $100.000$ & test \\
-query pattern & uniformly at random & test \\ \hline
-\end{tabular}
-\caption{List of performance variables tested.}\label{tab:test-variables}
-\end{table}
-
-In our implementation, we have made the following assumptions about the LCE problem:
-\begin{itemize}
-\item A character in a string is always eight bits.
-\item A string always ends with a special null character, which does not occur anywhere else in the string.
-This does not affect our results, since we do not measure practical preprocessing time.
-If an algorithm has faster query times when the string ends in a special character, it could just add that character to the end of the string in its preprocessing.
-\item We make no assumptions on the order of $i$ and $j$.
-In some applications of the LCE problem, for example approximate string searching~\cite{approx-search}, we know that $i<j$, and some algorithms might be able to take advantage of this knowledge to get faster query times.
-We do not measure this case.
-\item Whenever we use random characters or query pairs, we use the \texttt{rand()} function from C. This gives repeatable pseudo-random numbers, which makes it easier for us to compare test runs.
-\end{itemize}
-
-\fi % report
 

### 2011-08-25 article/article.tex
m

@@
 
-\newif\ifarticle
-\newif\ifreport
-
-\articletrue
-
 \maketitle
@@
 However, in practice a much simpler approach with linear query time, no extra space and no preprocessing achieves significantly better average case performance.
-We show a new algorithm, \fprintk\, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time, where $\sortt$ is the time it takes to sort $n$ numbers from $\sigma$.
+We show a new algorithm, \fprintk, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time, where $\sortt$ is the time it takes to sort $n$ numbers from $\sigma$.
 Though this solution is asymptotically strictly worse than the asymptotically best previously known algorithms, it outperforms them in practice in average case and is almost as fast as the simple linear time algorithm.
@@
 
-Ilie~et~al.~\cite{ilie-navarro-tinta} gave an algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and $O(|\LCE(i,j)|)$ query time.
+Ilie~et~al.~\cite{ilie-navarro-tinta} gave an algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and $O(\LCE(i,j))$ query time.
 For a query $\LCE(i,j)$, the algorithm compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
@@
 One method, \proc{SuffixNca}, uses constant time nearest common ancestor (NCA) queries~\cite{nca} on a suffix tree.
-The LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$.
-In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$), and no two child edge labels of the same node will have the same first character.
+The LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of the suffixes $s[i\twodots n]$ and $s[j\twodots n]$.
+In a suffix tree, the path from the root to $L_i$ has label $s[i\twodots n]$ (likewise for $j$), and no two child edge labels of the same node will have the same first character.
 The longest common prefix of the two suffixes will therefore be the path label from the root to the nearest common ancestor of $L_i$ and $L_j$.
@@
 
-%\begin{figure}[tp]
-%    \begin{center}
-%        \includegraphics[width=0.5\textwidth,page=1]{wc-avg.pdf}
-%    \end{center}
-%    \caption{\label{fig:wc-avg}Graphical comparison of query times for different LCE algorithms.}
-%\end{figure}
-
 In practice, existing state of the art solutions are either good in worst case, while poor in average case (\proc{LcpRmq}), or good in average case while poor in worst case (\proc{DirectComp}).
@@
 In the \emph{cache-oblivious model}, the algorithm has no knowledge of the values of $M$ and $B$.
-
-\ifreport
-
-\subsubsection{\proc{SuffixNca}: NCA on a Suffix Tree}
-
-LCE of two indexes $i$ and $j$ is defined as the length of the longest common prefix of $\suff_i$ and $\suff_j$.
-In a suffix tree, the path from the root to $L_i$ has label $\suff_i$ (likewise for $j$).
-The path label on the path from the root to the nearest common ancestor of $L_i$ and $L_j$ is therefore a common prefix of the two suffixes.
-Since two child edges of the same node never have the same first character on their edge labels, the path label on the path from the root to $\NCA_\mathcal{T}(L_i,L_j)$ is not only a common prefix, but the longest common prefix of $\suff_i$ and $\suff_j$.
-We can therefore find the LCE value as $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$, where $\mathcal{T}$ is the suffix tree for $s$, and $D[v]$ is the label depth of node $v$ in the tree.
-
-For a string of length $n$ with alphabet size $\sigma$, we can preprocess \proc{SuffixNca} in $O(\sortt)$ time by building the suffix tree, preprocessing it for NCA and building the array of label depths.
-The query time is $O(1)$, since looking up $L_i$ and $L_j$, doing an NCA query, and looking up the label depth in $D$ all takes constant time.
-
-\fi % report
 
@@
 The practical worst case query time could however be negatively affected.
-Our experiments have shown that the improvement in practical average case query time far outweighs any degraded worst case performance, and for small values of $k$, changes in worst case performance is barely measurable.
+Our experiments have shown that the improvement in practical average case query time far outweighs any degraded worst case performance, and for small values of $k$, changes in worst case performance are barely measurable.
 %Our experiments have shown that for small values of $k$, optimizing for the average case significantly improves average case query time, while any negative effect on worst case query time is barely measurable.
@@
 This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
-\ifarticle
 Our experiments have shown that whenever $k$ is small, keeping step one improves average case query time, while it does not have a measurable effect on worst case query times.
@@
 We want to optimize our LCE query time for the average case where the LCE value is small, so our results in \pref{sec:results} does not include the variant of \fprintk\ with $O(k)$ average case query time.
-\fi % article
 
@@
 In step one of our query algorithm we perform one comparison at each level. We could instead do up to $O(n^{1/k})$ comparisons at each level without affecting asymptotic times.
-\ifarticle
 Our experiments have shown that always doing one comparison at each level is the best in practice.
-\fi % article
 \end{comment}
@@
 We first construct the SA and LCP arrays, which takes $O(\sortt)$ time~\cite{sort-complexity}.
-We then preprocess each of the $k$ levels in $O(n)$ time, since we loop through $n$ substrings, and comparing neighboring substrings takes constant time when we use the LCE array.
+We then preprocess each of the $k$ levels in $O(n)$ time, since we loop through $n$ substrings, and comparing neighboring substrings takes constant time when we use the LCP array.
 The total preprocessing time becomes $O(\sortt\\ + k\cdot n)$.
@@
 \item[\fprintk\textless$t_{k-1}$, ..., $t_1$\textgreater ac] is the \fprintk\ algorithm using $k$ levels, where $k$ is $2$, $3$ and $\logceil$.
-The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater describe the exact size of fingerprinted substrings at each level.
+The numbers \textless$t_{k-1}$, ..., $t_1$\textgreater\ describe the exact size of fingerprinted substrings at each level.
 \item[\proc{RMQ}\textless$n$, $1$\textgreater] is the \proc{LcpRmq} algorithm using constant time RMQ.
@@
 
-\ifarticle
-
 \proc{LcpRmq} shows a significant jump in query times around $n=1.000.000$ on the plot with average case strings, but not on the plot with worst case strings.
@@
 The cache miss rate for average case strings is 14\% and 9\% for the L1 and L2 caches, and for worst case strings the miss rate is 17\% and 13\%, which is the opposite of what could explain the jump we see in the plot.
-
-\fi % article
 
@@
 On the plot with worst case strings however, we see a jump in \fprint[\log n] between $n=20.000$ and $n=50.000$, since it uses all its $\logceil$ levels when the LCE values are large.
-
-\ifreport
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Optimal Value of $t_\ell$}
-
-\pref{fig:plot-fingerprint-tval} shows \fprint[2] for different values of $t_1$.
-On average case strings the value of $t_1$ makes no difference, and on worst case strings a value of $t_1$ between $2\sqrt n$ and $4\sqrt n$ is the best.
-The improvement relative to $\sqrt n$ is however insignificant, so we have chosen to use $t_1=\sqrt n$.
-
-Our results indicate that reading values from $H_0$ is faster than reading values from $H_1$.
-Differences between these two tables include that elements of $H_0 = s$ use one byte, whereas elements of $H_1$ use four bytes, and that elements we read from $H_0$ are located next to each other in memory, whereas elements we read from $H_1$ are spaced $t_1$ elements apart in memory.
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-2l-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-2l-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-fingerprint-tval}Query times of \fprint[2] for different values of $t_1$ versus query times for \proc{DirectComp}.}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsubsection{Average Case Optimization\label{sec:test-average}}
-
-In \pref{sec:improved-avg} we discussed how our \fprintk\ algorithm is optimized for the average case to get $O(1)$ instead of $O(k)$ average case query time, while it keeps the same asymptotic worst case query time.
-But in practice the worst case query time is slightly affected.
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avglog-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avglog-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-fingerprint-avglog}Query times of \fprint[\logceil] with $O(1)$ versus $O(k)$ average case query time.}
-\end{figure}
-
-In \pref{fig:plot-fingerprint-avglog} we see that on worst case strings, the query time of \fprint[\logceil] doubles when we use the average case optimization.
-On average case strings, the query time of \fprint[\logceil] is up to 40 times faster when using average case optimization.
-All four cases in \pref{fig:plot-fingerprint-avglog} show their expected $O(1)$ and $O(\log n)$ query times.
-In \pref{fig:plot-fingerprint-avg3l}, we see the same results for \fprint[3], as we saw for \fprint[\logceil], but at a smaller scale.
-The negative effect of doing average case optimization is barely measurable on worst case strings, while the improvement on average case strings is up to 40\%.
-
-\begin{figure}[p]
-    \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avg3l-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-avg3l-alla.plt}
-    \end{center}
-    \caption{\label{fig:plot-fingerprint-avg3l}Query times of \fprint[3] with $O(1)$ versus $O(k)$ average case query time.}
-\end{figure}
-
-In \pref{fig:plot-fingerprint-avg3l} we see that doing one comparison at each level on the way up in \fprint[3] is better than doing $n^{1/k}$ comparisons.
-This is most visible on worst case strings, where the algorithm doing one comparison at each level uses less time to reach level $\ell=k-1$, which it will reach anyway.
-On average case strings, the code specialized in handling only one comparison at each level is simpler.
-
-We conclude that using average case optimization yields a query time improvement on average case strings, which is significantly greater than the query time increase on worst case strings, and that \fprintk\ therefore should use the average case optimization.
-
-\fi % report
 
@@
 
-\ifreport
-\ifarticle
-\fxfatal{Opskriv funktioner for plads.}
-\fi % article
-\fi % report
+%\fxfatal{Opskriv funktioner for plads.}
 

### 2011-08-29 presentation/presentation.tex
m

@@
 \section{Introduction}
-\subsection{The LCE Problem}
-\begin{frame}{The LCE Problem and the \proc{DirectComp} algorithm}
+\subsection{The \proc{DirectComp} algorithm}
+\begin{frame}{The \proc{DirectComp} algorithm}
     \begin{block}{Input}
@@
         \end{block}
-        \begin{block}{LCE problem}
-            Efficiently perform multiple queries $(i,j)$ on a static string $s$
-        \end{block}
     }
@@
 
-\subsection{The \proc{DirectComp} algorithm}
+\subsection{The LCE Problem}
+\begin{frame}{The LCE Problem}
+    \begin{description}
+        \item[LCE value] $\LCE_s(i,j)$ is the length of the longest common prefix of the two suffixes of $s$ starting at index $i$ and $j$
+        \item[LCE problem] Efficiently query multiple LCE values on a static string $s$
+    \end{description}
+\end{frame}
+
+%\subsection{The \proc{DirectComp} algorithm}
 \begin{frame}{Existing Algorithm: \proc{DirectComp}}
@@
 
-\subsection{The \proc{LcpRmq} Algorithm}
-\begin{frame}{Existing Algorithm: \proc{LcpRmq}}
-    \begin{multicols}{2}{
-        \begin{align*}
-            s&=abbababba\\
-            s[2\twodots n]&=bbababba\\
-            s[3\twodots n]&=bababba\\
-            \LCE_s(2,3)&=1
-        \end{align*}
-        \newpage
+\subsection{The \proc{SuffixNca} and \proc{LcpRmq} Algorithms}
+\begin{frame}{Existing Algorithms: \proc{SuffixNca} and \proc{LcpRmq}}
+    %\begin{multicols}{2}{
+    %    \begin{align*}
+    %        s&=abbababba\\
+    %        s[2\twodots n]&=bbababba\\
+    %        s[3\twodots n]&=bababba\\
+    %        \LCE_s(2,3)&=1
+    %    \end{align*}
+    %    \newpage
+    %    \begin{center}
+    %        %\uncover<2->{
+    %        \includegraphics[width=0.26\textwidth,page=1]{../doc/sa+lcp+min.pdf}
+    %        %}
+    %    \end{center}
+    %}
+    %\end{multicols}
+    %\uncover<3->{
+    Two algorithms with best known bounds:
+    \begin{description}
+        \item[\proc{SuffixNca}] Nearest common ancestor queries on a suffix tree
+        \item[\proc{LcpRmq}] Range minimum queries on a longest common prefix array
+    \end{description}
+    \vspace{1em}
     \begin{center}
-            %\uncover<2->{
-            \includegraphics[width=0.26\textwidth,page=1]{../doc/sa+lcp+min.pdf}
-            %}
-        \end{center}
-    }
-    \end{multicols}
-    %\uncover<3->{
-    \begin{center}
-        $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$\\
-        where $\SAinv[i] < \SAinv[j]$\\
-        \vspace{1em}
+    %    $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$\\
+    %    where $\SAinv[i] < \SAinv[j]$\\
+    %    \vspace{1em}
         \begin{tabular}{r l}
@@
 
+\begin{comment}
 \subsection{Preprocessing}
@@
 
-    \hfill\\$k=1$:
+    \vspace{2em}
 
-    ~~Same as \proc{DirectComp}
-
-%    \hfill\\$k=2$:
-
-%    \begin{tabular}{r l}
-%        Preprocessing & $O(\sortt)$ \\
-%        Space & $O(n)$\\
-%        Query & $O(\sqrt n)$ \\
-%        Average query & $O(1)$ \\
-%    \end{tabular}
-
-    \begin{multicols}{2}{
-        $k=2$:
-
-        \begin{tabular}{r l}
-            Preprocessing & $O(\sortt)$ \\
-            Space & $O(n)$\\
-            Query & $O(\sqrt n)$ \\
-            Average query & $O(1)$ \\
-            Query I/O & $O\Big(\frac{\sqrt n}{B}\Big)$ \\
+    \begin{tabular}{r l l l}
+        & $k=1$ & $k=2$ & $k=\logceil$ \\
+        Preprocessing & $O(\sortt)$ & $O(\sortt)$ & $O(n\log n)$ \\
+        Space & $O(n)$ & $O(n)$ & $O(n\log n)$ \\
+        Query & $O(n)$ & $O(\sqrt n)$ & $O(\log n)$ \\
+        Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
+        Query I/O & $O\Big(\frac{n}{B}\Big)$ & $O\Big(\frac{\sqrt n}{B}\Big)$ & $O(\log n)$ \\
     \end{tabular}
-        \newpage
-        $k=\logceil$:
-
-        \begin{tabular}{r l}
-            Preprocessing & $O(n\log n)$ \\
-            Space & $O(n\log n)$\\
-            Query & $O(\log n)$ \\
-            Average query & $O(1)$ \\
-            Query I/O & $O(\log n)$ \\
-        \end{tabular}
-    }
-    \end{multicols}
 \end{frame}
@@
     \textcolor{green}{shift operations} vs.\
-    \textcolor{blue}{multiplication and division}
+    \textcolor{blue}{multiplication, division and modulo}
     %by string length
@@
 
-\begin{comment}
+%\begin{comment}
 \section{LCE on trees}
@@
             \begin{itemize}
-                \item Construct a heavy tree decomposition
-                \item For each heavy path, store characters as a substring
+                \item Construct a heavy path decomposition
+                \item For each heavy path, store characters as a substring of $s$
             \end{itemize}
@@
             \begin{itemize}
-                \item Use constant time string LCE on each heavy path
+                \item Use constant time string LCE on $s$ for each heavy path
             \end{itemize}
@@
 \end{frame}
-\end{comment}
+%\end{comment}
 
@@
 
+\begin{frame}{Kommentarer til rapporten}
+    \begin{itemize}
+        \item Hvordan jeg fandt frem til $r=0.73n^{0.42}$
+        \item Der står \fprint[3] nogle steder i cache-afsnittet hvor der skal stå \fprint[2]
+    \end{itemize}
+\end{frame}
+
 \end{document}

### 2011-08-29 article/article.tex
m

@@
 
-\title{Longest Common Extension via Fingerprinting}
+\title{Longest Common Extensions via Fingerprinting}
 
-\author{Philip Bille\\\texttt{phbi@imm.dtu.dk}\and Inte Li Gørtz\\\texttt{ilg@imm.dtu.dk}\and Jesper Kristensen\\\texttt{mail@jesperkristensen.dk}}
+\author{Philip Bille\\\texttt{phbi@imm.dtu.dk}\and Inge Li G{\o}rtz \\\texttt{ilg@imm.dtu.dk}\and Jesper Kristensen\\\texttt{mail@jesperkristensen.dk}}
 
-\date{Technical University of Denmark\\\hfill\\August 1, 2011}
+\date{}
 
@@
 These theoretically good solutions are however not the best in practice, since they have large constant factors for both query time and space usage.
-A much simpler solution exists with average case constant time and no space or preprocessing required other than storing the input string.
-It has significantly better practical performance for average case input as well as for average case queries on some real world strings, when compared to the asymptotically best known algorithms~\cite{ilie-navarro-tinta}.
-However this algorithm has linear worst case query time, and is thus only ideal when worst case performance is irrelevant.
+Ilie et.
+al~\cite{ilie-navarro-tinta} introduced a much simpler solution with average case constant time and no space or preprocessing required other than storing the input string.
+This solution has significantly better practical performance for average case input as well as for average case queries on some real world strings, when compared to the asymptotically best known algorithms.
+However, this algorithm has linear worst case query time, and is thus only ideal when worst case performance is irrelevant.
 In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.
@@
 At the same time it must scan legitimate data fast.
+The main goal of this paper is to design an algorithm that performs well in both situations, that is, achieves a good worst-case guarantee while maintaining a fast average case performance.
 
-In this paper we give a new algorithm, which is a generalization of the simple linear time algorithm.
+% PB: har slettet nedenstående afsnit til fordel for ekstra sætning i ovenstående. Resultaterne kommer senere.
+%In this paper we give a new algorithm, which is a generalization of the simple linear time algorithm.
 The new algorithm is almost as simple, yet it has significantly better worst case query time.
-This algorithm also achieves significantly better average case practical query time compared to the asymptotically best known algorithms.
-The algorithm uses string fingerprinting inspired by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for $k$ between one and $\log n$, on a string of length $n$.
+This algorithm also achieves significantly better average case practical query time compared to the asymptotically best known worst case query time algorithms.
+The algorithm uses string fingerprinting inspired by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for any integer $k$, $1 \leq k \leq \log n$, on a string of length $n$.
 The preprocessing time for this algorithm is $O(kn+\sortt)$.
@@
 \subsection{Previous Results\label{sec:existing-results}}
+Thoughout the paper let $s$ be a string of length $n$ over alphabet of size $\sigma$.
+Ilie~et~al.~\cite{ilie-navarro-tinta} gave an algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and has $O(\LCE(i,j))$ query time.
+For a query $\LCE(i,j)$, the algorithm simply compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
+The worst case query time is thus $O(n)$.
+However, on random strings and many real-word texts Ilie~et~al.~\cite{ilie-navarro-tinta} showed that the average LCE is $O(1)$, where the average is over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
+Hence, in these scenarios \proc{DirectComp} achieves $O(1)$ query time.
 
-Ilie~et~al.~\cite{ilie-navarro-tinta} gave an algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and $O(\LCE(i,j))$ query time.
-For a query $\LCE(i,j)$, the algorithm compares $s[i]$ to $s[j]$, then $s[i+1]$ to $s[j+1]$ and so on, until the two characters differ, or the end of the string is reached.
-The worst case query time is $O(n)$ on a string of length $n$.
-Given a string length $n$ and an alphabet size $\sigma$, we define \emph{average case query time} as the average of query times over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
-The average case LCE value over all possible inputs of a given string of length $n$ and alphabet size $\sigma$ is $O(1/(\sigma-1))=O(1)$~\cite{ilie-navarro-tinta}, which gives \proc{DirectComp} average case query time $O(1)$.
-The LCE problem can be solved with $O(1)$ worst case query time, using $O(n)$ space and $O(\sortt)$ preprocessing time. Two different ways of doing this exists.
+%Given a string length $n$ and an alphabet size $\sigma$, we define \emph{average case query time} as the average of query times over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
+
+The LCE problem can also be solved with $O(1)$ worst case query time, using $O(n)$ space and $O(\sortt)$ preprocessing time.
+Essentially, two different ways of doing this exists.
 One method, \proc{SuffixNca}, uses constant time nearest common ancestor (NCA) queries~\cite{nca} on a suffix tree.
@@
 In a suffix tree, the path from the root to $L_i$ has label $s[i\twodots n]$ (likewise for $j$), and no two child edge labels of the same node will have the same first character.
-The longest common prefix of the two suffixes will therefore be the path label from the root to the nearest common ancestor of $L_i$ and $L_j$.
-I.e.
-$\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.
+The longest common prefix of the two suffixes will therefore be the path label from the root to the nearest common ancestor of $L_i$ and $L_j$, i.e., $\LCE_s(i,j) = D[\NCA_\mathcal{T}(L_i,L_j)]$.
 The other method, \proc{LcpRmq}, uses constant time range minimum queries (RMQ)~\cite{jf-rmq} on a longest common prefix (LCP) array.
 The LCP array contains the length of the longest common prefixes of each pair of neighbor suffixes in the suffix array (SA).
-The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering.
-I.e.
-$\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
-Ilie~et~al.~\cite{ilie-navarro-tinta} have looked at a number of real world texts as well as texts of randomly generated characters, and found that all the texts they examined each has an average LCE of at most one, over all $n^2$ input pairs.
-Therefore it is interesting to have LCE algorithms, which perform good on average when the LCE value is small.
+The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e., $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
+\pref{tab:article-algorithms} summaries the above theoretical bounds.
+
+%Hence, it is interesting to have LCE algorithms, which perform good on average when the LCE value is small.
 Both \proc{SuffixNca} and \proc{LcpRmq} have the same asymptotic space and times.
@@
 \subsection{Our Results}
-
 We present a new LCE algorithm, \fprintk, based on multiple levels of string fingerprinting.
@@
 The performance of the algorithm is summarized by the following theorem:
-
 \begin{theorem}
@@
 \end{theorem}
-
-By choosing $k$ we introduce the following tradeoffs:
-
+By choosing $k$ we can obtain the following interesting tradeoffs.
 \begin{corollary}
@@
 \end{corollary}
-
 \begin{corollary}
@@
 \end{corollary}
-
 To preprocess the $O(k\cdot n)$ fingerprints used by our algorithm, we can use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, which takes $O(n\log n)$ time.
@@
 \proc{LcpRmq} & $O(n)$ & $O(1)$ & $O(\sortt)$ \\ \hline
-\proc{DirectComp} & $O(1)$ & $O(n)$ & $O(1)$ \\ \hline
+\proc{DirectComp} & $O(1)$ & $O(n)$ & None \\ \hline
 \fprintk * & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
@@
 %\paragraph{RMQ.}
-The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, i.e.\ $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
+The range minimum of $i$ and $j$ on an array $A$ is the index of a minimum element in $A[i,j]$, i.e.,\ $\RMQ_A(i,j) = \arg \min_{k\in\{i,...,j\}}\{A[k]\}$.
 A \emph{range minimum query} (RMQ) on a static array of $n$ elements can be answered in $O(1)$ time with $O(n)$ space and preprocessing time~\cite{jf-rmq}.
@@
 The \fprintk\ query speeds up \proc{DirectComp} by comparing fingerprints of substrings of the input string instead of individual characters.
-The central part of the query algorithm moves down through the levels comparing fingerprints of progressively shorter substrings.
-This gives speed at the higher levels and then precision at the lower levels.
-To achieve constant average case query time, the algorithm starts at the lowest level and moves up through the levels before doing the central part.
+The query algorithm consists of two traversals of the hierarchy of fingerprints.
+In the first traversal the algorithm compares progressively larger fingerprints of substrings until a mismatch is found and in the second traversal the algorithm compares progressively smaller substrings to find the precise point of the mismatch.
+The combination of these two traversals ensures both a fast worst case and average case performance.
 
-To perform a LCE query using the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
+The details of the query algorithm are as follows. Given the \fprintk\ data structure, start with $v=0$ and $\ell=0$, then do the following steps:
 \begin{samepage}
@@
 \begin{proof}
-First we consider the worst case: Step one takes $O(k)$ time.
+First we consider the worst case.
+Step one takes $O(k)$ time.
 In step two and three, the number of remaining characters left to check at level $\ell$ is $O(n^{(\ell+1)/k})$, since the previous level found two differing substrings of that length (at the top level $\ell=k-1$ we have $O(n^{(\ell+1)/k}) = O(n)$).
@@
 
-Next we consider the average case: At each step except step three, the algorithm increments $v$.
+Next we consider the average case.
+At each step except step three, the algorithm increments $v$.
 Step three is executed the same number of times as step one, in which $v$ is incremented.
 The query time is therefore linear in the number of times $v$ is incremented, and it is thereby $O(v)$.
-From the proof of \pref{lem:query-correct} we have $v=\LCE(i,j)$, and in average case $\LCE(i,j)=O(1)$~\cite{ilie-navarro-tinta}.
-The query time is thus $O(1)$ in average case.
+From the proof of \pref{lem:query-correct} we have $v=\LCE(i,j)$.
+By Ilie et al.~\cite{ilie-navarro-tinta} the average $\LCE(i,j)$ is $O(1)$ and hence the average case query time is $O(1)$.
 \end{proof}
@@
 \paragraph{Average Case vs.\ Worst Case\label{sec:improved-avg}}
+The first traversal in the query algorithm gaurantees $O(1)$ average case performance.
+However without it the worst case bound would remain $O(k)$.
+Thus, for a better practical worst-case performance we could omit the first traversal entirely.
+We have extensively experimented with both variants and we found that in nearly all scenarios the first traversal improved the overall performance.
+In the cases where performance was not improved the first traversal only degraded the performance slightly.
+We therefore focus exclusively on the two traversal variant in the remainder of the paper.
 
-The query algorithm is optimized for average case.
+%query algorithm is optimized for average case.
 Without this, the average case query time would be $O(k)$, while the asymptotic worst case time would not be affected.
@@
 Our experiments have shown that the improvement in practical average case query time far outweighs any degraded worst case performance, and for small values of $k$, changes in worst case performance are barely measurable.
+
 %Our experiments have shown that for small values of $k$, optimizing for the average case significantly improves average case query time, while any negative effect on worst case query time is barely measurable.
@@
 \item[Medium LCE value strings] These strings have an average LCE value over all $n^2$ query pairs of $n/2r$, where $r=0.73n^{0.42}$.
-We use results on these strings to show that \fprint\ can be faster than both \proc{DirectComp} and \proc{LcpRmq} on some inputs between the average case and worst case.
-The value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
-We construct these strings by repeating a substring of $r$ characters, where each character is unique.
+These strings where designed to show that \fprint\ can be faster than both \proc{DirectComp} and \proc{LcpRmq} on some inputs between the average case and worst case.
+The strings consist of repeating substrings of $r$ unique characters.
+%The value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
 \end{description}
@@
 \begin{thebibliography}{9}
+\bibitem{sort-complexity} M. Farach-Colton, P. Ferragina, and S. Muthukrishnan. On the sorting-complexity of suffix tree construction. \textit{J. ACM} 47(6):987-1011, 2000.
 
-\bibitem{ilie-navarro-tinta} Lucian Ilie, Gonzalo Navarro, and Liviu Tinta.
+\bibitem{jf-rmq} J. Fischer, and V. Heun. Theoretical and Practical Improvements on the RMQ-Problem, with Applications to LCA and LCE. In \textit{Proc. 17th CPM}, pages 36-48, 2006.
+
+\bibitem{nca} D. Harel, R. E. Tarjan. Fast Algorithms for Finding Nearest Common Ancestors. \textit{SIAM J. Comput.}, 13(2):338-355, 1984.
+
+\bibitem{ilie-navarro-tinta} L.
+Ilie, G.
+Navarro, and L.
+Tinta.
 The longest common extension problem revisited and applications to approximate string searching.
-Journal of Discrete Algorithms, Volume 8, Issue 4, December 2010, pages 418-428.
+\textit{J.
+Disc.
+Alg.,} 8(4):418-428, 2010.
 
-% \bibitem{linear-lcp} Toru Kasai, Gunho Lee, Hiroki Arimura, Setsuo Arikawa, and Kunsoo Park.
-Linear-Time Longest-Common-Prefix Computation in Suffix Arrays and Its Applications.
-CPM 2001, LNCS 2089, 2001, pages 181-192.
+\bibitem{karp-miller-rosenberg} R. M. Karp, R. E. Miller, and A. L. Rosenberg. Rapid identification of repeated patterns in strings, trees and arrays. In \textit{Proc. 4th STOC}, pages 125-136, 1972.
 
-\bibitem{sort-complexity} Martin Farach-Colton, Paolo Ferragina, and S. Muthukrishnan. On the sorting-complexity of suffix tree construction. J. ACM Vol. 47, No. 6, November 2000, pages 987-1011.
-
-\bibitem{karp-miller-rosenberg} Richard M.
-Karp, Raymond E.
-Miller, and Arnold L.
-Rosenberg.
-1972.
-Rapid identification of repeated patterns in strings, trees and arrays.
-In Proceedings of the fourth annual ACM symposium on Theory of computing (STOC '72).
-ACM, New York, NY, USA, 125-136.
-
-\bibitem{nca} D. Harel, R. E. Tarjan. Fast Algorithms for Finding Nearest Common Ancestors. SIAM J. Comput., 1984.
-% S. Alstrup, C. Gavoille, H. Kaplan, and T. Rauhe. Nearest Common Ancestors: A Survey and a New Algorithm for a Distributed Environment. Theory of Comput. Sys., 2004.
-
-\bibitem{jf-rmq} Johannes Fischer, and Volker Heun.
-Theoretical and Practical Improvements on the RMQ-Problem, with Applications to LCA and LCE.
-Proceedings of the 17th Annual Symposium on Combinatorial Pattern Matching (CPM'06), Lecture Notes in Computer Science 4009, 36-48, Springer-Verlag, 2006.
-
-%\bibitem{predecessor} Dan E. Willard. Log-logarithmic worst-case range queries are possible in space $\Theta$(N). Information Processing Letters, Volume 17, Issue 2, 24 August 1983, pages 81-84.
-
-%\bibitem{level-ancestor} Michael A. Bendera, Martín Farach-Colton. The Level Ancestor Problem simplified. Theoretical Computer Science 321, 2004, pages 5-12.
-
-\bibitem{approx-search} Gad M. Landau and Uzi Vishkin. Introducing efficient parallelism into approximate string matching and a new serial algorithm. 18th ACM STOC, pages 220–230, 1986.
+\bibitem{approx-search} Gad M. Landau and Uzi Vishkin. Introducing efficient parallelism into approximate string matching and a new serial algorithm. In \textit{Proc. 18th STOC}, pages 220-230, 1986.
 

### 2011-08-29 presentation/presentation.tex
m

@@
 \end{frame}
+% Jeg vil beskrive LCE problemet
+% Eksisterende løsninger er gode enten asymptotisk worst case eller praktisk average case
+% Formålet er at finde en balance mellem de to
+% Sidst vil jeg generalisere LCE til trær, som man kan bruge til LCE på komprimerede strenge
 
@@
         %\setlength{\unitlength}{1cm}
+        \definecolor{ucolor}{rgb}{.8,.8,.8}
         \definecolor{matchcolor}{rgb}{.6,1,.6}
@@
             \put(180, 30){\framebox(20, 20){a}}
-            \uncover<1-2>{
+            \uncover<1-3>{
                 \put(90, 15){\vector(0,1){13}}
@@
             \uncover<2-2>{
+                \put(80,30){\framebox(20, 20){\colorbox{ucolor}{\makebox(13,14){a}}}}
+                \put(120,30){\framebox(20, 20){\colorbox{ucolor}{\makebox(13,14){a}}}}
+            }
+            \uncover<3-3>{
                 \put(80,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){a}}}}
@@
             }
-            \uncover<2-3>{
+            \uncover<3-4>{
                 \put(200, 30){\makebox(100, 20){1 match}}
             }
-            \uncover<3-4>{
+            \uncover<4-5>{
                 \put(110, 15){\vector(0,1){13}}
@@
             \uncover<4-4>{
+                \put(100,30){\framebox(20, 20){\colorbox{ucolor}{\makebox(13,14){b}}}}
+                \put(140,30){\framebox(20, 20){\colorbox{ucolor}{\makebox(13,14){b}}}}
+            }
+            \uncover<5-5>{
                 \put(100,30){\framebox(20, 20){\colorbox{matchcolor}{\makebox(13,14){b}}}}
@@
             }
-            \uncover<4->{
+            \uncover<5->{
                 \put(200, 30){\makebox(100, 20){2 matches}}
             }
-            \uncover<5-6>{
+            \uncover<6-7>{
                 \put(130, 15){\vector(0,1){13}}
@@
             \uncover<6-6>{
+                \put(120,30){\framebox(20, 20){\colorbox{ucolor}{\makebox(13,14){a}}}}
+                \put(160,30){\framebox(20, 20){\colorbox{ucolor}{\makebox(13,14){b}}}}
+            }
+            \uncover<7-7>{
                 \put(120,30){\framebox(20, 20){\colorbox{mismatchcolor}{\makebox(13,14){a}}}}
@@
         \item[LCE value] $\LCE_s(i,j)$ is the length of the longest common prefix of the two suffixes of $s$ starting at index $i$ and $j$
-        \item[LCE problem] Efficiently query multiple LCE values on a static string $s$
+        \item[LCE problem] Efficiently query multiple LCE values on a static string $s$ and varying pairs $(i,j)$
     \end{description}
@@
 
-\begin{frame}{Kommentarer til rapporten}
+\begin{frame}{Kommentarer og rettelser til rapporten}
     \begin{itemize}
-        \item Hvordan jeg fandt frem til $r=0.73n^{0.42}$
-        \item Der står \fprint[3] nogle steder i cache-afsnittet hvor der skal stå \fprint[2]
+        \item Hvor kommer $r=0.73n^{0.42}$ fra?
+        \begin{center}
+            \includegraphics[width=0.45\textwidth,page=1]{../medium-lce-values.pdf}
+        \end{center}
+        \item Afsnit 4.5.3 Cache Optimization: \fprint[3] skal være \fprint[2]
     \end{itemize}

### 2011-08-30 presentation/presentation.tex
m

@@
 
+\newcommand{\bibs}{\hfill\\\rule{7em}{0.5pt}\vspace{.4em}\\\footnotesize References}
+\newcommand{\bib}[1]{\hfill\\\vspace{.5em}\scriptsize #1}
+
 %\subsection{The \proc{DirectComp} algorithm}
@@
     \begin{tabular}{r l}
-        Preprocessing & $O(1)$ \\
-        Space & $O(1)$ \\
+        Preprocessing & None \\
+        Space & $O(1)+|s|$ \\
         Query & $O(|\LCE(i,j)|)=O(n)$ \\
         Average query & $O(1)$ \\
-        Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big) = O\big(\frac{n}{B}\big)$ \\
+        %Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big) = O\big(\frac{n}{B}\big)$ \\
     \end{tabular}
@@
     For a string length $n$ and alphabet size $\sigma$, the average LCE value over all $n^\sigma$ strings and $n^2$ query pairs is $O(1)$.
+    \bibs
+    \bib{L. Ilie, G. Navarro, and L. Tinta. The longest common extension problem revisited and applications to approximate string searching. \textit{J. Disc. Alg.,} 8(4):418-428, 2010.}
 \end{frame}
@@
     \end{description}
-    \vspace{1em}
-    \begin{center}
+    \vspace{.5em}
+    %\begin{center}
     %    $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$\\
@@
             Average query & $O(1)$ \\
-            Query I/O & $O(1)$ \\
+            %Query I/O & $O(1)$ \\
         \end{tabular}
-    \end{center}
+    %\end{center}
     %}
+    \vspace{1em}
+    \bibs
+    \bib{J. Fischer, and V. Heun. Theoretical and Practical Improvements on the RMQ-Problem, with Applications to LCA and LCE. In \textit{Proc. 17th CPM}, pages 36-48, 2006.}
+    \bib{D. Harel, R. E. Tarjan. Fast Algorithms for Finding Nearest Common Ancestors. \textit{SIAM J. Comput.}, 13(2):338-355, 1984.}
+
 \end{frame}
@@
     \begin{tabular}{r l}
+        & $1\leq k\leq\logceil$ \\
         Preprocessing & $O(k\cdot n + \sortt)$ \\
@@
 
+\begin{comment}
 \begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
@@
 \end {frame}
+\end{comment}
 
@@
     \end{columns}
+    \bibs
+    \bib{Michael A. Bendera, Martín Farach-Colton. The Level Ancestor Problem simplified. Theoretical Computer Science 321, 2004, pages 5-12.}
+\end{frame}
+
+\begin{frame}{Constant Time String LCE on Heavy Paths}
+    Data structure:
+    \begin{itemize}
+        \item A {\bf heavy path decomposition}
+        \item A string $s$ storing labels of each heavy path
+        \begin{itemize}
+        \item Constant time {\bf string LCE} on $s$
+        \end{itemize}
+        \item {\bf Pointers} from each node to
+        \begin{itemize}
+            \item its index in $s$
+            \item the end of its heavy path
+        \end{itemize}
+        \item Constant time {\bf nearest common ancestor}
+        \item Constant time {\bf level ancestor}
+    \end{itemize}
+    \vspace{1em}
+        \begin{tabular}{r l}
+            Preprocessing & $O(\sortt)$ \\
+            Space & $O(n)$\\
+            Query & $O(\log n)$ \\
+        \end{tabular}
 \end{frame}
@@
         \item Cache optimization of \fprintk\ improves query times at $k=2$ and worsens query times at $k\geq 3$
-        %\item We can generalize LCE to use on trees:
+        \item We can generalize LCE to use on trees with $O(\log n)$ query time in linear space
         %\begin{tabular}{r l}

### 2011-08-30 presentation/presentation.tex
m

@@
 
-\subsection{I/O}
-\begin{frame}{The \fprintk\ Algorithm: I/O}
+\begin{frame}{The \fprintk\ Algorithm}
+    \begin{tabular}{r l}
+        & $1\leq k\leq\logceil$ \\
+        Preprocessing & $O(k\cdot n + \sortt)$ \\
+        Space & $O(k\cdot n)$\\
+        Query & $O(k\cdot n^{1/k})$ \\
+        Average query & $O(1)$ \\
+        %Query I/O & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
+    \end{tabular}
+
+    \vspace{2em}
+
+    \begin{tabular}{r l l l}
+        & $k=1$ & $k=2$ & $k=\logceil$ \\
+        Preprocessing & $O(\sortt)$ & $O(\sortt)$ & $O(n\log n)$ \\
+        Space & $O(n)$ & $O(n)$ & $O(n\log n)$ \\
+        Query & $O(n)$ & $O(\sqrt n)$ & $O(\log n)$ \\
+        Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
+        %Query I/O & $O\Big(\frac{n}{B}\Big)$ & $O\Big(\frac{\sqrt n}{B}\Big)$ & $O(\log n)$ \\
+    \end{tabular}
+\end{frame}
+
+\subsection{Practical Results}
+\begin{frame}{Practical Results}
+    Query times of
+    \textcolor{red}{\proc{DirectComp}},
+    \textcolor{green}{\fprint[2]},
+    \textcolor{blue}{\fprint[3]},
+    \textcolor{magenta}{\fprint[\logceil]} and
+    \textcolor{cyan}{\proc{LcpRmq}}
+    by string length
+    \begin{multicols}{2}{
+        \begin{center}
+            Average case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
+            \newpage
+            Worst case
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-alla.plt}
+        \end{center}
+    }
+    \end{multicols}
+\end {frame}
+
+\begin{frame}{Practical Results}
+    Query times of
+    \textcolor{red}{\proc{DirectComp}},
+    \textcolor{green}{\fprint[2]},
+    \textcolor{blue}{\fprint[3]},
+    \textcolor{magenta}{\fprint[\logceil]} and
+    \textcolor{cyan}{\proc{LcpRmq}}
+    by string length
+    \begin{center}
+        Medium LCE size\\
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-repeat-pow.plt}
+    \end{center}
+\end {frame}
+
+\subsection{Cache Optimization}
+\begin{frame}{Cache Optimization of \fprintk}
     \begin{itemize}
@@
 
-\begin{frame}{The \fprintk\ Algorithm}
-    \begin{tabular}{r l}
-        & $1\leq k\leq\logceil$ \\
-        Preprocessing & $O(k\cdot n + \sortt)$ \\
-        Space & $O(k\cdot n)$\\
-        Query & $O(k\cdot n^{1/k})$ \\
-        Average query & $O(1)$ \\
-        Query I/O & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
-    \end{tabular}
-
-    \vspace{2em}
-
-    \begin{tabular}{r l l l}
-        & $k=1$ & $k=2$ & $k=\logceil$ \\
-        Preprocessing & $O(\sortt)$ & $O(\sortt)$ & $O(n\log n)$ \\
-        Space & $O(n)$ & $O(n)$ & $O(n\log n)$ \\
-        Query & $O(n)$ & $O(\sqrt n)$ & $O(\log n)$ \\
-        Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
-        Query I/O & $O\Big(\frac{n}{B}\Big)$ & $O\Big(\frac{\sqrt n}{B}\Big)$ & $O(\log n)$ \\
-    \end{tabular}
-\end{frame}
-
-\subsection{Practical Results}
-\begin{frame}{Practical Results}
-    Query times of
-    \textcolor{red}{\proc{DirectComp}},
-    \textcolor{green}{\fprint[2] (cache opt.)},
-    \textcolor{blue}{\fprint[3] (not cache opt.)},
-    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
-    \textcolor{cyan}{\proc{LcpRmq}}
-    by string length
-    \begin{multicols}{2}{
-        \begin{center}
-            Average case
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-rand10.plt}
-            \newpage
-            Worst case
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-alla.plt}
-        \end{center}
-    }
-    \end{multicols}
-\end {frame}
-
-\begin{frame}{Practical Results}
-    Query times of
-    \textcolor{red}{\proc{DirectComp}},
-    \textcolor{green}{\fprint[2] (cache opt.)},
-    \textcolor{blue}{\fprint[3] (not cache opt.)},
-    \textcolor{magenta}{\fprint[\logceil] (not cache opt.)} and
-    \textcolor{cyan}{\proc{LcpRmq}}
-    by string length
-    \begin{center}
-        Medium LCE size\\
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-slides-cache-repeat-pow.plt}
-    \end{center}
-\end {frame}
-
 \begin{frame}{Cache Optimization, Practical Results}
@@
 %\subsection{Practical I/O}
-\begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
+\begin{frame}{Cache Optimization, Practical Results}
     Query times of \fprint[2]
@@
 
-\begin{frame}{The \fprintk\ Algorithm: Practical Results, I/O}
+\begin{frame}{Cache Optimization, Practical Results}
     Query times of \fprint[3]

### 2011-11-23 article/article.tex
artikel fra philip

@@
 
-\documentclass[a4]{article}
+\documentclass{llncs}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{amsmath}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{textcomp}
@@
 
-\author{Philip Bille\\\texttt{phbi@imm.dtu.dk}\and Inge Li G{\o}rtz \\\texttt{ilg@imm.dtu.dk}\and Jesper Kristensen\\\texttt{mail@jesperkristensen.dk}}
+\author{Philip Bille \and Inge Li G{\o}rtz \and Jesper Kristensen}
 
-\date{}
+\institute{Technical University of Denmark, DTU Informatics}
 
-\newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
-\newtheorem{corollary}{Corollary}
+%\date{}
+
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{lemma}{Lemma}
+%\newtheorem{corollary}{Corollary}
 
@@
 However, in practice a much simpler approach with linear query time, no extra space and no preprocessing achieves significantly better average case performance.
-We show a new algorithm, \fprintk, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k\cdot n^{1/k})$ query time using $O(k\cdot n)$ space and $O(k\cdot n + \sortt)$ preprocessing time, where $\sortt$ is the time it takes to sort $n$ numbers from $\sigma$.
+We show a new algorithm, \fprintk, which for a parameter $k$, $1\leq k\leq\logceil$, on a string of length $n$ and alphabet size $\sigma$, gives $O(k n^{1/k})$ query time using $O(k n)$ space and $O(k n + \sortt)$ preprocessing time, where $\sortt$ is the time it takes to sort $n$ numbers from $\sigma$.
 Though this solution is asymptotically strictly worse than the asymptotically best previously known algorithms, it outperforms them in practice in average case and is almost as fast as the simple linear time algorithm.
@@
 These theoretically good solutions are however not the best in practice, since they have large constant factors for both query time and space usage.
-Ilie et.
-al~\cite{ilie-navarro-tinta} introduced a much simpler solution with average case constant time and no space or preprocessing required other than storing the input string.
+Ilie et~al.~\cite{ilie-navarro-tinta} introduced a much simpler solution with average case constant time and no space or preprocessing required other than storing the input string.
 This solution has significantly better practical performance for average case input as well as for average case queries on some real world strings, when compared to the asymptotically best known algorithms.
@@
 This algorithm also achieves significantly better average case practical query time compared to the asymptotically best known worst case query time algorithms.
-The algorithm uses string fingerprinting inspired by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to achieve a time/space-tradeoff with $O(k\cdot n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for any integer $k$, $1 \leq k \leq \log n$, on a string of length $n$.
+The algorithm uses string fingerprinting inspired by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg} to achieve a time/space-tradeoff with $O(k n^{1/k})$ worst case query time, constant average case query time, and $O(kn)$ space for any integer $k$, $1 \leq k \leq \log n$, on a string of length $n$.
 The preprocessing time for this algorithm is $O(kn+\sortt)$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Previous Results\label{sec:existing-results}}
+%\subsection{Previous Results\label{sec:existing-results}}
+\paragraph{Previous Results\label{sec:existing-results}}
 Thoughout the paper let $s$ be a string of length $n$ over alphabet of size $\sigma$.
@@
 The worst case query time is thus $O(n)$.
-However, on random strings and many real-word texts Ilie~et~al.~\cite{ilie-navarro-tinta} showed that the average LCE is $O(1)$, where the average is over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
+However, on random strings and many real-word texts Ilie~et~al.~\cite{ilie-navarro-tinta} showed that the average LCE is $O(1)$, where the average is over all $\sigma^n n^2$ combinations of strings and query inputs.
 Hence, in these scenarios \proc{DirectComp} achieves $O(1)$ query time.
 
-%Given a string length $n$ and an alphabet size $\sigma$, we define \emph{average case query time} as the average of query times over all $\sigma^n\cdot n^2$ combinations of strings and query inputs.
+%Given a string length $n$ and an alphabet size $\sigma$, we define \emph{average case query time} as the average of query times over all $\sigma^n n^2$ combinations of strings and query inputs.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Our Results}
+%\subsection{Our Results}
+\paragraph{Our Results}
 We present a new LCE algorithm, \fprintk, based on multiple levels of string fingerprinting.
@@
 \begin{theorem}
-For a string $s$ of length $n$ and alphabet size $\sigma$, the \fprintk\ algorithm, where $k$ is a parameter $1 \leq k \leq \logceil$, can solve the LCE problem in $O(k\cdot n^{1/k})$ worst case query time and $O(1)$ average case query time using $O(k\cdot n)$ space and $O(\sortt + k\cdot n)$ preprocessing time.
+For a string $s$ of length $n$ and alphabet size $\sigma$, the \fprintk\ algorithm, where $k$ is a parameter $1 \leq k \leq \logceil$, can solve the LCE problem in $O(k n^{1/k})$ worst case query time and $O(1)$ average case query time using $O(k n)$ space and $O(\sortt + k n)$ preprocessing time.
 \end{theorem}
@@
 \begin{corollary}
-\fprint[\logceil] uses $O(n\cdot\log n)$ space and $O(\log n)$ query time.
+\fprint[\logceil] uses $O(n\log n)$ space and $O(\log n)$ query time.
 The data structure is equivalent to the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, and a query only needs to do one comparison at each level.
 \end{corollary}
-To preprocess the $O(k\cdot n)$ fingerprints used by our algorithm, we can use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, which takes $O(n\log n)$ time.
-For $k=o(\log n)$, we can speed up preprocessing to $O(\sortt + k\cdot n)$ by using the SA and LCP arrays.
+To preprocess the $O(k n)$ fingerprints used by our algorithm, we can use Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}, which takes $O(n\log n)$ time.
+For $k=o(\log n)$, we can speed up preprocessing to $O(\sortt + k n)$ by using the SA and LCP arrays.
 
@@
 
-\begin{table}[tp]
+\begin{table}[tb]
 \centering
@@
 \proc{DirectComp} & $O(1)$ & $O(n)$ & None \\ \hline
-\fprintk * & $O(k\cdot n)$ & $O(k\cdot n^{1/k})$ & $O(\sortt+k\cdot n)$ \\
-~~~~$k=\logceil$ * & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline
+\fprintk * & $O(k n)$ & $O(k n^{1/k})$ & $O(\sortt+k n)$ \\
+~~~~$k=\logceil$ * & $O(n\log n)$ & $O(\log n)$ & $O(n\log n)$ \\ \hline \\
 \end{tabular}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Overview}
-In \pref{sec:fprintk} we present the \fprintk\ algorithm and analyze its theoretical complexity, and in \pref{sec:results} we present our findings from tests of a practical implementation of \fprintk, which we compare against the existing algorithms.
+%\subsection{Overview}
+%\paragraph{Overview}
+%In \pref{sec:fprintk} we present the \fprintk\ algorithm and analyze its theoretical complexity, and in \pref{sec:results} we present our findings from tests of a practical implementation of \fprintk, which we compare against the existing algorithms.
 
@@
 %\paragraph{NCA.}
-The \emph{nearest common ancestor} (NCA) of two nodes $u$ and $v$ is the node of greatest depth, which is an ancestor of both $u$ and $v$.
+The \emph{nearest common ancestor} (NCA) of two nodes $u$ and $v$ in a tree is the node of greatest depth, which is an ancestor of both $u$ and $v$.
 The ancestors of a node $u$ includes $u$ itself.
@@
 At the end of a string when $i+t-1>n$, we define $F_t[i]$ by adding extra characters to the end of the string as needed.
+The last character of the string must be a special character \$, which does not occur anywhere else in the string.
 
@@
 We always have $t_0 = n^{0/k} = 1$, such that $H_0$ is the original string $s$.
-The last character of the string must be a special character \$, which does not occur anywhere else in the string.
 An example is shown in \pref{fig:fingerprint-ds}.
+%
+Since each of the $k$ tables stores $n$ fingerprints of constant size, we get the following.
+\begin{lemma}
+The \fprintk\ data structure takes $O(k n)$ space.
+\end{lemma}
 
-\begin{figure}[tp]
+\begin{figure}[tb]
     \begin{center}
-        \includegraphics[width=0.98\textwidth,page=1]{../doc/fingerprint.pdf}
+        \includegraphics[width=0.98\textwidth,page=1]{fingerprint.pdf}
+        \vspace{-.5cm}
     \end{center}
@@
 \end{figure}
-
-\begin{lemma}
-The \fprintk\ data structure takes $O(k\cdot n)$ space.
-\end{lemma}
-\begin{proof}
-Each of the $k$ tables stores $n$ fingerprints of constant size.
-\end{proof}
 
@@
 
-\begin{figure}[tp]
+\begin{figure}[tb]
     \begin{center}
-        \includegraphics[width=0.98\textwidth,page=2]{../doc/fingerprint.pdf}
+        \includegraphics[width=0.98\textwidth,page=2]{fingerprint.pdf}
+        \vspace{-.5cm}
     \end{center}
@@
 \begin{lemma}
-The worst case query time for \fprintk\ is $O(k\cdot n^{1/k})$, and the average case query time is $O(1)$.
+The worst case query time for \fprintk\ is $O(k n^{1/k})$, and the average case query time is $O(1)$.
 \end{lemma}
@@
 Since we can check $t_\ell = \Theta(n^{\ell/k})$ characters in constant time at level $\ell$, the algorithm uses $O(n^{(\ell+1)/k})/\Theta(n^{\ell/k}) = O(n^{1/k})$ time at that level.
-Over all $k$ levels, $O(k\cdot n^{1/k})$ query time is used.
+Over all $k$ levels, $O(k n^{1/k})$ query time is used.
 
@@
 \paragraph{Average Case vs.\ Worst Case\label{sec:improved-avg}}
-The first traversal in the query algorithm gaurantees $O(1)$ average case performance.
-However without it the worst case bound would remain $O(k)$.
+The first traversal in the query algorithm guarantees $O(1)$ average case performance.
+Without it the average case query time would be $O(k)$.
+However, the worst case bound would remain $O(kn^{1/k})$.
 Thus, for a better practical worst-case performance we could omit the first traversal entirely.
@@
 We could have left out step one of the query algorithm and started with $\ell=k-1$.
-This would keep the asymptotic worst case query time of $O(k\cdot n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
+This would keep the asymptotic worst case query time of $O(k n^{1/k})$, while it might improve practical worst case query time, but it would increase our average case query time to $O(k)$.
 Our experiments have shown that whenever $k$ is small, keeping step one improves average case query time, while it does not have a measurable effect on worst case query times.
@@
 
-The tables of fingerprints use $O(k\cdot n)$ space.
+The tables of fingerprints use $O(k n)$ space.
 In the case with $k=\logceil$ levels, the data structure is the one generated by Karp-Miller-Rosenberg~\cite{karp-miller-rosenberg}.
@@
 
-We can preprocess the data structure in $O(\sortt + k\cdot n)$ time using the SA and LCP arrays. First create the SA and LCP arrays. Then preprocess each of the $k$ levels using the following steps:
+We can preprocess the data structure in $O(\sortt + k n)$ time using the SA and LCP arrays. First create the SA and LCP arrays. Then preprocess each of the $k$ levels using the following steps:
 \begin{samepage}
@@
 
-\begin{figure}[tp]
+\begin{figure}[tb]
     \begin{center}
-        \includegraphics[width=0.2\textwidth,page=1]{../doc/fingerprint-preproc.pdf}
+        \includegraphics[width=0.2\textwidth,page=1]{fingerprint-preproc.pdf}
+        \vspace{-.5cm}
     \end{center}
@@
 \begin{lemma}
-The preprocessing algorithm described above takes $O(\sortt + k\cdot n)$ time.
+The preprocessing algorithm described above takes $O(\sortt + k n)$ time.
 \end{lemma}
@@
 We then preprocess each of the $k$ levels in $O(n)$ time, since we loop through $n$ substrings, and comparing neighboring substrings takes constant time when we use the LCP array.
-The total preprocessing time becomes $O(\sortt\\ + k\cdot n)$.
+The total preprocessing time becomes $O(\sortt\\ + k n)$.
 \end{proof}
@@
 
-\begin{figure}[tp]
+\begin{figure}[tb]
     \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-alla.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-article-repeat-pow.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-article-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-article-alla.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-article-repeat-pow.plt}
     \end{center}
@@
 bible.txt & $3.9\cdot 2^{20}$ & 63 & 8.5 & 11.3 & 10.5 & 12.6 & 284.0 \\ \hline
-world192.txt & $2.3\cdot 2^{20}$ & 93 & 7.9 & 10.5 & 9.8 & 12.7 & 291.7 \\ \hline
+world192.txt & $2.3\cdot 2^{20}$ & 93 & 7.9 & 10.5 & 9.8 & 12.7 & 291.7 \\ \hline \\
 \end{tabular}
@@
 
-\begin{figure}[tp]
+\begin{figure}[tb]
     \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-alla.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-repeat-pow.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint-cache-rand2.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-fingerprint-cache-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-fingerprint-cache-alla.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-fingerprint-cache-repeat-pow.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-fingerprint-cache-rand2.plt}
     \end{center}
@@
 
-\begin{figure}[tp]
+\begin{figure}[tb]
     \begin{center}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint3-cache-rand10.plt}
-        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-fingerprint3-cache-alla.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-fingerprint3-cache-rand10.plt}
+        \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{length-fingerprint3-cache-alla.plt}
+        \vspace{-.5cm}
     \end{center}
@@
 
-The amount of I/O used by \fprintk\ is $O(k\cdot n^{1/k})$.
+The amount of I/O used by \fprintk\ is $O(k n^{1/k})$.
 However if we structure our tables of fingerprints differently, we can improve the number of I/O operations to $O(k(n^{1/k}/B+1))$ in the cache-oblivious model.
@@
 The cache optimized \fprint[2] variant with shift operations shows an increase in query times for large inputs, which we cannot explain.
-
+%
 The last plot on \pref{fig:plot-fingerprint-cache-horiz} shows a variant of average case where the alphabet size is changed to two.
@@
 In this bad case plot we see that the cache optimized variant of \fprint[2] has only slightly worse query time compared to the variant, which is not cache optimized.
-
-\pref{fig:plot-fingerprint3-cache} shows our measurements for \fprint[3].
-Unlike for \fprint[2], we see a degraded performance on worst case strings compared to the not cache optimized variant.
-Our results thus show that this cache optimization of \fprintk\ only gives a performance improvement for $k=2$.
+\pref{fig:plot-fingerprint3-cache} shows the measurements for \fprint[3].
+Unlike \fprint[2], the cache optimized variant is slightly slower than the unoptimized variant.
+Hence, our cache optimization is effective for $k=2$ but not $k=3$.
 
@@
 
-We have presented the \fprintk\ algorithm, where $k$ is a parameter $1 \leq k \leq \logceil$, which for a string $s$ of length $n$ and alphabet size $\sigma$, can solve the LCE problem in $O(k\cdot n^{1/k})$ worst case query time and $O(1)$ average case query time using $O(k\cdot n)$ space and $O(\sortt + k\cdot n)$ preprocessing time.
+We have presented the \fprintk\ algorithm, where $k$ is a parameter $1 \leq k \leq \logceil$, which for a string $s$ of length $n$ and alphabet size $\sigma$, can solve the LCE problem in $O(k n^{1/k})$ worst case query time and $O(1)$ average case query time using $O(k n)$ space and $O(\sortt + k n)$ preprocessing time.
 

### 2011-11-23 article/article.tex
m

@@
 The first traversal in the query algorithm guarantees $O(1)$ average case performance.
-Without it the average case query time would be $O(k)$.
+Without it, i.e.\ if the query started with $\ell=k-1$ and omitted step one, the average case query time would be $O(k)$.
 However, the worst case bound would remain $O(kn^{1/k})$.
@@
 The strings consist of repeating substrings of $r$ unique characters.
-%The value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
+The value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
 \end{description}

### 2011-11-23 article/article.tex
m

@@
 The LCE problem can be used in many algorithms for solving other algorithmic problems, e.g., the Landau-Vishkin algorithm for approximate string searching~\cite{approx-search}.
-Solutions with linear space, constant query time, and $O(\sortt)$ preprocessing time exists for the problem~\cite{nca,jf-rmq}.
+Solutions with linear space, constant query time, and $O(\sortt)$ preprocessing time exist for the problem~\cite{nca,jf-rmq}.
 Here $\sortt$ is the time it takes to sort $n$ numbers from an alphabet of size $\sigma$.
@@
 In situations where we need both average case and worst case performance to be good, none of the existing solutions are ideal.
-Such a situation could be use of approximate string searching in a firewall.
+An example could be a firewall, which needs to do approximate string searching.
 The firewall should not allow an attacker to significantly degrade its performance by sending it carefully crafted packages.
-At the same time it must scan legitimate data fast.
+At the same time it must scan legitimate data quickly.
 The main goal of this paper is to design an algorithm that performs well in both situations, that is, achieves a good worst-case guarantee while maintaining a fast average case performance.
@@
 \paragraph{Previous Results\label{sec:existing-results}}
-Thoughout the paper let $s$ be a string of length $n$ over alphabet of size $\sigma$.
+Thoughout the paper let $s$ be a string of length $n$ over an alphabet of size $\sigma$.
 Ilie~et~al.~\cite{ilie-navarro-tinta} gave an algorithm, \proc{DirectComp}, for solving the LCE problem, which uses no preprocessing and has $O(\LCE(i,j))$ query time.
@@
 The worst case query time is thus $O(n)$.
-However, on random strings and many real-word texts Ilie~et~al.~\cite{ilie-navarro-tinta} showed that the average LCE is $O(1)$, where the average is over all $\sigma^n n^2$ combinations of strings and query inputs.
+However, on random strings and many real-word texts, Ilie~et~al.~\cite{ilie-navarro-tinta} showed that the average LCE is $O(1)$, where the average is over all $\sigma^n n^2$ combinations of strings and query inputs.
 Hence, in these scenarios \proc{DirectComp} achieves $O(1)$ query time.
@@
 The length of the longest common prefix of two arbitrary suffixes in SA can be found as the minimum of all LCP values of neighbor suffixes between the two desired suffixes, because SA lists the suffixes in lexicographical ordering, i.e., $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$, where $\SAinv[i] < \SAinv[j]$.
-\pref{tab:article-algorithms} summaries the above theoretical bounds.
+\pref{tab:article-algorithms} summarizes the above theoretical bounds.
 
@@
 The strings consist of repeating substrings of $r$ unique characters.
-The value of $r$ is a power regression over $(r,n) = (5,100)$, $(15,1.000)$, $(35,10.000)$ and $(100,100.000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
+The value of $r$ is a power regression over $(r;n) = (5;100)$, $(15;1,000)$, $(35;10,000)$ and $(100;100,000)$, where each of the values of $n$ were chosen such that \fprintk\ was faster than both \proc{DirectComp} and \proc{LcpRmq} on a plot with the given constant value of $r$.
 \end{description}
@@
 On worst case strings, the \fprintk\ algorithms are significantly better than \proc{DirectComp} and somewhat worse than \proc{LcpRmq}.
-Up until $n=30.000$ the three measured \fprintk\ algorithms have nearly the same query times.
-Of the \fprintk\ algorithms, the $k=2$ variant has a slight advantage for small strings of length less than around $2.000$.
-For longer strings the $k=3$ variant performs the best up to strings of length $250.000$, at which point the $k=\logceil$ variant becomes the best.
+Up until $n=30,000$ the three measured \fprintk\ algorithms have nearly the same query times.
+Of the \fprintk\ algorithms, the $k=2$ variant has a slight advantage for small strings of length less than around $2,000$.
+For longer strings the $k=3$ variant performs the best up to strings of length $250,000$, at which point the $k=\logceil$ variant becomes the best.
 This indicates that for shorter strings, using fewer levels is better, and when the input size increases, the \fprintk\ variants with better asymptotic query times have better worst case times in practice.
@@
 
-\proc{LcpRmq} shows a significant jump in query times around $n=1.000.000$ on the plot with average case strings, but not on the plot with worst case strings.
+\proc{LcpRmq} shows a significant jump in query times around $n=1,000,000$ on the plot with average case strings, but not on the plot with worst case strings.
 We have run the tests in Cachegrind, and found that the number of instructions executed and the number of data reads and writes are exactly the same for both average case strings and worst case strings.

### 2012-02-09 article-presentation/presentation.tex
Opret konference slides

--- 2011-08-30 presentation/presentation.tex
@@
         Space & $O(1)+|s|$ \\
-        Query & $O(|\LCE(i,j)|)=O(n)$ \\
+        Query & $O(\LCE(i,j))=O(n)$ \\
         Average query & $O(1)$ \\
-        %Query I/O & $O\Big(\frac{|\LCE(i,j)|}{B}\Big) = O\big(\frac{n}{B}\big)$ \\
+        %Query I/O & $O\Big(\frac{\LCE(i,j)}{B}\Big) = O\big(\frac{n}{B}\big)$ \\
     \end{tabular}
@@
 
-%\begin{comment}
+\begin{comment}
 \section{LCE on trees}
@@
 \end{frame}
-%\end{comment}
+\end{comment}
 
@@
         \item Cache optimization of \fprintk\ improves query times at $k=2$ and worsens query times at $k\geq 3$
-        \item We can generalize LCE to use on trees with $O(\log n)$ query time in linear space
+        %\item We can generalize LCE to use on trees with $O(\log n)$ query time in linear space
         %\begin{tabular}{r l}
@@
         \end{center}
-        \item Afsnit 4.5.3 Cache Optimization: \fprint[3] skal være \fprint[2]
+        %\item Afsnit 4.5.3 Cache Optimization: \fprint[3] skal være \fprint[2]
     \end{itemize}

### 2012-02-09 article-presentation/presentation.tex
add preprocessing slide

@@
 
-\begin{comment}
+%\begin{comment}
 \subsection{Preprocessing}
@@
 
-\begin{comment}
-\section{LCE on trees}
-\subsection{Compression}
-\begin{frame}{LCE on Compressed Strings}
-    Goal
-    \begin{itemize}
-        \item Allow LCE queries without decompressing the string
-        \item Using Ziv-Lempel compression (LZ)
-    \end{itemize}
-    How
-    \begin{itemize}
-        \item LZ compression represents the string as a tree
-        \item An LCE query on a LZ compressed string is a number of LCE queries on a tree
-    \end{itemize}
-\end{frame}
-
-\begin{frame}{LCE on Trees}
-    \begin{itemize}
-        \item Trees:
-        \begin{itemize}
-            \item One character on each edge
-            \item LCE is the length of the longest common prefix of two strings along two paths
-        \end{itemize}
-        \item $p_i=bbc$ and $p_j=bbba$ gives $\LCE(p_i,p_j)=2$
-    \end{itemize}
-    \begin{center}
-        \includegraphics[width=0.4\textwidth,page=1]{tree-lce.pdf}
-    \end{center}
-\end{frame}
-
-\subsection{Constant Time String LCE on Heavy Paths}
-\begin{frame}{Constant Time String LCE on Heavy Paths}
-    \begin{multicols}{2}{
-        \begin{itemize}
-            \item Data structure:
-            \begin{itemize}
-                \item Construct a heavy path decomposition
-                \item For each heavy path, store characters as a substring of $s$
-            \end{itemize}
-            \item Query:
-            \begin{itemize}
-                \item Use constant time string LCE on $s$ for each heavy path
-            \end{itemize}
-        \end{itemize}
-        \begin{center}
-            \includegraphics[width=0.44\textwidth,page=4]{../doc/tree-fingerprint.pdf}
-        \end{center}
-        \begin{tabular}{r l}
-            Preprocessing & $O(\sortt)$ \\
-            Space & $O(n)$\\
-            Query & $O(\log n)$ \\
-        \end{tabular}
-    }
-    \end{multicols}
-\end{frame}
-
-\begin{frame}{Constant Time String LCE on Heavy Paths}
-    \begin{columns}[t]
-        \begin{column}{0.7\textwidth}
-            \begin{itemize}
-                \item How to find the indexes $(i,j)$ in the string:
-                \begin{itemize}
-                    \item Store an index at each node
-                \end{itemize}
-                \item How to know then the heavy path splits from the queried path:
-                \begin{itemize}
-                    \item Store a pointer to the end of the heavy path at each node
-                    \item Find NCA of the end of the heavy path and the end of the queried path
-                \end{itemize}
-                \item How to find a node on the queried path:
-                \begin{itemize}
-                    \item Use level ancestor
-                \end{itemize}
-            \end{itemize}
-        \end{column}
-        \begin{column}{0.3\textwidth}
-        \hfill\\
-            \includegraphics[width=1\textwidth,page=3]{../doc/tree-fingerprint.pdf}
-        \end{column}
-    \end{columns}
-    \bibs
-    \bib{Michael A. Bendera, Martín Farach-Colton. The Level Ancestor Problem simplified. Theoretical Computer Science 321, 2004, pages 5-12.}
-\end{frame}
-
-\begin{frame}{Constant Time String LCE on Heavy Paths}
-    Data structure:
-    \begin{itemize}
-        \item A {\bf heavy path decomposition}
-        \item A string $s$ storing labels of each heavy path
-        \begin{itemize}
-        \item Constant time {\bf string LCE} on $s$
-        \end{itemize}
-        \item {\bf Pointers} from each node to
-        \begin{itemize}
-            \item its index in $s$
-            \item the end of its heavy path
-        \end{itemize}
-        \item Constant time {\bf nearest common ancestor}
-        \item Constant time {\bf level ancestor}
-    \end{itemize}
-    \vspace{1em}
-        \begin{tabular}{r l}
-            Preprocessing & $O(\sortt)$ \\
-            Space & $O(n)$\\
-            Query & $O(\log n)$ \\
-        \end{tabular}
-\end{frame}
-\end{comment}
-
 \section{Summary}
@@
         \item Cache optimization of \fprintk\ improves query times at $k=2$ and worsens query times at $k\geq 3$
-        %\item We can generalize LCE to use on trees with $O(\log n)$ query time in linear space
-        %\begin{tabular}{r l}
-        %    Preprocessing & $O(\sortt)$ \\
-        %    Space & $O(n)$\\
-        %    Query & $O(\log n)$ \\
-        %\end{tabular}
     \end{itemize}
@@
 
+\begin{comment}
 \begin{frame}{Kommentarer og rettelser til rapporten}

### 2012-02-11 article-presentation/presentation.tex
lata presentation

@@
 \setbeamertemplate{footline}[frame number]
-\title{Algorithms for Longest Common Extensions}
-\author{Jesper Kristensen}
-\institute[DTU Informatics]{DTU Informatics\\Technical University of Denmark}
-\date{August 31, 2011}
+\title{Longest Common Extensions via Fingerprinting}
+\author{Philip Bille \and Inge Li G{\o}rtz \and Jesper Kristensen}
+\institute[DTU Informatics]{Technical University of Denmark\\DTU Informatics}
+\date{LATA, March 9, 2012}
 \begin{document}
@@
 \begin{frame}{Existing Algorithms: \proc{SuffixNca} and \proc{LcpRmq}}
-    %\begin{multicols}{2}{
-    %    \begin{align*}
-    %        s&=abbababba\\
-    %        s[2\twodots n]&=bbababba\\
-    %        s[3\twodots n]&=bababba\\
-    %        \LCE_s(2,3)&=1
-    %    \end{align*}
-    %    \newpage
-    %    \begin{center}
-    %        %\uncover<2->{
-    %        \includegraphics[width=0.26\textwidth,page=1]{../doc/sa+lcp+min.pdf}
-    %        %}
-    %    \end{center}
-    %}
-    %\end{multicols}
-    %\uncover<3->{
     Two algorithms with best known bounds:
@@
     \vspace{.5em}
-    %\begin{center}
-    %    $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$\\
-    %    where $\SAinv[i] < \SAinv[j]$\\
-    %    \vspace{1em}
         \begin{tabular}{r l}
@@
         \end{tabular}
-    %\end{center}
-    %}
     \vspace{1em}
@@
 \end{frame}
+
+\begin{comment}
+\begin{frame}{Existing Algorithm: \proc{LcpRmq}}
+    \begin{multicols}{2}{
+        \begin{align*}
+            s&=&abbababba\\
+            s[2\twodots n]&=&bbababba\\
+            s[3\twodots n]&=&bababba\\
+            \LCE_s(2,3)&=1
+        \end{align*}
+        \newpage
+        \begin{center}
+            %\uncover<2->{
+            \includegraphics[width=0.26\textwidth,page=1]{../doc/sa+lcp+min.pdf}
+            %}
+        \end{center}
+    }
+    \end{multicols}
+    \begin{center}
+        $\LCE(i,j)=\LCP[\RMQ_{\LCP}(\SAinv[i] + 1, \SAinv[j])]$\\
+        where $\SAinv[i] < \SAinv[j]$\\
+    \end{center}
+\end{frame}
+\end{comment}
 

### 2012-02-26 article-presentation/presentation.tex
move definition of LCE before DirectComp in article presentation
Shorten the article presentation

@@
 \section{Introduction}
+
+\subsection{The LCE Problem}
+\begin{frame}{The LCE Problem}
+    \begin{description}
+        \item[LCE value] $\LCE_s(i,j)$ is the length of the longest common prefix of the two suffixes of a string $s$ starting at index $i$ and $j$
+        \item[LCE problem] Efficiently query multiple LCE values on a static string $s$ and varying pairs $(i,j)$
+    \end{description}
+    Example:
+    \begin{description}
+        \item[input:] $s=abbababba$, $(i,j)=(4,6)$
+        \item[suffix $i$] = $ababba$
+        \item[suffix $j$] = $abba$
+        \item[longest common prefix $(i,j)$] = $ab$
+        \item[$\LCE_s(i,j)$] = $2$
+    \end{description}
+\end{frame}
+
+\section{Existing Results}
 \subsection{The \proc{DirectComp} algorithm}
-\begin{frame}{The \proc{DirectComp} algorithm}
+\begin{frame}{Existing algorithm: \proc{DirectComp}}
     \begin{block}{Input}
@@
 
-\subsection{The LCE Problem}
-\begin{frame}{The LCE Problem}
-    \begin{description}
-        \item[LCE value] $\LCE_s(i,j)$ is the length of the longest common prefix of the two suffixes of $s$ starting at index $i$ and $j$
-        \item[LCE problem] Efficiently query multiple LCE values on a static string $s$ and varying pairs $(i,j)$
-    \end{description}
-\end{frame}
-
 \newcommand{\bibs}{\hfill\\\rule{7em}{0.5pt}\vspace{.4em}\\\footnotesize References}
@@
     \begin{tabular}{r l}
-        Preprocessing & None \\
+        %Preprocessing & None \\
         Space & $O(1)+|s|$ \\
@@
 
-\section{Existing Results}
 %\subsection{The \proc{SuffixNca} Algorithm}
@@
         \begin{tabular}{r l}
-            Preprocessing & $O(\sortt)$ \\
+            %Preprocessing & $O(\sortt)$ \\
             Space & $O(n)$ \\
@@
 
-%\begin{comment}
+\begin{comment}
 \subsection{Preprocessing}
@@
         & $1\leq k\leq\logceil$ \\
-        Preprocessing & $O(k\cdot n + \sortt)$ \\
+        %Preprocessing & $O(k\cdot n + \sortt)$ \\
         Space & $O(k\cdot n)$\\
@@
         & $k=1$ & $k=2$ & $k=\logceil$ \\
-        Preprocessing & $O(\sortt)$ & $O(\sortt)$ & $O(n\log n)$ \\
+        %Preprocessing & $O(\sortt)$ & $O(\sortt)$ & $O(n\log n)$ \\
         Space & $O(n)$ & $O(n)$ & $O(n\log n)$ \\
@@
 
+\begin{comment}
 \begin{frame}{Practical Results}
@@
         & \proc{Comp} & \proc{SuffixNca} & \fprintk \\
-        Preprocessing & $O(1)$ & $O(\sortt)$ & $O(k\cdot n+\sortt)$ \\
+        %Preprocessing & $O(1)$ & $O(\sortt)$ & $O(k\cdot n+\sortt)$ \\
         Space & $O(1)$ & $O(n)$ & $O(k\cdot n)$\\

### 2012-03-02 article-presentation/presentation.tex
clearer example
updates based on feedback from presentation rehersal at DTU

@@
         \item[input:] $s=abbababba$, $(i,j)=(4,6)$
-        \item[suffix $i$] = $ababba$
-        \item[suffix $j$] = $abba$
-        \item[longest common prefix $(i,j)$] = $ab$
+        \item[suffix $i$ of $s$] = $ababba$
+        \item[suffix $j$ of $s$] = $abba$
+        \item[longest common prefix] = $ab$
         \item[$\LCE_s(i,j)$] = $2$
@@
             Average case
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-slide-rand10.plt}
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-slide-rand10.plt}\\
+            $s=$ random characters\\ $\sigma=10$
             \newpage
             Worst case
-            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-slide-alla.plt}
+            \includegraphics[width=0.49\textwidth,type=pdf,ext=.pdf,read=.pdf]{../src/results/length-rmq-directcomp-slide-alla.plt}\\
+            $s=$ aaaaa...
         \end{center}
@@
                 \item If the current substring $s[\SA[i]\twodots\SA[i]+t_\ell-1]$ is equal to the previous substring, give it the same fingerprint as the previous substring, otherwise give it a new unused fingerprint.
-The two substrings are equal when $\LCE[i] \geq t_\ell$.
+The two substrings are equal when $\LCP[i] \geq t_\ell$.
             \end{itemize}
@@
         \begin{center}
+            \includegraphics[width=0.2\textwidth,page=1]{../doc/fingerprint-preproc.pdf}
+        \end{center}
+        \newpage
+        %\begin{center}
         \hfill\\
@@
             $H_\ell =$ [3, 8, 6, 2, 6, 3, 8, 5, 1]
-        \end{center}
-        \newpage
-        \begin{center}
-            \includegraphics[width=0.2\textwidth,page=1]{../doc/fingerprint-preproc.pdf}
-        \end{center}
+        %\end{center}
     }
@@
     \end{tabular}
+
+    \vspace{2em}
+
+    Space for \fprintk\ is the same as for \proc{LcpRmq} when $k=6$.
 \end{frame}
@@
         Average query & $O(1)$ & $O(1)$ & $O(1)$ \\
+        & fast & slow & fast \\
         Query I/O & $O\big(\frac{n}{B}\big)$ & $O(1)$ & $O\Big(k\cdot\Big(\frac{n^{1/k}}{B}+1\Big)\Big)$ \\
+        Code complexity & very simple & complex & simple \\
     \end{tabular}
@@
     \begin{itemize}
-        \item In practice, the \fprintk\ algorithm is...
-        \begin{itemize}
-            \item ...almost as good as \proc{DirectComp} and significantly better than \proc{LcpRmq} in average case
-            \item ...significantly better than \proc{DirectComp} but worse than \proc{LcpRmq} in worst case
-        \end{itemize}
+        %\item In practice, the \fprintk\ algorithm is...
+        %\begin{itemize}
+        %    \item ...almost as good as \proc{DirectComp} and significantly better than \proc{LcpRmq} in average case
+        %    \item ...significantly better than \proc{DirectComp} but worse than \proc{LcpRmq} in worst case
+        %\end{itemize}
         \item Cache optimization of \fprintk\ improves query times at $k=2$ and worsens query times at $k\geq 3$
+        \item Space for \fprintk\ is the same as for \proc{LcpRmq} when $k=6$.
     \end{itemize}
